### 2017-12-27 write-up/proposal.tex
Initial commit

\documentclass[11pt]{article}

\usepackage[top=1.1in, bottom=1.25in, left=1.25in, right=1.25in]{geometry}
\usepackage{booktabs}
\usepackage{charter}
\usepackage{parskip}
\usepackage{setspace}
\usepackage[dvipsnames]{xcolor}
\usepackage[nottoc]{tocbibind}
\usepackage[utf8]{inputenc}
\usepackage[outputdir=build]{minted}
\usepackage{hyperref}

% \linespread{1.5}
\newcommand{\guidance}[1]{\textsl{\textcolor{Gray}{#1}}}

\begin{document}

\centerline{\Large Applications of Linear Types}
\vspace{2em}
\centerline{\Large \emph{A Part III project proposal}}
\vspace{2em}
\centerline{\large D. C. Makwana (\emph{dcm41}), Trinity College}
\vspace{1em}
\centerline{\large Project Supervisor: Dr.~N.~K.~Krishnaswami, Dr~S.~Dolan}
\vspace{1em}

\begin{abstract}
\textsl{%
    Complex resource management is a challenge for programmers.  Linear types
    allow the compiler and the user to statically keep track of the resources
    that an implementation uses, and offer a promising solution to
    resource management. However, they have not made their way into mainstream
    programming languages. I aim to take a step towards this with an
    incremental approach, by allowing users to use linear types via an OCaml
    library that offers them the ability to manipulate primitives of the Owl
    numerical library in a safe and efficient manner, by inferring and removing
    redundant copies.
}%

\end{abstract}

\section{Introduction, approach and outcomes}

% \guidance{%
%     Provide an introduction to your project or essay. In particular, try to
%     motivate the work and explain the relevant context (general background, as
%     well as sufficient detail about any related work).
% }

% \guidance{%
%     What's the basic idea and approach? What are you thinking of doing, and how
%     is it going to solve the problem (or need) you've identified. What are you
%     going to ``produce''?  A project will typically produce one (or perhaps
%     more) of the following: a piece of software, an evaluation of a published
%     result, a proof, or the design (and perhaps the construction of) a new piece
%     of hardware. An essay will typically either review and critique a particular
%     area of the academic literature, or evaluate a published result or proof.
%     Try to name the specific things (and describe them) in this part of the
%     proposal -- this will allow you to refer to them in the next.
% }

Simply stated, a linear value is a value that must be used once and only once.
More abstractly, it is the type-theory corresponding to Girard's Linear Logic.
Examples of linear types in the real world are the recent efforts to extend GHC
with Linear Types~\cite{retrofitting} for safe, pure-functional streaming: a
pointer into a stream must be processed before moving onto the next element (one
use) but crucially, you cannot keep this pointer and refer to it again (only one
use).  Rust implements linear types behind the scenes but presents \emph{affine}
types to the programmer: if a destructor is not called explicitly, the compiler
inserts ones when it infers the value's lifetime has ended. Apart from memory
(which enables zero-copy buffers and state-management in a pure functional
language), linear types can also be used to control other resources such as
files, network communications, pointers with capabilities and concurrent
channels.

My project will focus on memory, but not manipulating raw memory block
primitives, but ensuring the primitive types (such as large matrices) that
Owl~\cite{owl} manipulates are used efficiently. Owl provides an immutable
interface and so performs several redundant copies during a computation. It does
provide more efficient implementations of common numeric algorithms as
primitives which are hand-optimised and abstract away any unsafe calls involved.
This approach trades performance for conciseness, safety and ease of reasoning.

The opposite approach -- to rely on the programmer to note and remove redundant
copies -- is used extensively by Fortran's BLAS~\cite{blas} and Boost's
uBLAS~\cite{ublas}. They offer many different conventions for each operation to
meet every situation. For example, a function implementing a binary operation on
two matrices may return a newly-allocated matrix containing the result or modify
one of two or three input arguments to hold the result (if there is no aliasing,
like in \texttt{noalias(C)~=~prod(A,~B)} where the \texttt{=} operator is
overloaded) and even specify temporary intermediate results with annotations (as
with \texttt{prod(A,~prod<temp\_type>(B,C))}).

I hypothesise that it is possible to have the best of both worlds: write your
code as you wish and then have the computer statically determine the most
space-efficient sequence of copies and in-place writes that produces the
desired result. A small example to illuminate the point follows. Suppose we
have $n\times n$ matrices A, B and C. Knowing all those matrices are used
once and only once, we would then infer that we can evaluate the expression
\mintinline{ocaml}{(A + B) + C} \emph{in-place}, say, using the memory that A
occupies.

\begin{minted}{ocaml}
    let y  = (A + B) + C in    let y  = ((A + B) + C) * A in
    (* is the same as *)       (* is the same as *)
    let i1 = A + B  in         let (A1, A2) = dup(A) in
    let i2 = i1 + C in         let i1 = A1 + B       in
    i2                         let i2 = i1 + C       in
    (* is compiled to *)       let i4 = i2 * A2      in
    A := A + B;                i4
    A := A + C;                (* is compiled to *)
    return A                   B := B + A; B := B + C;
                               B := B * A; return B
\end{minted}

However, when an expression is used more than once, we can be more explicit and
invoke a function \emph{dup} for duplicate, to save A for use later.

It is important to note that \emph{dup} is not necessarily a copy: it is merely
a signal to the compiler via the type system that A's memory should not be
written to until both the results of the \emph{dup} (A1 and A2) are used.
Indeed, knowing that B is not used again, we could store (intermediate) results
in its memory.

\section{Workplan}
% \guidance{%
%     Project students have approximately 28 weeks between the submission of
%     the proposal, and the submission of the dissertation. Essay students have
%     approximately 14 weeks.  This section should account for what you intend
%     to do during that time. One approach would be to divide the time into
%     two-week chunks, and describe the work to be done (and, as relevant,
%     milestones to be achieved) in each chunk. You should leave one chunk for
%     writing an essay or two chunks for writing a project dissertation. You
%     should leave 1 chunk for contingencies.
% }

The output of this project will be a library that will, given a linear algebra
expression, produce OCaml code to compute the expression using mutability and
unsafe functions that will, thanks the static guarantees provided by linear
types, be safe, correct and perform well (less memory consumption, fewer
dynamic checks). Evaluation will consist of comparing test programs against
Owl's default immutable interface, its lazy interface, Julia and Python/Numpy.

During implementation, I will also explore other questions such as: if a matrix
does need to be copied, then, when during execution is it best to do so?  What
is necessary to support the class of numerical algorithms like the small
example above, such as Kalman filters that although in their mathematical
expression appear to use a value twice, can be implemented to have values
modified in-place? Numerical libraries and lazy interfaces (that use mutability
behind the scenes by reference counting) tend to perform poorly for allocating
lots of small matrices (for example, in robust linear regression); will
statically determining memory usage improve their performance?

\vspace*{\fill}
\begin{table}[H]
\centering
\begin{tabular}{ll}
    \toprule

    Date & Work \\

    \midrule

    % December
    04-12-2017 & Set-up build environments (using Nix) Git repository          \\
               & continuous-integration and testing framework. Start           \\
               & designing explicitly-typed DSL (domain-specific language).    \\
                                                                               \\

    18-12-2017 & Finish designing DSL and write a tree-interpreter for it      \\
               & that outputs matrices using Owl.                              \\
                                                                               \\

    % January
    01-01-2018 & Write a lexer/parser and \emph{compiler} for this language.   \\
                                                                               \\

    15-01-2018 & Review with supervisor/break for assignments.                 \\
                                                                               \\

    29-01-2018 & Start with basic elaboration and inference.                   \\
                                                                               \\

    % February
    12-02-2018 & Buffer.                                                       \\
                                                                               \\

    26-02-2018 & Project progress review. Begin discussion of advanced         \\
               & extensions such as staging or further inference and analysis. \\
                                                                               \\

    % March
    12-03-2018 & `One-minute madness: project `elevator pitch''.               \\
               &  Start with chosen extension. Submit to ICFP.                 \\
                                                                               \\

    26-03-2018 & Plan evaluation and start and write-up.                       \\
                                                                               \\

    % April
    09-04-2018 & Buffer.                                                       \\
                                                                               \\

    23-04-2018 & Perform evaluation.                                           \\
                                                                               \\

    % May
    07-05-2018 & Write-up.                                                     \\
                                                                               \\

    21-05-2018 & Project title changes (25\textsuperscript{th}).               \\
               & Final changes to code and thesis. Prepare for POPL.           \\
                                                                               \\

    % June
    01-06-2018 & Submission Deadline.                                          \\
                                                                               \\

    12-06-2018 & MPhil and Part III Project presentations.                     \\
                                                                               \\

  \bottomrule

\end{tabular}
\end{table}
\vspace*{\fill}
\newpage
\bibliographystyle{plain}
\bibliography{proposal.bib}

\end{document}

### 2017-12-27 src/utop.sh

@@ -1,0 +1,7 @@
+#/usr/bin/env bash
+# Because `jbuilder utop` crashes and burns
+# Run from within src directory
+
+set -euo pipefail
+jbuilder build lt4la.a
+jbuilder exec -- utop -init utop_init -I ./_build/default/src

### 2018-02-05 write-up/abstract.tex
Start Dissertation
At some point I should put this under continuous-integration as well?

\newpage
{\normalfont\Huge\sffamily\bfseries Abstract}
\vspace{24pt}

This is the abstract. Write a summary of the whole thing. Make 
sure it fits in one page.

\newpage
\vspace*{\fill}

### 2018-02-05 write-up/background.tex
Start Dissertation
At some point I should put this under continuous-integration as well?

\chapter{Background}

\begin{guidance}
    A more extensive coverage of what's required to understand your
    work. In general you should assume the reader has a good undergraduate
    degree in computer science, but is not necessarily an expert in
    the particular area you've been working on. Hence this chapter
    may need to summarize some ``text book'' material.

    This is not something you'd normally require in an academic paper,
    and it may not be appropriate for your particular circumstances.
    Indeed, in some cases it's possible to cover all of the ``background''
    material either in the introduction or at appropriate places in
    the rest of the dissertation.

\end{guidance}

\section{Tracking Resources with Linearity}

\section{Problem in Detail}

\section{Proposed Solution}

### 2018-02-05 write-up/conclusion.tex
Start Dissertation
At some point I should put this under continuous-integration as well?

\chapter{Conclusion}

\begin{guidance}
    As you might imagine: summarizes the dissertation, and draws
    any conclusions. Depending on the length of your work, and
    how well you write, you may not need a summary here.

    You will generally want to draw some conclusions, and point
    to potential future work.
\end{guidance}

### 2018-02-05 write-up/declaration.tex
Start Dissertation
At some point I should put this under continuous-integration as well?

\newpage
{\normalfont\Huge\sffamily\bfseries Declaration}

\vspace{24pt}

I \authorname of \authorcollege, being a candidate for Computer Science Tripos,
Part III, hereby declare that this report and the work described in it are my
own work, unaided except as may be specified below, and that the report does
not contain material that has already been used to any substantial extent for a
comparable purpose.

\vspace{24pt}
Total word count: \wordcount

\vspace{60pt}
\textbf{Signed}:

\vspace{12pt}
\textbf{Date}:

\vfill

This dissertation is copyright \copyright 2018 \authorname. 
\\
All trademarks used in this dissertation are hereby acknowledged.

\newpage
\vspace*{\fill}

### 2018-02-05 write-up/evaluation.tex
Start Dissertation
At some point I should put this under continuous-integration as well?

\chapter{Evaluation}

\begin{guidance}
    For any practical projects, you should almost certainly have
    some kind of evaluation, and it's often useful to separate
    this out into its own chapter.
\end{guidance}

### 2018-02-05 write-up/implementation.tex
Start Dissertation
At some point I should put this under continuous-integration as well?

\chapter{Implementation}

\begin{guidance}
This chapter may be called something else\ldots but in general
the idea is that you have one (or a few) ``meat'' chapters which
describe the work you did in technical detail.
\end{guidance}

\section{Core Language}

Here, I will describe the syntax and semantics.

\section{Elaboration and Inference}

\section{Compiling and Metaprogramming}

### 2018-02-05 write-up/introduction.tex
Start Dissertation
At some point I should put this under continuous-integration as well?

\chapter{Introduction}

% No resetting page numbers: ISO 7144
% \pagenumbering{arabic}
% \setcounter{page}{0}

\begin{guidance}
    This is the introduction where you should introduce your work.  In
    general the thing to aim for here is to describe a little bit of the
    context for your work --- why did you do it (motivation), what was the
    hoped-for outcome (aims) --- as well as trying to give a brief
    overview of what you actually did.

    It's often useful to bring forward some ``highlights'' into
    this chapter (e.g.\ some particularly compelling results, or
    a particularly interesting finding).

    It's also traditional to give an outline of the rest of the
    document, although without care this can appear formulaic
    and tedious. Your call.
\end{guidance}

\section{Overview of Problem}

\section{Contributions}

### 2018-02-05 write-up/related_work.tex
Start Dissertation
At some point I should put this under continuous-integration as well?

\chapter{Related Work}

\begin{guidance}

    This chapter covers relevant (and typically, recent) research
    which you build upon (or improve upon). There are two complementary
    goals for this chapter:
    \begin{enumerate}
      \item to show that you know and understand the state of the art; and
      \item to put your work in context
    \end{enumerate}

    Ideally you can tackle both together by providing a critique of
    related work, and describing what is insufficient (and how you do
    better!)

    The related work chapter should usually come either near the front or
    near the back of the dissertation. The advantage of the former is that
    you get to build the argument for why your work is important before
    presenting your solution(s) in later chapters; the advantage of the
    latter is that don't have to forward reference to your solution too
    much. The correct choice will depend on what you're writing up, and
    your own personal preference.

\end{guidance}

### 2018-04-19 write-up/background.tex
Start background and add matrices to semantics

@@
 
-\begin{guidance}
-    A more extensive coverage of what's required to understand your
-    work. In general you should assume the reader has a good undergraduate
-    degree in computer science, but is not necessarily an expert in
-    the particular area you've been working on. Hence this chapter
-    may need to summarize some ``text book'' material.
+% \begin{guidance}
+%     A more extensive coverage of what's required to understand your
+%     work. In general you should assume the reader has a good undergraduate
+%     degree in computer science, but is not necessarily an expert in
+%     the particular area you've been working on. Hence this chapter
+%     may need to summarize some ``text book'' material.
+%
+%     This is not something you'd normally require in an academic paper,
+%     and it may not be appropriate for your particular circumstances.
+%     Indeed, in some cases it's possible to cover all of the ``background''
+%     material either in the introduction or at appropriate places in
+%     the rest of the dissertation.
+% \end{guidance}
 
-    This is not something you'd normally require in an academic paper,
-    and it may not be appropriate for your particular circumstances.
-    Indeed, in some cases it's possible to cover all of the ``background''
-    material either in the introduction or at appropriate places in
-    the rest of the dissertation.
-
-\end{guidance}
+\prechapter{%
+    In this chapter, I will outline the concept of linear types and show how
+    they can be used to solve the problems faced by programmers writing code
+    using linear-algebra libraries. I will be emphasising the \emph{practical}
+    and intuitive explanations of linear types to keep this thesis accessible
+    to working programmers as well as academics not familiar with type-theory;
+    giving only a terse overview of the history and theory behind linear types
+    for the interested reader to pursue further.
+}%
 
 \section{Tracking Resources with Linearity}
+Familiar examples of using a type-system to express program-invariants are
+existential-types for abstraction and encapsulation, polymorphic types for
+parametricity and composition (a.k.a generics). Less-known examples include
+dependent-types (contracts or pre- and post-conditions). The advantages of
+using a type-system to express program invariants are summarised by saying the
+stronger the rules you follow, the better the guarantees you can get about your
+program, \emph{before} you run it. At first, the rules seems restrictive, but
+similar to how the rules of grammar, spelling and more generally writing help a
+writer make it easier and clearer to communicate the ideas they wish to express,
+so to do rules about typing make it easier to communicate the intent and
+assumptions under which a program is written. An added, but often overlooked
+benefit is automated-checking: a programmer can boldly refactor in certain
+ways and the compiler will \emph{assist} in ensuring the relevant invariants
+the type-system enforces are updated and kept consistent by pointing-out
+where they are violated.
+
+Linear types are a way to help a programmer track and manage resources. In
+practical programming terms, they enforce the restriction that a value may be
+used exactly once.\footnote{This definition may differ from more colloquial
+uses in discussions surrounding \emph{substructural} type systems and/or Rust.
+A brief explanation can be found in Section~\ref{sec:further}.} While this
+restriction may seem limiting at first, precisely these constraints can be used
+to express common invariants of the programs written by working programmers
+every day. For example: a file or a socket, once opened \emph{must} closed; all
+memory that is manually allocated must be freed.  C++'s destructors and Rust's
+Drop-trait (and more generally, its borrow-checker) attempt to enforce these
+constraints by basically doing the same thing: any resource that has not been
+moved is deallocated at the end of the current lexical scope. Notably, these
+languages also permit aliasing, alongside rules enforcing when it is acceptable
+to do so.  On face value, the above one-line description of linear types
+prevents aliasing or functions such as $\lambda x.\;x \times x$, such features
+are still allowed (albeit in a more restricted fashion) in a \emph{usable}
+linear type system designed for working with linear-algebra libraries.
 
 \section{Problem in Detail}
+Given this background, the most pertinent question at hand is: what problems do
+linear-algebra library users (and writers) typically face? The answer to this
+question depends on which of two buckets a programmer falls (or is forced by
+domain) into. On one side, we have users of high-level linear-algebra libraries such
+as Owl (for OCaml), Julia and Numpy (for Python); other the other, we have
+users of more manual, lower-level libraries such as BLAS (Basic Linear Algebra
+Subroutines) and Eigen for languages like C++ and FORTRAN.\footnote{I am not including
+Rust in this comparison because its linear-algebra libraries are under active
+development and not as well-known/used. Later on, given that it is a language
+with in-built support of substructural features to track resources, Rust will
+be compared and constrasted with this project to evaluate the classic
+(E)DSL-versus-language-feature debate as it applies to the domain of
+linear-algebra libraries.}
+% https://github.com/AtheMathmo/rulinalg/blob/e0d8b76ddf8bbb5547cb9c98172e594605ab5fef/src/matrix/impl_ops.rs
+% Which operand is overwritten determines scope. Would make sense to argue
+% pros-cons of implicitness vs explicitness, emulating Rust's Affine with
+% linear sugar.
+
+\subsection{One Too Many Copies and a Thousand Bytes Behind}
+% CODE EXAMPLES
+
+\subsection{IHNIWTNM}
+The title of this subsection\footnote{I Have No Idea What Those Names Mean.}
+illustrates the problem with the C++/FORTRAN side: legibility (and ease of
+reasoning) is sacrified at the alter of performance and efficiency.
+
+% CODE EXAMPLES
 
 \section{Proposed Solution}
+A (E)DSL!
+
+% CODE EXAMPLES
+
+\section{Further Reading and Theory}\label{sec:further}
+\begin{itemize}
+    \item History - Girard and so on
+    \item Congratulations, it's a co-monad
+    \item Co-effects
+    \item General direction of field - Haskell, Rust, RAML, Atkey/McBride \& Idris, Krishnaswami
+\end{itemize}

### 2018-04-19 write-up/evaluation.tex
Start background and add matrices to semantics

@@
 \end{guidance}
+
+\prechapter{%
+    In this chapter, I will argue the central premise of this thesis: linear
+    types are a practical and usable tool to help working programmers
+    write code that is both more legible and less resource-hungry than with
+    existing linear-algebra frameworks. My project illustrates how to do so in
+    a way that can be implemented as a usable \emph{library} for existing
+    languages and frameworks that leverages the already impressive amount of
+    work gone into optimising them so far.
+}%
+
+\section{Against Overly-Safe Copying}
+% Numpy, Julia, Owl
+
+\section{Against Poor Legibility}
+% C++, Fortran

### 2018-04-19 write-up/implementation.tex
Start background and add matrices to semantics

@@
 
+\prechapter{%
+    I implemented this project in OCaml, however I strongly believe the ideas
+    described in this chapter can be applied easily to other languages and also
+    are modular enough to extend the OCaml implementation to output to different
+    back-end languages. I will show how a small core language with a few features
+    can be the target of heavy-desugaring of typical linear-algebra programs. This
+    core language can then be elaborated and checked for linearity before
+    performing some simple and predictable optimisations and emitting (in this
+    particular implementation) OCaml code that is not obviously safe, correct
+    and performant.
+}%
+
 \section{Core Language}
-
-Here, I will describe the syntax and semantics.
 

### 2018-05-08 write-up/background.tex
Start implementation, related work and conclusion

@@
 writer make it easier and clearer to communicate the ideas they wish to express,
-so to do rules about typing make it easier to communicate the intent and
+so too do typing rules make it easier to communicate the intent and
 assumptions under which a program is written. An added, but often overlooked
@@
 used exactly once.\footnote{This definition may differ from more colloquial
-uses in discussions surrounding \emph{substructural} type systems and/or Rust.
-A brief explanation can be found in Section~\ref{sec:further}.} While this
-restriction may seem limiting at first, precisely these constraints can be used
-to express common invariants of the programs written by working programmers
-every day. For example: a file or a socket, once opened \emph{must} closed; all
-memory that is manually allocated must be freed.  C++'s destructors and Rust's
-Drop-trait (and more generally, its borrow-checker) attempt to enforce these
-constraints by basically doing the same thing: any resource that has not been
-moved is deallocated at the end of the current lexical scope. Notably, these
-languages also permit aliasing, alongside rules enforcing when it is acceptable
-to do so.  On face value, the above one-line description of linear types
-prevents aliasing or functions such as $\lambda x.\;x \times x$, such features
-are still allowed (albeit in a more restricted fashion) in a \emph{usable}
-linear type system designed for working with linear-algebra libraries.
+uses in discussions surrounding \emph{substructural} type systems and/or
+Rust.} While this restriction may seem limiting at first, precisely these
+constraints can be used to express common invariants of the programs written by
+working programmers every day. For example: a file or a socket, once opened
+\emph{must} closed; all memory that is manually allocated \emph{must} be freed.
+C++'s destructors and Rust's Drop-trait (and more generally, its
+borrow-checker) attempt to enforce these constraints by basically doing the
+same thing: any resource that has not been moved is deallocated at the end of
+the current lexical scope. Notably, these languages also permit aliasing,
+alongside rules enforcing when it is acceptable to do so.  On face value, the
+above one-line description of linear types prevents aliasing or functions such
+as $\lambda x.\;x \times x$, such features are still allowed (albeit in a more
+restricted fashion) in a \emph{usable} linear type system designed for working
+with linear-algebra libraries.
 
@@
 users of more manual, lower-level libraries such as BLAS (Basic Linear Algebra
-Subroutines) and Eigen for languages like C++ and FORTRAN.\footnote{I am not including
+Subroutines) for languages like C++ and FORTRAN.\footnote{I am not including
 Rust in this comparison because its linear-algebra libraries are under active
@@
 with in-built support of substructural features to track resources, Rust will
-be compared and constrasted with this project to evaluate the classic
+be compared and contrasted with this project to evaluate the classic
 (E)DSL-versus-language-feature debate as it applies to the domain of
-linear-algebra libraries.}
+linear-algebra libraries.} Most of what follows applies to \emph{dense}
+linear-algebra computations rather than \emph{sparse} because memory
+allocated for results typically depends on the sparsity of the inputs and
+so is not immediately amenable to the techniques proposed in this thesis.
+
+\subsection{One Too Many Copies and a Thousand Bytes Behind}
+\begin{figure}[tbp]
+    \centering
+    \begin{minted}{python}
+# Numpy (Python)
+import numpy.matlib 
+a = [[1,0],[0,1]] 
+b = [[4,1],[2,2]] 
+c = numpy.matmul(a,b)
+    \end{minted}
+    \begin{minted}{julia}
+# Julia
+c = [1 0; 0 1] * [4 1; 2 2]
+    \end{minted}
+    \begin{minted}{ocaml}
+
+(* Owl (OCaml) *)
+open Owl
+let a = Mat.of_arrays [| [| 1.; 0. |]; [| 0.; 1. |] |]
+let b = Mat.of_array [| [| 4.; 1. |]; [| 1.; 2. |] |]
+let c = Mat.(a *@ b)
+    \end{minted}
+    \caption{Matrix Multiplication in Numpy (Python), Julia and Owl (OCaml).}\label{fig:mat_mul_copies}
+\end{figure}
+
+In Figure~\ref{fig:mat_mul_copies}, we see that matrix-multiplication is fairly
+trivial to write and execute in Numpy, Julia and Owl. Let us call this approach
+\emph{value-semantic}, meaning that objects are \emph{values} just like
+integers and floating-point numbers. This approach confers two key advantages to the
+programmmer: it is easy to read (equational and algebraic declarations) and it
+is easy to reason about (as one would with a mathematical formula). Although this
+approach does permit \emph{aliasing}, the consequences are benign because the
+result of any computation is a \emph{new} value, distinct from any used during
+the calculation of that value.
+
+\begin{figure}[tbp]
+    \centering
+    \begin{minted}{ocaml}
+let mul x y =
+  if same_shape x y then
+    let y = copy y in
+    (_owl_mul (kind x) (numel x) x y y; y)
+  else
+    broadcast_op (_owl_broadcast_mul (kind x)) x y
+    \end{minted}
+    \caption{Implementation of Matrix Multiplication in Owl (OCaml).
+        Note the `copy' for the result and the unsafe `\_owl\_mul' operation
+    used to perform an in-place multiplication.}\label{fig:mat_mul_owl}
+\end{figure}
+
+However, these advantages come with some costs: constantly producing new values
+is wasteful on memory (although the example given in
+Figure~\ref{fig:mat_mul_copies} is only a $2 \times 2$ matrix, many real-world
+datasets can contain up to gigabytes of data). A complex expression may create
+many short-lived temporaries which would need to be reclaimed by a
+garbage-collector (see Figure~\ref{fig:mat_mul_owl}). Libraries taking a
+\emph{value-semantic} approach offer a dichotomy for a user wishing to
+implement a new algorithm: either use the existing and safe primitives to build
+an easy to reason about but slower, more memory-intensive algorithm, or use
+escape-hatches (typically provided by most libraries, which permit in-place
+modification of objects) to build faster, and more efficient algorithms which
+are harder to reason about.
+
+\subsection{IHNIWTLM}
+The title of this subsection\footnote{I Have No Idea What Those Letters Mean.}
+illustrates the problem with the C++/FORTRAN side: legibility (and ease of
+reasoning) is sacrificed at the altar of performance and efficiency.
+
+Although escape-hatches do exist in value-semantic libraries, their use is
+discouraged. Systematic consideration of performance requires lowering the
+level of abstraction a programmer is working on. At this level, several
+factors such as memory layout, allocation, re-use as well as cache behaviour
+and parallelism become apparent. Of these, memory allocation and re-use are
+of most relevance to linear-types and this thesis.
+
+\begin{figure}[tbp]
+    \begin{minted}{fortran}
+      program blasMatMul
+      implicit none
+      real*4 a(2,2), b(2,2), c(2,2)
+C     External from BLAS
+      external dgemm
+C     Initialize in column major storage of Fortran
+      data a/ 1,0,
+     *        0,1/
+      data b/ 4,1,
+     *        1,2/
+C                tfm  tfm  rowA colB K  alpha a lda  b  ldb beta c  ldc
+      call dgemm('N', 'N', 2,   2,   2, 1.0,  a,  2, b, 2,  0.0, c, 2)
+    \end{minted}
+    \caption{One of \emph{several} BLAS (Fortran) routines for Matrix
+    Multiplication.}\label{fig:fortran_blas}
+\end{figure}
+
+In Fortran (Figure~\ref{fig:fortran_blas}), data is typically allocated
+statically (at compile time) so temporary storage for all intermediate values
+must be managed by the programmer. While this approach leads to verbose and
+less readable code, the explicitness is good for understanding the memory
+concerns of the program, albeit at the expense of understanding what the
+program is actually calculating.
+
+On the other hand, C++ (with operator overloading) can end up looking fairly
+readable. For safety and correctness, expressions are typically handled with
+value-semantics. However, given \emph{extra} information about, aliasing
+(Eigen, Figure~\ref{fig:cpp_eigen}) or usage of intermediate expressions
+(uBLAS, Figure~\ref{fig:cpp_ublas}), the number of temporaries allocated can be
+reduced and increased \emph{implicitly} to improve performance (remove
+unnecessary allocations or re-calculations respectively). Further tricks to
+improve performance include expression templates (building up an
+expression-tree at compile time and then pattern-matching on it to produce
+code) and lazy evaluation (only calculating a result when it is needed). These
+will be discussed in more detail in Chapter~\ref{chap:related}.
+
+\begin{figure}[tbp]
+    \begin{minted}{c++}
+#include <iostream>
+#include <Eigen/Dense>
+using namespace std;
+int main()
+{
+    Eigen::Matrix2d a,b,c;
+    a << 1, 0,
+         0, 1;
+    b << 4, 1,
+         1, 2;
+    c << 0, 0,
+         0, 0;
+    a * b; // new matrix
+    c += a * b; // temporary for correctness in case of aliasing
+    c.noalias() += a * b; // no temporaries
+}
+    \end{minted}
+    \caption{Some examples of Matrix Multiplication in Eigen. Using expression
+        templates (to be discussed later) and \emph{explicitly provided} aliasing
+        information, Eigen can emit a single BLAS `dgemm'-like call for the last
+        line, mirroring the Fortran example of
+        Figure~\ref{fig:fortran_blas}.}\label{fig:cpp_eigen}
+\end{figure}
+
+\begin{figure}[tbp]
+    \begin{minted}{c++}
+noalias(C) = prod(A, B);
+// Preferable if T is preallocated
+temp_type T = prod(B,C); R = prod(A,T);
+prod(A, temp_type(prod(B,C));
+prod(A, prod<temp_type>(B,C));
+    \end{minted}
+    \caption{Boost uBLAS example of Matrix Multiplication. Temporaries need to
+        be marked as such to prevent unnecessary re-computation of
+        values.}\label{fig:cpp_ublas}
+\end{figure}
+
+\section{TO DO: Proposed Solution}
+A (E)DSL!
 % https://github.com/AtheMathmo/rulinalg/blob/e0d8b76ddf8bbb5547cb9c98172e594605ab5fef/src/matrix/impl_ops.rs
@@
 
-\subsection{One Too Many Copies and a Thousand Bytes Behind}
-% CODE EXAMPLES
-
-\subsection{IHNIWTNM}
-The title of this subsection\footnote{I Have No Idea What Those Names Mean.}
-illustrates the problem with the C++/FORTRAN side: legibility (and ease of
-reasoning) is sacrified at the alter of performance and efficiency.
-
-% CODE EXAMPLES
-
-\section{Proposed Solution}
-A (E)DSL!
-
 % CODE EXAMPLES
@@
 \section{Further Reading and Theory}\label{sec:further}
-\begin{itemize}
-    \item History - Girard and so on
-    \item Congratulations, it's a co-monad
-    \item Co-effects
-    \item General direction of field - Haskell, Rust, RAML, Atkey/McBride \& Idris, Krishnaswami
-\end{itemize}
+
+% Just as no Scottish MP's initial speech in the House of Commons would be
+% complete without some claim or mention to Rabbie Burns,
+
+No exposition of linear types would be complete without a mention of Girard's
+Linear Logic~\cite{girard}. As mentioned in the Stanford Encyclopedia of
+Philosophy, it is ``a refinement of classical and intuitionistic logic. Instead
+of emphasizing truth, as in classical logic, or proof, as in intuitionistic
+logic, linear logic emphasizes the role of formulas as resources.'' A walk from
+logic to programming along the well-trodden Curry-Howard bridge brings us to
+linear types.
+
+For the category theory inclined reader, the !-operator (sometimes, for reasons
+elided here, called \emph{exponentiation}) forms a co-monad; for the rest of
+us, this entails two (rather simple) facts about a value you can use any number
+of times: you can (1) can use it once (co-unit), and (2) can pass it to many
+contexts that will use it many times (co-multiply).
+
+More generally, by annotating variables in the context with their usage (when
+implementing a type-checker for a linearly typed language), we can express the
+rules of \emph{substructural} (including affine, relevant and ordered type
+systems) under the more general framework of \emph{co-effects}~\cite{petricek}.
+
+Stepping further back, both the practice and theory behind resource-aware
+programming has made visible progress in the past few years.  On the
+programming side, we have Linear Haskell, Rust and Idris (experimental). On the
+research side, we have Resource Aware ML~\cite{hoffmann} and the tantalising
+promise of integrating linear and dependent types~\cite{atkey}.

### 2018-05-08 write-up/conclusion.tex
Start implementation, related work and conclusion

@@
-\chapter{Conclusion}
+\chapter{TO DO: Conclusion}
 
@@
 \end{guidance}
+
+\textbf{What are the benefits of the type-based approach I have taken? Have I successfully 
+argued whatever I wrote in the introduction?}
+
+\section{Future Work}
+
+\textbf{Graph stuff... check with supervisors.}
+
+All of the matrix expression compilers mentioned in previous section construct
+some sort of dataflow graph to represent the computation being executed. While
+this seems intuitive, there is no formal argument for this  appr
+Some directions in which a type-based approach to efficient matrix expression compilation could
+be taken are:
+
+\begin{itemize}
+
+    \item \textbf{As a typed IR} for matrix expression compilers. This in turn
+        could enable
+        \begin{itemize}
+            \item existing matrix expression compilers to be less opaque about
+                what resources they are consuming.
+            \item open up opportunities for non-local sharing of temporary values
+                with some intra-procedural analysis.
+            \item allow the user to choose: use a matrix expression compiler when
+                desired and drop down to a usable typed-IR for finer control,
+                whilst still retaining safety guarantees.
+        \end{itemize}
+
+    \item \textbf{Formal verification of matrix expression compilers} by
+        precisely specifying source and target languages.
+
+    \item \textbf{Multi-stage programming} to use information only available
+        at runtime in many situations (such as sizes, matrix properties,
+        sparsity, control flow) can be effectively incorporated into code
+        generated.
+
+    \item \textbf{Dependent types} to have control over how resources can be
+        used and split. In addition to formal verification, dependent types
+        could be combined with linearity to express finer-grain conventions
+        surrounding blocking, slicing and writing to \emph{parts} of the matrix
+        instead of the whole. This is already prevalent with `dsymm' like BLAS
+        routines which only read and write to the lower or upper triangle of a
+        matrix.  This idea is inspired by Conor McBride's talk on writing to
+        terminals with ``Space Monads''~\cite{space_monads}.
+
+    \item \textbf{Compiling to hardware} is also an option - once we know
+        exactly when and where temporaries are required and what can be re-used
+        when, we come one (small, but useful) step closer to realising matrix
+        expressions directly on hardware.
+
+\end{itemize}

### 2018-05-08 write-up/implementation.tex

--- 2018-02-05 write-up/implementation.tex
@@
 
+\prechapter{%
+    I implemented LT4LA in OCaml, however I strongly believe the ideas
+    described in this chapter can be applied easily to other languages and also
+    are modular enough to extend the OCaml implementation to output to different
+    back-end languages. I will show how a small core language with a few features
+    can be the target of heavy-desugaring of typical linear-algebra programs. This
+    core language can then be elaborated and checked for linearity before
+    performing some simple and predictable optimisations and emitting (in this
+    particular implementation) OCaml code that is not obviously safe and correct
+    (with respect to linearity).
+}%
+
+\section{Structure of LT4LA}
+
+LT4LA follows the structure of a typical compiler for a (E)DSL\@. From the
+start, I made a concerted effort to (1) write pure-functional code (typically
+using a monadic-style) which helped immensely with modularity and debugging
+when tests showed errors (2) produce readable, useful and precise
+error-messages in the hope that someone who did not understand linear types
+could still use the LT4LA (3) write tests and set-up
+continuous-integration for all non-trivial functions so that I could spot and
+correct errors that were not caught by OCaml's type-system whenever I
+implemented new features or or refactored my code.
+
+\begin{enumerate}
+
+    \item \textbf{Parsing \& Desugaring}. A generated, LR(1) parser parses a text
+        file into a syntax tree, which is then desugared into a smaller, more
+        concise abstract syntax tree. The former aims to mimic OCaml syntax
+        with a few extensions and keywords so that it is familiar and thus easy
+        to pick-up for OCaml users. The latter allows for the type-checker to
+        be simpler to implement and easier to specify. In general, this part
+        will vary for different languages or can be dealt with differently
+        using combinators (the EDSL approach) or a syntax-extension if the host
+        language offers such support.
+
+    \item \textbf{Type-checking}. The abstract syntax tree is explicitly-typed,
+        with some inference to make it less verbose and more convenient to
+        write typical programs.
+
+    \item \textbf{Matrix Expressions}. During type-checking, if a matrix-expression
+        is encountered, it is either successfully elaborated into an expression
+        in the abstract syntax tree which is then consequently type-checked, or
+        fails to find suitable routines to calculate the given expression.
+
+    \item \textbf{Code Generation}. The abstract syntax tree is translated into
+        standard OCaml and a few-particular `optimisations' are made to produce
+        more readable code. This process is type-preserving: the linear type
+        system is embedded into OCaml's type system, and so when the OCaml compiler
+        compiles the generated code, it acts as a sanity check on the code produced.
+
+    \item \textbf{Executable Artifacts}. A transpiler and a REPL are the main artifacts
+        produced for this thesis.
+
+    \item \textbf{Tests}. As mentioned before, almost all non-trivial functions
+        have tests to check their behaviour. The output of the transpiler
+        was also tested by having the build system generate code at OCaml code
+        which could then be compiled and tested as usual.
+
+\end{enumerate}
+
 \section{Core Language}
 
-Here, I will describe the syntax and semantics.
+A full description of the core language can be found in
+Appendix~\ref{chap:ott_spec}. To make a linearly-typed language usable, we need
+some way of using values zero or more than once, as we would an intuitionistic
+value. For this, we have in the type-expressions the !-constructor and in the
+term-expressions the \textbf{many}-constructor. For soundness, I implemented
+\emph{value-restriction} for this type system, only \emph{values} (anything not
+an array, matrix or an expression that can be reduced) that use \emph{no linear
+values} can be marked with !. This means we can freely have integers, booleans
+and (appropriate) functions usable as normal. Recursive functions can only use
+linear values that are passed in as arguments to the function.
 
-\section{Elaboration and Inference}
+% TODO Once I've implemented the <- syntax, this should just include
+% ../test/examples/weighted_avg_infer.lt directly
+\begin{figure}[tb]
+    \begin{minted}[linenos]{ocaml}
+let rec f (!i : !int)
+        (!n : !int)
+        (!x0 : !elt)
+        (write : z arr)
+        ('x)
+        (weights : 'x arr)
+        : z arr * 'x arr =
+  if n = i then (write, weights) else
+  let !w0 <- weights[0] in
+  let !w1 <- weights[1] in
+  let !w2 <- weights[2] in
+  let !x1 <- write[i] in
+  let !x2 <- write[i + 1] in
+  let !x : !elt = w0 *. x0 +. w1 *. x1 +. w2 *. x2 in
+  f (i + 1) n x1 (write[i] := x) _ weights
+    \end{minted}
 
-\section{Compiling and Metaprogramming}
+    \caption{A recursive function to perform a \emph{in-place} (simplified)
+        1D-convolution (sliding weighted-average) using a read-only array of
+        weights and a writeable array, originally containing the input
+        array.}\label{fig:oned_conv}
+
+\end{figure}
+
+So, how does a typical programmer write code that uses a matrix or an array
+more than once? Figure~\ref{fig:oned_conv} shows an example program that
+executes a (simplified) 1D-convolution (sliding weighted-average) using a
+read-only array of three weights. On line 1, \ltfla{let rec} introduces a
+recursive function `f', followed by explicitly annotated formal parameters.
+Intuitively, \ltfla{!id} allows \ltfla{id} to be used intuitionistically. The
+types \ltfla{!int} and \ltfla{!elt} correspond to integers and element types
+(for example, 64-bit floating-point).
+
+Array and matrix types are parameterised by \emph{fractional-capabilities}.  A
+fraction of 1 ($2^0$) represents complete ownership of a value; in
+particular, this allows a programmer to write or free it. Creating an array
+gives you ownership of it; the function \ltfla{array : !int --o z arr}
+(where \ltfla{z} represents `0'). Once you have ownership of an
+array/matrix, you can free it: \ltfla{free : z arr --o unit}.
+Importantly, because a linear-value may only be used once, the array just freed
+is \emph{out of scope} for following expressions, preventing use-after-free.
+Ownership also enables you to write to the array: \ltfla{set : z arr
+--o !int --o !elt --o z arr} (the syntax \ltfla{w[i] := j} is just
+sugar for \ltfla{set w i j}). Here, linearity prevents accessing values
+which represented the array \emph{before} the mutation.
+
+Any fraction less than 1 (for simplicity, limited to $2^{-k}$ in this system,
+for a positive integer $k$) represents read-only access. So, the \ltfla{'x}
+represents a natural number (either a zero \ltfla{z}, variable \ltfla{'x} or a
+successor (+1) of a natural number). Hence, you can read from (index) any array
+\ltfla{get : 'x .  'x arr --o !int --o !elt * 'x arr} (the syntax \ltfla{let !v
+<- w[i]} is just sugar for \ltfla{let (!v, w) = get _ w i}). Transparently
+rebinding \ltfla{w} with the returned value means a program can appear to use
+\ltfla{w} multiple times; the underscore is how a programmer tells the compiler
+to automatically \emph{infer} the correct fractional-capability based on the
+other arguments passed to the function. Fractions also permit aliasing:
+\ltfla{share : 'x . 'x arr --o 'x s arr * 'x s arr}. However, given the types
+of the primitives provided, these aliases can only be read from and not written
+to. If you want to write to this array, you must \ltfla{unshare: 'x . 'x s arr
+--o 'x s arr --o 'x arr} it until you are left with a value of type \ltfla{z
+arr}, guaranteeing no other aliases exists.
+
+Given this set-up, we now \emph{statically} have \emph{perfect} information
+about aliasing and ownership of values in the program. We can only write to an
+array when we own it; ownership guarantees no other aliases exist in scope at
+the point of usage. In the next section, I show how this perfect information
+can be used to write more natural-looking code using value-semantic expressions
+which behave in precisely the way we intend it to. Now the programmer need not
+resort to manually tracking and inserting \texttt{noalias} annotations; instead
+they can let the loyal and tireless compiler do the heavy lifting.
+
+\section{Matrix Expressions}
+
+We have now arrived at the titular \emph{applications} of linear types. I will
+show how, in addition to preventing the errors explained in the previous
+section, we can take linear types one step further and apply it to the domain
+of efficient compilation of matrix-expressions.
+
+\begin{figure}[tb]
+    \inputminted[linenos]{fortran}{kalman.f90}
+    \caption{Kalman filter in Fortran 90 (Credit: Matthew
+        Rocklin~\cite{rocklin_gist}).}\label{fig:fortran_kalman}
+\end{figure}
+
+In Figure~\ref{fig:fortran_kalman}, we see the difficulty of efficiently
+implementing a \emph{Kalman filter}, a powerful and set of equations applicable
+to a wide variety of problems. From the comments, we see that every variable is
+annotated with the step/matrix expression that it will hold at some point
+during the computation (an equivalent alternative, say in C++, could be to have
+a meaningful name for each step/matrix expression and manually annotate/keep
+track of which names alias the same location).
+
+In contrast, Figure~\ref{fig:ltfla_kalman}, offers the advantages of
+\begin{itemize}
+    \item aliasing: labelling each step with a different, more meaningful variable name,
+    \item easily spotting which resources are being passed in and which are
+        allocated for the function (new/copy),
+    \item unambiguously seeing \emph{when} and what values are freed;
+\end{itemize}
+and have the compiler automatically ensure the safety of each of the above by respectively
+\begin{itemize}
+    \item making it impossible to refer to steps/values which are no longer usable,
+    \item ensuring all values are declared and \emph{initialised} correctly before they are used,
+    \item checking no values are used after they are freed \emph{and} ensuring no values are leaked.
+\end{itemize}
+
+Indeed, an inexperienced programmer could take the  na\"ive approach of just
+copying sub-expressions by default and then letting the compiler tell it which
+copies are never used and removing them systematically until it type checks.
+While it is not quite a black-box, push-button compilation of an expression, I
+would argue that, it is just as easy (if not easier) to become familiar with as
+Rust and its borrow-checker.
+
+\begin{figure}[tb]
+
+    \begin{minted}[linenos]{ocaml}
+let !kalman
+        ('s) (sigma : 's mat) (* n, n *)
+        ('h) (h : 'h mat)     (* k, n *)
+        ('m) (mu : 'm arr)    (* n    *)
+        (r_1 : z mat)         (* k, k *)
+        (data_1 : z arr)      (* k    *)
+        : ('s mat * ('h mat * 'm arr)) * (z arr * z mat) =
+    let ((!k, !n), h) = dimM _ h in
+    let zeros = zeroM n n in
+    let sigma_h = new [| sym (up, sigma) * h |] in
+    let r_2 = [| h * sigma_h + r_1 |] in
+    let mu_1 <- copy mu in
+    let sigma_1 <- copyM sigma in
+    let data_2 = [| h * mu + -1. * data_1 |] in
+    let h_1 <- copyM h in
+    let h_sol = [| sym (up, r) ^ -1 * h_1 ] in
+    let data_sol = [| sym (up, r) ^ -1 * data_2 |] in
+    let h_h1 = new [| h * h_sol |] in
+    let h_data = new [| h * data |] in
+    let h_h1_sigma = new [| sym (up, h_h1) * sigma |] in
+    let mu_2 = [| sym (up, sigma) * h_data * mu_1 |] in
+    (* non-linear use of sigma_c..? *)
+    let sigma_c = [|  -1. * sym (up, sigma_c) * h_h1_sigma + sigma_c |] in
+    (* frees... *)
+    (mu_c , sigma_c)
+    \end{minted}
+
+    \caption{Kalman filter in LT4LA.}\label{fig:ltfla_kalman}
+
+\end{figure}
+
+\subsection{TO DO: Elaboration}
+
+So-called `matrix \emph{expressions}' are still `side-effecting'/consuming
+linear values, and do not produce \emph{values} but a \emph{sequence of
+(re-)bindings} which in turn dictate what values are still available/in-scope
+after the computation.  As such, compilation cannot be done purely
+compositionally via structural-induction on the expression language; it is most
+concisely expressed via CPS, where the result of elaborating a matrix
+expression is a function that takes as its argument the rest of the computation
+expecting to use the result it has just made available under that requested
+name.
+
+\subsection{TO DO: Inference}
+
+\section{TO DO: Code Generation}
+
+High-level overview: primitives, translations, optimisations, difficulty, build system.
+
+\subsection{TO DO: Compiling Constructs to OCaml}
+
+\subsection{TO DO: Metaprogramming}
+
+\section{TO DO: Theory}
+
+Fractional permissions, not proven soundness etc\. but compilation preserves
+types as checked by OCaml.

### 2018-05-08 write-up/introduction.tex
Start background and add matrices to semantics

@@
-\chapter{Introduction}
+\chapter{TO DO: Introduction}
 
@@
 
+\prechapter{%
+    In this thesis, I will argue that linear types are an appropriate,
+    \emph{type-based formalism} for the problem of \emph{efficient}
+    matrix-expression compilation. I will show that framing the problem using
+    linear types can help \emph{reduce bugs} by making precise and explicit the
+    informal, ad-hoc practices typically employed by human experts and linear
+    algebra \emph{compilers} and automate checking them. As evidence for this
+    argument, I will show programs written with this safety, precision and
+    explicitness (1) can be just as pleasant and convenient for a programmer as
+    less efficient, but higher-level linear algebra libraries and (2) perform
+    just as \emph{efficiently and predictably} as lower-level, less readable
+    and more error-prone linear algebra libraries.
+}%
+
 \section{Overview of Problem}

### 2018-05-08 write-up/ott_spec.tex

\chapter{Ott Specification}\label{chap:ott_spec}

The following pages present a specification of the grammar and type-system used
by my project, produced using the Ott\cite{sewell_ott} tool. It is important to note
that the type-system described here is not how it is implemented: it is easier
and clearer to describe the system as below for explaining. However, for implementing,
I found it much more and user- and debugging-friendly to:

\begin{itemize}

    \item Implement it so that the type-environment \emph{changes} as a result
        of type-checking an expression; with this, the below semantics describe
        the \emph{difference} between the environment after and before checking
        an expression.

    \item \emph{Mark} variables as used instead of \emph{removing} them from the
        environment for better error messages.

    \item Have \emph{one} environment where variables were \emph{tagged} as
        linear and un-used, linear and used, and intuitionistic. This was
        definitely an implementation convenience so that variable binding could
        be handled uniformly for linear and intuitionistic variables and
        scoping/variable look-up could be handled implicitly thanks to the
        associative-list structure of the environment.

\end{itemize}

\clearpage%
\ottall%

### 2018-05-08 write-up/related_work.tex
Start background and add matrices to semantics

@@
-\chapter{Related Work}
+\chapter{Related Work}\label{chap:related}
 
-\begin{guidance}
+% \begin{guidance}
+% 
+%     This chapter covers relevant (and typically, recent) research
+%     which you build upon (or improve upon). There are two complementary
+%     goals for this chapter:
+%     \begin{enumerate}
+%       \item to show that you know and understand the state of the art; and
+%       \item to put your work in context
+%     \end{enumerate}
+% 
+%     Ideally you can tackle both together by providing a critique of
+%     related work, and describing what is insufficient (and how you do
+%     better!)
+% 
+%     The related work chapter should usually come either near the front or
+%     near the back of the dissertation. The advantage of the former is that
+%     you get to build the argument for why your work is important before
+%     presenting your solution(s) in later chapters; the advantage of the
+%     latter is that don't have to forward reference to your solution too
+%     much. The correct choice will depend on what you're writing up, and
+%     your own personal preference.
+% 
+% \end{guidance}
 
-    This chapter covers relevant (and typically, recent) research
-    which you build upon (or improve upon). There are two complementary
-    goals for this chapter:
-    \begin{enumerate}
-      \item to show that you know and understand the state of the art; and
-      \item to put your work in context
-    \end{enumerate}
+\prechapter{%
+    Now that I have described my contributions, I will explain how it relates
+    to existing work, leaving brief discussions on future work to the next
+    chapter. I strongly believe matrix expression compilation research would
+    benefit greatly from a comprehensive literature review but unfortunately
+    that is beyond the scope of this chapter.
+}%
 
-    Ideally you can tackle both together by providing a critique of
-    related work, and describing what is insufficient (and how you do
-    better!)
+\section{Matrix Expression Compilation}
 
-    The related work chapter should usually come either near the front or
-    near the back of the dissertation. The advantage of the former is that
-    you get to build the argument for why your work is important before
-    presenting your solution(s) in later chapters; the advantage of the
-    latter is that don't have to forward reference to your solution too
-    much. The correct choice will depend on what you're writing up, and
-    your own personal preference.
+% For decades, small gains in efficiency of \emph{executing} numerically
+% intensive code have and continue to translate into massive savings in costs and
+% time. This has spurned mountains of research into an almost bewildering variety
+% of hardware, software and compilation techniques, typically entrusted with a
+% (team of) domain-specific expert programmer(s) to apply in real world
+% situations.
+% 
+% Given the rise of big data and machine learning, there is now a movement to
+% improve \emph{programmer} efficiency in this domain after focusing for so long
+% on \emph{computational} efficiency. Matrix expression compilers promise to
+% generate code that is reasonably close to `optimal'/what a domain-expert may
+% write but without the need to acquire years of experience and depth of
+% knowledge (a similar trade-off to garbage-collection for memory management or
+% software-transactional memory for concurrency).
 
-\end{guidance}
+Most of the projects below try to be fully-automated black-boxes which model
+computing a matrix expression as some sort of graph with informal, ad-hoc rules
+about what can and should be copied or modified in-place. Allocations,
+temporaries and common sub-expressions are details invisible to the programmer,
+left to the compiler.
+
+The matrix expression `compiler' as implemented in LT4LA is intended to be a
+mere proof-of-concept of how linear types can and arguably should be used
+\emph{ergonomically}. I have taken the approach of attempting to help
+programmers precisely and explicitly capture, using types, the practices
+prevalent in code they already write.
+
+I believe the advantages of my approach are two-fold (1) \emph{predictable}
+performance and (2) more accurate modelling of how low-level kernels handle
+their resources. My confidence in the latter claim comes from finding at least
+two errors in the Fortran code output by SymPy to compute a Kalman filter, one
+to do with aliasing (resulting in code that was not legal Fortran) that would
+not have type-checked had it been translated via LT4LA as an intermediate
+representation (the other error to be explained later
+in~\ref{subsec:lin_dep_types}).
+
+\subsection{SymPy}
+
+SymPy is a symbolic computer algebra system for Python; its matrix expression
+compiler~\cite{rocklin_thesis} uses a term-rewrite system, with rules supplied
+by a BLAS expert (which must be strongly-normalising, that is, never cause a
+loop) but need not be confluent (there can be more than one solution per
+expression). Rules include expressions to match on, the expression it can
+produce, information about the expressions (such as whether the matrix is
+symmetric or full-rank) and information about which variable is updated
+in-place.
+
+\subsection{Clak and Cl1ck}
+
+Clak and Cl1ck~\cite{fabregat_thesis} were developed independently around the
+same time as SymPy's matrix expression compilation. Clak attempts to produce
+\emph{multiple} algorithms for a single matrix expression, by considering a
+wider matrix expression grammar and more matrix properties and inference rules.
+These algorithms assume basic building blocks such as products and
+factorisations. Cl1ck attempts to take on the challenge of writing BLAS/LAPACK
+like libraries too, by generating lower-level loop-based blocked routines for
+the aforementioned basic building blocks, in the spirit of the
+FLAME~\cite{gunnels_flame} project.
+
+\subsection{Linnea and Taco}
+
+Linnea~\cite{linnea} and Taco~\cite{taco} are two newer contenders to Clak and
+Cl1ck respectively. Linnea continues the work of Clak to producing real
+executable code for \emph{existing} libraries and kernels, as well as
+incorporating work on a \emph{generalised} matrix chain algorithm~\cite{gmc}.
+Taco (\emph{Tensor Algebra COmpiler}) focuses on emitting efficient routines
+for expressions in tensor index notation, with many optimisations for
+\emph{sparse} tensors.
+
+\section{Metaprogramming}
+
+Most of the compilers in the aforementioned projects usually built, analysed,
+compiled, ran (and in some cases, dynamically loaded) expressions (including
+functions) at runtime, similar to how regular expressions are handled in most
+languages -- in particular, even when the regular expression is known at
+compile time.
+
+In LT4LA, I took the approach of having a concrete syntax and expression
+language which was then translated and made available as a \emph{typed
+expression} to other modules \emph{at compile time} via the build system. Apart
+from convenience in programming and testing, there was nothing inherent in the
+approach that prevented me from using OCaml's PPX syntax-extensions so that I
+could write normal OCaml expressions from within OCaml and have them checked
+for linearity before compilation.
+
+Having a statically compiled language and a build system as so affords the
+advantage of eliminating the runtime overheads mentioned at the start of this
+section. However, there is some useful information (such as matrix dimensions
+for the matrix chain algorithm) which is sometimes known only at runtime (but
+once known, usually fixed). In these cases, using \emph{multi-stage
+programming} would be a better approach to implementing a matrix expression
+compiler.
+
+\subsection{MetaOCaml and Scala LMS}
+
+MetaOCaml~\cite{metaocaml} and Scala with Lightweight Modular
+Staging~\cite{scala_lms} are systems which support multi-stage programming. A
+typical example of this is the generation of Fast Fourier-Transform kernels,
+specialised to a desired array length.
+
+\subsection{Expression Templates in C++ Libraries}
+
+Expression templates are a commonly used compile-time metaprogramming
+technique, used by Eigen, uBLAS and Armadillo to name a few. If known at
+compile-time, matrix dimensions can also be passed in as template arguments to
+ensure operations match (otherwise checking at runtime). In Eigen, such
+features are combined with heuristics to enable \emph{lazy evaluation} and
+automatically determine whether a sub-expression is evaluated into a temporary
+variable or not.
+
+They perform rudimentary pattern-matching and in some cases, loop-fusion, to
+avoid evaluating expressions in a purely binary manner (invoking the bane of a
+C++ programmer: temporaries and unnecessary copies) when possible (either by
+translating to a library kernel call or, as an example, inlining a $v := a + b
++ c$ vector expression into one loop).
+
+As is the case with LT4LA, this approach shares \emph{some} elimination of
+runtime overheads, but not all, thanks to the heuristics surrounding lazy
+evaluation and evaluating sub-expressions into temporaries. This comes at the
+cost of a user being able to easily inspect the generated C++ code, losing
+explicitness.
+
+\section{Types}
+
+Apart from lazy evaluation, the following projects show how instead of a (E)DSL
+library approach, we could have type-level resource management provided far
+more conveniently and naturally at the \emph{language} level. The difference is
+that a library can be designed, shipped and used now whereas language features
+take time and can have unintended interactions with other language features. My
+hope is that once people are convinced of the utility type-level resource
+management by using a library, the impetus for integrating such features into
+the language follows.
+
+\subsection{Lazy Evaluation}
+
+A particularly strong advantage LT4LA has over other libraries that use lazy
+evaluation is, funnily enough, linear types, more precisely and the static and
+perfect information they guarantee about aliasing: with Owl, every graph-node
+has only one node in or out; with Eigen, every ``assignment'' has a
+\texttt{noalias} annotation.
+
+This simplifies the rules and exceptions a programmer reasoning about memory
+usage needs to remember. Of course, now the programmer has to figure out how
+they are using their temporaries, but because matrices are linearly-typed,
+redundant copies/missed frees can be pointed out by the compiler, guiding them
+towards a satisfying solution.
+
+\subsection{Futhark}
+
+Futhark~\cite{futhark} is a second-order (meaning it supports functions such as
+map and fold/reduce) array combinator (meaning array operations can be fused
+into streams to reduce temporaries) language designed for efficient parallel
+compilation. It supports ML-style modules, loops, limited parametric
+polymorphism, size types and uniqueness types. Intuitively, where linearity
+provides a local guarantee of ``this value is not aliased in this scope'',
+uniqueness types provide a global guarantee of ``this value is not aliased
+anywhere''. Its combinators are more expressive than typical linear algebra
+library kernels so encourages shorter, more declarative linear algebra code.
+
+\subsection{Substructural Features in Rust}
+
+Rust~\cite{rust} is a (relatively) new systems programming language aiming to
+bring the last two decades of programming language research to the masses in a
+usable and friendly manner. Its \emph{borrow-checker} is the feature most
+relevant to this thesis because it statically attempts to prevent many
+resource-related bugs. Although there are a few linear algebra libraries for
+Rust under development, careful use of its macro system and borrow-checker
+could make it the safest and easiest to use language for linear algebra
+projects to come. Its struggle is more likely to be against the inertia of the
+large amounts of C++/Fortran code already out there rather than its usability
+or benefits.
+
+\subsection{Linear Types in Haskell}
+
+Linear types have been incorporated into a branch of the Glasgow Haskell
+Compiler~\cite{retrofitting} in an attempt to provide safe, functional
+streaming and IO (after people saw the potential from libraries providing
+linearity features). Practical benefits include zero-copy buffers and
+eliminating garbage collection in certain situations by allowing the user to
+safely manage memory. The fact that it can and has been done gives me hope that
+other languages will also see the value and adopt some form of
+resource-management in their type systems.
+
+\subsection{Linear and Dependent Types in Idris}\label{subsec:lin_dep_types}
+
+Dimension mismatches are seen as an irritating but small inconvenience when
+writing linear algebra code. However, the second error I found in the Fortran
+code output by SymPy to compute a Kalman filter was a dimension/transposition
+error. Although we would not need full dependent types to solve dimension
+mismatches (symbolic size types would be sufficient), managing properties about
+matrices could be done at the type-level in a dependently typed setting.
+
+We could then express the usual properties and results of operations at the
+type-level, ensuring, for example, that certain functions are called only when
+the matrix is symmetric and can be written to. Idris (a Haskell inspired
+language with dependent types) has had experimental support for uniqueness
+types since its early days and now a linear types extension~\cite{idris_linear}
+is also being worked on based on new research around integrating linear and
+dependent types~\cite{atkey}.

### 2018-05-19 bin/kalman.h
WIP: Pure-functional benchmark and data processing

@@ -1,5 +1,5 @@
-#ifndef LT4LA_MEASURE_KALMAN
-#define LT4LA_MEASURE_KALMAN
+#ifndef LT4LA_KALMAN
+#define LT4LA_KALMAN
 
 #include <sys/time.h>
 #include <sys/resource.h>
@@ -51,15 +51,15 @@ double data[K][1] = {
 };
 
 static void kalman(
-    int n,
-    int k,
-    double *sigma,     /* n,n */
-    double *h,         /* k,n */
-    double *mu,        /* n,1 */
-    double *r,         /* k,k */
-    double *data,      /* k,1 */
-    double **ret_mu,   /* k,1 */
-    double **ret_sigma /* n,n */
+    const int n,
+    const int k,
+    const double *sigma, /* n,n */
+    const double *h,     /* k,n */
+    const double *mu,    /* n,1 */
+    double *r,           /* k,k */
+    double *data,        /* k,1 */
+    double **ret_mu,     /* k,1 */
+    double **ret_sigma   /* n,n */
 ) {
         double* k_by_n = (double *) malloc(k * n * sizeof(double));
 /*20*/  cblas_dsymm(CblasRowMajor, CblasRight, CblasUpper, k, n, 1., sigma, n, h, n, 0., k_by_n, n);
@@ -89,34 +89,87 @@ static void kalman(
         *ret_mu = new_mu;
 }
 
-double measure_kalman(int arg) {
+static double measure_kalman_no_free(
+    const int n,
+    const int k,
+    const double *sigma, /* n,n */
+    const double *h,     /* k,n */
+    const double *mu,    /* n,1 */
+    double *r,           /* k,k */
+    double *data         /* k,1 */
+) {
     struct rusage usage;
 
     getrusage(RUSAGE_SELF, &usage);
     struct timeval start = usage.ru_utime;
 
-    kalman(N, K, &sigma[0][0], &h[0][0], &mu[0][0], &r[0][0], &data[0][0], &new_mu, &new_sigma);
+    kalman(n, k, sigma, h, mu, r, data, &new_mu, &new_sigma);
 
     getrusage(RUSAGE_SELF, &usage);
     struct timeval end = usage.ru_utime;
 
+    // Execution time would have to reach about 300 years before (* 1000000) will cause problems
+    return (double) ((end.tv_sec - start.tv_sec) * 1000000 + (end.tv_usec - start.tv_usec));
+}
+
+double measure_kalman(
+    const int n,
+    const int k,
[…]

### 2018-05-21 write-up/background.tex
Finish Background, continue Implementation

@@
 open Owl
-let a = Mat.of_arrays [| [| 1.; 0. |]; [| 0.; 1. |] |]
-let b = Mat.of_array [| [| 4.; 1. |]; [| 1.; 2. |] |]
+let a = Mat.of_arrays [|[|1.;0.|];[|0.;1.|]|]
+let b = Mat.of_array [|[|4.;1.|];[|1.;2.|]|]
 let c = Mat.(a *@ b)
@@
 \begin{figure}[tbp]
-    \begin{minted}{fortran}
+    \begin{minted}[fontsize=\small]{fortran}
   program blasMatMul
@@
 C Initialize in column major storage of Fortran
-      data a/ 1,0,
-     *        0,1/
-      data b/ 4,1,
-     *        1,2/
+  data a/ 1,0,0,1/
+  data b/ 4,1,1,2/
 C            tfm  tfm  rowA colB K  alpha a lda  b  ldb beta c  ldc
@@
 
-\begin{figure}[tbp]
+It is important to note that should these annotations (in
+Figures~\ref{fig:cpp_eigen} and~\ref{fig:cpp_ublas}) be wrong, the program's
+behaviour is very likely to end up being undefined (like how \texttt{memcpy()}
+for overlapping regions of memory is explicitly undefined in the POSIX and C
+standards). Indeed, one of Fortran's strengths lies in assuming that references
+cannot be aliased (with certain caveats) in more cases than C permits (this
+informal, general statement comes with many nuances left for the interested
+reader to pursue).
+
+\begin{figure}[tp]
     \begin{minted}{c++}
@@
     Eigen::Matrix2d a,b,c;
-    a << 1, 0,
-         0, 1;
-    b << 4, 1,
-         1, 2;
-    c << 0, 0,
-         0, 0;
+    a << 1, 0, 0, 1; b << 4, 1, 1, 2; c << 0, 0, 0, 0;
     a * b; // new matrix
@@
 
-\section{TO DO: Proposed Solution}
-A (E)DSL!
-% https://github.com/AtheMathmo/rulinalg/blob/e0d8b76ddf8bbb5547cb9c98172e594605ab5fef/src/matrix/impl_ops.rs
-% Which operand is overwritten determines scope. Would make sense to argue
-% pros-cons of implicitness vs explicitness, emulating Rust's Affine with
-% linear sugar.
+\section{Proposed Solution}
 
-% CODE EXAMPLES
+My proposed solution to this dichotomy is a \emph{domain-specific language}
+(DSL), called LT4LA (Linear Types for Linear Algebra). Although for expository
+and testing purposes I have defined a concrete-syntax, a full implementation
+would make use of a language's syntax-extension features (such as PPX for
+OCaml) to \emph{embed} the DSL into the host language. Such an embedding is
+straightforward but fairly tedious to implement. As a half-way point, I used
+compile-time code generation to make the DSL's output available to OCaml for
+testing and evaluation.
+
+Let us have a look at a few examples of functions we can write with linear
+types.  We can define the canonical factorial function
+(Figure~\ref{fig:ltfla_factorial}) and sum over an array
+(Figure~\ref{fig:ltfla_sumarray}).
+
+\begin{figure}[tp]
+    \inputminted[linenos, fontsize=\small]{ocaml}{../test/examples/factorial.lt}
+    \caption{Factorial function in LT4LA.}\label{fig:ltfla_factorial}
+\end{figure}
+
+\begin{figure}[tp]
+    \inputminted[linenos, fontsize=\small]{ocaml}{../test/examples/sum_array.lt}
+    \caption{Summing over an array in LT4LA.}\label{fig:ltfla_sumarray}
+\end{figure}
+
+The syntax is intended to resemble OCaml's, apart from the spurious
+\ltfla{`!'}s found here and there (they are annotations to show that we can
+use a value more than once). In Figure~\ref{fig:ltfla_sumarray}, we see
+\ltfla{row} has type (\ltfla{'x arr}). More detailed explanations of
+\emph{what and why} will be given in Chapter~\ref{chap:impl}, but for now, it
+is enough to know it means we can only \emph{read} from \ltfla{row}, and not
+write to it. If we did try to write to or free \ltfla{row}, say by adding
+another line as shown in Figure~\ref{fig:ltfla_univ}, would get a helpful error
+message, as shown.
+
+\begin{figure}[tp]
+    \begin{minted}[linenos, firstnumber=7, fontsize=\small]{ocaml}
+let row = row[i] := x1 in (* or *) let () = free row in
+(* Could not show equality:                                        *)
+(*     z arr                                                       *)
+(* with                                                            *)
+(*     'x arr                                                      *)
+(*                                                                 *)
+(* Var 'x is universally quantified                                *)
+(* Are you trying to write to/free/unshare an array you don't own? *)
+(* In test/examples/sum_array.lt, at line: 7 and column: 19        *)
+    \end{minted}
+    \caption{Attempting to write to or free a read only array in
+    LT4LA.}\label{fig:ltfla_univ}
+\end{figure}
+
+Now suppose we were trying to square a matrix, using a `dgemm' like BLAS
+routine which takes two read-only matrices and a third matrix it can write to
+and performs $C := AB + C$. How would we use such a routine to square a matrix?
+Surely this would break linearity, since $A$ would be have to be given to the
+function twice, and we can only do so once?
+
+We can use a special primitive called \ltfla{share} to produce read-only
+\emph{aliases} of any matrix (including ones we cannot write to). We then pass
+this into the function, and it works as expected. Once the squared matrix has
+been obtained, we may not want the original any more, and thus decide to free
+it. Before we do so, we must first \ltfla{unshare} any read-only aliases that
+exist -- in this context, \ltfla{unshare} swallows them (the only primitive
+which can return fewer read-only values than it receives) and gives back a
+single, read-write handle.
+
+\begin{figure}[tp]
+    \begin{minted}[linenos, fontsize=\small]{ocaml}
+let (a1, a2) = share _ a in
+let ((a1, a2), c) = simple_dgemm _ a1 _ a2 c in
+let a = unshare _ a1 a2 in
+let () = free a in c
+    \end{minted}
+    \caption{Squaring a matrix in LT4LA.}\label{fig:ltfla_square}
+\end{figure}
+
+If we tried to free one of the read-only aliases instead of
+\ltfla{unshare}-ing, then we would get the error show in
+Figure~\ref{fig:ltfla_free}. Briefly, a \ltfla{z arr} is a read-write array,
+everything else (\ltfla{'x arr} or \ltfla{_ s arr}) is read-only. The types of
+\ltfla{free : z arr --o unit}, \ltfla{share : 'x arr --o 'x s arr * 'x s arr}
+and \ltfla{unshare : 'x . 'x s arr --o 'x s arr --o 'x arr} are set up so that
+you can only free something when you have read-write access to it, guaranteed
+by linearity to be the only name in scope with this capability (aliases can
+only be read-only).
+
+\begin{figure}[tp]
+    \begin{minted}[linenos, fontsize=\small]{ocaml}
+let (a1, a2) = share _ a in
+let ((a1, a2), c) = simple_dgemm _ a1 _ a2 c in
+let () = free a1 in c
+(* Error:                                                        *)
+(* Could not show equality:                                      *)
+(*     z arr                                                     *)
+(* with                                                          *)
+(*     z s arr                                                   *)
+(*                                                               *)
+(* Could not show z and z s are equal.                           *)
+(* Are you trying to write to/free an array before unsharing it? *)
+(* In test.lt, at line: 3 and column: 17                         *)
+    \end{minted}
+    \caption{Attempting to free a read-only alias of matrix.}\label{fig:ltfla_free}
+\end{figure}
 
@@
 promise of integrating linear and dependent types~\cite{atkey}.
+
+\section{Summary}
+
+I have given an \emph{intuitive} exposition of linear types with
+\emph{fractional-capabilities}, emphasising small, but illustrative and
+practical code examples, leaving \emph{what} is going on and \emph{why} it
+works as details for the next chapter. I have shown that it is possible to
+control aliasing and reading/writing capabilities \emph{statically} with a
+\emph{type system} -- a programmer can be prevented from writing over and
+freeing arrays they do not own whilst being able to safely create read-only
+aliases when it makes sense to do so. Unlike Rust's more full-featured
+borrow-checker, the next chapter will show that these features can be expressed
+and implemented in a concise, \emph{high-level} and language-independent manner
+to be packaged up, provided and used \emph{as a library}.

### 2018-05-21 write-up/implementation.tex
Finish Background, continue Implementation

@@
-\chapter{Implementation}
+\chapter{Implementation}\label{chap:impl}
 
@@
 
-    \item \textbf{Executable Artifacts}. A transpiler and a REPL are the main artifacts
-        produced for this thesis.
+    \item \textbf{Executable Artifacts}. A transpiler and a REPL are the main
+        artifacts produced for this thesis. For evaluation, I implemented
+        Kalman filters in Owl, LT4LA and CBLAS/LAPACKE and a benchmarking
+        program to measure execution times.
 
     \item \textbf{Tests}. As mentioned before, almost all non-trivial functions
-        have tests to check their behaviour. The output of the transpiler
-        was also tested by having the build system generate code at OCaml code
-        which could then be compiled and tested as usual.
+        have tests to check their behaviour. The output of the transpiler was
+        also tested by having the build system generate OCaml code at compile
+        time, which in turn could then be compiled and tested like handwritten
+        OCaml code.
 
@@
 A full description of the core language can be found in
-Appendix~\ref{chap:ott_spec}. To make a linearly-typed language usable, we need
-some way of using values zero or more than once, as we would an intuitionistic
-value. For this, we have in the type-expressions the !-constructor and in the
-term-expressions the \textbf{many}-constructor. For soundness, I implemented
-\emph{value-restriction} for this type system, only \emph{values} (anything not
-an array, matrix or an expression that can be reduced) that use \emph{no linear
-values} can be marked with !. This means we can freely have integers, booleans
-and (appropriate) functions usable as normal. Recursive functions can only use
-linear values that are passed in as arguments to the function.
+Appendix~\ref{chap:ott_spec}. Its main features are intuitionistic values,
+value-restriction, fractional-capabilities (inferrable at call sites),
+if-expressions and recursion.
 
-% TODO Once I've implemented the <- syntax, this should just include
-% ../test/examples/weighted_avg_infer.lt directly
-\begin{figure}[tb]
-    \begin{minted}[linenos]{ocaml}
-let rec f (!i : !int)
-        (!n : !int)
-        (!x0 : !elt)
-        (write : z arr)
-        ('x)
-        (weights : 'x arr)
-        : z arr * 'x arr =
-  if n = i then (write, weights) else
-  let !w0 <- weights[0] in
-  let !w1 <- weights[1] in
-  let !w2 <- weights[2] in
-  let !x1 <- write[i] in
-  let !x2 <- write[i + 1] in
-  let !x : !elt = w0 *. x0 +. w1 *. x1 +. w2 *. x2 in
-  f (i + 1) n x1 (write[i] := x) _ weights
-    \end{minted}
+\subsection{Intuitionistic Values}
 
-    \caption{A recursive function to perform a \emph{in-place} (simplified)
-        1D-convolution (sliding weighted-average) using a read-only array of
-        weights and a writeable array, originally containing the input
-        array.}\label{fig:oned_conv}
+% Why, runtime representation, trade-offs etc.
 
-\end{figure}
+To make a linearly-typed language usable, we need some way of using values zero
+or more than once, as we would an intuitionistic value. For this, we have in
+the type-expressions the !-constructor and in the term-expressions the
+\ltfla{Many}-constructor and the \ltfla{let Many <id> = .. in .. } eliminator.
+The idea behind the !-type is that the value uses no `resources'
+(linearly-typed expressions). To start off with, it is enough to say that
+anything which can be passed around by copying, will have a !-type. This
+includes integers, elements and booleans. So \ltfla{3 : !int} and \ltfla{3. +.
+4. : !elt}. However, all bindings are still linear by default, so to emulate
+intuitionism, I desugar \ltfla{let !x = <exp> in <body>} to  \ltfla{let Many x
+= <exp> in let Many x = Many (Many x) in <body>} (similarly for function
+argument bindings). The reader can check using the rules in the Appendix that
+this has the effect of moving \ltfla{x : !t} from the linear to the
+intuitionistic environments, only if \ltfla{<exp> : !t}.
 
-So, how does a typical programmer write code that uses a matrix or an array
-more than once? Figure~\ref{fig:oned_conv} shows an example program that
-executes a (simplified) 1D-convolution (sliding weighted-average) using a
-read-only array of three weights. On line 1, \ltfla{let rec} introduces a
-recursive function `f', followed by explicitly annotated formal parameters.
-Intuitively, \ltfla{!id} allows \ltfla{id} to be used intuitionistically. The
-types \ltfla{!int} and \ltfla{!elt} correspond to integers and element types
-(for example, 64-bit floating-point).
+However, just that desugaring alone is not enough to prevent a user from taking
+an array or matrix and moving it into the intuitionistic environments. Why?
+There are certain situations in which we \emph{should not} use the \ltfla{Many}
+constructor.  Consider the following code: \ltfla{let Many x = Many (array 5)
+in <body>}; the expression \ltfla{array 5} uses no linearly-typed variables
+from the linear environment. Although we could just reject types of the form
+\ltfla{!(_ arr)} to fix this simple example, what about pairs \ltfla{let Many
+xy = Many (3, array 5) in <body>}?  Ad-hoc pattern matching on the type cannot
+account for all possible situations. With the last case, we can use \ltfla{xy}
+as many times as we would like, destruct the pair to get the second component
+and thus create \emph{distinct} read-write aliases to the same array.  Alas,
+now arrays can be used intuitionistically and all the benefits of linearity are
+lost.  Or are they?
+
+\subsection{Value-Restriction}
+
+Not quite, but to understand how we can fix this problem, we need to question
+an assumption left implicit up until this point: what does \ltfla{Many} even
+mean? What does it do at runtime? One option is to go down the C++ route and
+make \ltfla{Many} act like a \ltfla{shared_ptr} and act as a runtime
+reference-count for arrays. I chose to not go for this option because it went
+against the \emph{explicitness} and \emph{predictability} that C and Fortran
+have. It would make analysing when and what is allocated and freed more like
+the higher-level languages I was trying to move away from.
+
+My aim is to show linear types can be simple to understand and apply to linear
+algebra, enough so that it can be grafted (in a limited way) on to existing
+languages as a \emph{library}. In that spirit, the simplest thing that
+\ltfla{Many} can mean at runtime is \emph{nothing}. This language construct is
+translated into a standard OCaml language constructor of the form \ltfla{type
+'a bang = Many of 'a [@@unboxed]}. The \emph{unboxed} annotation means
+that the type and its constructor only exist for the purpose of type-checking
+in OCaml; \emph{the runtime representation of values of type \ltfla{'a} and
+\ltfla{'a bang} is exactly the same}.
+
+With this understanding, our problem is that arrays and matrices are unlike
+other values such as integers and elements because (under the OCaml hood)
+calling a function with an array argument copies a \emph{pointer} to the array
+rather than the array itself, instead of the \emph{value} itself. So, we can
+start making a distinction, \emph{defining} elements, integers, booleans,
+intuitionistic variables, units and lambda-expressions that capture no linear
+variables as \emph{values} (since they cannot break referential-transparency)
+and anything else (arrays, matrices, expressions which can be reduced, such as
+function application and if-expressions) as not being `values'. If this sounds
+familiar, it is because this is the same \emph{value-restriction} `trick' from
+the world of polymorphic types applied to linearity instead. We then have the
+rule that we can only use \ltfla{Many} on expressions that are defined to be
+\emph{values} and \emph{use no linear variables}.
+
+\subsection{Fractional-Capabilities and Inference}
+
+Having understood and correctly implemented intuitionism, we now tackle
+the problem of how to implement safe aliasing, that is taking a read-write
+alias and splitting into read-only aliases, as well as the inverse.
 
 Array and matrix types are parameterised by \emph{fractional-capabilities}.  A
-fraction of 1 ($2^0$) represents complete ownership of a value; in
-particular, this allows a programmer to write or free it. Creating an array
-gives you ownership of it; the function \ltfla{array : !int --o z arr}
-(where \ltfla{z} represents `0'). Once you have ownership of an
-array/matrix, you can free it: \ltfla{free : z arr --o unit}.
-Importantly, because a linear-value may only be used once, the array just freed
-is \emph{out of scope} for following expressions, preventing use-after-free.
-Ownership also enables you to write to the array: \ltfla{set : z arr
---o !int --o !elt --o z arr} (the syntax \ltfla{w[i] := j} is just
-sugar for \ltfla{set w i j}). Here, linearity prevents accessing values
-which represented the array \emph{before} the mutation.
+fraction of 1 ($2^0$) represents complete ownership of a value; in particular,
+this allows a programmer to write or free it. Creating an array gives you
+ownership of it; the function \ltfla{array : !int --o z arr} (where \ltfla{z}
+represents `0'). Once you have ownership of an array, you can free it:
+\ltfla{free : z arr --o unit}.  Importantly, because a linear-value may only be
+used once, the array just freed is \emph{out of scope} for following
+expressions, preventing use-after-free.  Ownership also enables you to write to
+the array: \ltfla{set : z arr --o !int --o !elt --o z arr} (the syntax
+\ltfla{w[i] := j} is just sugar for \ltfla{set w i j}). Here, linearity
+prevents accessing aliases which represented the array \emph{before} the
+mutation.
 
@@
 successor (+1) of a natural number). Hence, you can read from (index) any array
-\ltfla{get : 'x .  'x arr --o !int --o !elt * 'x arr} (the syntax \ltfla{let !v
-<- w[i]} is just sugar for \ltfla{let (!v, w) = get _ w i}). Transparently
+\ltfla{get : 'x .  'x arr --o !int --o 'x arr * !elt} (the syntax \ltfla{let !v
+<- w[i]} is just sugar for \ltfla{let (w, !v) = get _ w i}). Transparently
 rebinding \ltfla{w} with the returned value means a program can appear to use
-\ltfla{w} multiple times; the underscore is how a programmer tells the compiler
+\ltfla{w} multiple times. The underscore is how a programmer tells the compiler
 to automatically \emph{infer} the correct fractional-capability based on the
-other arguments passed to the function. Fractions also permit aliasing:
-\ltfla{share : 'x . 'x arr --o 'x s arr * 'x s arr}. However, given the types
-of the primitives provided, these aliases can only be read from and not written
-to. If you want to write to this array, you must \ltfla{unshare: 'x . 'x s arr
---o 'x s arr --o 'x arr} it until you are left with a value of type \ltfla{z
-arr}, guaranteeing no other aliases exists.
+other arguments passed to the function. In conjunction with the requirement
+that functions declarations need type-annotations for their arguments, this
+allows a fractional-capability to be correctly inferred in any program.
+
+Fractions exist solely to enable safe aliasing, via the primitive \ltfla{share
+: 'x . 'x arr --o 'x s arr * 'x s arr}.  The two arrays returned (which happen
+to just be the given array) can now only be read from and not written to.  If
+you want to write to this array, you must \ltfla{unshare: 'x .  'x s arr --o
+'x s arr --o 'x arr} it with other read-only aliases until you are left with a
+value of type \ltfla{z arr}, guaranteeing no other aliases exists.
 
@@
 array when we own it; ownership guarantees no other aliases exist in scope at
-the point of usage. In the next section, I show how this perfect information
-can be used to write more natural-looking code using value-semantic expressions
-which behave in precisely the way we intend it to. Now the programmer need not
-resort to manually tracking and inserting \texttt{noalias} annotations; instead
-they can let the loyal and tireless compiler do the heavy lifting.
+the point of usage. In Figure~\ref{fig:ltfla_kalman}, I show how this perfect
+information can be used to write more natural-looking code using value-semantic
+expressions which behave in precisely the way we intend it to.  Now the
+programmer need not resort to manually tracking and inserting \texttt{noalias}
+annotations; instead they can let the loyal and tireless compiler do the heavy
+lifting.
+
+\subsection{If-Expressions}
+
+Because we do not know which way a condition will evaluate at run time, we must
+guarantee the both branches use the same set of linear variables. Writing the
+type-checker in a pure-functional monadic style paid off here because I could
+now sandwich monadic values with state-adjustments either side of it. Given two
+monadic values that represented type-checking two branches of an if-expression,
+I could use the code in Figure~\ref{fig:same_resources} to easily save, reset
+and compare the state either side of running those monadic values.
+
+\begin{figure}[tp]
+    \begin{minted}{ocaml}
+let same_resources (wf_a, loc_a) (wf_b, loc_b) =
+  let open Let_syntax in
+  (* Save state *)
+  let%bind {used_vars=prev; env=old_env; _} as state = get in
+  (* Reset, run a, save state *)
+  let%bind () = put { state with used_vars = empty_used } in
+  let%bind res_a = wf_a in
+  let%bind {used_vars=used_a; _} as state = get in
+  (* Reset, run b, save state *)
+  let%bind () = put { state with used_vars = empty_used; env = old_env } in
+  let%bind res_b = wf_b in
+  let%bind {used_vars=used_b; _} as state = get in
+  (* Check if same resources *)
+  let keys_a, keys_b = (* convert to (used_a, used_b) to sets *) in
+  if Set.equal keys_a keys_b then
+    (* merge used_vars and used_b environments *)
+  else
+    (* report differences *)
+    \end{minted}
+
+    \caption{Implementation of same\_resources helper method for type-checking
+        if-expressions. Note the monadic style helped compose computations that
+        affected the type-checker's state in a simple manner.}\label{fig:same_resources}
+
+\end{figure}
+
+\subsection{Functions and Recursion}
+
+A non-recursive function may be used more than once if it does not refer to any
+linear variables from the surrounding scope. So, we can desugar something like
+\ltfla{let x = 3 in let !f (y : !int) = x + y in <body>} to \ltfla{let x = 3 in
+let Many f =  Many (fun y : !int -> x + y) in  <body>}. Recursion is
+slightly more complicated: we can desugar the factorial function
+(Figure~\ref{fig:ltfla_factorial}) to \ltfla{let Many f = fix (f, x : !int,
+if (*..*) : !int) in <body> }. However, \ltfla{fix}, like \ltfla{Many},
+must also not use any linear variables from its surrounding scope.
 
@@
 We have now arrived at the titular \emph{applications} of linear types. I will
-show how, in addition to preventing the errors explained in the previous
-section, we can take linear types one step further and apply it to the domain
-of efficient compilation of matrix-expressions.
+show how, in addition to preventing the errors explained hitherto, we can take
+linear types one step further and apply it to the domain of efficient
+compilation of matrix-expressions.
 
-\begin{figure}[tb]
-    \inputminted[linenos]{fortran}{kalman.f90}
-    \caption{Kalman filter in Fortran 90 (Credit: Matthew
-        Rocklin~\cite{rocklin_gist}).}\label{fig:fortran_kalman}
-\end{figure}
+\begin{sidewaysfigure}
+    \inputminted[linenos, fontsize=\footnotesize]{fortran}{kalman.f90}
+    \caption{Kalman filter in Fortran 90.}\label{fig:fortran_kalman}
+\end{sidewaysfigure}
 
 In Figure~\ref{fig:fortran_kalman}, we see the difficulty of efficiently
-implementing a \emph{Kalman filter}, a powerful and set of equations applicable
+implementing a \emph{Kalman filter}, a powerful set of equations applicable
 to a wide variety of problems. From the comments, we see that every variable is
@@
     \item ensuring all values are declared and \emph{initialised} correctly before they are used,
-    \item checking no values are used after they are freed \emph{and} ensuring no values are leaked.
+    \item checking no values are used after they are freed \emph{or} leaked.
 \end{itemize}
@@
 copying sub-expressions by default and then letting the compiler tell it which
-copies are never used and removing them systematically until it type checks.
+copies are never used and removing them systematically until it type-checks.
 While it is not quite a black-box, push-button compilation of an expression, I
@@
 \begin{figure}[tb]
-
-    \begin{minted}[linenos]{ocaml}
-let !kalman
-        ('s) (sigma : 's mat) (* n, n *)
-        ('h) (h : 'h mat)     (* k, n *)
-        ('m) (mu : 'm arr)    (* n    *)
-        (r_1 : z mat)         (* k, k *)
-        (data_1 : z arr)      (* k    *)
-        : ('s mat * ('h mat * 'm arr)) * (z arr * z mat) =
-    let ((!k, !n), h) = dimM _ h in
-    let zeros = zeroM n n in
-    let sigma_h = new [| sym (up, sigma) * h |] in
-    let r_2 = [| h * sigma_h + r_1 |] in
-    let mu_1 <- copy mu in
-    let sigma_1 <- copyM sigma in
-    let data_2 = [| h * mu + -1. * data_1 |] in
-    let h_1 <- copyM h in
-    let h_sol = [| sym (up, r) ^ -1 * h_1 ] in
-    let data_sol = [| sym (up, r) ^ -1 * data_2 |] in
-    let h_h1 = new [| h * h_sol |] in
-    let h_data = new [| h * data |] in
-    let h_h1_sigma = new [| sym (up, h_h1) * sigma |] in
-    let mu_2 = [| sym (up, sigma) * h_data * mu_1 |] in
-    (* non-linear use of sigma_c..? *)
-    let sigma_c = [|  -1. * sym (up, sigma_c) * h_h1_sigma + sigma_c |] in
-    (* frees... *)
-    (mu_c , sigma_c)
-    \end{minted}
-
+    \inputminted[linenos, fontsize=\small]{ocaml}{../test/examples/kalman.lt}
     \caption{Kalman filter in LT4LA.}\label{fig:ltfla_kalman}
-
 \end{figure}
@@
 So-called `matrix \emph{expressions}' are still `side-effecting'/consuming
-linear values, and do not produce \emph{values} but a \emph{sequence of
+linear variables, and do not produce \emph{values} but a \emph{sequence of
 (re-)bindings} which in turn dictate what values are still available/in-scope
@@
 
-\subsection{TO DO: Inference}
-
 \section{TO DO: Code Generation}
@@
 \subsection{TO DO: Metaprogramming}
-
-\section{TO DO: Theory}
-
-Fractional permissions, not proven soundness etc\. but compilation preserves
-types as checked by OCaml.

### 2018-05-21 write-up/template.tex
Finish Background, continue Implementation

\chapter{Primitives}\label{chap:primitives}

The following signature gives an indication of how I embedded a linear type
system into the OCaml one and typed LT4LA's primitives accordingly. This helped
catch bugs and increase confidence in the correctness of the code produced.

\begin{minted}[fontsize=\small, linenos]{ocaml}
module Arr = Owl.Dense.Ndarray.D
type z = Z
type 'a s = Succ
type 'a arr = A of Arr.arr [@@unboxed]
type 'a mat = M of Arr.arr [@@unboxed]
type 'a bang = Many of 'a [@@unboxed]
module Prim :
sig
  val extract : 'a bang -> 'a
  (** Boolean *)
  val not_ : bool bang -> bool bang
  (** Arithmetic, some omitted for brevity *)
  val addI : int bang -> int bang -> int bang
  val ltI : int bang -> int bang -> bool bang
  (** Arrays *)
  val set : z arr -> int bang -> float bang -> z arr
  val get : 'a arr -> int bang -> 'a arr * float bang
  val share : 'a arr -> 'a s arr * 'a s arr
  val unshare : 'a s arr -> 'a s arr -> 'a arr
  val free : z arr -> unit
  (** Owl *)
  val array : int bang -> z arr
  val copy : 'a arr -> 'a arr * z arr
  val sin : z arr -> z arr
  val hypot : z arr -> 'a arr -> 'a arr * z arr
  (** Level 1 BLAS *)
  val asum : 'a arr -> 'a arr * float bang
  val axpy : float bang -> 'a arr -> z arr -> 'a arr * z arr
  val dot : 'a arr -> 'b arr -> ('a arr * 'b arr) * float bang
  val rotmg : float bang * float bang -> float bang * float bang ->
              (float bang * float bang) * (float bang * z arr)
  val scal : float bang -> z arr -> z arr
  val amax : 'a arr -> 'a arr * int bang
  (* Matrix, some omitted for brevity *)
  val matrix : int bang -> int bang -> z mat
  val copy_mat : 'a mat -> 'a mat * z mat
  val copy_mat_to : 'a mat -> z mat -> 'a mat * z mat
  val size_mat : 'a mat -> 'a mat * (int bang * int bang)
  val transpose : 'a mat -> 'a mat * z mat
  (* Level 3 BLAS/LAPACK *)
  val gemm : float bang -> ('a mat * bool bang) -> ('b mat * bool bang) ->
             float bang -> z mat -> ('a mat * 'b mat) * z mat
  val symm : bool bang -> float bang -> 'a mat -> 'b mat ->
             float bang -> z mat -> ('a mat * 'b mat) * z mat
  val posv : z mat -> z mat -> z mat * z mat
  val potrs : 'a mat -> z mat -> 'a mat * z mat
end
\end{minted}

\clearpage%
\ottall%

### 2018-05-21 write-up/kalman.f90

@@ -1,0 +1,37 @@
+subroutine f(mu, Sigma, H, INFO, R, Sigma_2, data, mu_2, k, n)
+implicit none
+
+integer, intent(in)    :: k
+integer, intent(in)    :: n
+real*8,  intent(in)    :: Sigma(n,n)   ! Sigma
+real*8,  intent(in)    :: H(k,n)       ! H
+real*8,  intent(in)    :: mu(n)        ! mu
+real*8,  intent(inout) :: data(k)      ! data, H*mu - data , (H*Sigma*H^T + R)^-1*(H*mu - data)
+real*8,  intent(inout) :: R(k, k)      ! R, H*Sigma*H^T + R
+
+integer, intent(out)   :: INFO         ! INFO
+real*8,  intent(out)   :: mu_2(n)      ! mu, Sigma*H^T*(H*Sigma*H^T + R)^-1*(H*mu - data) + mu
+real*8                 :: H_2(k,n)     ! H * Sigma , H , (H*Sigma*H^T + R)^-1*H
+real*8                 :: chol_R(k,k)  ! R, U where (H*Sigma*H^T + R)=U^T*U
+real*8,  intent(out)   :: Sigma_2(n,n) ! H^T*(H*Sigma*H^T + R)^-1*H, Sigma, Sigma - Sigma*H^T*(H*Sigma*H^T + R)^-1*H*Sigma
+real*8                 :: H_data(n)    ! H^T*(H*Sigma*H^T + R)^-1*(H*mu - data)
+real*8                 :: N_N_tmp(n,n) ! H^T*(H*Sigma*H^T + R)^-1*H*Sigma
+
+call dsymm('R', 'U', k, n, 1, Sigma, n, H, n, 0, H_2, n)            ! H_2     := 1. * H   * Sigma +  0. * H_2
+call dgemm('N', 'T', k, k, n, 1, H_2, n, H, n, 1, R, k)             ! R       := 1. * H_2 * H     +  1. * R
+call dgemm('N', 'N', k, 1, n, 1, H, n, mu, 1, -1, data, 1)          ! data    := 1. * H   * mu    + -1. * data
+call dcopy(k*n, H, 1, H_2, 1)                                       ! H_2     := H
+call dcopy(k*k, R, 1, chol_R, 1)                                    ! chol_R  := R
+call dposv('U', k, n, chol_R, k, H_2, n, INFO)                      ! chol_R  := U where R = U^T * U
[…]

### 2018-05-22 write-up/abstract.tex
Finish Implementation

@@
 \newpage
-{\normalfont\Huge\sffamily\bfseries Abstract}
+{\normalfont\Huge\sffamily\bfseries TO DO: Abstract}
 \vspace{24pt}
 
-This is the abstract. Write a summary of the whole thing. Make 
-sure it fits in one page.
+In this thesis, I will argue that linear types are an appropriate,
+\emph{type-based formalism} for expressing aliasing, read/write permissions,
+memory allocation, re-use and deallocation, first, in the context of the APIs
+of linear algebra libraries, consequently extending these principles to the
+problem of \emph{efficient} matrix-expression compilation.
+
+I will show that framing the problem using linear types can help \emph{reduce
+bugs} by making precise and explicit the informal, ad-hoc practices typically
+employed by human experts and linear algebra \emph{compilers} and automate
+checking them.
+
+As evidence for this argument, I will show programs written with this safety,
+precision and explicitness (1) can be just as pleasant and convenient for a
+programmer as less efficient, but higher-level linear algebra libraries and (2)
+perform just as \emph{efficiently and predictably} as lower-level, less
+readable and more error-prone interfaces linear algebra libraries. I will
+further show that support for such programs can be implemented \emph{as
+libraries} on top of existing programming languages and libraries.
+
+\textbf{Summarise evaluation.}
+
+\textbf{Summary future work.}
 

### 2018-05-22 write-up/background.tex
Finish Implementation

@@
     \centering
-    \begin{minted}{python}
+    \begin{minted}[fontsize=\small]{python}
 # Numpy (Python)
@@
     \end{minted}
-    \begin{minted}{julia}
+    \begin{minted}[fontsize=\small]{julia}
 # Julia
@@
     \end{minted}
-    \begin{minted}{ocaml}
+    \begin{minted}[fontsize=\small]{ocaml}
 
@@
     \centering
-    \begin{minted}{ocaml}
+    \begin{minted}[fontsize=\small]{ocaml}
 let mul x y =
@@
 \begin{figure}[tp]
-    \begin{minted}{c++}
+    \begin{minted}[fontsize=\small]{c++}
 #include <iostream>
@@
 \begin{figure}[tbp]
-    \begin{minted}{c++}
+    \begin{minted}[fontsize=\small]{c++}
 noalias(C) = prod(A, B);
@@
 My proposed solution to this dichotomy is a \emph{domain-specific language}
-(DSL), called LT4LA (Linear Types for Linear Algebra). Although for expository
-and testing purposes I have defined a concrete-syntax, a full implementation
-would make use of a language's syntax-extension features (such as PPX for
-OCaml) to \emph{embed} the DSL into the host language. Such an embedding is
-straightforward but fairly tedious to implement. As a half-way point, I used
-compile-time code generation to make the DSL's output available to OCaml for
-testing and evaluation.
+(DSL), called LT4LA (Linear Types for Linear Algebra). It offers readable,
+explicit management of aliasing, read/write permissions, memory allocation,
+re-use and deallocation all with automated checking: offering a safety net to
+catch the baby whilst swiftly disposing of the bath water.  Although for
+expository and testing purposes I have defined a concrete-syntax, a full
+implementation would make use of a language's syntax-extension features (such
+as PPX for OCaml) to \emph{embed} the DSL into the host language. Such an
+embedding is straightforward but fairly tedious to implement. As a half-way
+point, I used compile-time code generation to make the DSL's output available
+to OCaml for testing and evaluation.
 
@@
 everything else (\ltfla{'x arr} or \ltfla{_ s arr}) is read-only. The types of
-\ltfla{free : z arr --o unit}, \ltfla{share : 'x arr --o 'x s arr * 'x s arr}
-and \ltfla{unshare : 'x . 'x s arr --o 'x s arr --o 'x arr} are set up so that
-you can only free something when you have read-write access to it, guaranteed
-by linearity to be the only name in scope with this capability (aliases can
-only be read-only).
+\ltfla{free}, \ltfla{share} and \ltfla{unshare} (see
+Appendix~\ref{chap:primitives}) are set up so that you can only free something
+when you have read-write access to it, guaranteed by linearity to be the only
+name in scope with this capability (aliases can only be read-only). Conversely,
+if we \emph{forgot} to \ltfla{free a}, we would also get \texttt{Variable a not
+used} error.
 
@@
 logic to programming along the well-trodden Curry-Howard bridge brings us to
-linear types.
+linear types~\cite{wadler}.
 
@@
 practical code examples, leaving \emph{what} is going on and \emph{why} it
-works as details for the next chapter. I have shown that it is possible to
-control aliasing and reading/writing capabilities \emph{statically} with a
-\emph{type system} -- a programmer can be prevented from writing over and
-freeing arrays they do not own whilst being able to safely create read-only
-aliases when it makes sense to do so. Unlike Rust's more full-featured
-borrow-checker, the next chapter will show that these features can be expressed
-and implemented in a concise, \emph{high-level} and language-independent manner
-to be packaged up, provided and used \emph{as a library}.
+works as details for the next chapter. I have shown that it is possible to have
+high-level, explicit and automatically checked management of aliasing,
+read/write permissions, memory allocation, re-use and deallocation with a
+\emph{type system} -- preventing a whole class of errors that can occur with
+lower-level languages \emph{at compile time}, using the of context linear
+algebra libraries as as a specific example. In the next chapter, I will show
+that these features can be expressed and implemented in a language-independent
+manner to be provided and used \emph{as a library}.

### 2018-05-22 write-up/implementation.tex
Finish Implementation

@@
 
-\begin{guidance}
-This chapter may be called something else\ldots but in general
-the idea is that you have one (or a few) ``meat'' chapters which
-describe the work you did in technical detail.
-\end{guidance}
+% \begin{guidance}
+% This chapter may be called something else\ldots but in general
+% the idea is that you have one (or a few) ``meat'' chapters which
+% describe the work you did in technical detail.
+% \end{guidance}
 
@@
 
-% Why, runtime representation, trade-offs etc.
-
 To make a linearly-typed language usable, we need some way of using values zero
@@
 = <exp> in let Many x = Many (Many x) in <body>} (similarly for function
-argument bindings). The reader can check using the rules in the Appendix that
-this has the effect of moving \ltfla{x : !t} from the linear to the
-intuitionistic environments, only if \ltfla{<exp> : !t}.
+argument bindings). The reader can check (using the rules in
+Appendix~\ref{chap:ott_spec}) that this has the effect of moving \ltfla{x : !t}
+from the linear to the intuitionistic environments, only if \ltfla{<exp> : !t}.
 
@@
 
-Having understood and correctly implemented intuitionism, we now tackle
-the problem of how to implement safe aliasing, that is taking a read-write
-alias and splitting into read-only aliases, as well as the inverse.
+Having started off with linearity and understanding how it helps a programmer
+keep track of of memory allocation and deallocation, and then understood and
+correctly implemented intuitionism for the only values we wish to be
+intuitionistic, and, we now tackle the problem of how to implement safe
+aliasing and read/write permissions.
 
@@
 \ltfla{get : 'x .  'x arr --o !int --o 'x arr * !elt} (the syntax \ltfla{let !v
-<- w[i]} is just sugar for \ltfla{let (w, !v) = get _ w i}). Transparently
-rebinding \ltfla{w} with the returned value means a program can appear to use
-\ltfla{w} multiple times. The underscore is how a programmer tells the compiler
-to automatically \emph{infer} the correct fractional-capability based on the
-other arguments passed to the function. In conjunction with the requirement
-that functions declarations need type-annotations for their arguments, this
-allows a fractional-capability to be correctly inferred in any program.
+<- w[i]} is just sugar for \ltfla{let (w, !v) = get _ w i}). In general, a
+left-arrow \ltfla{<-} signifies transparent rebinding with returned values: it
+means a program can \emph{appear} to use a variable multiple times, important
+for keeping LT4LA usable and readable. The underscore is how a programmer tells
+the compiler to automatically \emph{infer} the correct fractional-capability
+based on the other arguments passed to the function. In conjunction with the
+requirement that functions declarations need type-annotations for their
+arguments, this allows a fractional-capability to be correctly inferred in any
+program.
 
-Fractions exist solely to enable safe aliasing, via the primitive \ltfla{share
-: 'x . 'x arr --o 'x s arr * 'x s arr}.  The two arrays returned (which happen
-to just be the given array) can now only be read from and not written to.  If
-you want to write to this array, you must \ltfla{unshare: 'x .  'x s arr --o
-'x s arr --o 'x arr} it with other read-only aliases until you are left with a
-value of type \ltfla{z arr}, guaranteeing no other aliases exists.
+Fractions exist to provide both safe aliasing and read/write permissions, via
+the primitives \ltfla{share : 'x . 'x arr --o 'x s arr * 'x s arr} and
+\ltfla{unshare: 'x .  'x s arr --o 'x s arr --o 'x arr}.  For the former, two
+arrays returned (which happen to just be the given array) can now only be read
+from and not written to. If you want to write to this array, you must use the
+latter to combine other read-only aliases until you are left with a value of
+type \ltfla{z arr}, guaranteeing no other aliases exists.
 
 Given this set-up, we now \emph{statically} have \emph{perfect} information
-about aliasing and ownership of values in the program. We can only write to an
-array when we own it; ownership guarantees no other aliases exist in scope at
-the point of usage. In Figure~\ref{fig:ltfla_kalman}, I show how this perfect
-information can be used to write more natural-looking code using value-semantic
-expressions which behave in precisely the way we intend it to.  Now the
-programmer need not resort to manually tracking and inserting \texttt{noalias}
-annotations; instead they can let the loyal and tireless compiler do the heavy
-lifting.
+about aliasing and ownership of values in the program. We can only write to or
+free an array only when we own it; ownership guarantees no other aliases exist
+in scope at the point of usage. In Figure~\ref{fig:ltfla_kalman}, I show how
+this perfect information can be used to write more natural-looking code using
+value-semantic expressions which behave in precisely the way we intend it to.
+Now the programmer need not resort to manually figuring out and inserting
+\texttt{noalias} annotations and worrying about what variables can and cannot
+be written to or freed; instead they can let the loyal and tireless compiler do
+the heavy lifting.
 
@@
 \begin{figure}[tp]
-    \begin{minted}{ocaml}
+    \begin{minted}[fontsize=\small]{ocaml}
 let same_resources (wf_a, loc_a) (wf_b, loc_b) =
@@
 
-We have now arrived at the titular \emph{applications} of linear types. I will
-show how, in addition to preventing the errors explained hitherto, we can take
-linear types one step further and apply it to the domain of efficient
-compilation of matrix-expressions.
+We have now arrived at an extended application of linear types. I will show
+how, we can apply the ability to automatically check aliasing, read/write
+permissions, memory allocation, re-use and deallocation, to the domain of
+matrix-expression compilation.
 
@@
     \item ensuring all values are declared and \emph{initialised} correctly before they are used,
-    \item checking no values are used after they are freed \emph{or} leaked.
+    \item checking values are neither used after they are freed \emph{nor} leaked.
 \end{itemize}
@@
 
-\begin{figure}[tb]
+\begin{figure}[tp]
     \inputminted[linenos, fontsize=\small]{ocaml}{../test/examples/kalman.lt}
@@
 
-\subsection{TO DO: Elaboration}
+\subsection{Elaboration}
 
-So-called `matrix \emph{expressions}' are still `side-effecting'/consuming
-linear variables, and do not produce \emph{values} but a \emph{sequence of
-(re-)bindings} which in turn dictate what values are still available/in-scope
-after the computation.  As such, compilation cannot be done purely
-compositionally via structural-induction on the expression language; it is most
-concisely expressed via CPS, where the result of elaborating a matrix
-expression is a function that takes as its argument the rest of the computation
-expecting to use the result it has just made available under that requested
-name.
+All of the syntax in Figure~\ref{fig:ltfla_kalman} can be unambiguously
+desugared \emph{before} type-checking, through fairly simple pattern matching.
+Matrix expression compilation is well-trodden territory in
+academia~\cite{rocklin_thesis, fabregat_thesis, gunnels_flame, linnea, taco}
+but this is, to my knowledge, the first type-based approach to it.
 
-\section{TO DO: Code Generation}
+An overview of the translations are in Figure~\ref{fig:mat_patterns}; details
+about choosing between `symm' or `gemm' are omitted for brevity. Before
+settling on this approach, I tried implementing a more general type-directed,
+nested matrix expression compiler; I will now highlight some of the
+difficulties inherent in the problem.
 
-High-level overview: primitives, translations, optimisations, difficulty, build system.
+\begin{figure}[tp]
+    \begin{minted}[fontsize=\small]{ocaml}
+let x <- [| y |] in ..              ==> let (y,x) = copyM_to _ y x in ..
+let x <- new [| y |] in ..          ==> let (y,x) = copyM _ y in ..
+let x <- [| a*b + c |] in ..        ==> let x <- [| 1.*a*b + 1.*c |] in ..
+let x <- [| i.*a*b + j.*c |] in ..  ==> let ((a,b), c) = (* BLAS *) in ..
+let x <- new (m, n) [| a*b |] in .. ==> let c = matrix m n in
+                                        let x <- [| 1.*a*b + 0.*c |] in ..
+    \end{minted}
+    \caption{Syntactic translations of matrix expressions to linearly-typed
+        matrix functions. Further annotations on the matrix variables
+        (\ltfla{sym} or \ltfla{^T}) determine which BLAS routine and what
+        parameters to call it with.}\label{fig:mat_patterns}
 
-\subsection{TO DO: Compiling Constructs to OCaml}
+\end{figure}
 
-\subsection{TO DO: Metaprogramming}
+One of the first hurdles I encountered was compositionality: to compile $AB +
+C$ it is typically better to use a BLAS routine directly rather than but first
+compiling $A$, then $B$, then adding a call to multiply them, then $C$ and
+finishing with a call to add the results.
+
+Another compositionality problem is that a call to a linear function does not
+just return a result, but a sequence of re-bindings that dictate which
+variables are still in scope/usable.  As such, to compile an expression, you
+need to provide a CPS-style function of type $var \rightarrow exp$ representing
+how you would use a variable representing the result in the rest of the
+expression.
+
+However, the type of that variable also determines how you can use it: can you
+write to it or must you copy it? This information depends on how the expression
+representing the variable was elaborated and adds to the complexity of the
+pattern-matching.
+
+Copying leads us into dealing with temporaries: do you first allocate all
+temporaries in new matrices and then analyse lifetimes and dimensions to figure
+out which slots of memory can be re-used? Or do you try and infer the available
+resources from the environment as you go?  Can, and should, you type arrays and
+matrices (or $n$-dimensional tensors) as the same?
+
+Adding in more and more considerations, the problem starts to resemble register
+allocation: there are registers of different types and sizes, many instructions
+to choose from, a cost model to take into account all whilst trying to balance
+the number of registers in use and the number of instructions emitted.
+
+\section{Code Generation}
+
+Code generation is a thankfully straightforward mapping from core LT4LA
+constructs to OCaml constructs, with the appropriate sprinkling of \ltfla{Many}
+constructors for integer, element and boolean literals.
+
+To make the code produced readable, I added a few `optimisations', to the
+compiler, to `re-sugar' some of the constructs translated where appropriate.
+Almost all of them involved the erasing the !-eliminator which is not needed in
+regular, intuitionistic OCaml. So, so we can simply replace any expression of
+the form:
+
+\begin{itemize}
+
+    \item \ltfla{let xy, z = <exp> in let x,y = xy in <body>} with
+        \ltfla{let (x,y), z = <exp in <body>},
+
+    \item \ltfla{let Many x = x in let Many x = Many (Many x) in <body>} with
+        \ltfla{<body>},
+
+    \item \ltfla{let Many x = <exp> in let Many x = Many (Many x) in <body>} with
+        \ltfla{let x = <exp> in <body>},
+
+    \item \ltfla{let Many x = Many <exp> in <body>} with \ltfla{let x = <exp> in <body},
+
+    \item \ltfla{let Many f = fix (f, x, .., <exp>, ..) in <body>} with
+        \ltfla{let rec f x = <exp> in <body>}.
+
+\end{itemize}
+
+The end result is visible in Figures~\ref{fig:ocaml_sumarray}
+and~\ref{fig:ocaml_kalman}: OCaml code that is not obviously safe and correct
+with respect to linearity, as promised at the start of the chapter.
+
+\begin{figure}[tp]
+    \begin{minted}[linenos, fontsize=\small]{ocaml}
+let rec f i n x0 row =
+  if Prim.extract @@ Prim.eqI i n then (row, x0)
+  else
+    let row, x1 = Prim.get row i in
+    f (Prim.addI i (Many 1)) n (Prim.addE x0 x1) row
+in
+f
+    \end{minted}
+    \caption{Recursive OCaml function for a summing an array, generated from
+        Figure~\ref{fig:ltfla_sumarray}, passed through \texttt{ocamlformat} for
+        presentation.}\label{fig:ocaml_sumarray}
+
+\end{figure}
+
+Even though Figure~\ref{fig:ocaml_kalman} is in OCaml, I see little reason that
+it cannot be C or Fortran. Translating LT4LA's core language constructs to
+those languages would not be as easy as translating to OCaml. However, it is
+easy to see that both Figures~\ref{fig:fortran_kalman}
+and~\ref{fig:ocaml_kalman} are realisations of a concise, linearly-typed Kalman
+filter~\ref{fig:ltfla_kalman} \emph{specification} that describes the whole
+behaviour and intent (with respect to read/writes permissions, memory
+management and aliasing) of the program and the BLAS primitives it uses more
+accurately than a Fortran, C or OCaml implementation could.
+
+\begin{figure}[p]
+    \begin{minted}[linenos, fontsize=\footnotesize]{ocaml}
+let kalman sigma h mu r_1 data_1 =
+  let h, _p_k_n_p_ = Prim.size_mat h in
+  let k, n = _p_k_n_p_ in
+  let sigma_h = Prim.matrix k n in
+  let (sigma, h), sigma_h =
+    Prim.symm (Many true) (Many 1.) sigma h (Many 0.) sigma_h
+  in
+  let (sigma_h, h), r_2 =
+    Prim.gemm (Many 1.) (sigma_h, Many false) (h, Many true) (Many 1.) r_1
+  in
+  let (h, mu), data_2 =
+    Prim.gemm (Many 1.) (h, Many false) (mu, Many false) (Many (-1.)) data_1
+  in
+  let h, new_h = Prim.copy_mat_to h sigma_h in
+  let r_2, new_r = Prim.copy_mat r_2 in
+  let chol_r, sol_h = Prim.posv new_r new_h in
+  let chol_r, sol_data = Prim.potrs chol_r data_2 in
+  let () = Prim.free_mat chol_r in
+  let h_sol_h = Prim.matrix n n in
+  let (h, sol_h), h_sol_h =
+    Prim.gemm (Many 1.) (h, Many true) (sol_h, Many false) (Many 0.) h_sol_h
+  in
+  let () = Prim.free_mat sol_h in
+  let h_sol_data = Prim.matrix n (Many 1) in
+  let (h, sol_data), h_sol_data =
+    Prim.gemm (Many 1.) (h, Many true) (sol_data, Many false) (Many 0.) h_sol_data
+  in
+  let mu, mu_copy = Prim.copy_mat mu in
+  let (sigma, h_sol_data), new_mu =
+    Prim.symm (Many false) (Many 1.) sigma h_sol_data (Many 1.) mu_copy
+  in
+  let () = Prim.free_mat h_sol_data in
+  let h_sol_h_sigma = Prim.matrix n n in
+  let (sigma, h_sol_h), h_sol_h_sigma =
+    Prim.symm (Many true) (Many 1.) sigma h_sol_h (Many 0.) h_sol_h_sigma
+  in
+  let sigma, sigma_copy = Prim.copy_mat_to sigma h_sol_h in
+  let (sigma, h_sol_h_sigma), new_sigma =
+    Prim.symm (Many false) (Many (-1.)) sigma h_sol_h_sigma (Many 1.) sigma_copy
+  in
+  let () = Prim.free_mat h_sol_h_sigma in
+  ((sigma, (h, (mu, (r_2, sol_data)))), (new_mu, new_sigma)) )
+in
+kalman
+    \end{minted}
+    \caption{OCaml code for a Kalman filter, generated from
+        Figure~\ref{fig:ltfla_kalman}, passed through \texttt{ocamlformat} for
+        presentation.}\label{fig:ocaml_kalman}
+
+\end{figure}
+
+\subsection{Build System}
+
+So once you have written your memory-optimised program with all the features
+and support provided by LT4LA and made it produce well-typed, compilable OCaml
+code, the question then becomes, how to use this code. This process will vary
+across ecosystems, but within OCaml, the new build-system on the block
+\emph{Dune} has support for generating, compiling and linking OCaml modules at
+\emph{compile time}. Indeed, this is how I have written tests and benchmarks
+for programs produced by LT4LA from within OCaml. In particular, I can use my
+linearly-typed Kalman filter implementation just like and with any other OCaml
+function.
+
+Advantages of generating code at compile time include avoiding runtime overhead
+as well as catching any interface changes between the generated code and the
+code that uses it via OCaml's type system (another benefit of embedding LT4LA's
+type system inside of OCaml).
+
+I suspect that this approach will be very valuable to not just users of
+libraries such as Numpy or Owl, but also their implementors: library functions
+which presented a safe, value-semantic interface but use unsafe, mutating
+operations on the inside could now be expressed using LT4LA and have an extra
+safety-net and automatic checking for their implementations.
+
+\section{Summary}
+
+I have explained how a few core features -- linearity, the \ltfla{Many}
+constructor, value-restriction, fractional-capabilities with inference,
+if-expressions and recursive functions -- are enough to \emph{statically
+capture and automatically check} aliasing, read/write permissions, memory
+allocation, re-use and deallocation of real world linear algebra library
+routines. I have also demonstrated that simple pattern-matching and desugaring
+provides the potential for a new, \emph{type-directed} approach to matrix
+expression compilation. Lastly I have shown that it is possible to use these
+features with \emph{existing} languages and frameworks.\footnote{As mentioned
+in the previous chapter, if the host language supports
+\emph{syntax-extensions}, like PPX for OCaml, it would be possible to use LT4LA
+features \emph{from within} OCaml instead of just alongside it.}

### 2018-05-22 write-up/introduction.tex
Finish Implementation

--- 2018-02-05 write-up/introduction.tex
@@
 
-\begin{guidance}
-    This is the introduction where you should introduce your work.  In
-    general the thing to aim for here is to describe a little bit of the
-    context for your work --- why did you do it (motivation), what was the
-    hoped-for outcome (aims) --- as well as trying to give a brief
-    overview of what you actually did.
+% \begin{guidance}
+%     This is the introduction where you should introduce your work.  In
+%     general the thing to aim for here is to describe a little bit of the
+%     context for your work --- why did you do it (motivation), what was the
+%     hoped-for outcome (aims) --- as well as trying to give a brief
+%     overview of what you actually did.
+% 
+%     It's often useful to bring forward some ``highlights'' into
+%     this chapter (e.g.\ some particularly compelling results, or
+%     a particularly interesting finding).
+% 
+%     It's also traditional to give an outline of the rest of the
+%     document, although without care this can appear formulaic
+%     and tedious. Your call.
+% \end{guidance}
 
-    It's often useful to bring forward some ``highlights'' into
-    this chapter (e.g.\ some particularly compelling results, or
-    a particularly interesting finding).
-
-    It's also traditional to give an outline of the rest of the
-    document, although without care this can appear formulaic
-    and tedious. Your call.
-\end{guidance}
+\prechapter{%
+}%
 

### 2018-05-22 write-up/related_work.tex
Finish Implementation

@@
     to existing work, leaving brief discussions on future work to the next
-    chapter. I strongly believe matrix expression compilation research would
+    chapter. I strongly believe the field of matrix expression compilation would
     benefit greatly from a comprehensive literature review but unfortunately
@@
 typical example of this is the generation of Fast Fourier-Transform kernels,
-specialised to a desired array length.
+specialised to a desired array length. Combining this with recent work on
+generalising and automatically deriving \emph{partially static} representations
+of data~\cite{yallop}, it may be possible to apply such techniques to
+\emph{tensor algebra} expressions.
 

### 2018-05-22 write-up/template.tex
Finish Implementation

@@
   val not_ : bool bang -> bool bang
-  (** Arithmetic, some omitted for brevity *)
+  (** Arithmetic, many omitted for brevity *)
   val addI : int bang -> int bang -> int bang
-  val ltI : int bang -> int bang -> bool bang
+  val ltE : float bang -> float bang -> bool bang
   (** Arrays *)

### 2018-05-24 write-up/abstract.tex
Start Evaluation

--- 2018-02-05 write-up/abstract.tex
@@
 \newpage
-{\normalfont\Huge\sffamily\bfseries Abstract}
+{\normalfont\Huge\sffamily\bfseries TO DO: Abstract}
 \vspace{24pt}
 
-This is the abstract. Write a summary of the whole thing. Make 
-sure it fits in one page.
+In this thesis, I argue that linear types are an appropriate, \emph{type-based
+formalism} for expressing aliasing, read/write permissions, memory allocation,
+re-use and deallocation, first, in the context of the APIs of linear algebra
+libraries and then in the context of matrix-expression compilation.
+
+I show that framing the problem using linear types can help \emph{reduce bugs}
+by making precise and explicit the informal, ad-hoc practices typically
+employed by human experts and linear algebra \emph{compilers} and automate
+checking them.
+
+As evidence for this argument, I show non-trivial, yet readable, linear algebra
+programs, that are safe (with respect to aliasing, read/write permissions,
+memory re-use and deallocation) and explicit (with respect to memory
+allocation) which (1) are more memory-efficient than equivalent programs
+written using purely high-level linear algebra libraries and (2) perform just
+as predictably as equivalent programs written using low-level linear-algebra
+libraries. I also argue \emph{the experience} of writing such programs with
+linear types is ``better'' in key-respects. In addition to all of this, I show
+that it is possible to provide such features \emph{as a library} on top of
+existing programming languages and linear algebra libraries.
+
+\textbf{Summarise evaluation.}
+
+\textbf{Summary future work.}
 

### 2018-05-24 write-up/background.tex
Start Evaluation

@@
 \prechapter{%
-    In this chapter, I will outline the concept of linear types and show how
-    they can be used to solve the problems faced by programmers writing code
-    using linear-algebra libraries. I will be emphasising the \emph{practical}
-    and intuitive explanations of linear types to keep this thesis accessible
-    to working programmers as well as academics not familiar with type-theory;
-    giving only a terse overview of the history and theory behind linear types
-    for the interested reader to pursue further.
+    I will outline the concept of linear types and show how they can be used to
+    solve the problems faced by programmers writing code using linear-algebra
+    libraries. I will emphasise \emph{practical} and intuitive explanations of
+    linear types to keep this thesis accessible to working programmers as well
+    as academics not familiar with type-theory and will give only a terse
+    overview of the history and theory behind linear types for the interested
+    reader to pursue further.
 }%

### 2018-05-24 write-up/eval_data.tex
Start Evaluation

\chapter{Evaluation Raw Data}\label{chap:eval_data}

Below is formatted output from a trace I obtained by recompiling Owl with print
statements inserted on the relevant primitives. I made two modifications: I
shortened `\_matrix\_transpose' to `\_mtrsp' for formatting and I split `posv'
into `potrf/potrs' for a fairer comparison.

\inputminted[linenos, fontsize=\small]{text}{trace.txt}

Below is the raw output from the benchmarking script I wrote.
\inputminted[linenos, fontsize=\small]{text}{timings.txt}

### 2018-05-24 write-up/evaluation.tex
Start implementation, related work and conclusion
Finish Background, continue Implementation
Start Evaluation

--- 2018-02-05 write-up/evaluation.tex
@@
 \chapter{Evaluation}
+
+% Least Squares Linear Regression
+% p73 onwards https://people.cs.uchicago.edu/~mrocklin/storage/dissertation.pdf
 
@@
 \end{guidance}
+
+\prechapter{%
+    I will evaluate the central premise of this thesis: linear types are a
+    practical and usable tool to help working programmers write readable, safe
+    (with respect to aliasing, read/write permissions, memory re-use and
+    deallocation) and explicit (with respect to memory allocation) code that
+    (1) is more memory-efficient than code written using purely high-level
+    linear algebra libraries and (2) performs just as predictably as code
+    written using low-level linear-algebra libraries. I will also elaborate on
+    the qualitative benefits of using linearly types to write linear-algebra
+    programs.
+}%
+
+\section{Set-up}
+
+I analysed four implmentations of a Kalman filter, increasing
+the abstraction level at each step:
+
+\begin{enumerate}
+
+    \item a CBLAS/LAPACKE implementation, handwritten in C
+        (Figure~\ref{fig:cblas_kalman}), with a minimal number of temporaries,
+        calls to `symm' for matrices known to be symmetric ahead of time,
+        transposition passed in to `gemm' as a flag and Cholesky decomposition
+        for multiplying by an inverse of a matrix
+
+    \item a LT4LA implementation (Figure~\ref{fig:ltfla_kalman}), also with a
+        minimal number of temporaries, calls to `symm' for matrices known to be
+        symmetric ahead of time, transposition passed in to `gemm' as a flag
+        and Cholesky decomposition for multiplying by an inverse of a matrix
+
+    \item an Owl/OCaml implementation using a Cholesky decomposition
+        (Figure~\ref{fig:chol_owl_kalman}) but not taking advantage of matrices
+        known to be symmetric ahead of time, and producing a new temporary
+        matrix for every operation (including inverse and tranpose)
+
+    \item an idiomatic Owl/OCaml implmentation
+        (Figure~\ref{fig:chol_owl_kalman}) with an explicit inverse (LU
+        decomposition), not taking advantage of matrices known to be symmetric
+        ahead of time, and producing a new temporary matrix for every operation
+        (including inverse and tranpose).
+
+\end{enumerate}
+
+\begin{sidewaysfigure}
+
+    \begin{minted}[linenos, fontsize=\small]{c}
+static void kalman(
+    const int n,               const int k,                const double *sigma, /* n,n */
+    const double *h, /* k,n */ const double *mu, /* n,1 */ double *r,           /* k,k */
+    double *data,    /* k,1 */ double **ret_mu,  /* k,1 */ double **ret_sigma   /* n,n */
+) {
+        double* k_by_n = (double *) malloc(k * n * sizeof(double));
+/*20*/  cblas_dsymm(CblasRowMajor, CblasRight, CblasUpper, k, n, 1., sigma, n, h, n, 0., k_by_n, n);
+/*21*/  cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasTrans, k, k, n, 1., k_by_n, n, h, n, 1., r, k);
+/*22*/  cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans, k, 1, n, 1., h, n, mu, 1, -1., data, 1);
+/*23*/  cblas_dcopy(k * n, h, 1, k_by_n, 1);
+        double* k_by_k = (double *) malloc(k * k * sizeof(double));
+/*24*/  cblas_dcopy(k * k, r, 1, k_by_k, 1);
+/*25*/  LAPACKE_dposv(LAPACK_ROW_MAJOR, 'U', k, n, k_by_k, k, k_by_n, n);
+/*27*/  LAPACKE_dpotrs(LAPACK_ROW_MAJOR, 'U', k, 1, k_by_k, k, data, 1);
+        free(k_by_k);
+        double* n_by_n = (double *) malloc(n * n * sizeof(double));
+/*28*/  cblas_dgemm(CblasRowMajor, CblasTrans, CblasNoTrans, n, n, k, 1., h, n, k_by_n, n, 0., n_by_n, n);
+        free(k_by_n);
+        double* n_by_1 = (double *) malloc(n * sizeof(double));
+/*29*/  cblas_dgemm(CblasRowMajor, CblasTrans, CblasNoTrans, n, 1, k, 1., h, n, data, 1, 0., n_by_1, 1);
+        double* new_mu = (double *) malloc(n * sizeof(double));
+/*30*/  cblas_dcopy(n, mu, 1, new_mu, 1);
+/*31*/  cblas_dsymm(CblasRowMajor, CblasLeft, CblasUpper, n, 1, 1., sigma, n, n_by_1, 1, 1., new_mu, 1);
+        free(n_by_1);
+        double* n_by_n2 = (double *) malloc(n * n * sizeof(double));
+/*32*/  cblas_dsymm(CblasRowMajor, CblasRight, CblasUpper, n, n, 1., sigma, n, n_by_n, n, 0., n_by_n2, n);
+/*33*/  cblas_dcopy(n*n, sigma, 1, n_by_n, 1);
+/*34*/  cblas_dsymm(CblasRowMajor, CblasLeft, CblasUpper, n, n, -1., sigma, n, n_by_n2, n, 1., n_by_n, n);
+        free(n_by_n2);
+        *ret_sigma = n_by_n;
+        *ret_mu = new_mu;
+}
+    \end{minted}
+    \caption{CBLAS/LAPACKE implementation of a Kalman filter. I used C instead
+        of Fortran because it is what Owl uses under the hood and OCaml FFI
+        support for C is better and easier to use than that for Fortran. A distinct
+        `measure\_kalman' function that sandwiches a call to this funtion with
+        \texttt{getrusage} is omitted for brevity.}\label{fig:cblas_kalman}
+
+\end{sidewaysfigure}
+
+\begin{figure}[tp]
+    \begin{minted}[linenos, fontsize=\small]{ocaml}
+let potrs ~uplo a b =
+  let b = Owl.Mat.copy b in
+  Owl.Lapacke.potrs ~uplo ~a ~b
+;;
+
+let chol_kalman ~sigma ~h ~mu ~r ~data =
+  let open Owl.Mat in
+  let ( * ) = dot in
+  let h' = transpose h in
+  let sigma_h' = sigma * h' in
+  let chol = Owl.Linalg.D.chol (r + h * sigma_h') in
+  let sigma_h'_inv rest = sigma_h' * potrs ~uplo:'U' chol rest in
+  let new_sigma = sigma - sigma_h'_inv (h * sigma) in
+  let new_mu = mu + sigma_h'_inv (h * mu - data) in
+  ((sigma, (h, (mu, (r, data)))), (new_mu, new_sigma))
+;;
+
+let owl_kalman ~sigma ~h ~mu ~r ~data =
+  let open Owl.Mat in
+  let ( * ) = dot in
+  let h' = transpose h in
+  let sigma_h' = sigma * h' in
+  let x = sigma_h' * (inv @@ r + h * sigma_h') in
+  let new_mu = mu + x * (h * mu - data) in
+  let new_sigma = sigma - x * h * sigma in
+  ((sigma, (h, (mu, (r, data)))), (new_mu, new_sigma))
+;;
+    \end{minted}
+    \caption{Implementations of a Kalman filter using Owl, top one using a
+        Cholesky decomposition, bottom one using idiomatic Owl. Owl does not
+        yet provide a non-mutating `potrs' function, so I wrote my own which
+        returns a mutated copy of its argument instead.}\label{fig:chol_owl_kalman}
+
+\end{figure}
+
+I evaluated the implmentations on two metrics: memory usage (via number and
+size of temporaries allocated) and execution time. For the former, I compiled
+Owl with print-statements on the relevant primitives to see exactly the number
+of temporaries allocated for a single call of each function. While I did also
+attempt to use gperftools and OCaml's profiling support with gprof for a more
+holistic view of memory usage in the presence of OCaml's garbage-collector, I
+ran into technical difficulties irrelevant to this thesis.
+
+I measured execution time, in micro-seconds, against an exponentially (powers
+of 5) increasing scaling factor for matrix size parameters $n=5$ and $k=3$.
+For small scaling factors (1, 5, 25), I used the Core\_bench micro-benchmarking
+library, for larger factors (125 and greater), I used the \texttt{getrusage}
+system call (called \ltfla{Unix.times} in OCaml).  Core\_bench performs a
+linear-regression (here, time against batch-size) so includes 95\%
+confidence-interval with $R^2$. Larger scales have errors reported to $\pm
+\sigma$ (one standard-deviation)
+
+\section{Results}
+
+Raw output of implementation traces and the benchmarking program (including
+sample sizes) for each of the implementations are in
+Appendix~\ref{chap:eval_data}.
+
+\subsection{Memory Usage}
+
+Inspecting the Owl trace shows generated 11 temporary matrices (13 calls to
+empty, 2 of which are the resulting matrices). Chol shows 13 temporaries (same
+as Chol plus two temporaries for the two calls to potrs), whereas both LT4LA
+and CBLAS have 4 temporaries. Analysing the sub-expressions of the Owl
+implementation shows the total amount of memory allocated for temporaries is $n
++ n^2 + 4nk + 3k^2 + 2k$ words; for Chol the total is that of Owl plus $n +
+nk$; for LT4LA and CBLAS the total is $n + n^2 + nk + k^2$.
+
+\subsection{Execution Time}
+
+A graph of the execution times (with error bars which are present but quite
+small) is show in Figure~\ref{fig:timings}.
+
+\begin{figure}[tp]
+    \centering
+    \input{timings.tex}
+    \caption{Comparison of execution times. Small matrices and timings $n \le
+        5^3$ were micro-benchmarked with the Core\_bench library. Larger ones used
+    Unix's \texttt{getrusage} functionality.}\label{fig:timings}
+\end{figure}
+
+For $n=5$, and a per implementation sample size of around 450, the CBLAS
+implementation is the fastest ($24\mu s$), followed by LT4LA ($41 \mu s$), then
+Cholesky and Owl (around $52 \mu s$).  However, for $n=25$, and around 350
+measurements per implementation, the idiomatic Owl implementation is the fastest
+($95 \mu s$), \emph{then} CBLAS ($104 \mu s$), with LT4LA and Chol last (around
+$130 \mu s$). The 95\% confidence-intervals around these measurements have been
+between 0 and 2.
+
+For $n=125$, and a per implementation sample size of around 110, the trends change:
+CBLAS implmentation is now the \emph{slowest} ($1803 \mu s\, [1746,\, 1867]$).
+This is followed by LT4LA ($1678 \mu s\, [1646,\, 1714]$) and Owl ($1488 \mu
+s\, [1464,\, 1515]$). Chol had a low $R^2$ value ($0.74$) and so its execution
+time is inconclusive here.
+
+For $n=625$ and a sample size of 1000, Chol was now the fastest ($125.5 \pm
+25\,ms$), then Owl ($146 \pm 32 \,ms$), followed by LT4LA ($180.5 \pm 38 \,ms$)
+and CBLAS ($188 \pm 36 \, ms$). Despite the large sample size, the
+standard-deviation is still quite high; however, calculating the pair-wise
+$p$-values (Welch's t-test) suggests that the difference in the means is meaningful.
+
+For $n=3125$, Chol takes approximately half the execution time ($11.4 \pm 2.7
+\,s$) of the rest (differences between LT4LA, Owl and CBLAS were not
+statistically significant, $p>.05$, using Welch's t-test) around $22s$.
+
+\section{TO DO: Analysis}
+
+Having access to primitives which allow a programmer to re-use memory means
+that memory usage for temporaries in LT4LA is on par with that of C. One caveat
+is that the `free\_mat' primitive is a no-op in LT4LA, which still relies on
+OCaml's garbage-collector. The difference between these two implementations
+and the idiomatic Owl implementation is $k(3n+2k+2)$ words, and
+
+\section{TO DO: Limitations}
+
+\section{TO DO: Qualitative Benefits}

### 2018-05-24 write-up/implementation.tex
Start Evaluation

@@
 \prechapter{%
-    I implemented LT4LA in OCaml, however I strongly believe the ideas
-    described in this chapter can be applied easily to other languages and also
-    are modular enough to extend the OCaml implementation to output to different
-    back-end languages. I will show how a small core language with a few features
-    can be the target of heavy-desugaring of typical linear-algebra programs. This
-    core language can then be elaborated and checked for linearity before
-    performing some simple and predictable optimisations and emitting (in this
-    particular implementation) OCaml code that is not obviously safe and correct
-    (with respect to linearity).
+    I will describe the structure of LT4LA and explain the features of its core
+    language. I will then show how typical linear-algebra programs can be
+    elaborated into the core language and checked for linearity. Finally,
+    I will explain how such programs can be translated (in this particular
+    implementation) to OCaml code that is not obviously safe (with
+    respect to aliasing, read/write permissions and memory management).
+    Although I implemented LT4LA in OCaml, I believe the ideas described in
+    this chapter can be applied easily to other languages and also are modular
+    enough to extend the OCaml implementation to output to different back-end
+    languages.
 }%
@@
 
-Code generation is a thankfully straightforward mapping from core LT4LA
-constructs to OCaml constructs, with the appropriate sprinkling of \ltfla{Many}
-constructors for integer, element and boolean literals.
+Code generation is a straightforward mapping from core LT4LA constructs to
+OCaml constructs, with the addition of \ltfla{Many} constructors to wrap
+integer, element and boolean literals.
 
@@
 
-Even though Figure~\ref{fig:ocaml_kalman} is in OCaml, I see little reason that
-it cannot be C or Fortran. Translating LT4LA's core language constructs to
-those languages would not be as easy as translating to OCaml. However, it is
-easy to see that both Figures~\ref{fig:fortran_kalman}
+It is clear that both Figures~\ref{fig:fortran_kalman}
 and~\ref{fig:ocaml_kalman} are realisations of a concise, linearly-typed Kalman
@@
 \emph{Dune} has support for generating, compiling and linking OCaml modules at
-\emph{compile time}. Indeed, this is how I have written tests and benchmarks
-for programs produced by LT4LA from within OCaml. In particular, I can use my
-linearly-typed Kalman filter implementation just like and with any other OCaml
-function.
+\emph{compile time} (Figure~\ref{fig:build}). Indeed, this is how I have written
+tests and benchmarks for programs produced by LT4LA from within OCaml. In
+particular, I can use my linearly-typed Kalman filter implementation just like
+and with any other OCaml function.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=\textwidth]{impl_build}
+    \caption{Using LT4LA functions from OCaml with the Dune
+        build-system.}\label{fig:build}
+\end{figure}
 
@@
 capture and automatically check} aliasing, read/write permissions, memory
-allocation, re-use and deallocation of real world linear algebra library
-routines. I have also demonstrated that simple pattern-matching and desugaring
-provides the potential for a new, \emph{type-directed} approach to matrix
-expression compilation. Lastly I have shown that it is possible to use these
-features with \emph{existing} languages and frameworks.\footnote{As mentioned
-in the previous chapter, if the host language supports
-\emph{syntax-extensions}, like PPX for OCaml, it would be possible to use LT4LA
-features \emph{from within} OCaml instead of just alongside it.}
+allocation, re-use and deallocation of non-trivial linear algebra programs. I
+have also demonstrated that simple pattern-matching and desugaring provides the
+potential for a new, \emph{type-directed} approach to matrix expression
+compilation. Lastly I have shown that it is possible to use these features with
+\emph{existing} languages and frameworks.\footnote{As mentioned in the previous
+    chapter, if the host language supports \emph{syntax-extensions}, like PPX
+for OCaml, it is possible to construct LT4LA expressions \emph{from within} the
+host language.}

### 2018-05-24 write-up/abstract.tex
Finish Evaluation

@@
 allocation) which (1) are more memory-efficient than equivalent programs
-written using purely high-level linear algebra libraries and (2) perform just
+written using high-level linear algebra libraries and (2) perform just
 as predictably as equivalent programs written using low-level linear-algebra
 libraries. I also argue \emph{the experience} of writing such programs with
-linear types is ``better'' in key-respects. In addition to all of this, I show
-that it is possible to provide such features \emph{as a library} on top of
-existing programming languages and linear algebra libraries.
+linear types is qualitatively better in key-respects. In addition to all of
+this, I show that it is possible to provide such features \emph{as a library}
+on top of existing programming languages and linear algebra libraries.
 

### 2018-05-24 write-up/evaluation.tex
Finish Evaluation

@@
     deallocation) and explicit (with respect to memory allocation) code that
-    (1) is more memory-efficient than code written using purely high-level
+    (1) is more memory-efficient than code written using high-level
     linear algebra libraries and (2) performs just as predictably as code
     written using low-level linear-algebra libraries. I will also elaborate on
-    the qualitative benefits of using linearly types to write linear-algebra
+    the qualitative benefits of using linear types to write linear-algebra
     programs.
@@
 
-I analysed four implmentations of a Kalman filter, increasing
+I analysed four implementations of a Kalman filter, increasing
 the abstraction level at each step:
@@
         known to be symmetric ahead of time, and producing a new temporary
-        matrix for every operation (including inverse and tranpose)
+        matrix for every operation (including inverse and transpose)
 
-    \item an idiomatic Owl/OCaml implmentation
+    \item an idiomatic Owl/OCaml implementation
         (Figure~\ref{fig:chol_owl_kalman}) with an explicit inverse (LU
@@
         ahead of time, and producing a new temporary matrix for every operation
-        (including inverse and tranpose).
+        (including inverse and transpose).
 
@@
         support for C is better and easier to use than that for Fortran. A distinct
-        `measure\_kalman' function that sandwiches a call to this funtion with
+        `measure\_kalman' function that sandwiches a call to this function with
         \texttt{getrusage} is omitted for brevity.}\label{fig:cblas_kalman}
@@
 
-I evaluated the implmentations on two metrics: memory usage (via number and
+I evaluated the implementations on two metrics: memory usage (via number and
 size of temporaries allocated) and execution time. For the former, I compiled
@@
 library, for larger factors (125 and greater), I used the \texttt{getrusage}
-system call (called \ltfla{Unix.times} in OCaml).  Core\_bench performs a
-linear-regression (here, time against batch-size) so includes 95\%
-confidence-interval with $R^2$. Larger scales have errors reported to $\pm
-\sigma$ (one standard-deviation)
+system call (called \ltfla{Unix.times} in OCaml), sandwiched between calls to
+\ltfla{Gc.full_major} to ensure no garbage collection took place during the
+measurements.  Core\_bench performs a linear-regression (here, time against
+batch-size) so includes 95\% confidence-interval with $R^2$. Larger scales have
+errors reported to $\pm \sigma$ (one standard-deviation)
 
@@
 
-Raw output of implementation traces and the benchmarking program (including
-sample sizes) for each of the implementations are in
-Appendix~\ref{chap:eval_data}.
+All implementations produce the same answers (within at most $2^{-52}$). Raw
+output of implementation traces and the benchmarking program (including sample
+sizes) for each of the implementations are in Appendix~\ref{chap:eval_data}.
 
@@
         5^3$ were micro-benchmarked with the Core\_bench library. Larger ones used
-    Unix's \texttt{getrusage} functionality.}\label{fig:timings}
+        Unix's \texttt{getrusage} functionality, sandwiched between calls to
+        \ltfla{Gc.full_major} for the OCaml implementations.}\label{fig:timings}
+
 \end{figure}
@@
 ($95 \mu s$), \emph{then} CBLAS ($104 \mu s$), with LT4LA and Chol last (around
-$130 \mu s$). The 95\% confidence-intervals around these measurements have been
+$130 \mu s$). The 95\% confidence-intervals around these measurements are
 between 0 and 2.
@@
 For $n=125$, and a per implementation sample size of around 110, the trends change:
-CBLAS implmentation is now the \emph{slowest} ($1803 \mu s\, [1746,\, 1867]$).
+CBLAS implementation is now the \emph{slowest} ($1803 \mu s\, [1746,\, 1867]$).
 This is followed by LT4LA ($1678 \mu s\, [1646,\, 1714]$) and Owl ($1488 \mu
@@
 and CBLAS ($188 \pm 36 \, ms$). Despite the large sample size, the
-standard-deviation is still quite high; however, calculating the pair-wise
-$p$-values (Welch's t-test) suggests that the difference in the means is meaningful.
+standard-deviation is still quite high; however, \emph{because} of the large
+sample size, the pair-wise $p$-values (Welch's t-test) are very small ($p <
+.05$), suggesting that the difference in the means is statistically
+significant.
 
-For $n=3125$, Chol takes approximately half the execution time ($11.4 \pm 2.7
-\,s$) of the rest (differences between LT4LA, Owl and CBLAS were not
-statistically significant, $p>.05$, using Welch's t-test) around $22s$.
+For $n=3125$ and a sample size of 15, Chol and Owl had similar mean execution
+times ($11.2 \pm 0.85 \,s$ and $12.1 \pm 0.47 \,s$ respectively) while LT4LA
+and CBLAS had similar, but \emph{slower} mean execution times ($16.1 \pm 0.19
+\,s$ and $15.7 \pm 0.53 \,s$ respectively). Pair-wise $p$-values (Welch's
+t-test) are all less than $.05$, so the differences in means are statistically
+significant.
 
-\section{TO DO: Analysis}
+\section{Analysis}
 
 Having access to primitives which allow a programmer to re-use memory means
-that memory usage for temporaries in LT4LA is on par with that of C. One caveat
-is that the `free\_mat' primitive is a no-op in LT4LA, which still relies on
-OCaml's garbage-collector. The difference between these two implementations
-and the idiomatic Owl implementation is $k(3n+2k+2)$ words, and
+that memory usage for temporaries in LT4LA is on par with that of CBLAS. One
+caveat is that the \ltfla{freeM} primitive is a no-op in LT4LA, so deallocation
+still relies on OCaml's garbage-collector. The difference between these two
+implementations and the idiomatic Owl implementation is $k(3n+2k+2)$ words.
 
-\section{TO DO: Limitations}
+For small matrix sizes, LT4LA and CBLAS execution times differ. I suspect this
+is due to large sample sizes causing more allocations and thus more
+garbage-collector unpredictability (which the linear-regression \emph{did not}
+take into account; multi-variate regressions are not fully supported
+Core\_bench yet).
 
-\section{TO DO: Qualitative Benefits}
+As matrix size increases, execution times of LT4LA become very similar to those
+of CBLAS. However, that does not necessarily mean faster.  Especially for
+$n=625$ and $n=3125$ (sizes for which \ltfla{Gc.full_major} was called before
+each measurement), mean execution times start to split clearly into two groups:
+Chol and Owl as one group (similar, faster times) and LT4LA and CBLAS as
+another (similar, slower times).
+
+Overall, as far as memory usage is concerned, it is possible to write linear
+algebra programs with the safety (with respect to aliasing and read/write
+permissions) of using high-level libraries whilst gaining the explicit control
+over memory allocation and re-use associated with low-level libraries.
+As far as execution times are concerned, even though the idiomatic OCaml
+implementations seem to be faster for this particular
+benchmark,\footnote{Attempting to understand \emph{why} the
+    Chol/Owl implementations tended to be faster for anything except the
+    smallest of matrices, despite allocating almost twice the number of
+    temporaries was difficult. Running the different implementations through
+    the Cachegrind cache simulator (part of Valgrind) showed that LT4LA/CBLAS
+    had a roughly 1\% higher cache miss rate than Chol/Owl (rising from around
+    11\% to around 12\%). This seemed insufficient to account for the
+    differences, so I investigated further. I added an extra, modified
+    implementation of LT4LA, which transposed the `h' parameter into a new
+    matrix rather than using the transpose flag to `gemm', to see if
+    row-vs-column access patterns could account for the differences (they did
+    not). Had I been able to use gprof or gperftools, I would have profiled the
+    remaining key difference: the `symm' routine (which performs \emph{half} as
+    many multiplications as `gemm' because it knows one of its
+    arguments is symmetric).}
+we can conclude that LT4LA performs similarly to using lower-level libraries
+when garbage-collection is controlled for and especially as matrix sizes increase.
+
+\section{Limitations}
+
+I chose the example of a Kalman filter because it is used in the real world,
+consists purely of a sequence of matrix expressions and produces many
+unnecessary temporary matrices when implemented idiomatically in a high-level
+linear-algebra library. It is good for isolating the key differences between
+not having and having linear types to help a programmer safely manage memory
+and aliasing, whilst excluding other aspects also important to real world
+linear-algebra programs such as control flow or blocking.
+
+\section{Qualitative Benefits}
+
+Prior to this project, I had no experience with linear-algebra libraries or the
+problem of matrix expression compilation. As such, I based my initial LT4LA
+implementation of a Kalman filter using BLAS and LAPACK, on a popular GitHub
+gist of a Fortran implementation, one that was automatically generated from
+SymPy's matrix expression compiler~\cite{rocklin_thesis}.
+
+Once I translated the implementation, I attempted to compile it and found that
+(to my surprise) it did not type-check. This was because the original
+implementation contained incorrect aliasing, did not adhere to Fortran's
+read/write permissions (with respect to \texttt{intent} annotations
+\texttt{in}, \texttt{out} and \texttt{inout}) and unused and unnecessary
+temporaries, all of which were now highlighted by LT4LA's type system.
+
+The original implementation used 6 temporaries, one of which was immediately
+spotted as not being used due to linearity. It also contained two variables
+which were marked as \texttt{intent(in)} but would have been written over by
+calls to `gemm', spotted by the fractional-capabilities feature. Furthermore,
+it used a matrix \emph{twice} in a call to `symm', once with a read permission
+but once with a \emph{write} permission.  Fortran assumes that any parameter
+being written to is not aliased and so this call was not only incorrect, but
+illegal according to the standard, both aspects of which were captured by
+linearity and fractional-capabilities. Lastly, it contained another unnecessary
+temporary, however one that was not obvious without linear types. To spot it, I
+first performed live-range splitting (checked by linearity) by hoisting calls
+to \ltfla{freeM} and then annotated the freed matrices with their dimensions.
+After doing so and spotting two disjoint live-ranges of the same size, I
+replaced a call to \ltfla{freeM} followed by allocating call to \ltfla{copy}
+with one, in-place call to \ltfla{copyM_to}.
+
+\section{Summary}
+
+Writing a linear-algebra program using LT4LA gives a programmer both
+\emph{explicit control} (over read/write permissions, aliasing and memory
+allocation, re-use and deallocation) and \emph{signals intent} to the compiler
+and reader clearly without sacrificing either of readability or
+value-semantics. The resulting programs perform similarly to equivalent
+programs written directly using lower-level linear-algebra libraries. In turn,
+the programs written can be \emph{checked automatically} against their intent,
+especially when \emph{refactoring} or \emph{rewriting} code. Although any
+expert \emph{could} have followed the same line of reasoning laid out above,
+and arrived at the same program, LT4LA's type system enables a non-expert
+(yours truly) to do the same and increase confidence in the result\footnote{As
+always, this does not preclude testing code by actually executing it, but
+definitely \emph{complements} it.} by checking said line of reasoning.

### 2018-05-24 write-up/implementation.tex
Finish Evaluation

@@
 Copying leads us into dealing with temporaries: do you first allocate all
-temporaries in new matrices and then analyse lifetimes and dimensions to figure
-out which slots of memory can be re-used? Or do you try and infer the available
-resources from the environment as you go?  Can, and should, you type arrays and
-matrices (or $n$-dimensional tensors) as the same?
+temporaries in new matrices (SSA-style) and then analyse dimensions to figure
+out which slots of memory can be re-used via copy coalescing? Or do you try and
+infer the live ranges of available resources from the environment as you go?
+Can, and should, you type arrays and matrices (or $n$-dimensional tensors) as
+the same?
 
 Adding in more and more considerations, the problem starts to resemble register
-allocation: there are registers of different types and sizes, many instructions
-to choose from, a cost model to take into account all whilst trying to balance
-the number of registers in use and the number of instructions emitted.
+allocation: there are registers of different types and sizes, many
+(non-orthogonal) instructions to choose from, a cost model to take into account
+all whilst trying to balance the number of registers in use and the number of
+instructions emitted.
 
@@
 code, the question then becomes, how to use this code. This process will vary
-across ecosystems, but within OCaml, the new build-system on the block
+across ecosystems, but within OCaml, the new build system on the block
 \emph{Dune} has support for generating, compiling and linking OCaml modules at
@@
     \caption{Using LT4LA functions from OCaml with the Dune
-        build-system.}\label{fig:build}
+        build system.}\label{fig:build}
 \end{figure}

### 2018-05-27 write-up/abstract.tex
Finish Conclusion

--- 2018-05-22 write-up/abstract.tex
@@
 \newpage
-{\normalfont\Huge\sffamily\bfseries TO DO: Abstract}
+{\normalfont\Huge\sffamily\bfseries Abstract}
 \vspace{24pt}
 
-In this thesis, I will argue that linear types are an appropriate,
-\emph{type-based formalism} for expressing aliasing, read/write permissions,
-memory allocation, re-use and deallocation, first, in the context of the APIs
-of linear algebra libraries, consequently extending these principles to the
-problem of \emph{efficient} matrix-expression compilation.
+In this thesis, I argue that linear types are an appropriate, \emph{type-based
+formalism} for expressing aliasing, read/write permissions, memory allocation,
+re-use and deallocation, first, in the context of the APIs of linear-algebra
+libraries and then in the context of matrix-expression compilation.
 
-I will show that framing the problem using linear types can help \emph{reduce
-bugs} by making precise and explicit the informal, ad-hoc practices typically
-employed by human experts and linear algebra \emph{compilers} and automate
+I show that framing the problem using linear types can help \emph{reduce bugs}
+by making precise and explicit the informal, ad-hoc practices typically
+employed by human experts and linear-algebra \emph{compilers} and automate
 checking them.
 
-As evidence for this argument, I will show programs written with this safety,
-precision and explicitness (1) can be just as pleasant and convenient for a
-programmer as less efficient, but higher-level linear algebra libraries and (2)
-perform just as \emph{efficiently and predictably} as lower-level, less
-readable and more error-prone interfaces linear algebra libraries. I will
-further show that support for such programs can be implemented \emph{as
-libraries} on top of existing programming languages and libraries.
-
-\textbf{Summarise evaluation.}
-
-\textbf{Summary future work.}
+As evidence for this argument, I show non-trivial, yet readable, linear-algebra
+programs, that are safe and explicit (with respect to aliasing, read/write
+permissions, memory allocation, re-use and deallocation) which (1) are more
+memory-efficient than equivalent programs written using high-level
+linear-algebra libraries and (2) perform just as predictably as equivalent
+programs written using low-level linear-algebra libraries. I also argue
+\emph{the experience} of writing such programs with linear types is
+qualitatively better in key-respects. In addition to all of this, I show that
+it is possible to provide such features \emph{as a library} on top of existing
+programming languages and linear-algebra libraries.
 

### 2018-05-27 write-up/background.tex
Finish Conclusion

@@
     libraries. I will emphasise \emph{practical} and intuitive explanations of
-    linear types to keep this thesis accessible to working programmers as well
-    as academics not familiar with type-theory and will give only a terse
-    overview of the history and theory behind linear types for the interested
-    reader to pursue further.
+    linear types to keep this thesis accessible to programmers as well as
+    academics not familiar with type-theory and will give only a terse overview
+    of the history and theory behind linear types for the interested reader to
+    pursue further.
 }%
@@
 \section{Tracking Resources with Linearity}
-Familiar examples of using a type-system to express program-invariants are
+Familiar examples of using a type-system to express program invariants are
 existential-types for abstraction and encapsulation, polymorphic types for
@@
 stronger the rules you follow, the better the guarantees you can get about your
-program, \emph{before} you run it. At first, the rules seems restrictive, but
+program, \emph{before} you run it. At first, the rules seem restrictive, but
 similar to how the rules of grammar, spelling and more generally writing help a
@@
 practical programming terms, they enforce the restriction that a value may be
-used exactly once.\footnote{This definition may differ from more colloquial
-uses in discussions surrounding \emph{substructural} type systems and/or
-Rust.} While this restriction may seem limiting at first, precisely these
-constraints can be used to express common invariants of the programs written by
-working programmers every day. For example: a file or a socket, once opened
-\emph{must} closed; all memory that is manually allocated \emph{must} be freed.
-C++'s destructors and Rust's Drop-trait (and more generally, its
-borrow-checker) attempt to enforce these constraints by basically doing the
-same thing: any resource that has not been moved is deallocated at the end of
-the current lexical scope. Notably, these languages also permit aliasing,
-alongside rules enforcing when it is acceptable to do so.  On face value, the
-above one-line description of linear types prevents aliasing or functions such
-as $\lambda x.\;x \times x$, such features are still allowed (albeit in a more
-restricted fashion) in a \emph{usable} linear type system designed for working
-with linear-algebra libraries.
+used exactly once\footnote{This definition may differ from more colloquial uses
+in discussions surrounding \emph{substructural} type systems and/or Rust.}
+(Figure~\ref{fig:example_rules} shows two example typing rules for a typical,
+linearly typed language). While this restriction may seem limiting at first,
+precisely these constraints can be used to express common invariants of the
+programs written by programmers every day. For example: a file or a
+socket, once opened \emph{must} closed; all memory that is manually allocated
+\emph{must} be freed.  C++'s destructors and Rust's Drop-trait (and more
+generally, its borrow-checker) attempt to enforce these constraints by
+basically doing the same thing: any resource that has not been moved is
+deallocated at the end of the current lexical scope. Notably, these languages
+also permit aliasing, alongside rules enforcing when it is acceptable to do so.
+On face value, the above one-line description of linear types prevents aliasing
+or functions such as $\lambda x.\; (x, x)$; such features are still allowed
+(albeit in a more restricted fashion) in a \emph{usable} linear type system
+designed for working with linear-algebra libraries.
+
+\begin{figure}
+    \centering
+    \begin{minipage}{0.3\textwidth}
+        \centering
+    \begin{prooftree}
+        \AxiomC{}
+        \RightLabel{\textsc{Var}}
+        \UnaryInfC{$\Gamma, x : t \vdash x : t ; \Gamma$}
+    \end{prooftree}
+    \end{minipage}%
+    \begin{minipage}{0.6\textwidth}
+        \centering
+    \begin{prooftree}
+        \AxiomC{$\Gamma_1 \vdash e_1 : t_1 ; \Gamma_2$}
+        \AxiomC{$\Gamma_2 \vdash e_2 : t_2 ; \Gamma_3$}
+        \RightLabel{\textsc{Pair}}
+        \BinaryInfC{$\Gamma_1 \vdash (e_1, e_2) : t_1 \otimes t_2 ; \Gamma_3$}
+    \end{prooftree}
+    \end{minipage}
+    \caption{A typing rule has the general form $\Gamma_{\textrm{in}} \vdash e ; \Gamma_{\textrm{out}}$.
+    Typing a variable \emph{removes} it from the environment. Typing a pair
+    requires that each component of the pair uses \emph{different} resources
+    from the environment.}\label{fig:example_rules}
+\end{figure}
 
 \section{Problem in Detail}
+
 Given this background, the most pertinent question at hand is: what problems do
@@
     \centering
+    \begin{minipage}{0.45\textwidth}
+    \centering
     \begin{minted}[fontsize=\small]{python}
@@
     \end{minted}
+    \end{minipage}%
+    \begin{minipage}{0.45\textwidth}
+    \centering
     \begin{minted}[fontsize=\small]{ocaml}
-
 (* Owl (OCaml) *)
 open Owl
-let a = Mat.of_arrays [|[|1.;0.|];[|0.;1.|]|]
-let b = Mat.of_array [|[|4.;1.|];[|1.;2.|]|]
+let a = Mat.of_arrays
+  [|[|1.;0.|];[|0.;1.|]|]
+let b = Mat.of_arrays
+  [|[|4.;1.|];[|1.;2.|]|]
 let c = Mat.(a *@ b)
     \end{minted}
+    \end{minipage}
     \caption{Matrix Multiplication in Numpy (Python), Julia and Owl (OCaml).}\label{fig:mat_mul_copies}
@@
 integers and floating-point numbers. This approach confers two key advantages to the
-programmmer: it is easy to read (equational and algebraic declarations) and it
+programmmer: it is easy to read (equational and algebraic expressions) and it
 is easy to reason about (as one would with a mathematical formula). Although this
@@
 escape-hatches (typically provided by most libraries, which permit in-place
-modification of objects) to build faster, and more efficient algorithms which
-are harder to reason about.
+modification of objects) to build faster, and more memory-efficient algorithms
+which are harder to reason about.
 
@@
 The title of this subsection\footnote{I Have No Idea What Those Letters Mean.}
-illustrates the problem with the C++/FORTRAN side: legibility (and ease of
-reasoning) is sacrificed at the altar of performance and efficiency.
+illustrates the problem with the C++/FORTRAN side: readability (equational
+and algebraic expressions) and ease of reasoning is sacrificed at
+the altar of performance and efficiency.
 
@@
 factors such as memory layout, allocation, re-use as well as cache behaviour
-and parallelism become apparent. Of these, memory allocation and re-use are
-of most relevance to linear-types and this thesis.
+and parallelism become apparent. Of these, \textbf{memory allocation and
+re-use} are relevant to linear-types and this thesis.
 
@@
 
-\section{Proposed Solution}
+\section{Proposed Solution}\label{sec:prop_sol}
 
-My proposed solution to this dichotomy is a \emph{domain-specific language}
-(DSL), called LT4LA (Linear Types for Linear Algebra). It offers readable,
-explicit management of aliasing, read/write permissions, memory allocation,
-re-use and deallocation all with automated checking: offering a safety net to
-catch the baby whilst swiftly disposing of the bath water.  Although for
-expository and testing purposes I have defined a concrete-syntax, a full
-implementation would make use of a language's syntax-extension features (such
-as PPX for OCaml) to \emph{embed} the DSL into the host language. Such an
-embedding is straightforward but fairly tedious to implement. As a half-way
-point, I used compile-time code generation to make the DSL's output available
-to OCaml for testing and evaluation.
+My proposed solution to this dichotomy (readability, ease of reasoning and
+safety versus memory-efficiency) is a \emph{domain-specific language} (DSL),
+called LT4LA (Linear Types for Linear Algebra). It is written in OCaml and
+transpiles to OCaml. It offers readable, explicit management of aliasing,
+read/write permissions, memory allocation, re-use and deallocation all with
+automated checking: offering a safety net to catch the baby whilst swiftly
+disposing of the bath water.  Although for expository and testing purposes I
+have defined a concrete-syntax, a full implementation would make use of a
+language's syntax-extension features (such as PPX for OCaml) to \emph{embed}
+the DSL into the host language. Such an embedding is straightforward but fairly
+tedious to implement. As a half-way point, I used compile-time code generation
+to make the DSL's output available to OCaml for testing and evaluation.
 
@@
 
-\begin{figure}[tp]
+\begin{figure}[p]
     \begin{minted}[linenos, firstnumber=7, fontsize=\small]{ocaml}
@@
 Now suppose we were trying to square a matrix, using a `dgemm' like BLAS
-routine which takes two read-only matrices and a third matrix it can write to
-and performs $C := AB + C$. How would we use such a routine to square a matrix?
-Surely this would break linearity, since $A$ would be have to be given to the
-function twice, and we can only do so once?
+routine (called `simple\_dgemm'\footnote{For the purposes of this example, I
+assume `simple\_dgemm' has type \ltfla{'x . 'x mat --o 'y .  'y mat --o z mat
+--o ('x mat * 'y mat) * z mat}.} in Figures~\ref{fig:ltfla_square}
+and~\ref{fig:ltfla_free}) which takes two read-only matrices and a third matrix
+it can write to and performs $C := AB + C$. How would we use such a routine to
+square a matrix, assuming $C=0$ initially?  Surely this would break linearity,
+since $A$ would be have to be given to the function twice, and we can only do
+so once?
 
-We can use a special primitive called \ltfla{share} to produce read-only
-\emph{aliases} of any matrix (including ones we cannot write to). We then pass
-this into the function, and it works as expected. Once the squared matrix has
-been obtained, we may not want the original any more, and thus decide to free
-it. Before we do so, we must first \ltfla{unshare} any read-only aliases that
-exist -- in this context, \ltfla{unshare} swallows them (the only primitive
-which can return fewer read-only values than it receives) and gives back a
-single, read-write handle.
+We can use a special primitive called \ltfla{share} to produce (more) read-only
+\emph{aliases} of any matrix. We then pass these into the function, and it
+works as expected. Once the squared matrix has been obtained, we may not want
+the original any more, and thus decide to free it. Before we do so, we must
+first \ltfla{unshare} any read-only aliases that exist -- in this context,
+\ltfla{unshare} swallows them and gives back a single, read-write handle.
 
-\begin{figure}[tp]
+\begin{figure}[p]
     \begin{minted}[linenos, fontsize=\small]{ocaml}
@@
     \end{minted}
-    \caption{Squaring a matrix in LT4LA.}\label{fig:ltfla_square}
+    \caption{Squaring a matrix in LT4LA (assuming \ltfla{c} is initially
+        0).}\label{fig:ltfla_square}
 \end{figure}
@@
 If we tried to free one of the read-only aliases instead of
-\ltfla{unshare}-ing, then we would get the error show in
+\ltfla{unshare}-ing, then we would get the error shown in
 Figure~\ref{fig:ltfla_free}. Briefly, a \ltfla{z arr} is a read-write array,
@@
 
-\begin{figure}[tp]
+Another way in which `simple\_dgemm' may be misused is by instantiating its
+informal description of $C := AB + C$ with $B = C = A$ and so mistakenly
+concluding that it computes $A := A^2 + A$ in-place. However, the type of
+\ltfla{share} prevents this as well -- \ltfla{let (a11, a12) = share _ a1 in
+simple_dgemm _ a11 _ a12 a2} would result in an error similar to the one shown
+in Figure~\ref{fig:ltfla_free}.
+
+\begin{figure}[p]
     \begin{minted}[linenos, fontsize=\small]{ocaml}
@@
 I have given an \emph{intuitive} exposition of linear types with
-\emph{fractional-capabilities}, emphasising small, but illustrative and
-practical code examples, leaving \emph{what} is going on and \emph{why} it
-works as details for the next chapter. I have shown that it is possible to have
-high-level, explicit and automatically checked management of aliasing,
-read/write permissions, memory allocation, re-use and deallocation with a
-\emph{type system} -- preventing a whole class of errors that can occur with
-lower-level languages \emph{at compile time}, using the of context linear
-algebra libraries as as a specific example. In the next chapter, I will show
-that these features can be expressed and implemented in a language-independent
-manner to be provided and used \emph{as a library}.
+\emph{fractional-capabilities}~\cite{boyland}, emphasising small, but
+illustrative and practical code examples, leaving \emph{what} is going on and
+\emph{why} it works as details for the next chapter. I have shown that it is
+possible to solve the dichotomy of readability, ease of reasoning and safety
+versus memory-efficiency by providing explicit and automatically checked
+management of aliasing, read/write permissions, memory allocation, re-use and
+deallocation with a \emph{type system}. This prevents a whole class of errors
+that can occur with lower-level languages \emph{at compile time}. I
+demonstrated these features using the context of linear-algebra libraries as as
+a specific example. In the next chapter, I will show that these features can be
+further applied to the problem of \emph{matrix expression compilation}. I will
+also explain how to express and implement them so that they can be provided and
+used \emph{as a library}.

### 2018-05-27 write-up/conclusion.tex
Finish Background, continue Implementation
Start Evaluation
Finish Evaluation
Finish Conclusion

@@
-\chapter{TO DO: Conclusion}
+\chapter{Conclusion}
 
-\begin{guidance}
-    As you might imagine: summarizes the dissertation, and draws
-    any conclusions. Depending on the length of your work, and
-    how well you write, you may not need a summary here.
+% \begin{guidance}
+%     As you might imagine: summarizes the dissertation, and draws
+%     any conclusions. Depending on the length of your work, and
+%     how well you write, you may not need a summary here.
+%
+%     You will generally want to draw some conclusions, and point
+%     to potential future work.
+% \end{guidance}
 
-    You will generally want to draw some conclusions, and point
-    to potential future work.
-\end{guidance}
+In this thesis, I presented \textbf{linear types with fractional-capabilities}
+as a type-based formalism for expressing \textbf{aliasing, read/write
+permissions, memory allocation, re-use and deallocation}; I provided a detailed
+description of each feature of the type-system, its intent and implementation
+in Section~\ref{sec:core_lang}.
 
-\textbf{What are the benefits of the type-based approach I have taken? Have I successfully 
-argued whatever I wrote in the introduction?}
+I used that formalism to make \textbf{precise and explicit the intensional
+behaviour and assumptions} (about aliasing, read/write permissions and memory
+re-use) of linear-algebra library APIs, first with the simplified example of
+the fictional `simple\_dgemm' primitive in Section~\ref{sec:prop_sol} -- where
+I showed how \textbf{the types made it impossible to use `simple\_dgemm'
+incorrectly} -- and then, more realistically with the types of LT4LA primitives
+(including actual CBLAS/LAPACKE routines) listed in
+Appendix~\ref{chap:primitives}.
+
+I also used that formalism to implement a simple, yet effective matrix
+expression `compiler' based on pattern-matching. I was able to use to that
+compiler to write, a \textbf{non-trivial, yet readable} (equational and
+algebraic) linear-algebra program (Kalman filter, as described in
+Section~\ref{sec:matrix_exps}). I shared how, in doing so, I could use the
+type-checking as a tool to \textbf{automatically highlight bugs} regarding
+aliasing, read/write permissions and memory re-use of a reference
+implementation (Section~\ref{sec:qual_benefits}) and the matrix expression
+compiler that generated it (Section~\ref{sec:mat_exp_comp}).
+
+% This suggests that although it is tempting to think of matrix expression
+% compilation as a data-flow problem in terms of graphs, such an approach may
+% struggle to accurately capture the intensional behaviour and assumptions of
+% linear-algebra kernels which seem to be expressed nicely with linear types and
+% fractional-capabilities.
+
+I showed that this implementation was \textbf{more memory-efficient} than one
+written using a high-level linear-algebra libraries and performed just as
+\textbf{predictably} as one written using a low-level linear-algebra library
+(Chapter~\ref{chap:eval}). The type system ensured that the implementation was
+safe and explicit with respect to aliasing, read/write permissions, memory
+allocation, re-use and deallocation.
+
+Lastly, I was able to provide all of this functionality as a \textbf{usable
+OCaml library}, by generating \textbf{readable and not obviously safe} OCaml
+code (1) which looks like it is written by a human/expert and (2) whose types
+reflect its intensional behaviour and assumptions (interpreting types into
+OCaml, Figure~\ref{fig:type_grammar}). This library, LT4LA, produces
+\textbf{helpful type-errors} (Section~\ref{sec:prop_sol}), comes with an
+interactive REPL and is documented and well-tested
+(Section~\ref{sec:structure_lt4la}).
 
@@
 
-\textbf{Graph stuff... check with supervisors.}
+LT4LA is a proof-of-concept design for linearly-typed DSL for writing
+linear-algebra programs in a way that can work with existing languages and
+libraries. It is made of up of the core language and its matrix expression
+`compiler'; both could be extended in many ways.
 
-All of the matrix expression compilers mentioned in previous section construct
-some sort of dataflow graph to represent the computation being executed. While
-this seems intuitive, there is no formal argument for this  appr
-Some directions in which a type-based approach to efficient matrix expression compilation could
-be taken are:
+\subsection{Core Language}
+
+Although I use well-established type system features, it may be worthwhile to
+formally state and verify my claimed safety properties.  The logical conclusion
+of LT4LA's core language would be a language expressive enough to provide a
+usable, linearly-typed interface to the unsafe parts of the Owl library for
+OCaml, perhaps as a functor like its current, `Lazy' interface.  To do that,
+the language implementation would need to support more than just 64-bit
+floating-point numbers for elements (for example, 16-bit, 32-bit or complex),
+polymorphism, a unified way of dealing with arrays and matrices (perhaps
+through sub-typing). It may also be useful to explore localised linearity
+\emph{inference}, to automate the reasoning I did manually when implementing a
+memory-efficient Kalman filter.
+
+Another, equally valid but opposing direction may be to decrease the amount of
+expressivity, to bring it closer to Fortran and C so that it would be easier to
+emit code in those languages. A full set of linearly-typed BLAS/LAPACK bindings
+(to encode information in documentation at the type level) would be a good
+first step, perhaps leading into more research on formal semantics for Fortran.
+
+\subsection{Matrix Expression Compilation}
+
+All of the matrix expression compilers mentioned in the previous chapter
+construct some sort of data-flow graph to represent the computation being
+executed. While this seems intuitive, there is no formal argument for this
+approach.  Some directions in which a type-based approach to efficient matrix
+expression compilation could be taken are:
 

### 2018-05-27 write-up/evaluation.tex
Finish Conclusion

--- 2018-04-19 write-up/evaluation.tex
@@
-\chapter{Evaluation}
+\chapter{Evaluation}\label{chap:eval}
 
-\begin{guidance}
-    For any practical projects, you should almost certainly have
-    some kind of evaluation, and it's often useful to separate
-    this out into its own chapter.
-\end{guidance}
+% Least Squares Linear Regression
+% p73 onwards https://people.cs.uchicago.edu/~mrocklin/storage/dissertation.pdf
+
+% \begin{guidance}
+%     For any practical projects, you should almost certainly have
+%     some kind of evaluation, and it's often useful to separate
+%     this out into its own chapter.
+% \end{guidance}
 
 \prechapter{%
-    In this chapter, I will argue the central premise of this thesis: linear
-    types are a practical and usable tool to help working programmers
-    write code that is both more legible and less resource-hungry than with
-    existing linear-algebra frameworks. My project illustrates how to do so in
-    a way that can be implemented as a usable \emph{library} for existing
-    languages and frameworks that leverages the already impressive amount of
-    work gone into optimising them so far.
+    I will evaluate the central premise of this thesis: linear types are a
+    practical and usable tool to help programmers write readable (equational
+    and algebraic), safe (with respect to aliasing, read/write permissions,
+    memory re-use and deallocation) and explicit (with respect to memory
+    allocation) code that (1) is more memory-efficient than code written using
+    high-level linear algebra libraries and (2) performs just as predictably as
+    code written using low-level linear-algebra libraries. I will also
+    elaborate on the qualitative benefits of using linear types to write
+    linear-algebra programs.
 }%
 
-\section{Against Overly-Safe Copying}
-% Numpy, Julia, Owl
+\section{Compared to C}
 
-\section{Against Poor Legibility}
-% C++, Fortran
+To test whether code written using LT4LA performs just as predictably as code
+written using low-level linear-algebra libraries, I wrote two implementations
+of a Kalman filter:
+
+\begin{enumerate}
+
+    \item A CBLAS/LAPACKE implementation, written in C
+        (Figure~\ref{fig:cblas_kalman}), with a minimal number of temporaries,
+        calls to `symm' for matrices known to be symmetric ahead of time,
+        transposition passed in to `gemm' as a flag and Cholesky decomposition
+        for multiplying by an inverse of a matrix.
+
+    \item A LT4LA implementation (Figure~\ref{fig:ltfla_kalman}), also with a
+        minimal number of temporaries, calls to `symm' for matrices known to be
+        symmetric ahead of time, transposition passed in to `gemm' as a flag
+        and Cholesky decomposition for multiplying by an inverse of a matrix.
+
+\end{enumerate}
+
+Both implementations produce the same answers (within at most $2^{-52}$). Raw
+output of implementation traces and the benchmarking program (including sample
+sizes) are in Appendix~\ref{chap:eval_data}.
+
+I compared the implementations on two metrics: memory usage (via number and
+size of temporaries allocated) and execution time. For the former, I compiled
+Owl with print-statements on the relevant primitives to see exactly the number
+of temporaries allocated for a single call of each function. While I did also
+attempt to use gperftools and OCaml's profiling support with gprof for a more
+holistic view of memory usage in the presence of OCaml's garbage-collector, I
+ran into technical difficulties irrelevant to this thesis.
+
+I measured execution time, in micro-seconds, against an exponentially (powers
+of 5) increasing scaling factor for matrix size parameters $n=5$ and $k=3$.
+For small scaling factors (1, 5, 25), I used the Core\_bench micro-benchmarking
+library, for larger factors (125 and greater), I used the \texttt{getrusage}
+system call (called \ltfla{Unix.times} in OCaml), sandwiched between calls to
+\ltfla{Gc.full_major} to ensure no garbage collection took place during the
+measurements.  Core\_bench performs a linear-regression (here, time against
+batch-size) so includes 95\% confidence-interval with $R^2$. Larger scales have
+errors reported to $\pm \sigma$ (one standard-deviation).
+
+\begin{landscape}
+\begin{figure}[p]
+    \centering
+    \begin{minipage}{1.45\textheight}
+    \begin{minted}[linenos, fontsize=\small]{c}
+static void kalman( const int n,               const int k,                const double *sigma, /* n,n */
+                    const double *h, /* k,n */ const double *mu, /* n,1 */ double *r,           /* k,k */
+                    double *data,    /* k,1 */ double **ret_mu,  /* k,1 */ double **ret_sigma   /* n,n */ ) {
+        double* k_by_n = (double *) malloc(k * n * sizeof(double));
+/*16*/  cblas_dsymm(CblasRowMajor, CblasRight, CblasUpper, k, n, 1., sigma, n, h, n, 0., k_by_n, n);
+/*17*/  cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasTrans, k, k, n, 1., k_by_n, n, h, n, 1., r, k);
+/*18*/  cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans, k, 1, n, 1., h, n, mu, 1, -1., data, 1);
+/*19*/  cblas_dcopy(k * n, h, 1, k_by_n, 1);
+        double* k_by_k = (double *) malloc(k * k * sizeof(double));
+/*20*/  cblas_dcopy(k * k, r, 1, k_by_k, 1);
+/*21*/  LAPACKE_dposv(LAPACK_ROW_MAJOR, 'U', k, n, k_by_k, k, k_by_n, n);
+/*23*/  LAPACKE_dpotrs(LAPACK_ROW_MAJOR, 'U', k, 1, k_by_k, k, data, 1);
+        free(k_by_k);
+        double* n_by_n = (double *) malloc(n * n * sizeof(double));
+/*24*/  cblas_dgemm(CblasRowMajor, CblasTrans, CblasNoTrans, n, n, k, 1., h, n, k_by_n, n, 0., n_by_n, n);
+        free(k_by_n);
+        double* n_by_1 = (double *) malloc(n * sizeof(double));
+/*25*/  cblas_dgemm(CblasRowMajor, CblasTrans, CblasNoTrans, n, 1, k, 1., h, n, data, 1, 0., n_by_1, 1);
+        double* new_mu = (double *) malloc(n * sizeof(double));
+/*26*/  cblas_dcopy(n, mu, 1, new_mu, 1);
+/*27*/  cblas_dsymm(CblasRowMajor, CblasLeft, CblasUpper, n, 1, 1., sigma, n, n_by_1, 1, 1., new_mu, 1);
+        free(n_by_1);
+        double* n_by_n2 = (double *) malloc(n * n * sizeof(double));
+/*28*/  cblas_dsymm(CblasRowMajor, CblasRight, CblasUpper, n, n, 1., sigma, n, n_by_n, n, 0., n_by_n2, n);
+/*29*/  cblas_dcopy(n*n, sigma, 1, n_by_n, 1);
+/*30*/  cblas_dsymm(CblasRowMajor, CblasLeft, CblasUpper, n, n, -1., sigma, n, n_by_n2, n, 1., n_by_n, n);
+        free(n_by_n2);
+        *ret_sigma = n_by_n;
+        *ret_mu = new_mu; }
+    \end{minted}
+    \caption{CBLAS/LAPACKE implementation of a Kalman filter. I used C instead
+        of Fortran because it is what Owl uses under the hood and OCaml FFI
+        support for C is better and easier to use than that for Fortran. A distinct
+        `measure\_kalman' function that sandwiches a call to this function with
+        \texttt{getrusage} is omitted for brevity.}\label{fig:cblas_kalman}
+
+    \end{minipage}
+\end{figure}
+\end{landscape}
+
+\subsection{Memory Usage}
+
+Inspecting the trace for LT4LA shows 6 calls to `empty' (function for
+allocating new matrices): 4 temporaries plus 2 for storing the results. From
+the CBLAS implementation (Figure~\ref{fig:cblas_kalman}), we can see that it
+too has 6 calls to malloc: 4 temporaries plus 2 to storing the results.
+
+\subsection{Execution Time}
+
+A graph of the execution times (with error bars which are present but quite
+small) is shown in Figure~\ref{fig:timings}.
+
+\begin{figure}[t]
+    \centering
+    \input{timings.tex}
+    \caption{Comparison of execution times (error bars are present but quite
+        small). Small matrices and timings $n \le 5^3$ were micro-benchmarked
+        with the Core\_bench library. Larger ones used Unix's
+        \texttt{getrusage} functionality, sandwiched between calls to
+        \ltfla{Gc.full_major} for the OCaml implementations.}\label{fig:timings}
+\end{figure}
+
+For $n=5$, CBLAS was faster ($24\mu s$, 526 samples) than LT4LA ($41 \mu s$,
+466 samples).  For $n=25$, and around 350 samples each, CBLAS was again faster
+($104 \mu s$) than LT4LA ($133 \mu s$). The 95\% confidence-intervals for these
+measurements differ from the means by at most $2 \mu s$. For $n=125$, and
+around 110 samples, CBLAS was \emph{slower} ($1803 \mu s\, [1746,\, 1867]$)
+than LT4LA ($1678 \mu s\, [1646,\, 1714]$). For $n=625$ and 1000 samples each,
+they took a \emph{very similar} amount of time: LT4LA took $180.5 \pm 38 \,ms$
+and CBLAS took $188 \pm 36 \, ms$. Despite the large sample size, the
+standard-deviation is still quite high; however, \emph{because} of the large
+sample size, the $p$-value (Welch's t-test) is very small ($p < .05$),
+suggesting that the difference in the means is statistically significant.
+Lastly, for $n=3125$ and 15 samples each, LT4LA and CBLAS continued to take a
+similar amount of time ($16.1 \pm 0.19 \,s$ and $15.7 \pm 0.53 \,s$
+respectively, $p<.05$ using Welch's t-test).
+
+\subsection{Analysis}
+
+Having access to primitives which allow a programmer to re-use memory means
+that memory usage for temporaries in LT4LA is on par with that of CBLAS. One
+caveat is that the \ltfla{freeM} primitive is a no-op in LT4LA, so deallocation
+still relies on OCaml's garbage-collector.
+
+For small matrix sizes, LT4LA and CBLAS execution times differ. I suspect this
+is due to large sample sizes causing more allocations and thus more
+garbage-collector unpredictability (which the linear-regression \emph{did not}
+take into account; multi-variate regressions are not fully supported by
+Core\_bench yet). As matrix size increases, execution times of LT4LA become
+very similar to those of CBLAS.
+
+Therefore, by using LT4LA, it is possible to write linear-algebra programs that
+perform just as predictably (with regards to memory usage and execution time)
+as code written using low-level linear algebra libraries, but gain readability
+(equational and algebraic expressions) and safety (with respect to aliasing
+and read/write permissions).
+
+\section{Compared to OCaml}
+
+To test whether code written using LT4LA is more memory efficient than code
+written using high-level linear-algebra libraries, I wrote two\footnote{I
+attempted a third implementation using Owl's \emph{Lazy} module but it did
+not support matrix inversion at the time of writing.} more implementations
+of a Kalman filter:
+
+\begin{enumerate}
+    \item An Owl/OCaml implementation using a Cholesky decomposition
+        (Figure~\ref{fig:chol_owl_kalman}) but not taking advantage of matrices
+        known to be symmetric ahead of time, and producing a new temporary
+        matrix for every operation (including inverse and transpose).
+
+    \item An idiomatic Owl/OCaml implementation
+        (Figure~\ref{fig:chol_owl_kalman}) with an explicit inverse (LU
+        decomposition), not taking advantage of matrices known to be symmetric
+        ahead of time, and producing a new temporary matrix for every operation
+        (including inverse and transpose).
+\end{enumerate}
+
+\begin{figure}[tp]
+    \begin{minted}[linenos, fontsize=\small]{ocaml}
+let potrs ~uplo a b =
+  let b = Owl.Mat.copy b in
+  Owl.Lapacke.potrs ~uplo ~a ~b
+;;
+
+let chol_kalman ~sigma ~h ~mu ~r ~data =
+  let open Owl.Mat in
+  let ( * ) = dot in
+  let h' = transpose h in
+  let sigma_h' = sigma * h' in
+  let chol = Owl.Linalg.D.chol (r + h * sigma_h') in
+  let sigma_h'_inv rest = sigma_h' * potrs ~uplo:'U' chol rest in
+  let new_sigma = sigma - sigma_h'_inv (h * sigma) in
+  let new_mu = mu + sigma_h'_inv (h * mu - data) in
+  ((sigma, (h, (mu, (r, data)))), (new_mu, new_sigma))
+;;
+
+let owl_kalman ~sigma ~h ~mu ~r ~data =
+  let open Owl.Mat in
+  let ( * ) = dot in
+  let h' = transpose h in
+  let sigma_h' = sigma * h' in
+  let x = sigma_h' * (inv @@ r + h * sigma_h') in
+  let new_mu = mu + x * (h * mu - data) in
+  let new_sigma = sigma - x * h * sigma in
+  ((sigma, (h, (mu, (r, data)))), (new_mu, new_sigma))
+;;
+    \end{minted}
+    \caption{Implementations of a Kalman filter using Owl, top one using a
+        Cholesky decomposition, bottom one using idiomatic Owl. Owl does not
+        yet provide a non-mutating `potrs' function, so I wrote my own which
+        returns a mutated copy of its argument instead.}\label{fig:chol_owl_kalman}
+
+\end{figure}
+
+These implementations also produce the same answers (within at most $2^{-52}$)
+as their LT4LA and CBLAS counterparts and their data is also included in
+Appendix~\ref{chap:eval_data}.
+
+\subsection{Memory Usage}
+
+Inspecting the Owl trace shows it used 11 temporary matrices (13 calls to
+empty, 2 of which are the resulting matrices). The same for Chol shows it used
+13 temporaries (same as Owl plus two temporaries for the two calls to potrs).
+Analysing the sub-expressions of the Owl implementation shows the total amount
+of memory allocated for temporaries is $n + n^2 + 4nk + 3k^2 + 2k$ words; for
+Chol the total is that of Owl plus $n + nk$.
+
+\subsection{Analysis}
+
+For LT4LA and CBLAS the total amount of memory allocated for temporaries is $n
++ n^2 + nk + k^2$. The difference between these two implementations and the
+idiomatic Owl implementation is $k(3n+2k+2)$ words. Hence, by using LT4LA, it
+is possible to have the readability (equational and algebraic expressions) and
+safety of high-level linear algebra libraries, whilst gaining precise and
+explicit control over memory allocation and re-use.
+
+\section{Limitations}
+
+I chose the example of a Kalman filter because it is used in the real world,
+consists purely of a sequence of matrix expressions and produces many
+unnecessary temporary matrices when implemented idiomatically in a high-level
+linear-algebra library. It is good for isolating the key differences between
+not having and having linear types to help a programmer safely manage memory
+and aliasing, whilst excluding other aspects also important to real world
+linear-algebra programs such as control flow or blocking.
+
+\subsection{Curious Behaviour}
+
+A graph of the execution times (with error bars which are present but quite
+small) \emph{all four} implementations is show in Figure~\ref{fig:timings_all}.
+
+\begin{figure}[tp]
+    \centering
+    \input{timings_all.tex}
+    \caption{Comparison of execution times (error bars are present but quite
+        small). Small matrices and timings $n \le 5^3$ were micro-benchmarked
+        with the Core\_bench library. Larger ones used Unix's
+        \texttt{getrusage} functionality, sandwiched between calls to
+        \ltfla{Gc.full_major} for the OCaml implementations.}\label{fig:timings_all}
+\end{figure}
+
+For $n=25$ and $n=125$, the idiomatic Owl implementation is the \emph{fastest}
+of them all at $95 \mu s$ and $1488 \mu s\, [1464, 1515]$ respectively. But
+then for $n=625$ and $n=3125$, Chol is the fastest, at $125.5 \pm 25\, ms$ and
+$11.2 \pm 0.85\, s$ respectively (with Owl a close second for the latter at
+$12.1 \pm 0.47 \, s$, a statistically significant difference with $p<.05$
+using Welch's t-test).
+
+The trend here is that for $n=625$ and $n=3125$ (sizes for which
+\ltfla{Gc.full_major} was called before each measurement), mean execution times
+start to split into two groups: Chol and Owl as one group (similar,
+\emph{faster} times) and LT4LA and CBLAS as another (similar, \emph{slower}
+times). \emph{Although the goals of this thesis have been met}, this is
+unexpected behaviour, and attempting to understand \emph{why} the Chol/Owl
+implementations tended to be faster for anything except the smallest of
+matrices, baffled several people I consulted, including experts at OCamlLabs.
+
+Here are several reasons why this behaviour is surprising:
+
+\begin{itemize}
+
+    \item More temporaries means more garbage-collection pressure meaning
+        more frequent garbage collection, especially as matrix sizes grow.
+
+    \item CBLAS/LT4LA use less memory so in principle should have a smaller
+        working set and better temporal and spatial locality for cache.
+
+    \item CBLAS/LT4LA use routines that combine multiplication and addition
+        of matrices (such as gemm, and symm).
+
+    \item CBLAS/LT4LA have \emph{four} calls to the `symm' routine, which
+        performs \emph{half} as many multiplications as `gemm' under the
+        assumption that one of its arguments is a symmetric matrix.
+
+    \item All of the implementations either directly or indirectly (via Owl's
+        bindings) call the same set of CBLAS/LAPACKE bindings.
+
+\end{itemize}
+
+Here are things I checked:
+
+\begin{itemize}
+
+    \item \textbf{Cache miss-rates.}  Running the different implementations
+        through the Cachegrind cache simulator (part of Valgrind) showed that
+        LT4LA/CBLAS had a roughly 1\% higher cache miss rate than Chol/Owl
+        (rising from around 11\% to around 12\%). This seemed insufficient to
+        account for the differences, so I investigated further.
+
+    \item \textbf{Data access patterns.} I added an extra, modified
+        implementation of LT4LA, which transposed the `h' parameter into a new
+        matrix rather than using the transpose flag to `gemm', to see if
+        row-vs-column access patterns could account for the differences. They
+        did not.
+
+\end{itemize}
+
+Had I been able to use gprof or gperftools, I would have profiled the remaining
+key difference: the `symm' routine, just to be able to eliminate it as a
+suspect.
+
+A \textbf{positive take-away} from all of this is that LT4LA at least gives a
+programmer \emph{choice} and \emph{control} about how to optimise their program
+to their needs because (1) it met its goal of enabling a programmer to write
+code that performs just as predictably as code written using lower-level
+linear-algebra libraries (2) it does so \emph{without} the associated dangers
+of unenforced and uncheckable rules around aliasing, read/write permissions,
+memory allocation, re-use and deallocation that would come with using C,
+Fortran, or the unsafe primitives of Owl.
+
+\section{Qualitative Benefits}\label{sec:qual_benefits}
+
+We have already seen (last chapter and this chapter) that linear types help
+programmers write readable, safe (with respect to aliasing, read/write
+permissions, memory re-use and deallocation) and explicit (with respect to
+memory allocation) code. To justify their why I think they are a
+\emph{practical} and \emph{usable} way to do so, I will elaborate on some of
+the qualitative benefits I experienced whilst using LT4LA.
+
+Prior to this project, I had no experience with linear-algebra libraries or the
+problem of matrix expression compilation. As such, I based my initial LT4LA
+implementation of a Kalman filter using BLAS and LAPACK, on a popular GitHub
+gist of a Fortran implementation, one that was automatically generated from
+SymPy's matrix expression compiler~\cite{rocklin_thesis}.
+
+Once I translated the implementation, I attempted to compile it and found that
+(to my surprise) it did not type-check. This was because the original
+implementation contained incorrect aliasing, did not adhere to Fortran's
+read/write permissions (with respect to \texttt{intent} annotations
+\texttt{in}, \texttt{out} and \texttt{inout}) and unused and unnecessary
+temporaries, all of which were now highlighted by LT4LA's type system.
+
+The original implementation used 6 temporaries, one of which was immediately
+spotted as not being used due to linearity. It also contained two variables
+which were marked as \texttt{intent(in)} but would have been written over by
+calls to `gemm', spotted by the fractional-capabilities feature. Furthermore,
+it used a matrix \emph{twice} in a call to `symm', once with a read permission
+but once with a \emph{write} permission.  Fortran assumes that any parameter
+being written to is not aliased and so this call was not only incorrect, but
+illegal according to the standard, both aspects of which were captured by
+linearity and fractional-capabilities. Lastly, it contained another unnecessary
+temporary, however one that was not obvious without linear types. To spot it, I
+first performed live-range splitting (checked by linearity) by hoisting calls
+to \ltfla{freeM} and then annotated the freed matrices with their dimensions.
+After doing so and spotting two disjoint live-ranges of the same size, I
+replaced a call to \ltfla{freeM} followed by allocating call to \ltfla{copy}
+with one, in-place call to \ltfla{copyM_to}. I believe the ability to boldly
+refactor code which manages memory is good evidence of the usefulness of
+linearity as a tool for programming.
+
+\section{Summary}
+
+Writing a linear-algebra program using LT4LA combines the best of low- and
+high-level linear-algebra libraries: it gives a programmer \emph{explicit
+control} (over read/write permissions, aliasing and memory allocation, re-use
+and deallocation), readability (equational and algebraic expressions)
+\emph{and} safety (automatic checking). Programs written using it \emph{signal
+precise intent} to the reader and compiler and perform similarly to equivalent
+programs written directly using lower-level linear-algebra libraries. In turn,
+such programs can be \emph{checked automatically} against their intent,
+especially when \emph{refactoring} or \emph{rewriting} code. Although any
+expert \emph{could} have followed the same line of reasoning laid out above,
+and arrived at the same program, LT4LA's type system enables a non-expert
+(yours truly) to do the same with increased confidence in the
+result\footnote{This does not preclude testing code by actually
+\emph{executing} it, but definitely \emph{complements} it.} by checking said
+reasoning.

### 2018-05-27 write-up/implementation.tex
Finish Conclusion

@@
 \prechapter{%
-    I will describe the structure of LT4LA and explain the features of its core
-    language. I will then show how typical linear-algebra programs can be
-    elaborated into the core language and checked for linearity. Finally,
-    I will explain how such programs can be translated (in this particular
-    implementation) to OCaml code that is not obviously safe (with
-    respect to aliasing, read/write permissions and memory management).
-    Although I implemented LT4LA in OCaml, I believe the ideas described in
-    this chapter can be applied easily to other languages and also are modular
-    enough to extend the OCaml implementation to output to different back-end
-    languages.
+    I will describe the structure of LT4LA: a DSL (domain specific language)
+    \textbf{library} written in OCaml that transpiles to OCaml. I will explain
+    the features of its core language and show how readable (equational and
+    algebraic expressions) and easy to reason about linear-algebra programs
+    can be elaborated into memory-efficient ones in the core language that are
+    then checked for safety (with respect to aliasing, read/write permissions,
+    memory allocation, re-use and de-allocation). Finally, I will explain how
+    such programs can be transpiled to OCaml code that is not obviously safe.
+    Although I refer to OCaml-specific features of the type-checker and
+    transpiler, I believe the ideas described in this chapter can easily be
+    implemented in other languages and are also general enough to transpile to
+    different back-end languages, such as C or Fortran.
 }%
 
-\section{Structure of LT4LA}
+\section{Structure of LT4LA}\label{sec:structure_lt4la}
 
@@
 continuous-integration for all non-trivial functions so that I could spot and
-correct errors that were not caught by OCaml's type-system whenever I
-implemented new features or or refactored my code.
+correct errors that were not caught by OCaml's type system whenever I
+implemented new features or refactored my code.
 
@@
 
-\section{Core Language}
+\section{Core Language}\label{sec:core_lang}
 
 A full description of the core language can be found in
-Appendix~\ref{chap:ott_spec}. Its main features are intuitionistic values,
+Appendix~\ref{chap:ott_spec}. For convenience, I have reproduced its type
+grammar and defined how I interpret those types in OCaml's type system in
+Figure~\ref{fig:type_grammar}.  The point of LT4LA is to keep the same values
+and types we are familiar with, but \emph{restrict their usage} in certain
+ways. LT4LA's core language's main features are: intuitionistic values,
 value-restriction, fractional-capabilities (inferrable at call sites),
@@
 
+\begin{figure}[tp]
+    \centering
+    \begin{minipage}{.3\textwidth}
+        \centering
+        \begin{grammar}
+            <f> ::= `'
+            \alt <fc>
+            \alt `Z'
+            \alt `S' <f>
+
+            <t> ::= `'
+            \alt `unit'
+            \alt `bool'
+            \alt `int'
+            \alt `elt'
+            \alt <f> `arr'
+            \alt <f> `mat'
+            \alt `!' <t>
+            \alt \synt{$\forall$} <fc.> <t>
+            \alt <t> \lit{$\otimes$} \synt{$t'$}
+            \alt <t> \lit{$\multimap$} \synt{$t'$}
+        \end{grammar}
+    \end{minipage}
+    \begin{minipage}{.3\textwidth}
+        \centering
+        \begin{minted}[fontsize=\small, linenos]{ocaml}
+module Arr =
+  Owl.Dense.Ndarray.D
+
+type z = Z
+type 'a s = Succ
+
+type 'a arr =
+  A of Arr.arr
+  [@@unboxed]
+
+type 'a mat =
+  M of Arr.arr
+  [@@unboxed]
+
+type 'a bang =
+  Many of 'a
+  [@@unboxed]
+        \end{minted}
+    \end{minipage}
+    \begin{minipage}{.3\textwidth}
+        \begin{align*}
+            [\![ f\!c ]\!] &= \texttt{'fc} \\
+            [\![ \textbf{Z} ]\!] &= \texttt{z}\\
+            [\![ \textbf{S} \, f ]\!] &= [\![ f ]\!]\, \texttt{s}\\
+            [\![ \textbf{unit} ]\!] &= \texttt{unit}\\
+            [\![ \textbf{bool} ]\!] &= \texttt{bool}\\
+            [\![ \textbf{int} ]\!] &= \texttt{int}\\
+            [\![ \textbf{elt} ]\!] &= \texttt{float}\\
+            [\![ f\, \textbf{arr} ]\!] &= [\![ f ]\!]\, \texttt{arr}\\
+            [\![ f\, \textbf{mat} ]\!] &= [\![ f ]\!]\, \texttt{mat}\\
+            [\![ \textbf{!} \, t ]\!] &= [\![ t ]\!]\, \texttt{bang}\\
+            [\![ \forall f\!c.\, t ]\!] &= [\![ t ]\!]\\
+            [\![ t \otimes t' ]\!] &= [\![ t ]\!] \texttt{*} [\![ t' ]\!]\\
+            [\![ t \multimap t' ]\!] &= [\![ t ]\!] \rightarrow [\![ t' ]\!]
+        \end{align*}
+    \end{minipage}
+    \caption{Type grammar of LT4LA (left) and an informal interpretation of it
+        into OCaml types (right).}\label{fig:type_grammar}
+\end{figure}
+
 \subsection{Intuitionistic Values}
 
-To make a linearly-typed language usable, we need some way of using values zero
-or more than once, as we would an intuitionistic value. For this, we have in
-the type-expressions the !-constructor and in the term-expressions the
-\ltfla{Many}-constructor and the \ltfla{let Many <id> = .. in .. } eliminator.
+To make LT4LA a usable DSL, I needed some way of using values zero or more than
+once, as a programmer would an intuitionistic value. For this, I added the
+!-constructor at the type-level of the DSL and the \ltfla{Many}-constructor and
+the \ltfla{let Many <id> = .. in .. } eliminator at the term-level of the DSL.
 The idea behind the !-type is that the value uses no `resources'
@@
 includes integers, elements and booleans. So \ltfla{3 : !int} and \ltfla{3. +.
-4. : !elt}. However, all bindings are still linear by default, so to emulate
-intuitionism, I desugar \ltfla{let !x = <exp> in <body>} to  \ltfla{let Many x
-= <exp> in let Many x = Many (Many x) in <body>} (similarly for function
-argument bindings). The reader can check (using the rules in
-Appendix~\ref{chap:ott_spec}) that this has the effect of moving \ltfla{x : !t}
-from the linear to the intuitionistic environments, only if \ltfla{<exp> : !t}.
+4. : !elt}. However, all bindings are still linear by default, so to
+emulate intuitionism, I desugar \ltfla{let !x = <exp> in <body>} to
+\ltfla{let Many x = <exp> in let Many x = Many (Many x) in <body>}
+(similarly for function argument bindings). The reader can check (using the
+rules in Appendix~\ref{chap:ott_spec}) that this has the effect of moving
+\ltfla{x : !t} from the linear to the intuitionistic environments, only if
+\ltfla{<exp> : !t}.
 
@@
 an assumption left implicit up until this point: what does \ltfla{Many} even
-mean? What does it do at runtime? One option is to go down the C++ route and
-make \ltfla{Many} act like a \ltfla{shared_ptr} and act as a runtime
-reference-count for arrays. I chose to not go for this option because it went
-against the \emph{explicitness} and \emph{predictability} that C and Fortran
-have. It would make analysing when and what is allocated and freed more like
-the higher-level languages I was trying to move away from.
+mean in the DSL? How is it translated to OCaml? What is the OCaml runtime
+behaviour of the DSL's \ltfla{Many}-constructor? One option is to go down the
+C++ route and make \ltfla{Many} act like a \ltfla{shared_ptr} and act as a
+runtime reference-count for arrays. I chose to not go for this option because
+it went against the \emph{explicitness} and \emph{predictability} that C and
+Fortran have. It would make analysing when and what is allocated and freed more
+like the higher-level languages I was trying to move away from.
 
-My aim is to show linear types can be simple to understand and apply to linear
-algebra, enough so that it can be grafted (in a limited way) on to existing
-languages as a \emph{library}. In that spirit, the simplest thing that
-\ltfla{Many} can mean at runtime is \emph{nothing}. This language construct is
-translated into a standard OCaml language constructor of the form \ltfla{type
-'a bang = Many of 'a [@@unboxed]}. The \emph{unboxed} annotation means
-that the type and its constructor only exist for the purpose of type-checking
-in OCaml; \emph{the runtime representation of values of type \ltfla{'a} and
-\ltfla{'a bang} is exactly the same}.
+My aim is to show linear types are simple to understand and apply to
+linear-algebra libraries, enough so that it can be grafted (in a limited way)
+on to existing languages as a \emph{library}. In that spirit, the simplest
+thing that the DSL's \ltfla{Many}-constructor can mean in the OCaml runtime is
+\emph{nothing}.  This language construct is translated into a standard OCaml
+type-and-constructor of the form \ltfla{type 'a bang = Many of 'a [@@unboxed]}.
+The \emph{unboxed} annotation means that the type and its constructor only
+exist for the purpose of type-checking in OCaml; \emph{the OCaml runtime
+representation of values of type \ltfla{'a} and \ltfla{'a bang} is exactly the
+same}.
 
@@
 the world of polymorphic types applied to linearity instead. We then have the
-rule that we can only use \ltfla{Many} on expressions that are defined to be
-\emph{values} and \emph{use no linear variables}.
+rule that in LT4LA, we can only use \ltfla{Many} on expressions that are
+defined to be \emph{values} and \emph{use no linear variables}.
 
@@
 
-Having started off with linearity and understanding how it helps a programmer
-keep track of of memory allocation and deallocation, and then understood and
-correctly implemented intuitionism for the only values we wish to be
-intuitionistic, and, we now tackle the problem of how to implement safe
-aliasing and read/write permissions.
+I started off with linearity and showing how it helps a programmer keep track
+of of memory allocation and deallocation, and then demonstrated how to add
+intuitionism for the only values we wish to be intuitionistic. I will now
+explain how I enforce safe aliasing and read/write permissions.
 
-Array and matrix types are parameterised by \emph{fractional-capabilities}.  A
-fraction of 1 ($2^0$) represents complete ownership of a value; in particular,
-this allows a programmer to write or free it. Creating an array gives you
-ownership of it; the function \ltfla{array : !int --o z arr} (where \ltfla{z}
-represents `0'). Once you have ownership of an array, you can free it:
-\ltfla{free : z arr --o unit}.  Importantly, because a linear-value may only be
-used once, the array just freed is \emph{out of scope} for following
-expressions, preventing use-after-free.  Ownership also enables you to write to
-the array: \ltfla{set : z arr --o !int --o !elt --o z arr} (the syntax
-\ltfla{w[i] := j} is just sugar for \ltfla{set w i j}). Here, linearity
-prevents accessing aliases which represented the array \emph{before} the
-mutation.
+Array and matrix types are parameterised by
+\emph{fractional-capabilities}~\cite{boyland}.  A fraction of 1 ($2^0$)
+represents complete ownership of a value; in particular, this allows a
+programmer to write or free it. Creating an array gives you ownership of it;
+the function \ltfla{array : !int --o z arr} (where \ltfla{z} represents `0').
+Once you have ownership of an array, you can free it: \ltfla{free : z arr --o
+unit}.  Importantly, because a linear-value may only be used once, the array
+just freed is \emph{out of scope} for following expressions, preventing
+use-after-free.  Ownership also enables you to write to the array: \ltfla{set :
+z arr --o !int --o !elt --o z arr} (the syntax \ltfla{w[i] := j} is just sugar
+for \ltfla{set w i j}). Here, linearity prevents accessing aliases which
+represented the array \emph{before} the mutation.
 
@@
 based on the other arguments passed to the function. In conjunction with the
-requirement that functions declarations need type-annotations for their
-arguments, this allows a fractional-capability to be correctly inferred in any
-program.
+requirement that function declarations need type-annotations for their
+arguments, \textbf{this allows a fractional-capability to be \emph{correctly
+inferred} in \emph{any} LT4LA expression}.
 
 Fractions exist to provide both safe aliasing and read/write permissions, via
-the primitives \ltfla{share : 'x . 'x arr --o 'x s arr * 'x s arr} and
-\ltfla{unshare: 'x .  'x s arr --o 'x s arr --o 'x arr}.  For the former, two
+the primitives \ltfla{share : 'x . 'x arr --o ('x s) arr * ('x s) arr} and
+\ltfla{unshare: 'x .  ('x s) arr --o ('x s) arr --o 'x arr}.  For the former, two
 arrays returned (which happen to just be the given array) can now only be read
@@
 
-Given this set-up, we now \emph{statically} have \emph{perfect} information
-about aliasing and ownership of values in the program. We can only write to or
-free an array only when we own it; ownership guarantees no other aliases exist
-in scope at the point of usage. In Figure~\ref{fig:ltfla_kalman}, I show how
-this perfect information can be used to write more natural-looking code using
-value-semantic expressions which behave in precisely the way we intend it to.
-Now the programmer need not resort to manually figuring out and inserting
-\texttt{noalias} annotations and worrying about what variables can and cannot
-be written to or freed; instead they can let the loyal and tireless compiler do
-the heavy lifting.
+Given this set-up, a programmer now \emph{statically} has \emph{perfect}
+information about aliasing and ownership of values in the program. They can
+only write to or free an array only when they own it; ownership guarantees no
+other aliases exist. In Figure~\ref{fig:ltfla_kalman}, I show how this perfect
+information can be used to write more readable (equational and algebraic) code
+using easy to reason about, value-semantic expressions whose \emph{intensional}
+behaviour (regarding aliasing, read/write permissions and memory allocation,
+re-use and de-allocation)  is also explicitly and precisely described and
+\emph{automatically checkable}. Now the programmer need not resort to manually
+figuring out and inserting \texttt{noalias} annotations and worrying about what
+variables can and cannot be written to, re-used or freed; instead they can let
+the loyal and tireless compiler do the heavy lifting.
 
@@
 
-Because we do not know which way a condition will evaluate at run time, we must
-guarantee the both branches use the same set of linear variables. Writing the
-type-checker in a pure-functional monadic style paid off here because I could
-now sandwich monadic values with state-adjustments either side of it. Given two
-monadic values that represented type-checking two branches of an if-expression,
-I could use the code in Figure~\ref{fig:same_resources} to easily save, reset
-and compare the state either side of running those monadic values.
+An if-expression's condition may evaluate either way at run time, so a
+programmer must guarantee that both branches use the same set of linear
+variables. Writing the type-checker in a pure-functional monadic style paid off
+here because I could now sandwich monadic values with state-adjustments either
+side of it. Given two monadic values that represented type-checking two
+branches of an if-expression, I could use the code in
+Figure~\ref{fig:same_resources} to easily save, reset and compare the state
+either side of running those monadic values.
 
@@
 \ltfla{let x = 3 in let !f (y : !int) = x + y in <body>} to \ltfla{let x = 3 in
-let Many f =  Many (fun y : !int -> x + y) in  <body>}. Recursion is
-slightly more complicated: we can desugar the factorial function
-(Figure~\ref{fig:ltfla_factorial}) to \ltfla{let Many f = fix (f, x : !int,
-if (*..*) : !int) in <body> }. However, \ltfla{fix}, like \ltfla{Many},
-must also not use any linear variables from its surrounding scope.
+let Many f =  Many (fun y : !int -> x + y) in  <body>}.  Recursion is slightly
+more complicated: we can desugar the factorial function
+(Figure~\ref{fig:ltfla_factorial}) to \ltfla{let Many f = fix (f, x : !int, if
+(*..*) : !int) in <body> }. However, \ltfla{fix}, like \ltfla{Many}, must also
+not use any linear variables from its surrounding scope, because calling a
+recursive function may evaluate its body multiple times.
 
-\section{Matrix Expressions}
+\begin{landscape}
+    \begin{figure}[p]
+    \centering
+    \begin{minipage}{1.47\textheight}
+    \inputminted[linenos, fontsize=\footnotesize]{fortran}{kalman.f90}
+    \end{minipage}
+    \caption{Kalman filter in Fortran 90.}\label{fig:fortran_kalman}
+    \end{figure}
+\end{landscape}
 
-We have now arrived at an extended application of linear types. I will show
-how, we can apply the ability to automatically check aliasing, read/write
+\section{Matrix Expressions}\label{sec:matrix_exps}
+
+We have now arrived at an extended application of linear types. I will show how
+we can apply the ability to automatically check aliasing, read/write
 permissions, memory allocation, re-use and deallocation, to the domain of
-matrix-expression compilation.
-
-\begin{sidewaysfigure}
-    \inputminted[linenos, fontsize=\footnotesize]{fortran}{kalman.f90}
-    \caption{Kalman filter in Fortran 90.}\label{fig:fortran_kalman}
-\end{sidewaysfigure}
+matrix-expression compilation (automatically generating code like
+Figure~\ref{fig:fortran_kalman} from mathematical expressions like
+Figure~\ref{fig:maths_kalman}).
 
 In Figure~\ref{fig:fortran_kalman}, we see the difficulty of efficiently
-implementing a \emph{Kalman filter}, a powerful set of equations applicable
-to a wide variety of problems. From the comments, we see that every variable is
-annotated with the step/matrix expression that it will hold at some point
-during the computation (an equivalent alternative, say in C++, could be to have
-a meaningful name for each step/matrix expression and manually annotate/keep
-track of which names alias the same location).
+implementing a \emph{Kalman filter}~\cite{kalman}, a powerful set of equations
+(Figure~\ref{fig:maths_kalman}) applicable to a wide variety of problems. From
+the comments, we see that every variable is annotated with the step/matrix
+expression that it will hold at some point during the computation (an
+equivalent alternative, say in C++, could be to have a meaningful name for each
+step and manually keep track of which names alias the same location).
+
+\begin{figure}[tp]
+    \begin{align*}
+        \mu' &= \mu + \Sigma H^T (R + H \Sigma H^T)^{-1} (H \mu - \textrm{data})\\
+        \Sigma' &= \Sigma ( I - H^T (R + H \Sigma H^T)^{-1} H \Sigma )
+    \end{align*}
+    \caption{Kalman filter equations (credit:
+    \href{http://matthewrocklin.com/blog/work/2012/11/24/Kalman-Filter}{matthewrocklin.com}).}\label{fig:maths_kalman}
+\end{figure}
 
@@
         allocated for the function (new/copy),
-    \item unambiguously seeing \emph{when} and what values are freed;
+    \item unambiguously seeing \emph{when} and what values are freed or written-over;
 \end{itemize}
@@
     \item ensuring all values are declared and \emph{initialised} correctly before they are used,
-    \item checking values are neither used after they are freed \emph{nor} leaked.
+    \item checking values are neither used after they are freed/written-over \emph{nor} leaked.
 \end{itemize}
@@
     \inputminted[linenos, fontsize=\small]{ocaml}{../test/examples/kalman.lt}
-    \caption{Kalman filter in LT4LA.}\label{fig:ltfla_kalman}
+    \caption{Kalman filter in LT4LA. In contrast with the Fortran
+        (Figure~\ref{fig:fortran_kalman}) or C (Figure~\ref{fig:cblas_kalman})
+        implementations, this one is readable (equational and algebraic),
+        using easy to reason about, value-semantic expressions. In addition to
+        that, its \emph{intensional} behaviour(regarding aliasing, read/write
+        permissions and memory allocation, re-use and de-allocation) is also
+        explicitly and precisely described and (unlike with Fortran or C)
+        \emph{automatically checkable}.}\label{fig:ltfla_kalman}
 \end{figure}
@@
 academia~\cite{rocklin_thesis, fabregat_thesis, gunnels_flame, linnea, taco}
-but this is, to my knowledge, the first type-based approach to it.
+but this is, to my knowledge, the \textbf{first type-based approach} to it.
 
@@
 let x <- [| a*b + c |] in ..        ==> let x <- [| 1.*a*b + 1.*c |] in ..
-let x <- [| i.*a*b + j.*c |] in ..  ==> let ((a,b), c) = (* BLAS *) in ..
+let x <- [| i*a*b + j*c |] in ..    ==> let ((a,b), c) = (* BLAS *) in ..
 let x <- new (m, n) [| a*b |] in .. ==> let c = matrix m n in
@@
         matrix functions. Further annotations on the matrix variables
-        (\ltfla{sym} or \ltfla{^T}) determine which BLAS routine and what
-        parameters to call it with.}\label{fig:mat_patterns}
+        (\ltfla{sym} or \ltfla{^T}) determine which BLAS routine is called and
+        what parameters it is passed.}\label{fig:mat_patterns}
 
@@
 C$ it is typically better to use a BLAS routine directly rather than but first
-compiling $A$, then $B$, then adding a call to multiply them, then $C$ and
-finishing with a call to add the results.
+compiling $A$, then $B$, then adding a call to multiply them, then compiling
+$C$ and finishing with a call to add the results.
 
@@
 
-    \item \ltfla{let Many x = Many <exp> in <body>} with \ltfla{let x = <exp> in <body},
+    \item \ltfla{let Many x = Many <exp> in <body>} with \ltfla{let x = <exp> in <body>},
 
@@
 The end result is visible in Figures~\ref{fig:ocaml_sumarray}
-and~\ref{fig:ocaml_kalman}: OCaml code that is not obviously safe and correct
-with respect to linearity, as promised at the start of the chapter.
+and~\ref{fig:ocaml_kalman}:  (surprisingly beautiful) OCaml code that is not
+obviously safe, as promised at the start of the chapter.
 
 \begin{figure}[tp]
+    \centering
     \begin{minted}[linenos, fontsize=\small]{ocaml}
@@
     \end{minted}
-    \caption{Recursive OCaml function for a summing an array, generated from
-        Figure~\ref{fig:ltfla_sumarray}, passed through \texttt{ocamlformat} for
-        presentation.}\label{fig:ocaml_sumarray}
+    \caption{Recursive OCaml function for a summing an array, generated (at
+        \emph{compile time}) from the code in Figure~\ref{fig:ltfla_sumarray},
+        passed through \texttt{ocamlformat} for presentation.}\label{fig:ocaml_sumarray}
 
@@
     \end{minted}
-    \caption{OCaml code for a Kalman filter, generated from
-        Figure~\ref{fig:ltfla_kalman}, passed through \texttt{ocamlformat} for
-        presentation.}\label{fig:ocaml_kalman}
+    \caption{OCaml code for a Kalman filter, generated (at \emph{compile time})
+        from the code in Figure~\ref{fig:ltfla_kalman}, passed through
+        \texttt{ocamlformat} for presentation.}\label{fig:ocaml_kalman}
 
@@
 Advantages of generating code at compile time include avoiding runtime overhead
-as well as catching any interface changes between the generated code and the
-code that uses it via OCaml's type system (another benefit of embedding LT4LA's
-type system inside of OCaml).
+as well as catching any interface changes (type-signatures now include
+\emph{intensional} information about which parameters are read from or written
+to) between the generated code and the code that uses it via OCaml's type
+system, \emph{another benefit} (in addition to a sanity check on LT4LA's
+type-checker implementation) of \emph{embedding} LT4LA's type system inside of
+OCaml's.
 
-I suspect that this approach will be very valuable to not just users of
-libraries such as Numpy or Owl, but also their implementors: library functions
-which presented a safe, value-semantic interface but use unsafe, mutating
-operations on the inside could now be expressed using LT4LA and have an extra
-safety-net and automatic checking for their implementations.
+I suspect that this approach will be very valuable to not just \emph{users} of
+libraries such as Numpy or Owl, but also their \emph{implementors}: library
+functions which present a safe, value-semantic interface but use unsafe,
+mutating operations on the inside could now be expressed using LT4LA and have
+an extra safety-net and automatic checking for their implementations.
 
@@
 capture and automatically check} aliasing, read/write permissions, memory
-allocation, re-use and deallocation of non-trivial linear algebra programs. I
+allocation, re-use and deallocation of non-trivial linear-algebra programs. I
 have also demonstrated that simple pattern-matching and desugaring provides the

### 2018-05-27 write-up/ott_spec.tex
Finish Conclusion

@@
 The following pages present a specification of the grammar and type-system used
-by my project, produced using the Ott\cite{sewell_ott} tool. It is important to note
+by my project, produced using the Ott~\cite{sewell_ott} tool. It is important to note
 that the type-system described here is not how it is implemented: it is easier
@@
     \item Implement it so that the type-environment \emph{changes} as a result
-        of type-checking an expression; with this, the below semantics describe
+        of type-checking an expression, similar to the rules shown in
+        Figure~\ref{fig:example_rules}; with this, the below semantics describe
         the \emph{difference} between the environment after and before checking
-        an expression.
+        an expression. For example, in the pair-introduction rule, $\Gamma =
+        \Gamma_2 - \Gamma_1$ and $\Gamma' = \Gamma_3 - \Gamma_2$, for an
+        appropriate definition of $(-)$.
 
@@
     \item Have \emph{one} environment where variables were \emph{tagged} as
-        linear and un-used, linear and used, and intuitionistic. This was
+        linear and unused, linear and used, and intuitionistic. This was
         definitely an implementation convenience so that variable binding could
@@
         scoping/variable look-up could be handled implicitly thanks to the
-        associative-list structure of the environment.
+        associative-list structure of the environment. So, the variable
+        rule would most accurately look like:
+        \begin{prooftree}
+            \AxiomC{}
+            \RightLabel{\textsc{Ty\_Var}}
+            \UnaryInfC{$\Theta; \Gamma, x \overset{n}{:} t \vdash x : t ; \Gamma, x \overset{n-1}{:} t$}
+        \end{prooftree}
+        for $n \in \{0\,\textrm{(used)}, 1\,\textrm{(unused)},
+        \omega\,\textrm{(intuitionistic)}\}$, $\omega - 1 = \omega$ and  $1 - 1 = 0$.
 

### 2018-05-27 write-up/related_work.tex
Finish Conclusion

@@
 
-\section{Matrix Expression Compilation}
-
-% For decades, small gains in efficiency of \emph{executing} numerically
-% intensive code have and continue to translate into massive savings in costs and
-% time. This has spurned mountains of research into an almost bewildering variety
-% of hardware, software and compilation techniques, typically entrusted with a
-% (team of) domain-specific expert programmer(s) to apply in real world
-% situations.
-% 
-% Given the rise of big data and machine learning, there is now a movement to
-% improve \emph{programmer} efficiency in this domain after focusing for so long
-% on \emph{computational} efficiency. Matrix expression compilers promise to
-% generate code that is reasonably close to `optimal'/what a domain-expert may
-% write but without the need to acquire years of experience and depth of
-% knowledge (a similar trade-off to garbage-collection for memory management or
-% software-transactional memory for concurrency).
+\section{Matrix Expression Compilation}\label{sec:mat_exp_comp}
 
@@
 I believe the advantages of my approach are two-fold (1) \emph{predictable}
-performance and (2) more accurate modelling of how low-level kernels handle
-their resources. My confidence in the latter claim comes from finding at least
-two errors in the Fortran code output by SymPy to compute a Kalman filter, one
-to do with aliasing (resulting in code that was not legal Fortran) that would
-not have type-checked had it been translated via LT4LA as an intermediate
-representation (the other error to be explained later
-in~\ref{subsec:lin_dep_types}).
+performance (as defined in Chapter~\ref{chap:eval}) and (2) more accurate
+modelling of how low-level kernels handle their resources. My confidence in the
+latter claim comes from finding errors in Fortran code output by SymPy to
+compute a Kalman filter (as described in Section~\ref{sec:qual_benefits});
+errors that would not have passed type-checking had it been translated via
+LT4LA as an intermediate representation.
 
@@
 A particularly strong advantage LT4LA has over other libraries that use lazy
-evaluation is, funnily enough, linear types, more precisely and the static and
-perfect information they guarantee about aliasing: with Owl, every graph-node
-has only one node in or out; with Eigen, every ``assignment'' has a
-\texttt{noalias} annotation.
+evaluation is, funnily enough, linear types; more precisely, it is the static
+and perfect information they guarantee about aliasing: with Owl, every
+graph-node has only one node in or out; with Eigen, every ``assignment'' has a
+\texttt{noalias} annotation, without the danger of getting it wrong and
+invoking undefined behaviour.
 
@@
 compilation. It supports ML-style modules, loops, limited parametric
-polymorphism, size types and uniqueness types. Intuitively, where linearity
-provides a local guarantee of ``this value is not aliased in this scope'',
-uniqueness types provide a global guarantee of ``this value is not aliased
-anywhere''. Its combinators are more expressive than typical linear algebra
-library kernels so encourages shorter, more declarative linear algebra code.
+polymorphism, size types and uniqueness types (an idea closely related to
+linear types). Its combinators are more expressive than typical linear-algebra
+library kernels so encourage shorter, more declarative linear-algebra code.
 
@@
 relevant to this thesis because it statically attempts to prevent many
-resource-related bugs. Although there are a few linear algebra libraries for
+resource-related bugs. Although there are a few linear-algebra libraries for
 Rust under development, careful use of its macro system and borrow-checker
-could make it the safest and easiest to use language for linear algebra
+could make it the safest and easiest to use language for linear-algebra
 projects to come. Its struggle is more likely to be against the inertia of the
 large amounts of C++/Fortran code already out there rather than its usability
-or benefits.
+or benefits. Given this inertia, I believe there is value in taking the DSL
+library approach of LT4LA which will work with existing systems.
 
@@
 Dimension mismatches are seen as an irritating but small inconvenience when
-writing linear algebra code. However, the second error I found in the Fortran
-code output by SymPy to compute a Kalman filter was a dimension/transposition
-error. Although we would not need full dependent types to solve dimension
-mismatches (symbolic size types would be sufficient), managing properties about
-matrices could be done at the type-level in a dependently typed setting.
+writing linear-algebra code. However, another error I found in the Fortran code
+output by SymPy to compute a Kalman filter was a dimension/transposition error.
+Although we would not need full dependent types to solve dimension mismatches
+(symbolic size types would be sufficient), managing properties about matrices
+could be done at the type-level in a dependently typed setting.
 

### 2018-05-27 write-up/kalman.f90

@@ -1,19 +1,15 @@
-subroutine f(mu, Sigma, H, INFO, R, Sigma_2, data, mu_2, k, n)
+subroutine kalman(mu, Sigma, H, INFO, R, Sigma_2, data, mu_2, k, n)
 implicit none
 
-integer, intent(in)    :: k
-integer, intent(in)    :: n
-real*8,  intent(in)    :: Sigma(n,n)   ! Sigma
-real*8,  intent(in)    :: H(k,n)       ! H
-real*8,  intent(in)    :: mu(n)        ! mu
+integer, intent(in)    :: k, n
+real*8,  intent(in)    :: Sigma(n,n), H(k,n), mu(n)
 real*8,  intent(inout) :: data(k)      ! data, H*mu - data , (H*Sigma*H^T + R)^-1*(H*mu - data)
 real*8,  intent(inout) :: R(k, k)      ! R, H*Sigma*H^T + R
-
 integer, intent(out)   :: INFO         ! INFO
+real*8,  intent(out)   :: Sigma_2(n,n) ! H^T*(H*Sigma*H^T + R)^-1*H, Sigma, Sigma*(I - H^T*(H*Sigma*H^T + R)^-1*H*Sigma)
 real*8,  intent(out)   :: mu_2(n)      ! mu, Sigma*H^T*(H*Sigma*H^T + R)^-1*(H*mu - data) + mu
 real*8                 :: H_2(k,n)     ! H * Sigma , H , (H*Sigma*H^T + R)^-1*H
 real*8                 :: chol_R(k,k)  ! R, U where (H*Sigma*H^T + R)=U^T*U
-real*8,  intent(out)   :: Sigma_2(n,n) ! H^T*(H*Sigma*H^T + R)^-1*H, Sigma, Sigma - Sigma*H^T*(H*Sigma*H^T + R)^-1*H*Sigma
 real*8                 :: H_data(n)    ! H^T*(H*Sigma*H^T + R)^-1*(H*mu - data)
 real*8                 :: N_N_tmp(n,n) ! H^T*(H*Sigma*H^T + R)^-1*H*Sigma
 
@@ -23,7 +19,7 @@ call dgemm('N', 'N', k, 1, n, 1, H, n, mu, 1, -1, data, 1)          ! data    :=
 call dcopy(k*n, H, 1, H_2, 1)                                       ! H_2     := H
 call dcopy(k*k, R, 1, chol_R, 1)                                    ! chol_R  := R
[…]

### 2018-05-27 write-up/conclusion.tex
Finish Introduction

@@
 written using a high-level linear-algebra libraries and performed just as
-\textbf{predictably} as one written using a low-level linear-algebra library
+\textbf{predictably} as one written using a low-level linear-algebra library,
+enabling a programmer to have the best of both worlds
 (Chapter~\ref{chap:eval}). The type system ensured that the implementation was
-safe and explicit with respect to aliasing, read/write permissions, memory
-allocation, re-use and deallocation.
+\textbf{safe and explicit} with respect to aliasing, read/write permissions,
+memory allocation, re-use and deallocation.
 

### 2018-05-27 write-up/introduction.tex
Finish Introduction

@@
-\chapter{TO DO: Introduction}
+\chapter{Introduction}
 
@@
 
-\prechapter{%
-}%
+Linear types allow the compiler and programmer to statically keep track of the
+resources that an implementation uses, thus offering a promising solution to
+resource management. However, they have not made their way into mainstream
+programming languages. To demonstrate their usefulness, I implemented an OCaml
+library that allows users to learn about, use and become familiar with linear
+types, specifically in the context of linear-algebra programs.
 
-\section{Overview of Problem}
+The main contributions of this thesis are:
 
-\section{Contributions}
+\begin{itemize}
+
+    \item An \textbf{original, usable implementation} of a type system that can express
+        aliasing, read/write permissions, memory allocation, re-use and deallocation.
+
+    \item An \textbf{original} demonstration of how that type system can be
+        \textbf{applied to the APIs} of linear algebra libraries and the
+        \textbf{benefits} of doing so.
+
+    \item Many \textbf{new} examples of how the type system can \textbf{automatically
+        check} for common aliasing, read/write permissions, memory allocation,
+        re-use and deallocation \textbf{errors} in the context of
+        linear-algebra programs.
+
+    \item An \textbf{original} demonstration of how that type system can be
+        \textbf{useful} for \textbf{matrix expression compilation}.
+
+    \item \textbf{New and readable implementations} of \textbf{non-trivial}
+        linear-algebra programs that \textbf{take advantage} of said type
+        system.
+
+    \item A \textbf{new solution} to the \textbf{dichotomy} of
+        \textbf{readability}, \textbf{ease of reasoning and safety} of
+        high-level linear-algebra libraries versus the
+        \textbf{memory-efficiency} of low-level linear-algebra libraries.
+
+    \item A \textbf{new library design} to provide said solution in a way that
+        \textbf{integrates well} with existing OCaml code and linear-algebra
+        libraries.
+
+\end{itemize}

### 2018-05-28 write-up/abstract.tex
Add Acknowledgements
Fix many small writing errors

--- 2018-05-24 write-up/abstract.tex
@@
 \newpage
-{\normalfont\Huge\sffamily\bfseries TO DO: Abstract}
+{\normalfont\Huge\sffamily\bfseries Abstract}
 \vspace{24pt}
@@
 re-use and deallocation, first, in the context of the APIs of linear algebra
-libraries and then in the context of matrix-expression compilation.
+libraries and then in the context of matrix expression compilation.
 
-I show that framing the problem using linear types can help \emph{reduce bugs}
-by making precise and explicit the informal, ad-hoc practices typically
-employed by human experts and linear algebra \emph{compilers} and automate
+I show that framing the problem using linear types can \emph{reduce bugs}
+by making precise and explicit, the informal, ad-hoc practices typically
+employed by experts and matrix expression compilers \emph{and} automate
 checking them.
@@
 As evidence for this argument, I show non-trivial, yet readable, linear algebra
-programs, that are safe (with respect to aliasing, read/write permissions,
-memory re-use and deallocation) and explicit (with respect to memory
-allocation) which (1) are more memory-efficient than equivalent programs
-written using purely high-level linear algebra libraries and (2) perform just
-as predictably as equivalent programs written using low-level linear-algebra
-libraries. I also argue \emph{the experience} of writing such programs with
-linear types is ``better'' in key-respects. In addition to all of this, I show
-that it is possible to provide such features \emph{as a library} on top of
-existing programming languages and linear algebra libraries.
+programs, that are safe and explicit (with respect to aliasing, read/write
+permissions, memory allocation, re-use and deallocation) which (1) are more
+memory-efficient than equivalent programs written using high-level
+linear algebra libraries and (2) perform just as predictably as equivalent
+programs written using low-level linear algebra libraries. I also argue
+\emph{the experience} of writing such programs with linear types is
+qualitatively better in key respects. In addition to all of this, I show that
+it is possible to provide such features \emph{as a library} on top of existing
+programming languages and linear algebra libraries.
 
-\textbf{Summarise evaluation.}
+\vspace*{\fill}
+{\normalfont\Huge\sffamily\bfseries Acknowledgements}
 
-\textbf{Summary future work.}
+I would like to thank my family and closest friends for supporting me through
+my time at Cambridge: you have been my strength when I have had none. I would
+also like to thank my supervisors, Dr\@.~Neelankantan Krishnaswami and
+Dr\@.~Stephen Dolan, for teaching me so much this year. I owe a debt of
+gratitude to both the 2017 Part III Appeals Committee and those who wrote
+letters in support of my appeal, for believing I deserved the chance to stay
+on: never have I valued the privilege of learning what I enjoy and enjoying
+what I learn more.  Lastly, I would like to thank the many staff and research
+members of the Computer Lab who have been approachable, patient and attentive
+during my time here.
 

### 2018-05-28 write-up/background.tex
Fix minor layout issues
Fix many small writing errors

@@
     I will outline the concept of linear types and show how they can be used to
-    solve the problems faced by programmers writing code using linear-algebra
+    solve the problems faced by programmers writing code that uses linear algebra
     libraries. I will emphasise \emph{practical} and intuitive explanations of
     linear types to keep this thesis accessible to programmers as well as
-    academics not familiar with type-theory and will give only a terse overview
+    academics not familiar with type theory and will give only a terse overview
     of the history and theory behind linear types for the interested reader to
@@
 
-\section{Tracking Resources with Linearity}
-Familiar examples of using a type-system to express program invariants are
-existential-types for abstraction and encapsulation, polymorphic types for
-parametricity and composition (a.k.a generics). Less-known examples include
-dependent-types (contracts or pre- and post-conditions). The advantages of
-using a type-system to express program invariants are summarised by saying the
-stronger the rules you follow, the better the guarantees you can get about your
-program, \emph{before} you run it. At first, the rules seem restrictive, but
-similar to how the rules of grammar, spelling and more generally writing help a
-writer make it easier and clearer to communicate the ideas they wish to express,
-so too do typing rules make it easier to communicate the intent and
-assumptions under which a program is written. An added, but often overlooked
-benefit is automated-checking: a programmer can boldly refactor in certain
-ways and the compiler will \emph{assist} in ensuring the relevant invariants
-the type-system enforces are updated and kept consistent by pointing-out
-where they are violated.
+\section{Tracking Resources with Linearity} Familiar examples of using a type
+system to express program invariants are existential types for abstraction and
+encapsulation, and polymorphic types for parametricity and composition (a.k.a
+generics). Less-known examples include dependent types for contracts or pre-
+and post-conditions. The advantages of using a type system to express program
+invariants are summarised by saying: the stronger the rules you follow, the
+better the guarantees you get about your program \emph{before} you run it. At
+first, the rules may seem restrictive, but similar to how the rules of grammar,
+spelling and more generally writing help a writer communicate their ideas more
+clearly, so too do typing rules make it easier for the programmer to
+communicate the intent and assumptions of their programs. An added, but often
+overlooked benefit is automated-checking: a programmer can boldly refactor
+their programs and the type checker will \emph{assist} in enforcing the
+invariants it was given by highlighting where they were violated.
 
@@
 linearly typed language). While this restriction may seem limiting at first,
-precisely these constraints can be used to express common invariants of the
-programs written by programmers every day. For example: a file or a
-socket, once opened \emph{must} closed; all memory that is manually allocated
-\emph{must} be freed.  C++'s destructors and Rust's Drop-trait (and more
-generally, its borrow-checker) attempt to enforce these constraints by
-basically doing the same thing: any resource that has not been moved is
-deallocated at the end of the current lexical scope. Notably, these languages
-also permit aliasing, alongside rules enforcing when it is acceptable to do so.
-On face value, the above one-line description of linear types prevents aliasing
-or functions such as $\lambda x.\; (x, x)$; such features are still allowed
-(albeit in a more restricted fashion) in a \emph{usable} linear type system
-designed for working with linear-algebra libraries.
+precisely this constraint can be used to express common invariants of typical
+programs. For example: a file or a socket, once opened \emph{must} closed; all
+memory that is manually allocated \emph{must} be freed.  C++'s destructors and
+Rust's Drop-trait (and more generally, its borrow-checker) attempt to enforce
+these constraints by basically doing the same thing: any resource that has not
+been moved is deallocated at the end of the current lexical scope. Notably,
+these languages also permit aliasing, alongside rules enforcing when it is
+acceptable to do so.  On face value, the above one-line description of linear
+types prevents aliasing or functions such as $\lambda x.\; (x, x)$; such
+features are still allowed (albeit in a more restricted fashion) in a
+\emph{usable} linear type system designed for working with linear algebra
+libraries.
 
@@
 Given this background, the most pertinent question at hand is: what problems do
-linear-algebra library users (and writers) typically face? The answer to this
-question depends on which of two buckets a programmer falls (or is forced by
-domain) into. On one side, we have users of high-level linear-algebra libraries such
-as Owl (for OCaml), Julia and Numpy (for Python); other the other, we have
-users of more manual, lower-level libraries such as BLAS (Basic Linear Algebra
-Subroutines) for languages like C++ and FORTRAN.\footnote{I am not including
-Rust in this comparison because its linear-algebra libraries are under active
-development and not as well-known/used. Later on, given that it is a language
-with in-built support of substructural features to track resources, Rust will
-be compared and contrasted with this project to evaluate the classic
-(E)DSL-versus-language-feature debate as it applies to the domain of
-linear-algebra libraries.} Most of what follows applies to \emph{dense}
-linear-algebra computations rather than \emph{sparse} because memory
-allocated for results typically depends on the sparsity of the inputs and
-so is not immediately amenable to the techniques proposed in this thesis.
+linear algebra library users (and implementors) typically face? The answer to
+this question depends on which of two buckets a programmer falls (or is forced
+by domain) into. On one side, we have users of high-level frameworks such as
+Owl (for OCaml), Julia and Numpy (for Python); on the other, we have users
+of more manual, low-level libraries such as BLAS (Basic Linear Algebra
+Subroutines) for languages like C, C++ and Fortran.\footnote{I am not including
+Rust in this comparison because its linear algebra libraries are under
+active development and not as well-known/used. Later on, I will compare and
+contrast Rust (a language with in-built support of substructural features
+to track resources) with this project to evaluate the classic
+(E)DSL-versus-language-feature debate as it applies to linear types.} Most of
+what follows applies to \emph{dense} linear algebra computations rather than
+\emph{sparse} ones because for the latter, memory allocated for results depends
+on the sparsity of the inputs and so is not immediately amenable to the
+techniques proposed in this thesis.
 
@@
 
-In Figure~\ref{fig:mat_mul_copies}, we see that matrix-multiplication is fairly
+In Figure~\ref{fig:mat_mul_copies}, we see that matrix multiplication is fairly
 trivial to write and execute in Numpy, Julia and Owl. Let us call this approach
@@
 escape-hatches (typically provided by most libraries, which permit in-place
-modification of objects) to build faster, and more memory-efficient algorithms
-which are harder to reason about.
+modification of objects) to build faster, less readable and more
+memory-efficient algorithms which are harder to reason about.
 
 \subsection{IHNIWTLM}
+
 The title of this subsection\footnote{I Have No Idea What Those Letters Mean.}
-illustrates the problem with the C++/FORTRAN side: readability (equational
-and algebraic expressions) and ease of reasoning is sacrificed at
-the altar of performance and efficiency.
+illustrates the problem with low-level libraries: readability (equational and
+algebraic expressions) and ease of reasoning is sacrificed at the altar of
+performance and (memory) efficiency.
 
-Although escape-hatches do exist in value-semantic libraries, their use is
+Although escape-hatches do exist in high-level libraries, their use is
 discouraged. Systematic consideration of performance requires lowering the
-level of abstraction a programmer is working on. At this level, several
+level of abstraction a programmer is working on. At this level, \textbf{several}
 factors such as memory layout, allocation, re-use as well as cache behaviour
-and parallelism become apparent. Of these, \textbf{memory allocation and
-re-use} are relevant to linear-types and this thesis.
+and parallelism are important. Of these, \textbf{only memory allocation and
+re-use} are relevant to linear types and this thesis.
 
@@
 
-In Fortran (Figure~\ref{fig:fortran_blas}), data is typically allocated
-statically (at compile time) so temporary storage for all intermediate values
-must be managed by the programmer. While this approach leads to verbose and
-less readable code, the explicitness is good for understanding the memory
-concerns of the program, albeit at the expense of understanding what the
-program is actually calculating.
+In Fortran (Figure~\ref{fig:fortran_blas}) and C, temporary storage for all
+intermediate values must be managed by the programmer. While this approach
+leads to verbose, less readable and harder to reason about code, the
+explicitness is good for understanding the memory concerns of the program.
 
@@
 readable. For safety and correctness, expressions are typically handled with
-value-semantics. However, given \emph{extra} information about, aliasing
-(Eigen, Figure~\ref{fig:cpp_eigen}) or usage of intermediate expressions
-(uBLAS, Figure~\ref{fig:cpp_ublas}), the number of temporaries allocated can be
-reduced and increased \emph{implicitly} to improve performance (remove
-unnecessary allocations or re-calculations respectively). Further tricks to
-improve performance include expression templates (building up an
-expression-tree at compile time and then pattern-matching on it to produce
-code) and lazy evaluation (only calculating a result when it is needed). These
-will be discussed in more detail in Chapter~\ref{chap:related}.
+value-semantics (the result of any computation is a \emph{new} value, distinct
+from any used during the calculation of that value). However, given
+\emph{extra} information about aliasing (Eigen, Figure~\ref{fig:cpp_eigen}) or
+usage of intermediate expressions (uBLAS, Figure~\ref{fig:cpp_ublas}), the
+number of temporaries allocated can be reduced and increased \emph{implicitly}
+to improve performance (remove unnecessary allocations or re-calculations
+respectively). Further tricks to improve performance include expression
+templates (building up an expression-tree at compile time and then
+pattern-matching on it to produce code) and lazy evaluation (only calculating a
+result when it is needed). These will be discussed in more detail in
+Chapter~\ref{chap:related}.
 
-It is important to note that should these annotations (in
-Figures~\ref{fig:cpp_eigen} and~\ref{fig:cpp_ublas}) be wrong, the program's
-behaviour is very likely to end up being undefined (like how \texttt{memcpy()}
-for overlapping regions of memory is explicitly undefined in the POSIX and C
-standards). Indeed, one of Fortran's strengths lies in assuming that references
-cannot be aliased (with certain caveats) in more cases than C permits (this
-informal, general statement comes with many nuances left for the interested
-reader to pursue).
+It is important to note that should a \texttt{noalias} annotation be wrong, the
+program's behaviour is very likely to end up being undefined (like how
+\texttt{memcpy()} for overlapping regions of memory is explicitly undefined in
+the POSIX and C standards). Indeed, one of Fortran's strengths lies in assuming
+that references cannot be aliased (with certain caveats) in more cases than C
+permits (this informal, general statement comes with many nuances left for the
+interested reader to pursue).
 
@@
 Let us have a look at a few examples of functions we can write with linear
-types.  We can define the canonical factorial function
+types.  We can define the factorial function
 (Figure~\ref{fig:ltfla_factorial}) and sum over an array
@@
 is enough to know it means we can only \emph{read} from \ltfla{row}, and not
-write to it. If we did try to write to or free \ltfla{row}, say by adding
-another line as shown in Figure~\ref{fig:ltfla_univ}, would get a helpful error
-message, as shown.
+write to it. If we did try to write to or free \ltfla{row}, we would get a
+helpful error message, as shown in Figure~\ref{fig:ltfla_univ}.
 
-\begin{figure}[p]
+\begin{figure}[tp]
     \begin{minted}[linenos, firstnumber=7, fontsize=\small]{ocaml}
@@
 
-Now suppose we were trying to square a matrix, using a `dgemm' like BLAS
+Now suppose we are trying to square a matrix, using a `dgemm' like BLAS
 routine (called `simple\_dgemm'\footnote{For the purposes of this example, I
@@
 square a matrix, assuming $C=0$ initially?  Surely this would break linearity,
-since $A$ would be have to be given to the function twice, and we can only do
-so once?
+since $A$ would have to be passed in to the function twice, and we can only 
+use non-!-annotated variables once?
 
-We can use a special primitive called \ltfla{share} to produce (more) read-only
-\emph{aliases} of any matrix. We then pass these into the function, and it
-works as expected. Once the squared matrix has been obtained, we may not want
-the original any more, and thus decide to free it. Before we do so, we must
-first \ltfla{unshare} any read-only aliases that exist -- in this context,
-\ltfla{unshare} swallows them and gives back a single, read-write handle.
+To solve this, we use a special primitive called \ltfla{share} to produce
+(more) read-only \emph{aliases} of any matrix. We then pass these into the
+function, and it works as expected. Once the squared matrix has been obtained,
+we may not want the original anymore, and thus decide to free it. Before we do
+so, we must first \ltfla{unshare} any read-only aliases that exist, to get back
+a single, read-write handle.
 
-\begin{figure}[p]
+\begin{figure}[tp]
     \begin{minted}[linenos, fontsize=\small]{ocaml}
@@
 
-If we tried to free one of the read-only aliases instead of
-\ltfla{unshare}-ing, then we would get the error shown in
+If we tried to free one of the read-only aliases before or instead of
+\ltfla{unshare}-ing them, then we would get the error shown in
 Figure~\ref{fig:ltfla_free}. Briefly, a \ltfla{z arr} is a read-write array,
@@
 
-\begin{figure}[p]
+\begin{figure}[tp]
     \begin{minted}[linenos, fontsize=\small]{ocaml}
@@
 More generally, by annotating variables in the context with their usage (when
-implementing a type-checker for a linearly typed language), we can express the
+implementing a type checker for a linearly typed language), we can express the
 rules of \emph{substructural} (including affine, relevant and ordered type
@@
 management of aliasing, read/write permissions, memory allocation, re-use and
-deallocation with a \emph{type system}. This prevents a whole class of errors
-that can occur with lower-level languages \emph{at compile time}. I
-demonstrated these features using the context of linear-algebra libraries as as
-a specific example. In the next chapter, I will show that these features can be
-further applied to the problem of \emph{matrix expression compilation}. I will
-also explain how to express and implement them so that they can be provided and
-used \emph{as a library}.
+deallocation with a \emph{type system}. This prevents a \textbf{whole class} of
+errors that can occur with low-level languages \emph{at compile time}. For
+example, \textbf{the types make it impossible to use `simple\_dgemm'
+incorrectly}. I demonstrated these features using the context of linear algebra
+libraries as as a specific example. In the next chapter, I will show that these
+features can be further applied to the problem of \emph{matrix expression
+compilation}. I will also explain how to express and implement them so that
+they can be provided and used \emph{as a library}.

### 2018-05-28 write-up/conclusion.tex
Fix many small writing errors

@@
 permissions, memory allocation, re-use and deallocation}; I provided a detailed
-description of each feature of the type-system, its intent and implementation
+description of each feature of the type system, its intent and implementation
 in Section~\ref{sec:core_lang}.
@@
 compiler to write, a \textbf{non-trivial, yet readable} (equational and
-algebraic) linear-algebra program (Kalman filter, as described in
+algebraic) linear algebra program (Kalman filter, as described in
 Section~\ref{sec:matrix_exps}). I shared how, in doing so, I could use the
@@
 I showed that this implementation was \textbf{more memory-efficient} than one
-written using a high-level linear-algebra libraries and performed just as
-\textbf{predictably} as one written using a low-level linear-algebra library,
+written using a high-level library and performed just as
+\textbf{predictably} as one written using a low-level library,
 enabling a programmer to have the best of both worlds
@@
 OCaml library}, by generating \textbf{readable and not obviously safe} OCaml
-code (1) which looks like it is written by a human/expert and (2) whose types
+code (1) which looks like it is written by an expert and (2) whose types
 reflect its intensional behaviour and assumptions (interpreting types into
@@
 
-LT4LA is a proof-of-concept design for linearly-typed DSL for writing
+LT4LA is a proof-of-concept design for linearly typed DSL for writing
 linear-algebra programs in a way that can work with existing languages and
@@
 of LT4LA's core language would be a language expressive enough to provide a
-usable, linearly-typed interface to the unsafe parts of the Owl library for
+usable, linearly typed interface to the unsafe parts of the Owl library for
 OCaml, perhaps as a functor like its current, `Lazy' interface.  To do that,
@@
 \emph{inference}, to automate the reasoning I did manually when implementing a
-memory-efficient Kalman filter.
+memory-efficient Kalman filter (Section~\ref{sec:qual_benefits}).
 
@@
 expressivity, to bring it closer to Fortran and C so that it would be easier to
-emit code in those languages. A full set of linearly-typed BLAS/LAPACK bindings
+emit code in those languages. A full set of linearly typed BLAS/LAPACK bindings
 (to encode information in documentation at the type level) would be a good
@@
             \item allow the user to choose: use a matrix expression compiler when
-                desired and drop down to a usable typed-IR for finer control,
+                desired and drop down to a usable typed IR for finer control,
                 whilst still retaining safety guarantees.
@@
     \item \textbf{Formal verification of matrix expression compilers} by
-        precisely specifying source and target languages.
+        precisely specifying the intensional and extensional properties of the
+        source and target languages.
 
-    \item \textbf{Multi-stage programming} to use information only available
-        at runtime in many situations (such as sizes, matrix properties,
-        sparsity, control flow) can be effectively incorporated into code
-        generated.
+    \item \textbf{Multi-stage programming} to use information only available at
+        runtime (such as sizes, matrix properties, sparsity, control flow) for
+        generating more specialised code.
 
@@
         instead of the whole. This is already prevalent with `dsymm' like BLAS
-        routines which only read and write to the lower or upper triangle of a
-        matrix.  This idea is inspired by Conor McBride's talk on writing to
+        routines which only read the lower or upper triangle of a matrix.  This
+        idea is inspired by Conor McBride's talk on writing to
         terminals with ``Space Monads''~\cite{space_monads}.

### 2018-05-28 write-up/evaluation.tex
Remove onehalfspacing
Fix many small writing errors

@@
     high-level linear algebra libraries and (2) performs just as predictably as
-    code written using low-level linear-algebra libraries. I will also
+    code written using low-level linear algebra libraries. I will also
     elaborate on the qualitative benefits of using linear types to write
-    linear-algebra programs.
+    linear algebra programs.
 }%
@@
 To test whether code written using LT4LA performs just as predictably as code
-written using low-level linear-algebra libraries, I wrote two implementations
+written using low-level linear algebra libraries, I wrote two implementations
 of a Kalman filter:
@@
 system call (called \ltfla{Unix.times} in OCaml), sandwiched between calls to
-\ltfla{Gc.full_major} to ensure no garbage collection took place during the
-measurements.  Core\_bench performs a linear-regression (here, time against
-batch-size) so includes 95\% confidence-interval with $R^2$. Larger scales have
+\ltfla{Gc.full_major} to minimise the effects of garbage-collection.
+Core\_bench performs a linear regression (here, time against batch-size) so
+includes a 95\% confidence-interval with $R^2$. Larger scales have
 errors reported to $\pm \sigma$ (one standard-deviation).
+
+\subsection{Memory Usage}
+
+Inspecting the trace for LT4LA shows 6 calls to `empty' (function for
+allocating new matrices): 4 temporaries plus 2 for storing the results. From
+the CBLAS implementation (Figure~\ref{fig:cblas_kalman}), we can see that it
+too has 6 calls to malloc: 4 temporaries plus 2 for storing the results.
 
@@
 
-\subsection{Memory Usage}
-
-Inspecting the trace for LT4LA shows 6 calls to `empty' (function for
-allocating new matrices): 4 temporaries plus 2 for storing the results. From
-the CBLAS implementation (Figure~\ref{fig:cblas_kalman}), we can see that it
-too has 6 calls to malloc: 4 temporaries plus 2 to storing the results.
-
 \subsection{Execution Time}
@@
 measurements differ from the means by at most $2 \mu s$. For $n=125$, and
-around 110 samples, CBLAS was \emph{slower} ($1803 \mu s\, [1746,\, 1867]$)
+around 110 samples each, CBLAS was \emph{slower} ($1803 \mu s\, [1746,\, 1867]$)
 than LT4LA ($1678 \mu s\, [1646,\, 1714]$). For $n=625$ and 1000 samples each,
@@
 is due to large sample sizes causing more allocations and thus more
-garbage-collector unpredictability (which the linear-regression \emph{did not}
+garbage-collector unpredictability (which the linear regression \emph{did not}
 take into account; multi-variate regressions are not fully supported by
@@
 
-Therefore, by using LT4LA, it is possible to write linear-algebra programs that
+Therefore, by using LT4LA, it is possible to write linear algebra programs that
 perform just as predictably (with regards to memory usage and execution time)
-as code written using low-level linear algebra libraries, but gain readability
-(equational and algebraic expressions) and safety (with respect to aliasing
+as code written using low-level libraries, but gain readability (equational and
+algebraic expressions) and safety (with respect to aliasing
 and read/write permissions).
@@
 
-To test whether code written using LT4LA is more memory efficient than code
-written using high-level linear-algebra libraries, I wrote two\footnote{I
-attempted a third implementation using Owl's \emph{Lazy} module but it did
-not support matrix inversion at the time of writing.} more implementations
-of a Kalman filter:
+To test whether code written using LT4LA is more memory-efficient than code
+written using high-level libraries, I wrote two\footnote{I attempted a third
+implementation using Owl's \emph{Lazy} module but it did not support matrix
+inversion at the time of writing.} more implementations of a Kalman filter:
 
@@
         yet provide a non-mutating `potrs' function, so I wrote my own which
-        returns a mutated copy of its argument instead.}\label{fig:chol_owl_kalman}
+        returns a mutated \emph{copy} of its argument
+        instead.}\label{fig:chol_owl_kalman}
 
@@
 is possible to have the readability (equational and algebraic expressions) and
-safety of high-level linear algebra libraries, whilst gaining precise and
-explicit control over memory allocation and re-use.
+safety of high-level libraries, and gain precise and explicit control over
+memory allocation and re-use.
 
@@
 unnecessary temporary matrices when implemented idiomatically in a high-level
-linear-algebra library. It is good for isolating the key differences between
-not having and having linear types to help a programmer safely manage memory
-and aliasing, whilst excluding other aspects also important to real world
-linear-algebra programs such as control flow or blocking.
+library. It is good for isolating the key differences between not having and
+having linear types to help a programmer safely manage memory, aliasing, and
+read/write permissions, whilst excluding other aspects also important to real
+world linear algebra programs such as control flow or blocking.
 
@@
 A graph of the execution times (with error bars which are present but quite
-small) \emph{all four} implementations is show in Figure~\ref{fig:timings_all}.
+small) of \emph{all four} implementations is show in Figure~\ref{fig:timings_all}.
 
@@
 
+    \item CBLAS/LT4LA use Cholesky decomposition: it performs \emph{half} as
+        many operations as LU decomposition (as used by Owl) under the
+        assumption that its argument is a symmetric matrix.
+
     \item All of the implementations either directly or indirectly (via Owl's
@@
 to their needs because (1) it met its goal of enabling a programmer to write
-code that performs just as predictably as code written using lower-level
-linear-algebra libraries (2) it does so \emph{without} the associated dangers
-of unenforced and uncheckable rules around aliasing, read/write permissions,
-memory allocation, re-use and deallocation that would come with using C,
-Fortran, or the unsafe primitives of Owl.
+code that performs just as predictably as code written using low-level
+libraries (2) it does so \emph{without} the associated dangers of unenforced
+and uncheckable rules regarding aliasing, read/write permissions, memory
+allocation, re-use and deallocation that would come with using C, Fortran, or
+the unsafe primitives of Owl.
 
@@
 memory allocation) code. To justify their why I think they are a
-\emph{practical} and \emph{usable} way to do so, I will elaborate on some of
+\emph{practical} and \emph{usable} way to do so, I will elaborate on
 the qualitative benefits I experienced whilst using LT4LA.
 
-Prior to this project, I had no experience with linear-algebra libraries or the
+Prior to this project, I had no experience with linear algebra libraries or the
 problem of matrix expression compilation. As such, I based my initial LT4LA
@@
 
-Once I translated the implementation, I attempted to compile it and found that
-(to my surprise) it did not type-check. This was because the original
-implementation contained incorrect aliasing, did not adhere to Fortran's
-read/write permissions (with respect to \texttt{intent} annotations
-\texttt{in}, \texttt{out} and \texttt{inout}) and unused and unnecessary
-temporaries, all of which were now highlighted by LT4LA's type system.
+Once I translated the implementation from Fortran to LT4LA, I attempted to
+compile it and found that (to my surprise) it did not type-check. This was
+because the original implementation contained incorrect aliasing, unused and
+unnecessary temporaries, and did not adhere to Fortran's read/write permissions
+(with respect to \texttt{intent} annotations \texttt{in}, \texttt{out} and
+\texttt{inout}) all of which were now highlighted by LT4LA's type system.
 
 The original implementation used 6 temporaries, one of which was immediately
-spotted as not being used due to linearity. It also contained two variables
+spotted as never being used due to linearity. It also contained two variables
 which were marked as \texttt{intent(in)} but would have been written over by
@@
 
-Writing a linear-algebra program using LT4LA combines the best of low- and
-high-level linear-algebra libraries: it gives a programmer \emph{explicit
+Writing a linear algebra program using LT4LA combines the best of low- and
+high-level libraries: it gives a programmer \emph{explicit
 control} (over read/write permissions, aliasing and memory allocation, re-use
@@
 precise intent} to the reader and compiler and perform similarly to equivalent
-programs written directly using lower-level linear-algebra libraries. In turn,
+programs written directly using low-level libraries. In turn,
 such programs can be \emph{checked automatically} against their intent,

### 2018-05-28 write-up/implementation.tex
Fix minor layout issues
Fix many small writing errors

@@
     the features of its core language and show how readable (equational and
-    algebraic expressions) and easy to reason about linear-algebra programs
+    algebraic expressions) and easy to reason about linear algebra programs
     can be elaborated into memory-efficient ones in the core language that are
@@
     such programs can be transpiled to OCaml code that is not obviously safe.
-    Although I refer to OCaml-specific features of the type-checker and
+    Although I refer to OCaml-specific features of the type checker and
     transpiler, I believe the ideas described in this chapter can easily be
     implemented in other languages and are also general enough to transpile to
-    different back-end languages, such as C or Fortran.
+    other back-end languages, such as C or Fortran.
 }%
@@
 error-messages in the hope that someone who did not understand linear types
-could still use the LT4LA (3) write tests and set-up
-continuous-integration for all non-trivial functions so that I could spot and
-correct errors that were not caught by OCaml's type system whenever I
-implemented new features or refactored my code.
+could still use LT4LA (3) write tests and set-up continuous-integration for all
+non-trivial functions so that I could spot and correct errors that were not
+caught by OCaml's type system whenever I implemented new features or refactored
+my code.
 
@@
 
-    \item \textbf{Parsing \& Desugaring}. A generated, LR(1) parser parses a text
-        file into a syntax tree, which is then desugared into a smaller, more
-        concise abstract syntax tree. The former aims to mimic OCaml syntax
-        with a few extensions and keywords so that it is familiar and thus easy
-        to pick-up for OCaml users. The latter allows for the type-checker to
-        be simpler to implement and easier to specify. In general, this part
-        will vary for different languages or can be dealt with differently
-        using combinators (the EDSL approach) or a syntax-extension if the host
-        language offers such support.
+    \item \textbf{Parsing}. A generated, LR(1) parser parses a text file into a
+        syntax tree. I tried to mimic OCaml syntax modulo a few extensions and
+        keywords so that it is familiar and thus easy to pick-up for OCaml
+        users. In general, this part will vary for different languages and can
+        also be dealt with using combinators or syntax-extensions (the EDSL
+        approach) if the host language offers such support .
 
-    \item \textbf{Type-checking}. The abstract syntax tree is explicitly-typed,
-        with some inference to make it less verbose and more convenient to
-        write typical programs.
+    \item \textbf{Desugaring}. The syntax tree is then desugared into a
+        smaller, more concise, abstract syntax tree. This allows for the type
+        checker to be simpler to specify and easier to implement.
 
-    \item \textbf{Matrix Expressions}. During type-checking, if a matrix-expression
-        is encountered, it is either successfully elaborated into an expression
-        in the abstract syntax tree which is then consequently type-checked, or
-        fails to find suitable routines to calculate the given expression.
+    \item \textbf{Matrix Expressions} are also desugared into the abstract
+        syntax tree through simple, yet effective pattern-matching.
+
+    \item \textbf{Type checking}. The abstract syntax tree is explicitly typed,
+        with some inference to make writing typical programs more convenient.
 
     \item \textbf{Code Generation}. The abstract syntax tree is translated into
-        standard OCaml and a few-particular `optimisations' are made to produce
-        more readable code. This process is type-preserving: the linear type
-        system is embedded into OCaml's type system, and so when the OCaml compiler
-        compiles the generated code, it acts as a sanity check on the code produced.
+        OCaml, with a few `optimisations' to produce more readable code. This
+        process is type-preserving: LT4LA's type system is embedded into
+        OCaml's (Figure~\ref{fig:type_grammar}), and so the OCaml type checker
+        acts as a sanity check on the generated code.
 
@@
 Appendix~\ref{chap:ott_spec}. For convenience, I have reproduced its type
-grammar and defined how I interpret those types in OCaml's type system in
-Figure~\ref{fig:type_grammar}.  The point of LT4LA is to keep the same values
-and types we are familiar with, but \emph{restrict their usage} in certain
-ways. LT4LA's core language's main features are: intuitionistic values,
-value-restriction, fractional-capabilities (inferrable at call sites),
-if-expressions and recursion.
+grammar and its translation into OCaml in Figure~\ref{fig:type_grammar}.  The
+point of LT4LA is to keep the same values and types we are familiar with from
+OCaml, but \emph{restrict their usage}. LT4LA's core language's main features
+are: intuitionistic values, value-restriction, fractional-capabilities
+(inferrable at call sites), if-expressions and recursion.
 
@@
     \end{minipage}
-    \caption{Type grammar of LT4LA (left) and an informal interpretation of it
-        into OCaml types (right).}\label{fig:type_grammar}
+    \caption{Type grammar of LT4LA (left) and my translation of it
+        into OCaml (right).}\label{fig:type_grammar}
 \end{figure}
@@
 
-To make LT4LA a usable DSL, I needed some way of using values zero or more than
-once, as a programmer would an intuitionistic value. For this, I added the
-!-constructor at the type-level of the DSL and the \ltfla{Many}-constructor and
-the \ltfla{let Many <id> = .. in .. } eliminator at the term-level of the DSL.
-The idea behind the !-type is that the value uses no `resources'
+To make LT4LA a usable DSL, I needed a way of allowing values to either never
+be used or be used more than once. For this, I added the \mbox{!-constructor}
+at the type-level of the DSL and its corresponding introduction (the
+\ltfla{Many}-constructor) and elimination (the \ltfla{let Many <id> = .. in ..
+} destructor) forms at the term-level of the DSL.  The idea behind the
+\mbox{!-constructor} is that a value of that type uses no `resources'
 (linearly-typed expressions). To start off with, it is enough to say that
-anything which can be passed around by copying, will have a !-type. This
+anything which can be passed around by copying, will have a \mbox{!-type}. This
 includes integers, elements and booleans. So \ltfla{3 : !int} and \ltfla{3. +.
-4. : !elt}. However, all bindings are still linear by default, so to
-emulate intuitionism, I desugar \ltfla{let !x = <exp> in <body>} to
-\ltfla{let Many x = <exp> in let Many x = Many (Many x) in <body>}
-(similarly for function argument bindings). The reader can check (using the
-rules in Appendix~\ref{chap:ott_spec}) that this has the effect of moving
-\ltfla{x : !t} from the linear to the intuitionistic environments, only if
-\ltfla{<exp> : !t}.
+4. : !elt}. However, all bindings are still linear by default, so to emulate
+intuitionism, I desugar \ltfla{let !x = <exp> in <body>} to \ltfla{let Many x =
+<exp> in let Many x = Many (Many x) in <body>} (similarly for function argument
+bindings). The reader can check (using the rules in
+Appendix~\ref{chap:ott_spec}) that this has the effect of moving \ltfla{x : !t}
+from the linear to the intuitionistic environments, only if \ltfla{<exp> : !t}.
 
@@
 in <body>}; the expression \ltfla{array 5} uses no linearly-typed variables
-from the linear environment. Although we could just reject types of the form
+from the environment. Although we could just reject types of the form
 \ltfla{!(_ arr)} to fix this simple example, what about pairs \ltfla{let Many
@@
 behaviour of the DSL's \ltfla{Many}-constructor? One option is to go down the
-C++ route and make \ltfla{Many} act like a \ltfla{shared_ptr} and act as a
+C++ route and make \ltfla{Many} behave like a \ltfla{shared_ptr} and act as a
 runtime reference-count for arrays. I chose to not go for this option because
@@
 
-My aim is to show linear types are simple to understand and apply to
-linear-algebra libraries, enough so that it can be grafted (in a limited way)
-on to existing languages as a \emph{library}. In that spirit, the simplest
-thing that the DSL's \ltfla{Many}-constructor can mean in the OCaml runtime is
-\emph{nothing}.  This language construct is translated into a standard OCaml
-type-and-constructor of the form \ltfla{type 'a bang = Many of 'a [@@unboxed]}.
-The \emph{unboxed} annotation means that the type and its constructor only
-exist for the purpose of type-checking in OCaml; \emph{the OCaml runtime
-representation of values of type \ltfla{'a} and \ltfla{'a bang} is exactly the
-same}.
+My aim is to show linear types are simple to understand and apply, enough so
+that they can be grafted (in a limited way) on to existing languages as a
+\emph{library}. In that spirit, the simplest thing that the DSL's
+\ltfla{Many}-constructor can mean in the OCaml runtime is \emph{nothing}.
+LT4LA's !/\ltfla{Many} constructors are translated into OCaml
+\ltfla{bang}/\ltfla{Many} constructors declared as \ltfla{type 'a bang = Many
+of 'a [@@unboxed]}.  The \emph{unboxed} annotation means that the type and its
+constructor only exist for the purpose of type checking in OCaml; \emph{the
+OCaml runtime representation of values of type \ltfla{'a} and \ltfla{'a bang}
+is exactly the same}.
 
@@
 calling a function with an array argument copies a \emph{pointer} to the array
-rather than the array itself, instead of the \emph{value} itself. So, we can
-start making a distinction, \emph{defining} elements, integers, booleans,
-intuitionistic variables, units and lambda-expressions that capture no linear
-variables as \emph{values} (since they cannot break referential-transparency)
-and anything else (arrays, matrices, expressions which can be reduced, such as
-function application and if-expressions) as not being `values'. If this sounds
-familiar, it is because this is the same \emph{value-restriction} `trick' from
-the world of polymorphic types applied to linearity instead. We then have the
-rule that in LT4LA, we can only use \ltfla{Many} on expressions that are
-defined to be \emph{values} and \emph{use no linear variables}.
+rather than the array itself. So, we can start making a distinction,
+\emph{defining} elements, integers, booleans, intuitionistic variables, units
+and lambda-expressions that capture no linear variables as \emph{values} (since
+they cannot break referential-transparency) and anything else (arrays,
+matrices, expressions which can be reduced, such as function application and
+if-expressions) as not being `values'. If this sounds familiar, it is because
+this is the same \emph{value-restriction} `trick' from the world of polymorphic
+types applied to linearity instead. We then have the rule that in LT4LA, we can
+only use \ltfla{Many} on expressions that are defined to be \emph{values} and
+\emph{use no linear variables}.
 
@@
 
-I started off with linearity and showing how it helps a programmer keep track
-of of memory allocation and deallocation, and then demonstrated how to add
+I started off with linearity and showed how it helps a programmer keep track
+of memory allocation and deallocation, and then demonstrated how to add
 intuitionism for the only values we wish to be intuitionistic. I will now
@@
 \emph{fractional-capabilities}~\cite{boyland}.  A fraction of 1 ($2^0$)
-represents complete ownership of a value; in particular, this allows a
-programmer to write or free it. Creating an array gives you ownership of it;
-the function \ltfla{array : !int --o z arr} (where \ltfla{z} represents `0').
-Once you have ownership of an array, you can free it: \ltfla{free : z arr --o
-unit}.  Importantly, because a linear-value may only be used once, the array
-just freed is \emph{out of scope} for following expressions, preventing
-use-after-free.  Ownership also enables you to write to the array: \ltfla{set :
-z arr --o !int --o !elt --o z arr} (the syntax \ltfla{w[i] := j} is just sugar
-for \ltfla{set w i j}). Here, linearity prevents accessing aliases which
-represented the array \emph{before} the mutation.
+represents complete ownership of a value. Creating an array gives you ownership
+of it; the function \ltfla{array : !int --o z arr} (where \ltfla{z} represents
+`0').  Once you have ownership of an array, you can free it: \ltfla{free : z
+arr --o unit}.  Importantly, because a linear value may only be used once, the
+array just freed is \emph{out of scope} for following expressions,
+\textbf{preventing use-after-free}.  Ownership also enables you to write to the
+array: \ltfla{set : z arr --o !int --o !elt --o z arr} (the syntax \ltfla{w[i]
+:= j} is just sugar for \ltfla{set w i j}). Here, linearity prevents accessing
+aliases which represented the array \emph{before} the mutation.
 
@@
 requirement that function declarations need type-annotations for their
-arguments, \textbf{this allows a fractional-capability to be \emph{correctly
+arguments, \textbf{this allows fractional-capabilities to be \emph{correctly
 inferred} in \emph{any} LT4LA expression}.
@@
 the primitives \ltfla{share : 'x . 'x arr --o ('x s) arr * ('x s) arr} and
-\ltfla{unshare: 'x .  ('x s) arr --o ('x s) arr --o 'x arr}.  For the former, two
-arrays returned (which happen to just be the given array) can now only be read
-from and not written to. If you want to write to this array, you must use the
-latter to combine other read-only aliases until you are left with a value of
-type \ltfla{z arr}, guaranteeing no other aliases exists.
+\ltfla{unshare: 'x .  ('x s) arr --o ('x s) arr --o 'x arr}.  For the former,
+the two arrays returned (which are just aliases of the given array) can now
+only be read from and not written to. If you want to write to this array, you
+must use the latter to combine other read-only aliases until you are left with
+only one value of type \ltfla{z arr}, guaranteeing no other aliases exists.
 
@@
 using easy to reason about, value-semantic expressions whose \emph{intensional}
-behaviour (regarding aliasing, read/write permissions and memory allocation,
-re-use and de-allocation)  is also explicitly and precisely described and
-\emph{automatically checkable}. Now the programmer need not resort to manually
-figuring out and inserting \texttt{noalias} annotations and worrying about what
-variables can and cannot be written to, re-used or freed; instead they can let
-the loyal and tireless compiler do the heavy lifting.
+behaviour and assumptions (regarding aliasing, read/write permissions and
+memory allocation, re-use and de-allocation) are also explicitly and precisely
+described and \emph{automatically checkable}. Now the programmer need not
+resort to manually figuring out and inserting \texttt{noalias} annotations and
+worrying about what variables can and cannot be written to, re-used or freed;
+instead they can let the loyal and tireless compiler do the heavy lifting.
 
@@
 
-An if-expression's condition may evaluate either way at run time, so a
+An if-expression's condition may evaluate either way during execution, so a
 programmer must guarantee that both branches use the same set of linear
-variables. Writing the type-checker in a pure-functional monadic style paid off
-here because I could now sandwich monadic values with state-adjustments either
-side of it. Given two monadic values that represented type-checking two
+variables. Writing the type checker in a pure-functional monadic style paid off
+here because I could now sandwich monadic values with state adjustments either
+side of it. Given two monadic values that represented type checking two
 branches of an if-expression, I could use the code in
@@
 
-    \caption{Implementation of same\_resources helper method for type-checking
+    \caption{Implementation of same\_resources helper method for type checking
         if-expressions. Note the monadic style helped compose computations that
-        affected the type-checker's state in a simple manner.}\label{fig:same_resources}
+        affected the type checker's state in a simple manner.}\label{fig:same_resources}
 
@@
 (*..*) : !int) in <body> }. However, \ltfla{fix}, like \ltfla{Many}, must also
-not use any linear variables from its surrounding scope, because calling a
+not use any linear variables from its surrounding scope, because a
 recursive function may evaluate its body multiple times.
@@
 (Figure~\ref{fig:maths_kalman}) applicable to a wide variety of problems. From
-the comments, we see that every variable is annotated with the step/matrix
+the comments, we see that every variable is annotated with the matrix
 expression that it will hold at some point during the computation (an
@@
 \begin{itemize}
-    \item making it impossible to refer to steps/values which are no longer usable,
+    \item making it impossible to refer to values which are no longer usable,
     \item ensuring all values are declared and \emph{initialised} correctly before they are used,
@@
 Indeed, an inexperienced programmer could take the  na\"ive approach of just
-copying sub-expressions by default and then letting the compiler tell it which
-copies are never used and removing them systematically until it type-checks.
-While it is not quite a black-box, push-button compilation of an expression, I
-would argue that, it is just as easy (if not easier) to become familiar with as
-Rust and its borrow-checker.
+copying sub-expressions by default and then letting the compiler tell them
+which copies are never used and removing them systematically until their
+program type checks.  While it is not quite a black-box, push-button
+compilation of an expression, I would argue it is just as easy (if not easier)
+to learn how to work with as Rust's borrow checker.
 
@@
         using easy to reason about, value-semantic expressions. In addition to
-        that, its \emph{intensional} behaviour(regarding aliasing, read/write
-        permissions and memory allocation, re-use and de-allocation) is also
-        explicitly and precisely described and (unlike with Fortran or C)
+        that, its \emph{intensional} behaviour and assumptions (regarding
+        aliasing, read/write permissions and memory allocation, re-use and
+        de-allocation) are also explicitly and precisely described and
         \emph{automatically checkable}.}\label{fig:ltfla_kalman}
@@
 All of the syntax in Figure~\ref{fig:ltfla_kalman} can be unambiguously
-desugared \emph{before} type-checking, through fairly simple pattern matching.
+desugared \emph{before} type checking, through fairly simple pattern matching.
 Matrix expression compilation is well-trodden territory in
@@
 One of the first hurdles I encountered was compositionality: to compile $AB +
-C$ it is typically better to use a BLAS routine directly rather than but first
+C$ it is typically better to use a BLAS routine directly rather than first
 compiling $A$, then $B$, then adding a call to multiply them, then compiling
@@
 Another compositionality problem is that a call to a linear function does not
-just return a result, but a sequence of re-bindings that dictate which
-variables are still in scope/usable.  As such, to compile an expression, you
+just return a result, but a sequence of re-bindings which dictate which
+variables are still usable (in scope).  As such, to compile an expression, you
 need to provide a CPS-style function of type $var \rightarrow exp$ representing
 how you would use a variable representing the result in the rest of the
-expression.
+computation.
 
@@
 compiler, to `re-sugar' some of the constructs translated where appropriate.
-Almost all of them involved the erasing the !-eliminator which is not needed in
-regular, intuitionistic OCaml. So, so we can simply replace any expression of
+Almost all of them involved the erasing the \mbox{!-eliminator} which is not needed in
+regular, intuitionistic OCaml. So, we can simply replace any expression of
 the form:
@@
     \item \ltfla{let xy, z = <exp> in let x,y = xy in <body>} with
-        \ltfla{let (x,y), z = <exp in <body>},
+        \ltfla{let (x,y), z = <exp> in <body>},
 
@@
     \end{minted}
-    \caption{Recursive OCaml function for a summing an array, generated (at
+    \caption{Recursive OCaml function for a summing over an array, generated (at
         \emph{compile time}) from the code in Figure~\ref{fig:ltfla_sumarray},
@@
 and~\ref{fig:ocaml_kalman} are realisations of a concise, linearly-typed Kalman
-filter~\ref{fig:ltfla_kalman} \emph{specification} that describes the whole
-behaviour and intent (with respect to read/writes permissions, memory
-management and aliasing) of the program and the BLAS primitives it uses more
-accurately than a Fortran, C or OCaml implementation could.
+filter~\ref{fig:ltfla_kalman} \emph{specification} that describes the intensional
+behaviour and assumptions (with respect to read/writes permissions, memory
+management and aliasing) of the program and the BLAS primitives it uses \textbf{more
+accurately} than a Fortran, C or OCaml implementation could.
 
@@
 
-Advantages of generating code at compile time include avoiding runtime overhead
-as well as catching any interface changes (type-signatures now include
-\emph{intensional} information about which parameters are read from or written
-to) between the generated code and the code that uses it via OCaml's type
-system, \emph{another benefit} (in addition to a sanity check on LT4LA's
-type-checker implementation) of \emph{embedding} LT4LA's type system inside of
-OCaml's.
+Generating code at compile time has two advantages. First, it avoids runtime
+overhead. Second, it catches interface and \textbf{relevant implementation
+changes} between the generated code and the code that uses it.  The second
+advantage is \emph{another benefit} of \emph{embedding} LT4LA's type system
+inside of OCaml's. Relevant implementation changes are caught because LT4LA's
+types express some of the intensional behaviour and assumptions of its
+programs.
 
@@
 functions which present a safe, value-semantic interface but use unsafe,
-mutating operations on the inside could now be expressed using LT4LA and have
-an extra safety-net and automatic checking for their implementations.
+mutating operations on the inside could now be expressed using LT4LA and gain
+\textbf{safety, value-semantics and automatic checking for their implementations}.
 
+\clearpage%
 \section{Summary}
 
-I have explained how a few core features -- linearity, the \ltfla{Many}
+I explained how a few core features -- linearity, the \ltfla{Many}
 constructor, value-restriction, fractional-capabilities with inference,
@@
 capture and automatically check} aliasing, read/write permissions, memory
-allocation, re-use and deallocation of non-trivial linear-algebra programs. I
-have also demonstrated that simple pattern-matching and desugaring provides the
-potential for a new, \emph{type-directed} approach to matrix expression
+allocation, re-use and deallocation of non-trivial linear algebra programs. I
+also demonstrated that simple pattern-matching and desugaring provides the
+potential for a \textbf{new, \emph{type-directed}} approach to matrix expression
 compilation. Lastly I have shown that it is possible to use these features with

### 2018-05-28 write-up/ott_spec.tex
Fix many small writing errors

@@
 
-The following pages present a specification of the grammar and type-system used
-by my project, produced using the Ott~\cite{sewell_ott} tool. It is important to note
-that the type-system described here is not how it is implemented: it is easier
-and clearer to describe the system as below for explaining. However, for implementing,
-I found it much more and user- and debugging-friendly to:
+The following pages present a specification of the grammar and type system used
+in LT4LA, produced using the Ott~\cite{sewell_ott} tool. It is important to
+note that the type system is not implemented how it is described in the coming
+pages.  For explaining (and using Ott) it was easier to set it out as below.
+However, for implementing, I found it much more and user- and
+debugging-friendly to:
 
@@
 
-    \item Implement it so that the type-environment \emph{changes} as a result
+    \item Have the type environment \emph{change} as a result
         of type-checking an expression, similar to the rules shown in
-        Figure~\ref{fig:example_rules}; with this, the below semantics describe
+        Figure~\ref{fig:example_rules}; with this, the below rules describe
         the \emph{difference} between the environment after and before checking
@@
         be handled uniformly for linear and intuitionistic variables and
-        scoping/variable look-up could be handled implicitly thanks to the
-        associative-list structure of the environment. So, the variable
-        rule would most accurately look like:
+        scoping/variable look-up could be handled implicitly with the
+        associative-list structure of the environment. So, it would be more
+        (but still not completely) accurate to define the variable rule as:
         \begin{prooftree}

### 2018-05-28 write-up/abstract.tex
Separate out acknowledgements onto its own sheet

@@
-\newpage
-{\normalfont\Huge\sffamily\bfseries Abstract}
-\vspace{24pt}
+% {\normalfont\Huge\sffamily\bfseries Abstract}
+% \vspace{24pt}
+\begin{abstract}
 
@@
 
+\end{abstract}
+
 \vspace*{\fill}
+\cleardoublepage%
+
 {\normalfont\Huge\sffamily\bfseries Acknowledgements}
+\vspace{24pt}
 
 I would like to thank my family and closest friends for supporting me through
-my time at Cambridge: you have been my strength when I have had none. I would
-also like to thank my supervisors, Dr\@.~Neelankantan Krishnaswami and
-Dr\@.~Stephen Dolan, for teaching me so much this year. I owe a debt of
-gratitude to both the 2017 Part III Appeals Committee and those who wrote
-letters in support of my appeal, for believing I deserved the chance to stay
-on: never have I valued the privilege of learning what I enjoy and enjoying
-what I learn more.  Lastly, I would like to thank the many staff and research
-members of the Computer Lab who have been approachable, patient and attentive
-during my time here.
+my time at Cambridge: you were my strength when I had none. I would also like
+to thank my supervisors, Dr\@.~Neelakantan Krishnaswami and Dr\@.~Stephen
+Dolan, for teaching me so much this year. I owe a debt of gratitude to both the
+2017 Part III Appeals Committee and those who wrote letters in support of my
+appeal, for believing I deserved the chance to stay on: never have I valued the
+privilege of learning what I enjoy and enjoying what I learn more.  Lastly, I
+would like to thank the many staff and research members of the Computer Lab who
+have been approachable, patient and attentive during my time here.
 
-\newpage
 \vspace*{\fill}
+\cleardoublepage%

### 2018-05-29 write-up/background.tex
Center source code figures

@@
 \subsection{One Too Many Copies and a Thousand Bytes Behind}
-\begin{figure}[tbp]
+\begin{figure}[tp]
     \centering
+
     \begin{minipage}{0.45\textwidth}
@@
 c = numpy.matmul(a,b)
-    \end{minted}
-    \begin{minted}[fontsize=\small]{julia}
 # Julia
@@
     \end{minipage}
-    \caption{Matrix Multiplication in Numpy (Python), Julia and Owl (OCaml).}\label{fig:mat_mul_copies}
+
+    \caption{Matrix Multiplication in Numpy (Python), Julia and Owl
+        (OCaml).}\label{fig:mat_mul_copies}
+
 \end{figure}
@@
 
-\begin{figure}[tbp]
+\begin{figure}[tp]
     \centering
@@
 
-\begin{figure}[tbp]
+\begin{figure}[tp]
+    \centering
     \begin{minted}[fontsize=\small]{fortran}
@@
 \begin{figure}[tp]
+    \centering
     \begin{minted}[fontsize=\small]{c++}
@@
 
-\begin{figure}[tbp]
+\begin{figure}[tp]
+    \centering
     \begin{minted}[fontsize=\small]{c++}
@@
 \begin{figure}[tp]
-    \inputminted[linenos, fontsize=\small]{ocaml}{../test/examples/factorial.lt}
+    \centering
+    \inputminted[fontsize=\small]{ocaml}{../test/examples/factorial.lt}
     \caption{Factorial function in LT4LA.}\label{fig:ltfla_factorial}
@@
 \begin{figure}[tp]
-    \inputminted[linenos, fontsize=\small]{ocaml}{../test/examples/sum_array.lt}
+    \centering
+    \inputminted[fontsize=\small]{ocaml}{../test/examples/sum_array.lt}
     \caption{Summing over an array in LT4LA.}\label{fig:ltfla_sumarray}
@@
 \begin{figure}[tp]
-    \begin{minted}[linenos, firstnumber=7, fontsize=\small]{ocaml}
+    \centering
+    \begin{minted}[fontsize=\small]{ocaml}
 let row = row[i] := x1 in (* or *) let () = free row in
@@
 \begin{figure}[tp]
-    \begin{minted}[linenos, fontsize=\small]{ocaml}
+    \centering
+    \begin{minted}[fontsize=\small]{ocaml}
 let (a1, a2) = share _ a in
@@
 
+\newpage%
 \section{Summary}

### 2018-05-29 write-up/implementation.tex
Center source code figures

@@
         \centering
-        \begin{minted}[fontsize=\small, linenos]{ocaml}
+        \begin{minted}[fontsize=\small]{ocaml}
 module Arr =
@@
 \begin{figure}[tp]
+    \centering
     \begin{minted}[fontsize=\small]{ocaml}
@@
     \centering
-    \begin{minipage}{1.47\textheight}
-    \inputminted[linenos, fontsize=\footnotesize]{fortran}{kalman.f90}
-    \end{minipage}
+    \inputminted[fontsize=\footnotesize]{fortran}{kalman.f90}
     \caption{Kalman filter in Fortran 90.}\label{fig:fortran_kalman}
@@
 \begin{figure}[tp]
-    \inputminted[linenos, fontsize=\small]{ocaml}{../test/examples/kalman.lt}
+    \centering
+    \inputminted[fontsize=\small]{ocaml}{../test/examples/kalman.lt}
     \caption{Kalman filter in LT4LA. In contrast with the Fortran
@@
     \centering
-    \begin{minted}[linenos, fontsize=\small]{ocaml}
+    \begin{minted}[fontsize=\small]{ocaml}
 let rec f i n x0 row =
@@
 \begin{figure}[p]
-    \begin{minted}[linenos, fontsize=\footnotesize]{ocaml}
+    \centering
+    \begin{minted}[fontsize=\footnotesize]{ocaml}
 let kalman sigma h mu r_1 data_1 =

### 2018-10-11 write-up/dissertation/evaluation.tex
Organise write-ups

--- 2018-05-28 write-up/evaluation.tex
@@
 
-\subsection{Memory Usage}
-
-Inspecting the trace for LT4LA shows 6 calls to `empty' (function for
-allocating new matrices): 4 temporaries plus 2 for storing the results. From
-the CBLAS implementation (Figure~\ref{fig:cblas_kalman}), we can see that it
-too has 6 calls to malloc: 4 temporaries plus 2 for storing the results.
-
 \begin{landscape}
@@
     \centering
-    \begin{minipage}{1.45\textheight}
-    \begin{minted}[linenos, fontsize=\small]{c}
+    \begin{minted}[fontsize=\footnotesize]{c}
 static void kalman( const int n,               const int k,                const double *sigma, /* n,n */
@@
 
-    \end{minipage}
 \end{figure}
 \end{landscape}
+
+\subsection{Memory Usage}
+
+Inspecting the trace for LT4LA shows 6 calls to `empty' (function for
+allocating new matrices): 4 temporaries plus 2 for storing the results. From
+the CBLAS implementation (Figure~\ref{fig:cblas_kalman}), we can see that it
+too has 6 calls to malloc: 4 temporaries plus 2 for storing the results.
 
@@
 \begin{figure}[tp]
-    \begin{minted}[linenos, fontsize=\small]{ocaml}
+    \centering
+    \begin{minted}[fontsize=\small]{ocaml}
 let potrs ~uplo a b =

### 2018-10-11 write-up/dissertation/introduction.tex
Organise write-ups

--- 2018-05-22 write-up/introduction.tex
@@
 
-\prechapter{%
-}%
+Linear types allow the compiler and programmer to statically keep track of the
+resources that a program uses, thus offering a promising solution to the
+problems associated with complex resource management. However, they have not
+made their way into many mainstream programming languages, in the same way
+parametrically polymorphic types have. To illustrate their simplicity and
+power, I implemented an OCaml library that allows users to learn about and
+become familiar with linear types, specifically in the context of linear
+algebra programs.
 
-\section{Overview of Problem}
+The main contributions of this thesis are:
 
-\section{Contributions}
+\begin{itemize}
+
+    \item An \textbf{original, usable implementation} of a type system that can express
+        aliasing, read/write permissions, memory allocation, re-use and deallocation.
+
+    \item An \textbf{original} demonstration of how that type system can be
+        \textbf{applied to the APIs} of linear algebra libraries and the
+        \textbf{benefits} of doing so.
+
+    \item Many \textbf{new} examples of how that type system can \textbf{automatically
+        check} for common aliasing, read/write permission, memory allocation,
+        re-use and deallocation \textbf{errors} in the context of
+        linear algebra programs.
+
+    \item An \textbf{original} demonstration of how that type system can be
+        \textbf{used} for \textbf{matrix expression compilation}.
+
+    \item \textbf{New and readable implementations} of \textbf{non-trivial}
+        linear algebra programs that \textbf{take advantage} of said type
+        system.
+
+    \item A \textbf{new solution} to the \textbf{dichotomy} of
+        \textbf{readability}, \textbf{ease of reasoning and safety} of
+        high-level linear algebra libraries versus the
+        \textbf{memory-efficiency} of low-level linear algebra libraries.
+
+    \item A \textbf{new library design} to provide said solution in a way that
+        \textbf{integrates well} with existing OCaml code and linear algebra
+        libraries.
+
+\end{itemize}

### 2018-10-11 write-up/dissertation/primitives.tex
Organise write-ups

\chapter{Primitives}\label{chap:primitives}

The following signature gives an indication of how I embedded LT4LA's type
system into OCaml's and typed its primitives accordingly. This helped catch
bugs and increase confidence in the correctness of the code produced.

\begin{minted}[fontsize=\small]{ocaml}
module Arr = Owl.Dense.Ndarray.D
type z = Z
type 'a s = Succ
type 'a arr = A of Arr.arr [@@unboxed]
type 'a mat = M of Arr.arr [@@unboxed]
type 'a bang = Many of 'a [@@unboxed]
module Prim :
sig
  val extract : 'a bang -> 'a
  (** Boolean *)
  val not_ : bool bang -> bool bang
  (** Arithmetic, many omitted for brevity *)
  val addI : int bang -> int bang -> int bang
  val ltE : float bang -> float bang -> bool bang
  (** Arrays *)
  val set : z arr -> int bang -> float bang -> z arr
  val get : 'a arr -> int bang -> 'a arr * float bang
  val share : 'a arr -> 'a s arr * 'a s arr
  val unshare : 'a s arr -> 'a s arr -> 'a arr
  val free : z arr -> unit
  (** Owl *)
  val array : int bang -> z arr
  val copy : 'a arr -> 'a arr * z arr
  val sin : z arr -> z arr
  val hypot : z arr -> 'a arr -> 'a arr * z arr
  (** Level 1 BLAS *)
  val asum : 'a arr -> 'a arr * float bang
  val axpy : float bang -> 'a arr -> z arr -> 'a arr * z arr
  val dot : 'a arr -> 'b arr -> ('a arr * 'b arr) * float bang
  val rotmg : float bang * float bang -> float bang * float bang ->
              (float bang * float bang) * (float bang * z arr)
  val scal : float bang -> z arr -> z arr
  val amax : 'a arr -> 'a arr * int bang
  (* Matrix, some omitted for brevity *)
  val matrix : int bang -> int bang -> z mat
  val copy_mat : 'a mat -> 'a mat * z mat
  val copy_mat_to : 'a mat -> z mat -> 'a mat * z mat
  val size_mat : 'a mat -> 'a mat * (int bang * int bang)
  val transpose : 'a mat -> 'a mat * z mat
  (* Level 3 BLAS/LAPACK *)
  val gemm : float bang -> ('a mat * bool bang) -> ('b mat * bool bang) ->
             float bang -> z mat -> ('a mat * 'b mat) * z mat
  val symm : bool bang -> float bang -> 'a mat -> 'b mat ->
             float bang -> z mat -> ('a mat * 'b mat) * z mat
  val posv : z mat -> z mat -> z mat * z mat
  val potrs : 'a mat -> z mat -> 'a mat * z mat
end
\end{minted}

### 2018-10-11 write-up/dissertation/related_work.tex
Organise write-ups

--- 2018-05-27 write-up/related_work.tex
@@
 These algorithms assume basic building blocks such as products and
-factorisations. Cl1ck attempts to take on the challenge of writing BLAS/LAPACK
-like libraries too, by generating lower-level loop-based blocked routines for
-the aforementioned basic building blocks, in the spirit of the
+factorisations. Cl1ck attempts to take on the challenge of writing
+BLAS/LAPACK-like libraries too, by generating low-level loop-based blocked
+routines for the aforementioned basic building blocks, in the spirit of the
 FLAME~\cite{gunnels_flame} project.
@@
 In LT4LA, I took the approach of having a concrete syntax and expression
-language which was then translated and made available as a \emph{typed
-expression} to other modules \emph{at compile time} via the build system. Apart
-from convenience in programming and testing, there was nothing inherent in the
+language which was then translated to OCaml and made available to other modules
+\emph{at compile time} via the build system. There was nothing inherent in the
 approach that prevented me from using OCaml's PPX syntax-extensions so that I
@@
 
-They perform rudimentary pattern-matching and in some cases, loop-fusion, to
-avoid evaluating expressions in a purely binary manner (invoking the bane of a
+Libraries that use expression templates usually perform rudimentary pattern-matching and in some cases, loop-fusion, to
+avoid evaluating expressions in a purely binary manner (preventing the bane of a
 C++ programmer: temporaries and unnecessary copies) when possible (either by
@@
 
-As is the case with LT4LA, this approach shares \emph{some} elimination of
-runtime overheads, but not all, thanks to the heuristics surrounding lazy
-evaluation and evaluating sub-expressions into temporaries. This comes at the
-cost of a user being able to easily inspect the generated C++ code, losing
-explicitness.
+Therefore, expression templates eliminate some runtime overheads, but not all
+(unlike LT4LA), because the heuristics they use rely on statically imperfect
+information. They also do not allow a user to easily inspect the generated code
+(also unlike LT4LA), thus losing explicitness.
 
@@
 management by using a library, the impetus for integrating such features into
-the language follows.
+a language follows.
 
@@
 evaluation is, funnily enough, linear types; more precisely, it is the static
-and perfect information they guarantee about aliasing: with Owl, every
-graph-node has only one node in or out; with Eigen, every ``assignment'' has a
-\texttt{noalias} annotation, without the danger of getting it wrong and
-invoking undefined behaviour.
+and perfect information they guarantee about aliasing: with Owl, this would be
+comparable to assuming every vertex of its computation graph has only one edge
+in and out; with Eigen, this would be comparable to having a \texttt{noalias}
+annotation on every ``assignment'', \textbf{without the danger of getting it
+wrong} and invoking undefined behaviour.
 
@@
 usage needs to remember. Of course, now the programmer has to figure out how
-they are using their temporaries, but because matrices are linearly-typed,
-redundant copies/missed frees can be pointed out by the compiler, guiding them
-towards a satisfying solution.
+they are using their temporaries, but because matrices are linearly typed,
+redundant copies and missed frees can be pointed out by the compiler, thus
+guiding the programmer towards a satisfying solution.
 
@@
 polymorphism, size types and uniqueness types (an idea closely related to
-linear types). Its combinators are more expressive than typical linear-algebra
-library kernels so encourage shorter, more declarative linear-algebra code.
+linear types). Its combinators are more expressive than typical linear algebra
+library kernels; hence it encourages shorter, more declarative code.
 
@@
 bring the last two decades of programming language research to the masses in a
-usable and friendly manner. Its \emph{borrow-checker} is the feature most
-relevant to this thesis because it statically attempts to prevent many
-resource-related bugs. Although there are a few linear-algebra libraries for
-Rust under development, careful use of its macro system and borrow-checker
-could make it the safest and easiest to use language for linear-algebra
+usable and friendly manner. Its \emph{borrow checker} is the feature most
+relevant to this thesis because it also statically attempts to prevent many
+resource-related bugs. There are a few linear algebra libraries for
+Rust under development; careful use of its macro system and borrow checker
+could make it the safest and easiest to use language for linear algebra
 projects to come. Its struggle is more likely to be against the inertia of the
-large amounts of C++/Fortran code already out there rather than its usability
-or benefits. Given this inertia, I believe there is value in taking the DSL
-library approach of LT4LA which will work with existing systems.
+large amounts of C, C++ and Fortran code already out there rather than its
+usability or benefits. Given this inertia, I believe there is value in taking
+the DSL library approach of LT4LA, which will work with existing systems.
 
@@
 Dimension mismatches are seen as an irritating but small inconvenience when
-writing linear-algebra code. However, another error I found in the Fortran code
+writing linear algebra code. However, another error I found in the Fortran code
 output by SymPy to compute a Kalman filter was a dimension/transposition error.

### 2018-10-16 write-up/samplepaper/intro.tex
Sketch out structure of paper

\section{Introduction}

\subsection{Contributions}
\begin{itemize}
    \item We have designed the \LANG\ programming language
    \item We illustrate that the design is sensible with many matrix-y examples
    \item We give a soundness proof for \LANG, using a step-indexed logical relation
    \item Incredibly simple type inference algorihtm for polymorphic fractional permissions
        \begin{itemize}
            \item Compare to Bierhof et al's \emph{Fraction Polymorphic Permission Inference},
                which uses a fancy dataflow analysis
            \item We use exactly the same unification algorithm type polymorphism does
        \end{itemize}
    \item We have an implementation - compatible with and usable from existing code!
\end{itemize}

### 2018-10-16 write-up/samplepaper/lang_and_examples.tex
Sketch out structure of paper

\section{Language Overview and Examples}

\subsection{Short language description}

This should be a high-level description of the language. We should
aim to spend about a page on this section, highlighting the key
properties of \LANG{}:
\begin{itemize}
\item Linearity, obviously
\item Fractional permissions: sharing and unsharing
\item Dynamic errors: matrix dimensions and unsharing (matrix identity checks)
  \begin{itemize}
  \item In the related work, we should point to L3 statically tracking
    pointer identities, but we want to keep the number of type indices
    under control
  \end{itemize}
\end{itemize}

\subsection{Examples}

The more examples, the better. In fact, it's actually impossible to
have too many examples -- it's okay (indeed, desirable) to spend 5-6
pages on examples.

\begin{itemize}

    \item Simple: factorial, shows recursion, !-annotations etc.

    \item Less simple: summing over an array, indexing and safety

    \item Medium: one-dimensioal convolution, permissions

    \item Harder: linear regression, pattern-matching and apparent non-linearity

    \item Harder: L1 norm-minimisation, some frees, re-using memory

    \item Big finish: Kalman filter

\end{itemize}

### 2018-10-17 write-up/samplepaper/lang_and_examples.tex
Add language overview

@@
-\section{Language Overview and Examples}
+\section{\lang\ Overview and Examples}\label{sec:lang_and_examples}
 
-\subsection{Short language description}
+\subsection{Overview}
 
-This should be a high-level description of the language. We should
-aim to spend about a page on this section, highlighting the key
-properties of \LANG{}:
-\begin{itemize}
-\item Linearity, obviously
-\item Fractional permissions: sharing and unsharing
-\item Dynamic errors: matrix dimensions and unsharing (matrix identity checks)
-  \begin{itemize}
-  \item In the related work, we should point to L3 statically tracking
-    pointer identities, but we want to keep the number of type indices
-    under control
-  \end{itemize}
-\end{itemize}
+Linearity is at the heart of \lang. Linearity allows us to express a
+pure-functional API for numerical library routines that mutate arrays and
+matrices. Linearity also restricts aliasing of (values which represent)
+pointers.
+
+However, linearity by itself is not sufficient to produce an expressive enough
+programming language. For values such as booleans, integers, floating-point
+numbers as well as pure functions, we need to be able to use them
+\emph{intuitionistically}, that is, more than once or not at all. For this
+reason, we have the !-constructor at the type level and its corresponding
+\texttt{Many}-constructor and \texttt{let Many <id> = .. in ..}-eliminator at
+the term level. Because we want to restrict how a programmer can alias pointers
+and prevent a programmer from ignoring them (a memory leak), \lang\ enforces
+simple syntactic restrictions on which values can be wrapped up in a
+\texttt{Many} constructor (details in Section \ref{sec:formal_system}).
+
+There are also valid cases in which we would want to alias pointers to a
+matrix. The most common is exemplified by the BLAS routine \texttt{gemm}, which
+(rather tersely) stands for \emph{GEneric Matrix Multiplication}.  A
+\emph{simplified} definition of \texttt{gemm($\alpha$, A, B, $\beta$, C)} is $C
+:= \alpha AB + \beta C$. In this case, \texttt{A} and \texttt{B} may alias each
+other but neither may alias \texttt{C}, because it is being written to.
+Related to \emph{mutating} arrays and matrices is \emph{freeing} them. Here, we
+would also wish to restrict aliasing so that we do not free one alias and then
+attempt to use another. Although linearity on its own suffices to prevent
+use-after-free errors when values are \emph{not} aliased (a freed value is
+\emph{out of scope} for the rest of the expression), we still need another
+simple, yet powerful concept to provide us with the extra expressivity of
+aliasing \emph{without} losing any of the benefits of linearity.
+
+Fractional permissions provide exactly this. Concretely, types of (pointers to)
+arrays and matrices are \emph{parameterised} by a \emph{fraction}. A fraction
+is either $1$ ($2^0$) or exactly \emph{half} of another fraction ($2^{-k}$, for
+natural $k$). The former represents complete ownership of that value: the
+programmer may mutate or free that value as they choose; the latter represents
+read-only access or a \emph{borrow}: the programmer may read from the value but
+not write to or free it. Creating an array/matrix gives you ownership of it, so
+too does having one (with a fractional permission of $2^0$) passed in as an
+argument.
+
+In \lang, we can produce two aliases of a single array/matrix, by
+\emph{sharing} it. If the original alias had a fractional permission of
+$2^{-k}$ then the two new aliases of it will have a fractional permission of
+$2^{-(k+1)}$ each. Thanks to linearity, the original array/matrix with a
+fractional permission of $2^{-k}$ will be out of scope after the sharing.  When
+an array/matrix is shared as such, we can prevent the programmer from freeing
+or mutating it by making the types of \texttt{free} and \texttt{set} (for
+mutation) require a \emph{whole} ($2^0$) permission.
+
+If we have two aliases \emph{to the same matrix} with \emph{identical}
+fractional permissions ($2^{-(k+1)}$), we can recombine or \emph{unshare} them
+back into a single one, with a larger $2^{-k}$ permission. As before, thanks to
+linearity, the original two aliases will be out of scope after unsharing.
+
+Matrix unsharing is one of only \emph{two} operations that can fail at runtime
+(the other being dimension checks, such as for \texttt{gemm}). The check being
+performed is a simple sanity check that the two aliasing pointers passed to
+\texttt{unshare} point to the same array/matrix. Section
+\ref{sec:discussion_related_work} contains an overview of how we could remove
+the need for this by tracking pointer identities statically by augmenting the
+type system further.
+
+The final feature of \lang\ which makes it sufficiently expressive is recursion
+(and of course, conditional branches to ensure termination). Conditional
+branches are implemented by ensuring that both branches use the same set of
+linear values. A function can be recursive if it captures no linear values from
+its environment. Like with \texttt{Many}, this is enforced via simple syntactic
+restrictions on the definition of recursive functions.
 

### 2018-10-17 write-up/samplepaper/lang_and_examples.tex
Add factorial example

@@
 
+\subsubsection{Intuitionism: ! and \highl{Many}}
+
 However, linearity by itself is not sufficient to produce an expressive enough
@@
 reason, we have the !-constructor at the type level and its corresponding
-\texttt{Many}-constructor and \texttt{let Many <id> = .. in ..}-eliminator at
+\highl{Many}-constructor and \highl{let Many <id> = .. in ..}-eliminator at
 the term level. Because we want to restrict how a programmer can alias pointers
@@
 simple syntactic restrictions on which values can be wrapped up in a
-\texttt{Many} constructor (details in Section \ref{sec:formal_system}).
+\highl{Many} constructor (details in Section~\ref{sec:formal_system}).
+
+\subsubsection{Fractional Permissions}
 
@@
 an array/matrix is shared as such, we can prevent the programmer from freeing
-or mutating it by making the types of \texttt{free} and \texttt{set} (for
+or mutating it by making the types of \highl{free} and \highl{set} (for
 mutation) require a \emph{whole} ($2^0$) permission.
@@
 
+\subsubsection{Runtime Errors}
+
 Matrix unsharing is one of only \emph{two} operations that can fail at runtime
-(the other being dimension checks, such as for \texttt{gemm}). The check being
+(the other being dimension checks, such as for \highl{gemm}). The check being
 performed is a simple sanity check that the two aliasing pointers passed to
-\texttt{unshare} point to the same array/matrix. Section
-\ref{sec:discussion_related_work} contains an overview of how we could remove
-the need for this by tracking pointer identities statically by augmenting the
-type system further.
+\highl{unshare} point to the same array/matrix.
+Section~\ref{sec:discussion_related_work} contains an overview of how we could
+remove the need for this by tracking pointer identities statically by
+augmenting the type system further.
+
+\subsubsection{Recursion}
 
@@
 linear values. A function can be recursive if it captures no linear values from
-its environment. Like with \texttt{Many}, this is enforced via simple syntactic
+its environment. Like with \highl{Many}, this is enforced via simple syntactic
 restrictions on the definition of recursive functions.
@@
 \end{itemize}
+
+\subsubsection{Introductory: Factorial}
+
+Although a factorial function (Figure~\ref{fig:lang_factorial}) may seem like
+an aggresively pedestrian first example, in a linearly typed language such as
+\lang\ it represents the culmination of a rather number of features.
+
+To simplifiy the design and implementation of \lang's type system, recursive
+functions must have full type annotations (non-recursive functions need only
+their argument types annotated). Its body is a closed expression (with respect
+to the function's arguments), so it type-checks (since it does not capture any
+linear values from its environment).
+
+The only argument is \highl{!x : !int}. The !-annotation on \highl{x} is a
+syntactic convenience for declaring the value to used intuitionistically, its
+full and precise meaning is described in Section~\ref{sec:implementation}.
+
+The condition for an \highl{if} may or may not use linear values (here, with
+\highl{x < 0} \highl{ || x = 0}, it does not). Any linear values used by the conditional
+would not be in scope in either branch of the \highl{if}-expression.  Both
+branches use \highl{x} differently: one ignores it completely and the other
+uses it twice.
+
+All numeric and boolean literals are implicitly wrapped in a \highl{Many} and
+all primitives involving them return a \highl{!int}, \highl{!bool} or
+\highl{!elt} (types of elements of arrays/matrices, typically 64-bit
+floating-point numbers). The short-circuting \highl{||} behaves in exactly the
+same way as a boolean-valued \highl{if}-expression.
+
+\begin{figure}[tp]
+    \centering
+    \inputminted[fontsize=\small, linenos]{ocaml}{../../test/examples/factorial.lt}
+    \caption{Factorial function in \lang.}\label{fig:lang_factorial}
+\end{figure}
+
+\subsubsection{Introductory: Summing over an Array}
+
+\begin{figure}[tp]
+    \centering
+    \inputminted[fontsize=\small]{ocaml}{../../test/examples/sum_array.lt}
+    \caption{Summing over an array in \lang.}\label{fig:lang_sumarray}
+\end{figure}

### 2018-10-20 write-up/paper/discussion.tex
Rename files

\section{Discussion and Related Work}\label{sec:discussion_related_work}

The main point we want to make is that using linear types for BLAS
is an ``obvious'' idea, but is surprisingly under-explored.

\begin{itemize}
\item Rust
\item ATS
\item Single-assignment C 
\item Linear Haskell
\item Bernardy and Sveningsson
\item L3
\item Boyland fractional permissions
\end{itemize}

### 2018-10-20 write-up/paper/formal_system.tex
Rename files

\section{Formal System}\label{sec:formal_system}

% TODO value restriction

\subsection{The Core Type Theory and Dynamic Semantics}

Describe the typing rules and operational semantics here

\subsection{The logical relation}

Describe the step-indexed logical relation and its main properties

\subsection{Soundness Theorem}

State the fundamental lemma, and sketch the proof a little

### 2018-10-20 write-up/paper/implementation.tex
Rename files

\section{Implementation}\label{sec:implementation}
\subsection{Implementation Strategy}

% TODO desugaring

Talk about how you implemented \lang\ and the general
architecture. Talk about how simple everything is, and also
about how implementing inference for fractions is.

\subsection{Performance Metrics}

Here, evaluate the performance of the examples from the second
section.  Compare with your C implementations, and perhaps as well as
the straightforward math transcribed into (Matlab/R/Numpy?).

### 2018-10-20 write-up/paper/intro.tex
Rename files

--- 2018-10-16 write-up/samplepaper/intro.tex
@@
 \section{Introduction}
+
+\lang\ is a functional programming language designed to express the APIs of
+low-level linear algebra libraries (such as BLAS/LAPACK) safely and explicitly.
+It does so by combining linear types, fractional permissions, runtime errors
+and recursion into a small, easily understandable, yet expressive set of core
+constructs. In addition to this, \lang's implementation supports several
+syntactic conveniences as well as a usable integration with real-world OCaml
+code.
 
@@
 \begin{itemize}
-    \item We have designed the \LANG\ programming language
+    \item We have designed the \lang\ programming language
     \item We illustrate that the design is sensible with many matrix-y examples
-    \item We give a soundness proof for \LANG, using a step-indexed logical relation
+    \item We give a soundness proof for \lang, using a step-indexed logical relation
     \item Incredibly simple type inference algorihtm for polymorphic fractional permissions

### 2018-10-20 write-up/paper/lang_and_examples.tex
Add array sum example

--- 2018-10-17 write-up/samplepaper/lang_and_examples.tex
@@
 \emph{intuitionistically}, that is, more than once or not at all. For this
-reason, we have the !-constructor at the type level and its corresponding
-\highl{Many}-constructor and \highl{let Many <id> = .. in ..}-eliminator at
+reason, we have the ! constructor at the type level and its corresponding
+\highl{Many} constructor and \highl{let Many <id> = .. in ..} eliminator at
 the term level. Because we want to restrict how a programmer can alias pointers
@@
 
-Matrix unsharing is one of only \emph{two} operations that can fail at runtime
-(the other being dimension checks, such as for \highl{gemm}). The check being
-performed is a simple sanity check that the two aliasing pointers passed to
-\highl{unshare} point to the same array/matrix.
+Aside from out-of-bounds indexing, matrix unsharing is one of only \emph{two}
+operations that can fail at runtime (the other being dimension checks, such as
+for \highl{gemm}). The check being performed is a simple sanity check that the
+two aliasing pointers passed to \highl{unshare} point to the same array/matrix.
 Section~\ref{sec:discussion_related_work} contains an overview of how we could
@@
 an aggresively pedestrian first example, in a linearly typed language such as
-\lang\ it represents the culmination of a rather number of features.
+\lang\ it represents the culmination of many features.
 
@@
 
+Now we can add fractional permissions to the mix:
+Figure~\ref{fig:lang_sumarray} shows a simple, tail-recursive implementation of
+summing all the elements in an array. There are many new features; first among
+them is \highl{!x0 : !elt}, the type of array/matrix elements (64-bit floating
+point).
+
 \begin{figure}[tp]
@@
 \end{figure}
+
+Second is \highl{('x) (row : 'x arr)} which is an array with a
+universally-quantified fractional permission. In particular, this means the
+body of the function cannot mutate or free the input array, only read from it.
+If the programmer did try to mutate or free \highl{row}, then they would get a
+helpful error message (Figure~\ref{fig:lang_errormsg}).
+
+\begin{figure}[tp]
+    \centering
+    \begin{minted}[fontsize=\small]{ocaml}
+let row = row[i] := x1 in (* or *) let () = free row in
+(* Could not show equality:                                        *)
+(*     z arr                                                       *)
+(* with                                                            *)
+(*     'x arr                                                      *)
+(*                                                                 *)
+(* Var 'x is universally quantified                                *)
+(* Are you trying to write to/free/unshare an array you don't own? *)
+(* In test/examples/sum_array.lt, at line: 7 and column: 19        *)
+    \end{minted}
+    \caption{Attempting to write to or free a read only array in
+    \lang.}\label{fig:lang_errormsg}
+\end{figure}
+
+Alongside taking a \highl{row : 'x arr}, the function also returns an array
+with exactly the same fractional permission as the \highl{row} (which can only
+be \highl{row}).  This is necessary because of linearity: for the caller, the
+original array passed in as an argument would be out of scope for the rest of
+the expression, so it needs to be returned and then rebound to be used for the
+rest of the function.
+
+An example of this consuming and rebinding is in \highl{let (row, !x1) =
+row[i]}. Indexing is implemented as a primitive \texttt{get: 'x. 'x arr --o
+!int --o 'x arr * !elt}. Although fractional permissions can be passed around
+explicitly  (as done in the recursive call), they can also be
+\emph{automatically inferred at call sites}, \texttt{row[i] == get \_ row i}
+takes advantage of this convenience.

### 2018-10-21 write-up/paper/lang_and_examples.tex
Add all examples except matrix squaring

@@
 There are also valid cases in which we would want to alias pointers to a
-matrix. The most common is exemplified by the BLAS routine \texttt{gemm}, which
+matrix. The most common is exemplified by the BLAS routine \highl{gemm}, which
 (rather tersely) stands for \emph{GEneric Matrix Multiplication}.  A
@@
 an array/matrix is shared as such, we can prevent the programmer from freeing
-or mutating it by making the types of \highl{free} and \highl{set} (for
+or mutating it by making the types of \texttt{free} and \texttt{set} (for
 mutation) require a \emph{whole} ($2^0$) permission.
@@
 operations that can fail at runtime (the other being dimension checks, such as
-for \highl{gemm}). The check being performed is a simple sanity check that the
-two aliasing pointers passed to \highl{unshare} point to the same array/matrix.
+for \texttt{gemm}). The check being performed is a simple sanity check that the
+two aliasing pointers passed to \texttt{unshare} point to the same array/matrix.
 Section~\ref{sec:discussion_related_work} contains an overview of how we could
@@
 pages on examples.
-
 \begin{itemize}
 
-    \item Simple: factorial, shows recursion, !-annotations etc.
+    \item Simple: factorial, shows recursion, ! annotations etc.
 
@@
 
-\subsubsection{Introductory: Factorial}
+\subsubsection{Factorial}
 
@@
 
-The only argument is \highl{!x : !int}. The !-annotation on \highl{x} is a
+The only argument is \highl{!x : !int}. The ! annotation on \texttt{x} is a
 syntactic convenience for declaring the value to used intuitionistically, its
@@
 The condition for an \highl{if} may or may not use linear values (here, with
-\highl{x < 0} \highl{ || x = 0}, it does not). Any linear values used by the conditional
+\highl{x < 0 || x = 0}, it does not). Any linear values used by the conditional
 would not be in scope in either branch of the \highl{if}-expression.  Both
-branches use \highl{x} differently: one ignores it completely and the other
+branches use \texttt{x} differently: one ignores it completely and the other
 uses it twice.
@@
 \highl{!elt} (types of elements of arrays/matrices, typically 64-bit
-floating-point numbers). The short-circuting \highl{||} behaves in exactly the
+floating-point numbers). The short-circuting \texttt{||} behaves in exactly the
 same way as a boolean-valued \highl{if}-expression.
 
-\begin{figure}[tp]
+\begin{figure}[t]
     \centering
@@
 
-\subsubsection{Introductory: Summing over an Array}
+\subsubsection{Summing over an Array}
 
@@
 
-\begin{figure}[tp]
+\begin{figure}[t]
     \centering
@@
 body of the function cannot mutate or free the input array, only read from it.
-If the programmer did try to mutate or free \highl{row}, then they would get a
+If the programmer did try to mutate or free \texttt{row}, then they would get a
 helpful error message (Figure~\ref{fig:lang_errormsg}).
 
-\begin{figure}[tp]
+\begin{figure}[t]
     \centering
@@
 Alongside taking a \highl{row : 'x arr}, the function also returns an array
-with exactly the same fractional permission as the \highl{row} (which can only
-be \highl{row}).  This is necessary because of linearity: for the caller, the
+with exactly the same fractional permission as the \texttt{row} (which can only
+be \texttt{row}).  This is necessary because of linearity: for the caller, the
 original array passed in as an argument would be out of scope for the rest of
@@
 
-An example of this consuming and rebinding is in \highl{let (row, !x1) =
-row[i]}. Indexing is implemented as a primitive \texttt{get: 'x. 'x arr --o
+An example of this consuming and re-binding is in \highl{let (row, !x1) =
+row[i]}. Indexing is implemented as a primitive \highl{get: 'x. 'x arr --o
 !int --o 'x arr * !elt}. Although fractional permissions can be passed around
 explicitly  (as done in the recursive call), they can also be
-\emph{automatically inferred at call sites}, \texttt{row[i] == get \_ row i}
+\emph{automatically inferred at call sites}, \highl{row[i] == get _ row i}
 takes advantage of this convenience.
+
+\subsubsection{One-dimensional Convolution}
+
+\begin{figure}[t]
+    \centering
+    \inputminted[fontsize=\small]{ocaml}{../../test/examples/weighted_avg_infer.lt}
+    \caption{\emph{Simplified} one-dimensional convolution.}\label{fig:lang_oned_conv}
+\end{figure}
+
+Figure~\ref{fig:lang_oned_conv} extends the set of features demonstrated by the
+previous examples by mutating one of the input arrays. A one-dimensional
+convolution involves two arrays: a read-only kernel (array of weights) and an
+input vector. It modifies the input vector \emph{in-place} by replacing each
+\highl{write[i]} with a weighted (as per the values in the kernel) sum of it
+and its neighbours; intuitively, sliding a dot-product with the kernel across
+the vector.
+
+What's implemented in Figure~\ref{fig:lang_oned_conv} is a \emph{simplified}
+version of this idea, so as to not distract from the features of \lang. The
+simplifications are:
+\begin{itemize}
+    \item the kernel has a fixed length (of 3), as such, only the value of
+        \highl{write[i-1]} prior to modification in the previous iteration
+        needs to be carried forward using \highl{x0}
+    \item \highl{write} is assumed to have length \highl{n+1}
+    \item \texttt{i}'s initial value is assumed to be \texttt{1}
+    \item \highl{x0}'s initial value is assumed to be \highl{write[0]}
+    \item the first and last values of \texttt{write} are ignored.
+\end{itemize}
+
+Mutating an array is implemented similarly to indexing one: a primitive
+\highl{set: z arr --o !int --o !elt --o z arr}.  It consumes the original
+array and returns a new array with the updated value.  \highl{let written =
+write[i] := <exp> } is just syntactic sugar for \highl{let written = set write i <exp>}.
+
+Since \highl{write: z arr} (where \texttt{z} stands for $k=0$, representing a
+fractional permission of $2^{-k} = 2^{-0} = 1$), we may mutate it, but since we
+only need to read from \texttt{weights}, its fractional permission index can be
+universally-quantified. In the recursive call, we see \texttt{\_} being used
+explicitly to tell the compiler to \emph{infer} the correct fractional
+permission based on the given arguments.
+
+\subsubsection{Linear Regression}
+
+\begin{figure}[t]
+    \centering
+    \inputminted[fontsize=\small]{ocaml}{../../test/examples/lin_reg.lt}
+    \caption{Linear regression (OLS): $\hat\beta =
+        \mathbf{(X^T X)^{-1} X^T y}$}\label{fig:lang_lin_reg}
+\end{figure}
+
+\emph{The most pertinent aspect of \lang's type system are the types of its
+primitives.} While the types of operations such as \texttt{get} and
+\texttt{set} might be borderline obvious, they types of BLAS/LAPACK routines
+become an \emph{incredibly useful, automated check for using the API correctly.}
+
+Figure~\ref{fig:lang_lin_reg} shows a \emph{non-recursive} function declaration (the
+return type is inferred). Since we would like to be able to use a function like
+\texttt{lin\_reg} more than once, it is marked with a \texttt{!} annotation
+(which also ensure it captures no linear values from the surrounding
+environment).
+
+We wish to compute $\hat\beta = \mathbf{(X^T X)^{-1} X^T y}$. First, we extract
+the dimensions of matrix \texttt{x}. Then, we say we would like a new matrix, of
+dimension $m \times 1$, which contains the result of $\mathbf{X^T y}$. The line
+\highl{let xy <- new (m,1) [| x^T * y |]} is syntactic sugar for first creating
+a new matrix \highl{let xy = matrix m 1 in} and then storing the result of
+\highl{(x, true (* transp'd *))} multiplied with \highl{(y, false (* not
+transp'd *))} into \texttt{xy}: \highl{let (xy, (x, y)) = gemm 1.  _ (x,
+true) _ (y, false) 0. xy}, appropriately re-binding the identifiers.  The line
+\highl{let x_T_x <- new (m,m) [| x^T * x |]}, works similarly, but uses a more
+appropriate, \highl{syrk true 1. x 0. x_T_x} BLAS call.
+
+By using some simple pattern-matching and syntactic sugar, we can:
+\begin{itemize}
+    \item write normal-looking, apparently non-linear code
+    \item use matrix expressions directly and have an appropriate, efficient
+        call to a BLAS/LAPACK routine inserted
+    \item retain the safety of linear types with fractional permissions by
+        having the compiler statically enforce the aliasing and read/write rules
+        implicitly assumed by BLAS/LAPACK routines.
+\end{itemize}
+
+After computing \texttt{x\_T\_x}, we need to invert it and then multiply it by
+\texttt{xy}. The BLAS routine \highl{posv: z mat --o z mat --o z mat * z mat}
+does exactly that: assuming the first argument is symmetric, \texttt{posv}
+mutates its second argument to contain the desired value. Its first argument is
+also mutated to contain the (upper triangular) Cholesky decomposition factor of
+the original matrix. Since we do not need that matrix (or its memory) again, we
+\texttt{free} it. If we forgot to, we would get a \texttt{Variable to\_del not
+used} error. Lastly, we return the \texttt{answer} alongside the untouched
+input matrices \texttt{(x,y)}.
+
+\subsubsection{L1-Norm Minimisation on Manifolds}
+
+\begin{figure}[t]
+    \centering
+    \inputminted[fontsize=\small]{ocaml}{../../test/examples/l1_norm_min.lt}
+    \caption{L1-norm minimisation on manifolds:
+        $\mathbf{Q^{-1}U(I+U^TQ^{-1}U)^{-1}U^T}$}\label{fig:lang_l1_norm_min}
+\end{figure}
+
+Figure~\ref{fig:lang_l1_norm_min} shows even more pattern-matching. Patterns of
+the form \highl{let <id> <- [| beta * c + alpha * a * b |]} are desugared to
+\texttt{gemm} calls. Primitives like \highl{transpose: 'x. 'x mat --o 'x mat *
+z mat} and \highl{eye: !int --o z mat} allocate new matrices;
+\texttt{transpose} returns the transpose of a given matrix and \texttt{eye}
+returns a $k \times k$ identity matrix.
+
+We also see our first example of re-using memory for different matrices: like
+with \texttt{to\_del} and \texttt{posv} in the previous example, we do not need
+the value stored in \texttt{tmp\_5\_5} after the call to \texttt{gesv}. However,
+we can use the memory much later to store \texttt{answer} with \highl{let
+answer <- [| 0. * tmp_5_5 + q_inv_u * inv_u_T |]}. Again, thanks to linearity,
+the identifiers \texttt{q}, \texttt{tmp\_5\_5} are out of scope by the time
+\texttt{answer} is bound. Although during execution, all three refer to the
+same piece of memory, logically they represent different values throughout the
+computation.
+
+\subsubsection{Kalman Filter}
+
+\begin{figure}[t]
+    \centering
+    \inputminted[fontsize=\small]{ocaml}{../../test/examples/kalman.lt}
+    \caption{Kalman filter: see Figure~\ref{fig:maths_kalman} for the
+        equations this code implements.}\label{fig:lang_kalman}
+\end{figure}
+
+\begin{figure}[t]
+    \begin{align*}
+        \mu' &= \mu + \Sigma H^T (R + H \Sigma H^T)^{-1} (H \mu - \textrm{data})\\
+        \Sigma' &= \Sigma ( I - H^T (R + H \Sigma H^T)^{-1} H \Sigma )
+    \end{align*}
+    \caption{Kalman filter equations (credit:
+    \href{http://matthewrocklin.com/blog/work/2012/11/24/Kalman-Filter}{matthewrocklin.com}).}\label{fig:maths_kalman}
+\end{figure}
+
+Figure~\ref{fig:lang_kalman} shows a \lang\ implementation of a Kalman filter
+(equations in Figure~\ref{fig:maths_kalman}). A few new features and techniques
+are used in this implementation:
+\begin{itemize}
+
+    \item \texttt{sym} annotations in matrix expressions: when this is used, a 
+        call to \texttt{symm} (the equivalent of \texttt{gemm} but for
+        symmetric matrices so that only half the operations are performed) is
+        inserted (instead of one to \texttt{gemm})
+
+    \item \texttt{copyM\_to} is used to re-use memory by \emph{overwriting} the
+        contents of its second argument to that of its first (erroring if
+        dimensions do not match)
+
+    \item \highl{let new_r <- new [| r_2 |]} creates a copy of \texttt{r\_2}
+
+    \item \texttt{potrs \_ chol\_r data\_2} uses a pre-computed Cholesky decomposition
+        to multiply \texttt{data\_2} by \texttt{r\_2} $= (R + H \Sigma H^T)^{-1}$
+
+    \item a lot of memory re-use; the following sets of identifiers alias each other:
+        \begin{itemize}
+            \item \texttt{r\_1} and \texttt{r\_2}
+            \item \texttt{data\_1}, \texttt{data\_2} and \texttt{sol\_data}
+            \item \texttt{new\_h} and \texttt{sol\_h}
+            \item \texttt{h\_sol\_h}, \texttt{sigma\_copy} and \texttt{new\_sigma}
+            \item \texttt{mu\_copy} and \texttt{new\_mu}
+        \end{itemize}
+
+\end{itemize}

### 2018-10-21 write-up/paper/paper.tex
Add all examples except matrix squaring

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
\usepackage{hyperref}
\hypersetup{%
    colorlinks=true,
    linkcolor={black},
    citecolor={green}
}
\renewcommand\UrlFont{\color{blue}\rmfamily}

% Source code highlighting
\usepackage[outputdir=../build]{minted}
  % Magic incantation to stop minted from putting red boxes around shit
  \usepackage{etoolbox}
  \makeatletter
  \AtBeginEnvironment{minted}{\dontdofcolorbox}
  \def\dontdofcolorbox{\renewcommand\fcolorbox[4][]{##4}}
  \makeatother
\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}

\newmintinline[highl]{ocaml}{breaklines}

\usepackage{amsmath}

\begin{document}

\title{NumLin: Linear Types for Numerical Libraries}

% \titlerunning{NumLin: Linear Types for Numerical Libraries}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here

\newcommand{\lang}{\textsc{NumLin}}

\author{Dhruv C.~Makwana\inst{1}\orcidID{orcidID} \and
Neelakantan R.~Krishnaswami\inst{2}\orcidID{orcidID}}
%
\authorrunning{D.C. Makwana and N.R. Krishnaswami}

\institute{\email{dcm41@cam.ac.uk}
\url{dhruvmakwana.com} \and
Department of Computer Science, University of Cambridge\\
\email{nk480@cl.cam.ac.uk}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Briefly summarize the contents of the paper in 150--250 words.

\keywords{numerical, linear, algebra, types, permissions, OCaml}
\end{abstract}
%
%
%
\input{intro}

\input{lang_and_examples}

\input{formal_system}

\input{implementation}

\input{discussion}

\bibliographystyle{splncs04}

\bibliography{ourbib}

\end{document}

### 2018-10-21 write-up/paper/lang_and_examples.tex
Add matrix squaring example

@@
 
-The more examples, the better. In fact, it's actually impossible to
-have too many examples -- it's okay (indeed, desirable) to spend 5-6
-pages on examples.
-\begin{itemize}
-
-    \item Simple: factorial, shows recursion, ! annotations etc.
-
-    \item Less simple: summing over an array, indexing and safety
-
-    \item Medium: one-dimensioal convolution, permissions
-
-    \item Harder: linear regression, pattern-matching and apparent non-linearity
-
-    \item Harder: L1 norm-minimisation, some frees, re-using memory
-
-    \item Big finish: Kalman filter
-
-\end{itemize}
-
 \subsubsection{Factorial}
@@
 The condition for an \highl{if} may or may not use linear values (here, with
-\highl{x < 0 || x = 0}, it does not). Any linear values used by the conditional
+\highl{x < 0 || x = 0}, it does not). Any linear values used by the condition
 would not be in scope in either branch of the \highl{if}-expression.  Both
@@
 
-Second is \highl{('x) (row : 'x arr)} which is an array with a
+Second is \highl{('x) (row: 'x arr)} which is an array with a
 universally-quantified fractional permission. In particular, this means the
@@
 
-Alongside taking a \highl{row : 'x arr}, the function also returns an array
+Alongside taking a \highl{row: 'x arr}, the function also returns an array
 with exactly the same fractional permission as the \texttt{row} (which can only
@@
 explicitly  (as done in the recursive call), they can also be
-\emph{automatically inferred at call sites}, \highl{row[i] == get _ row i}
+\emph{automatically inferred at call sites}: \highl{row[i] == get _ row i}
 takes advantage of this convenience.
@@
 \begin{itemize}
-    \item the kernel has a fixed length (of 3), as such, only the value of
-        \highl{write[i-1]} prior to modification in the previous iteration
-        needs to be carried forward using \highl{x0}
+    \item the kernel has a length 3, so only the value of \highl{write[i-1]}
+        (prior to modification in the previous iteration) needs to be carried
+        forward using \highl{x0}
     \item \highl{write} is assumed to have length \highl{n+1}
@@
 
+\subsubsection{Squaring a Matrix}
+
+\begin{figure}[t]
+    \centering
+    \inputminted[fontsize=\small]{ocaml}{../../test/examples/square.lt}
+    \caption{Linear regression (OLS): $\hat\beta =
+        \mathbf{(X^T X)^{-1} X^T y}$}\label{fig:lang_square}
+\end{figure}
+
+\emph{The most pertinent aspect of \lang\ is the types of its primitives.}
+While the types of operations such as \texttt{get} and \texttt{set} might be
+borderline obvious, the types of BLAS/LAPACK routines become an
+\emph{incredibly useful, automated check for using the API correctly.}
+
+Figure~\ref{fig:lang_square} shows how a linearly-typed matrix squaring
+function may be written in \lang. It is a \emph{non-recursive} function
+declaration (the return type is inferred). Since we would like to be able to
+use a function like \texttt{square} more than once, it is marked with a
+\texttt{!} annotation (which also ensures it captures no linear values from the
+surrounding environment).
+
+To square a matrix, first, we extract the dimensions of the argument
+\texttt{x}. Then, because we need to use \texttt{x} twice (so that we can
+multiply it by itself) but linearity only allows one use, we use \highl{shareM:
+'x. 'x mat --o 'x s mat * 'x s mat} to split the permission \texttt{'x} (which
+represents $2^{-x}$) into two halves (\texttt{'x s}, which represents $2^{-(x+1)}$).
+
+Even if \texttt{x} had type \texttt{z mat}, sharing it now enforces the
+assumption of all BLAS/LAPACK routines that any matrix which is written to
+(which, in \lang, is always of type \texttt{z mat}) does not alias any other
+matrix in scope. So if we did try to use one of the aliases in mutating way,
+the expression would not type check, and we would get an error similar to the
+one in Figure~\ref{fig:lang_errormsg}.
+
+The line \highl{let answer <- new (m,n) [| x1 * x2 |]} is syntactic sugar for
+first creating a new $m \times n$ matrix (\highl{let answer = matrix m n}) and
+then storing the result of the multiplication in it (\highl{let ((x1, x2),
+answer) = gemm 1. _ (x1, false) _ (x2, false) 0. answer}). \highl{false} means
+the matrix should not be accessed with indices transposed.
+
+By using some simple pattern-matching and syntactic sugar, we can:
+\begin{itemize}
+    \item write normal-looking, apparently non-linear code
+    \item use matrix expressions directly and have a call to an efficient
+        call to a BLAS/LAPACK routine inserted with appropriate re-bindings
+    \item retain the safety of linear types with fractional permissions by
+        having the compiler statically enforce the aliasing and read/write rules
+        implicitly assumed by BLAS/LAPACK routines.
+\end{itemize}
+
 \subsubsection{Linear Regression}
@@
 
-\emph{The most pertinent aspect of \lang's type system are the types of its
-primitives.} While the types of operations such as \texttt{get} and
-\texttt{set} might be borderline obvious, they types of BLAS/LAPACK routines
-become an \emph{incredibly useful, automated check for using the API correctly.}
+In Figure~\ref{fig:lang_lin_reg}, we wish to compute $\hat\beta = \mathbf{(X^T
+X)^{-1} X^T y}$. To do that, first, we extract the dimensions of matrix
+\texttt{x}. Then, we say we would like \texttt{xy} to be a new matrix, of
+dimension $m \times 1$, which contains the result of $\mathbf{X^T y}$ (using
+syntactic sugar for \texttt{matrix} and \texttt{gemm} calls similar to that
+used in Figure~\ref{fig:lang_square}, with a \highl{^T} annotation on
+\texttt{x} to set \texttt{x}'s `transpose indices'-flag to \highl{true}).
 
-Figure~\ref{fig:lang_lin_reg} shows a \emph{non-recursive} function declaration (the
-return type is inferred). Since we would like to be able to use a function like
-\texttt{lin\_reg} more than once, it is marked with a \texttt{!} annotation
-(which also ensure it captures no linear values from the surrounding
-environment).
-
-We wish to compute $\hat\beta = \mathbf{(X^T X)^{-1} X^T y}$. First, we extract
-the dimensions of matrix \texttt{x}. Then, we say we would like a new matrix, of
-dimension $m \times 1$, which contains the result of $\mathbf{X^T y}$. The line
-\highl{let xy <- new (m,1) [| x^T * y |]} is syntactic sugar for first creating
-a new matrix \highl{let xy = matrix m 1 in} and then storing the result of
-\highl{(x, true (* transp'd *))} multiplied with \highl{(y, false (* not
-transp'd *))} into \texttt{xy}: \highl{let (xy, (x, y)) = gemm 1.  _ (x,
-true) _ (y, false) 0. xy}, appropriately re-binding the identifiers.  The line
-\highl{let x_T_x <- new (m,m) [| x^T * x |]}, works similarly, but uses a more
-appropriate, \highl{syrk true 1. x 0. x_T_x} BLAS call.
-
-By using some simple pattern-matching and syntactic sugar, we can:
-\begin{itemize}
-    \item write normal-looking, apparently non-linear code
-    \item use matrix expressions directly and have an appropriate, efficient
-        call to a BLAS/LAPACK routine inserted
-    \item retain the safety of linear types with fractional permissions by
-        having the compiler statically enforce the aliasing and read/write rules
-        implicitly assumed by BLAS/LAPACK routines.
-\end{itemize}
+However, the line \highl{let x_T_x <- new (m,m) [| x^T * x |]}, works for a
+slightly different reason: that pattern is matched to a BLAS call to
+(\highl{syrk true 1.  x 0. x_T_x}), which only uses \texttt{x} once. Hence
+\texttt{x} can appear \emph{twice} in the \emph{pattern} without any calls to
+\texttt{share}.
 
@@
 Figure~\ref{fig:lang_l1_norm_min} shows even more pattern-matching. Patterns of
-the form \highl{let <id> <- [| beta * c + alpha * a * b |]} are desugared to
+the form \highl{let <id> <- [| beta * c + alpha * a * b |]} are also desugared to
 \texttt{gemm} calls. Primitives like \highl{transpose: 'x. 'x mat --o 'x mat *
 z mat} and \highl{eye: !int --o z mat} allocate new matrices;
-\texttt{transpose} returns the transpose of a given matrix and \texttt{eye}
-returns a $k \times k$ identity matrix.
+\texttt{transpose} returns the transpose of a given matrix and \texttt{eye k}
+evaluates to a $k \times k$ identity matrix.
 
@@
 with \texttt{to\_del} and \texttt{posv} in the previous example, we do not need
-the value stored in \texttt{tmp\_5\_5} after the call to \texttt{gesv}. However,
-we can use the memory much later to store \texttt{answer} with \highl{let
-answer <- [| 0. * tmp_5_5 + q_inv_u * inv_u_T |]}. Again, thanks to linearity,
-the identifiers \texttt{q}, \texttt{tmp\_5\_5} are out of scope by the time
-\texttt{answer} is bound. Although during execution, all three refer to the
-same piece of memory, logically they represent different values throughout the
-computation.
+the value stored in \texttt{tmp\_5\_5} after the call to \texttt{gesv} (a
+primitive similar to \texttt{posv} but for a non-symmetric first argument).
+However, we can re-use its memory much later to store \texttt{answer} with
+\highl{let answer <- [| 0. * tmp_5_5 + q_inv_u * inv_u_T |]}. Again, thanks to
+linearity, the identifiers \texttt{q} and \texttt{tmp\_5\_5} are out of scope
+by the time \texttt{answer} is bound. Although during execution, all three
+refer to the same piece of memory, logically they represent different values
+throughout the computation.
 
@@
     \inputminted[fontsize=\small]{ocaml}{../../test/examples/kalman.lt}
-    \caption{Kalman filter: see Figure~\ref{fig:maths_kalman} for the
-        equations this code implements.}\label{fig:lang_kalman}
+    \caption{Kalman filter: see Figure~\ref{fig:kalman_eqns} for the
+        equations this code implements. Line numbers in comments refer to
+        equivalent lines in a C implementation.}\label{fig:lang_kalman}
 \end{figure}
@@
     \caption{Kalman filter equations (credit:
-    \href{http://matthewrocklin.com/blog/work/2012/11/24/Kalman-Filter}{matthewrocklin.com}).}\label{fig:maths_kalman}
+    \href{http://matthewrocklin.com/blog/work/2012/11/24/Kalman-Filter}{matthewrocklin.com}).}\label{fig:kalman_eqns}
 \end{figure}
@@
 Figure~\ref{fig:lang_kalman} shows a \lang\ implementation of a Kalman filter
-(equations in Figure~\ref{fig:maths_kalman}). A few new features and techniques
+(equations in Figure~\ref{fig:kalman_eqns}). A few new features and techniques
 are used in this implementation:
@@
         symmetric matrices so that only half the operations are performed) is
-        inserted (instead of one to \texttt{gemm})
+        inserted
 
@@
             \item \texttt{h\_sol\_h}, \texttt{sigma\_copy} and \texttt{new\_sigma}
-            \item \texttt{mu\_copy} and \texttt{new\_mu}
+            \item \texttt{mu\_copy} and \texttt{new\_mu}.
         \end{itemize}
@@
 \end{itemize}
+
+With expressions as complex as those in Figure~\ref{fig:kalman_eqns}, it can
+get very difficult to ensure memory is not being re-used incorrectly or re-used
+enough whilst also enusring the aliasing assumptions for the BLAS/LAPACK
+routine calls are guaranteed. However, using \lang\ makes incorrect aliasing
+and memory re-use impossible. Additionally, \lang\ makes it very easy to spot
+where memory is not being re-used enough \emph{precisely because} of its
+explicit handling of matrix allocations, copies and frees.

### 2018-10-21 write-up/paper/implementation.tex
Sketch out Implementation section

@@
 \section{Implementation}\label{sec:implementation}
-\subsection{Implementation Strategy}
-
-% TODO desugaring
 
@@
 about how implementing inference for fractions is.
+
+\subsection{Implementation Strategy}
+
+\lang\ transpiles to OCaml and its implementation follows the structure of a
+typical domain-specific language compiler. Although \lang's current
+implementation is not as embedded DSL, its the general design is simple enough
+to adapt to being so and also to target other languages.
+
+Alongside the transpiler, a `Read-Check-Translate' loop, benchmarking program
+and a test suite are included in executable artifacts accompanying this paper.
+
+\begin{enumerate}
+
+    \item \textbf{Parsing}. A generated, LR(1) parser parses a text file into a
+        syntax tree. In general, this part will vary for different languages
+        and can also be dealt with using combinators or syntax-extensions (the
+        EDSL approach) if the host language offers such support.
+
+    \item \textbf{Desugaring}. The syntax tree is then desugared into a
+        smaller, more concise, abstract syntax tree. This allows for the type
+        checker to be simpler to specify and easier to implement.
+
+    \item \textbf{Matrix Expressions} are also desugared into the abstract
+        syntax tree through some simple pattern-matching.
+
+    \item \textbf{Type checking}. The abstract syntax tree is explicitly typed,
+        with some inference to make writing typical programs more convenient.
+
+    \item \textbf{Code Generation}. The abstract syntax tree is translated into
+        OCaml, with a few `optimisations' to produce more readable code. This
+        process is type-preserving: \lang's type system is embedded into
+        OCaml's (Figure~\ref{fig:type_grammar}), and so the OCaml type checker
+        acts as a sanity check on the generated code.
+
+\end{enumerate}
+
+A very pleasant way to use \lang\ is to have the build system generate code at
+\emph{compile-time} and then have it be used by other modules like normal OCaml
+functions. This makes it possible and even easy to use \lang\ alongside
+existing OCaml libraries; in fact, this is exactly how the benchmarking program
+and test-suite use code written in \lang.
+
+\begin{figure}[tp]
+    \centering
+    \begin{minipage}{.3\textwidth}
+        \centering
+        \begin{grammar}
+            <f> ::= `'
+            \alt <fc>
+            \alt `Z'
+            \alt `S' <f>
+
+            <t> ::= `'
+            \alt `unit'
+            \alt `bool'
+            \alt `int'
+            \alt `elt'
+            \alt <f> `arr'
+            \alt <f> `mat'
+            \alt `!' <t>
+            \alt \synt{$\forall$} <fc.> <t>
+            \alt <t> \lit{$\otimes$} \synt{$t'$}
+            \alt <t> \lit{$\multimap$} \synt{$t'$}
+        \end{grammar}
+    \end{minipage}
+    \begin{minipage}{.3\textwidth}
+        \centering
+        \begin{minted}[fontsize=\small]{ocaml}
+module Arr =
+  Owl.Dense.Ndarray.D
+
+type z = Z
+type 'a s = Succ
+
+type 'a arr =
+  A of Arr.arr
+  [@@unboxed]
+
+type 'a mat =
+  M of Arr.arr
+  [@@unboxed]
+
+type 'a bang =
+  Many of 'a
+  [@@unboxed]
+        \end{minted}
+    \end{minipage}
+    \begin{minipage}{.3\textwidth}
+        \begin{align*}
+            [\![ f\!c ]\!] &= \texttt{'fc} \\
+            [\![ \textbf{Z} ]\!] &= \texttt{z}\\
+            [\![ \textbf{S} \, f ]\!] &= [\![ f ]\!]\, \texttt{s}\\
+            [\![ \textbf{unit} ]\!] &= \texttt{unit}\\
+            [\![ \textbf{bool} ]\!] &= \texttt{bool}\\
+            [\![ \textbf{int} ]\!] &= \texttt{int}\\
+            [\![ \textbf{elt} ]\!] &= \texttt{float}\\
+            [\![ f\, \textbf{arr} ]\!] &= [\![ f ]\!]\, \texttt{arr}\\
+            [\![ f\, \textbf{mat} ]\!] &= [\![ f ]\!]\, \texttt{mat}\\
+            [\![ \textbf{!} \, t ]\!] &= [\![ t ]\!]\, \texttt{bang}\\
+            [\![ \forall f\!c.\, t ]\!] &= [\![ t ]\!]\\
+            [\![ t \otimes t' ]\!] &= [\![ t ]\!] \texttt{*} [\![ t' ]\!]\\
+            [\![ t \multimap t' ]\!] &= [\![ t ]\!] \rightarrow [\![ t' ]\!]
+        \end{align*}
+    \end{minipage}
+    \caption{\lang's type grammar (left) and its embedding into OCaml
+        (right).}\label{fig:type_grammar}
+\end{figure}
+
+\subsubsection{Desugaring}\label{subsubsec:desugaring}
+
+Straightforward enough.
+
+\subsubsection{Matrix Expressions}
+
+Pattern matching!
+
+\subsubsection{Type Checking and Fractional Permission Inference}
+
+Unification!
+
+\subsubsection{Code Generation}
+
+A few examples?
 

### 2018-10-21 write-up/paper/lang_and_examples.tex
Sketch out Implementation section

@@
 syntactic convenience for declaring the value to used intuitionistically, its
-full and precise meaning is described in Section~\ref{sec:implementation}.
+full and precise meaning is described in Section~\ref{subsubsec:desugaring}.
 
@@
 With expressions as complex as those in Figure~\ref{fig:kalman_eqns}, it can
-get very difficult to ensure memory is not being re-used incorrectly or re-used
-enough whilst also enusring the aliasing assumptions for the BLAS/LAPACK
-routine calls are guaranteed. However, using \lang\ makes incorrect aliasing
-and memory re-use impossible. Additionally, \lang\ makes it very easy to spot
-where memory is not being re-used enough \emph{precisely because} of its
-explicit handling of matrix allocations, copies and frees.
+get very difficult to ensure memory is not being re-used incorrectly/enough
+whilst also enusring the aliasing assumptions for the BLAS/LAPACK routine calls
+are guaranteed. However, using \lang\ makes incorrect aliasing and memory
+re-use impossible. Additionally, \lang\ makes it possible to spot where memory
+is not being re-used enough \emph{precisely because} of its explicit handling
+of allocations, aliases and frees.

### 2018-10-23 write-up/semantics/lemmas.tex
Add Ty_Gen case

\section{Lemmas}

\subsection{$
    \forall \sigma_s, \sigma_r, e.\ \varsigma_s \star \varsigma_r \textrm{ defined }
    \Rightarrow \forall n.\ \langle \sigma_s, e \rangle \rightarrow^n =
    \langle \sigma_s + \sigma_r , e \rangle \rightarrow^n
$}\label{frame}

\begin{proof}

    \suffices{By induction on $n$, consider only the cases $\langle \sigma_s, e
        \rangle \rightarrow \langle \sigma_f, e_f \rangle$ where $\sigma_s \neq
        \sigma_f$.\\}

    \pfsketch{~Only \textsc{Op\_\{Free, Matrix, Share, Unshare\_Eq,
        Gemm\_Match\}} change the heap: the rest are either parametric in the
        heap or step to an $\ottkw{err}$.\\}

    \prove{$\langle \sigma_s + \sigma_r, e \rangle \rightarrow
        \langle \sigma_f + \sigma_r, e_f \rangle$.\\}

    \step{}{\case{\textsc{Op\_Free}, $\sigma_s \equiv \sigma' + \{ l
        \mapsto_1 m \}$, $\sigma_f = \sigma'$.}{\pf~Instantiate
        \textsc{Op\_Free} with $(\sigma' + \sigma_r) + \{ l \mapsto_1 m
        \}$,\\valid because $l \notin \dom(\varsigma_r)$ by $\varsigma' \star
        [l \mapsto_1 m] \star \varsigma_r$ defined (assumption).}}

    \step{}{\case{\textsc{Op\_Matrix}}{
        \pf~Rule has no requirements on $\sigma_s$ so will also work with
        $\sigma_s + \sigma_r$.
    }}

    \step{}{\case{\textsc{Op\_Share}, $\sigma_s \equiv \sigma' + \{ l \mapsto_f
        m \}$, $\sigma_f = \sigma' + \{ l \mapsto_{\frac{1}{2} \cdot f} m \} +
        \{ l \mapsto_{\frac{1}{2} \cdot f} m \}$.}{
        \pf~Union-ing $\sigma_r$ does not remove $l \mapsto_f m$, so that can
        be split out of $ \sigma_s + \sigma_r$ as before.
    }}

    \step{}{\case{\textsc{Op\_Unshare\_Eq}, $\sigma_s \equiv \sigma' + \{ l
        \mapsto_{\frac{1}{2} \cdot f} m \} + \{ l \mapsto_{\frac{1}{2} \cdot f}
        m \}$, $\sigma_f = \sigma' + \{ l \mapsto_f m \}$.}{}}

    \begin{proof}

        \step{}{Union-ing $\sigma_r$ does not remove $l \mapsto_{\frac{1}{2}
            \cdot f} m$, so that can still be split out of $ \sigma_s + \sigma_r$.}

        \step{}{There may also be other valid splits introduced by $\sigma_r$.}

        \step{}{However, by assumption of $\varsigma_s \star \varsigma_r$
            defined, any splitting of $\sigma_s + \sigma_r$ will
            satisfy $f \leq 1$.}

    \end{proof}

    \step{}{\case{\textsc{Op\_Gemm\_Match}}{}}

    \begin{proof}

        \step{}{By assumption of $\varsigma_s \star \varsigma_r$ defined,
            either $l_1$ (or $l_2$, or both) are not in $\sigma_r$, or they are
            and the matrix values they point to are the same.}

        \step{}{The permissions (of $l_1$ and/or $l_2$) may differ, but
            \textsc{Op\_Gemm\_Match} universally quantifies over them and
            leaves them unchanged, so they are irrelevant.}

        \step{}{Only the pointed to matrix value at $l_3$ changes.}

        \step{}{\suffices{$l_3 \notin \pi_1 [ \sigma_r ]$.}}

        \step{}{By assumption of $\varsigma_s \star \varsigma_r$ defined, $l_3
            \notin \dom(\varsigma_r)$.}

        \step{}{Hence $l_3 \notin \pi_1 [ \sigma_r ]$.}

    \end{proof}

\end{proof}

\subsection{$\forall k, t.\ \V{k}{t} \subseteq \C{k}{t}$}\label{subsetVC}
Follows from definition of $\C{k}{t}$, $\rightarrow^j$ ($\forall n.\ \langle
\sigma , v \rangle \rightarrow^n \langle \sigma , v \rangle$) for arbitrary
$j \leq k$ and \ref{frame}.

\subsection{$\forall \theta, \delta, \gamma, v.\ \theta(\delta(\gamma(v)))\textrm{ is a value.}$}\label{valueSub}

$\theta$ is irrelevant because it only maps fractional capability variables to
fractional capabilities. By construction, $\delta$ and $\gamma$ only map
variables to values, and values are closed under substitution.

\subsection{$
    \forall k, \sigma, \sigma', e, e', t.  \ (\varsigma', e') \in \C{k}{t} \wedge
    \langle \sigma, e \rangle \rightarrow \langle \sigma', e' \rangle
    \Rightarrow (\varsigma, e) \in \C{k+1}{t}
$}\label{stepInC}

\begin{proof}

    \assume{arbitrary $j < k + 1$, and $\sigma_r$ such that $\varsigma
    \star \varsigma_r$ defined.\\}

    \step{}{\case{$j=0$. Clearly $\sigma_f = \sigma_s + \sigma_r$ and $e' = e$.}{
        Remains to show that if $e$ is a value then $(\varsigma_s \star
        \varsigma_r, e) \in \V{k}{t}$.\\
        This is true vacuously, because by assumption, $e$ is not a value.}}

    \step{}{\case{$j \geq 1$. We have $\langle \sigma, e \rangle
        \rightarrow^j\, = \langle \sigma', e' \rangle \rightarrow^{j-1}$.}{
        Instantiate $(\varsigma', e') \in \C{k}{t}$, with $j-1 < k$ and
        $\sigma_r$ to conclude the required conditions.}}

\end{proof}

\subsection{$j \leq k \Rightarrow \den{\_\ }{k}{\cdot} \subseteq \den{\_\ }{j}{\cdot}$}\label{subsetKJ}

Lemma~\ref{stepInC} is the inductive step for this lemma for the $\C{}{}$ case.\\
Need to prove for $\V{}{}$, by induction on $t$ and then index.

\begin{proof}

    \suffices{Consider only $t \multimap t'$ case, rest use $k$ directly on structure of type.}

    \assume{Arbitrary $j \leq k$ and $(\varsigma_{v'}, v') \in \V{k}{t
        \multimap t'}$.}

    \prove{$(\varsigma_{v'}, v') \in \V{j}{t \multimap t'}$.\\}

    \step{}{$v'$ is of the correct syntactic form (lambda or fixpoint) by
        assumption.}

    \step{}{\assume{arbitrary $j' \leq j$ and $(\varsigma_v, v) \in \V{j'}{t}$
        such that $\varsigma_{v'} \star \varsigma_v$ is defined.}}

    \step{}{\suffices{to show $(\varsigma_{v'} \star \varsigma_v, v' v) \in
        \C{j'}{t'}$.}}

    \step{}{This is true by instantiating $(\varsigma_{v'}, v') \in \V{k}{t
        \multimap t'}$ with $j' \leq k$ and $(\varsigma_v, v) \in \V{j'}{t}$.}

\end{proof}

\subsection{$\forall \Delta, \Gamma, t, k, \theta, \delta, \gamma.\ %
    \delta \in \den{I}{k}{\Delta}\theta \wedge \gamma \in \pi_2[\den{L}{k}{\Gamma}\theta]
    \Rightarrow \dom(\Delta) = \dom(\delta)$ and $\dom(\Gamma) = \dom(\gamma)$}\label{samedom}

\pf~By induction on $\Delta$ and $\Gamma$.

\subsection{$\forall k, \Gamma, \Gamma', \theta, \sigma_+, \gamma_+.\ %
    (\varsigma_+, \gamma) \in \den{L}{k}{ \Gamma, \Gamma' }\theta
    \wedge \Gamma, \Gamma' \textrm{ disjoint } \Rightarrow\\
    \exists \sigma, \gamma, \sigma' , \gamma' .\ \sigma_+ = \sigma + \sigma'
    \wedge \gamma, \gamma' \textrm{ disjoint }
    \wedge \gamma_+ = \gamma \cup \gamma'\\
    \wedge (\varsigma, \gamma) \in \den{L}{k}{\Gamma}
    \wedge (\varsigma', \gamma') \in \den{L}{k}{\Gamma'} $}\label{restriction}

\pf~By induction on $\Gamma'$.

### 2018-10-25 write-up/paper/implementation.tex
Start desugaring

@@
 \lang\ transpiles to OCaml and its implementation follows the structure of a
-typical domain-specific language compiler. Although \lang's current
+typical domain-specific language (DSL) compiler. Although \lang's current
 implementation is not as embedded DSL, its the general design is simple enough
@@
 Alongside the transpiler, a `Read-Check-Translate' loop, benchmarking program
-and a test suite are included in executable artifacts accompanying this paper.
+and a test suite are included in the artifacts accompanying this paper.
 
@@
 A very pleasant way to use \lang\ is to have the build system generate code at
-\emph{compile-time} and then have it be used by other modules like normal OCaml
-functions. This makes it possible and even easy to use \lang\ alongside
-existing OCaml libraries; in fact, this is exactly how the benchmarking program
-and test-suite use code written in \lang.
+\emph{compile-time} and then have the generated code be used by other modules
+like normal OCaml functions. This makes it possible and even easy to use \lang\
+alongside existing OCaml libraries; in fact, this is exactly how the
+benchmarking program and test-suite use code written in \lang.
+
+\subsubsection{Desugaring}\label{subsubsec:desugaring}
+
+Desugaring is conventional.
+
+\begin{align*}
+    \mathbf{fun}\ (x : t)\ 'r\ (y : {'r}\ \mathbf{mat}) \rightarrow e
+    & \Rightarrow
+    \mathbf{fun}\ (x : t) \rightarrow \mathbf{fun}\ {'r} \rightarrow
+    \mathbf{fun}\ (y : {'r}\ \mathbf{mat}) \rightarrow e
+\\
+    x[e]
+    & \Rightarrow
+    \mathbf{get}\ \_\ x\ (e) \qquad \textrm{(similarly for matrices)}
+\\
+    x[e_1] := e_2
+    & \Rightarrow
+    \mathbf{set}\ x\ (e_1)\ (e_2) \qquad \textrm{(similarly for matrices)}
+\\
+    \mathbf{let}\ !x = e_1\ \mathbf{in}\ e_2
+    & \Rightarrow
+    \mathbf{let\ Many}\ x = e_1\ \mathbf{in\ }
+    \mathbf{let\ Many}\ x = \mathbf{Many}\ (\mathbf{Many}\ x)\ \mathbf{in}\ e_2
+\\
+    \mathbf{let}\ f\ {\langle args \rangle} = e_1\ \mathbf{in}\ e_2
+    & \Rightarrow
+    \mathbf{let}\ f = \mathbf{fun}\ {\langle args \rangle} \rightarrow e_1\
+    \mathbf{in}\ e_2
+\\
+    \mathbf{let}\ !f\ {\langle args \rangle} = e_1\ \mathbf{in}\ e_2
+    & \Rightarrow
+    \mathbf{let\ Many}\ f = \mathbf{Many}\ (\mathbf{fun}\ {\langle args \rangle}
+    \rightarrow e_1)\ \mathbf{in}\ e_2
+\\
+    \mathbf{let\ rec}\ f\ (x : t)\ {\langle args \rangle} : {t'} = e_1\ \mathbf{in}\ e_2
+    & \Rightarrow
+    \mathbf{let\ Many}\ f = \mathbf{Many}\ (\mathbf{fix}\ (f, x : t,
+    \mathbf{fun}\ {\langle args \rangle} \rightarrow e_1 : {t'} ))\ \mathbf{in}\ e_2
+\end{align*}
+
+\subsubsection{Matrix Expressions}
+
+Pattern matching!
+
+\subsubsection{Type Checking and Fractional Permission Inference}
+
+Unification!
+
+\subsubsection{Code Generation}
 
@@
 \end{figure}
-
-\subsubsection{Desugaring}\label{subsubsec:desugaring}
-
-Straightforward enough.
-
-\subsubsection{Matrix Expressions}
-
-Pattern matching!
-
-\subsubsection{Type Checking and Fractional Permission Inference}
-
-Unification!
-
-\subsubsection{Code Generation}
-
 A few examples?

### 2018-10-25 write-up/paper/lang_and_examples.tex
Start desugaring

@@
     \item \texttt{potrs \_ chol\_r data\_2} uses a pre-computed Cholesky decomposition
-        to multiply \texttt{data\_2} by \texttt{r\_2} $= (R + H \Sigma H^T)^{-1}$
+        to multiply \texttt{data\_2} $ = H \mu - \textrm{data}$ by
+        \texttt{r\_2} $= (R + H \Sigma H^T)^{-1}$
 
@@
 re-use impossible. Additionally, \lang\ makes it possible to spot where memory
-is not being re-used enough \emph{precisely because} of its explicit handling
-of allocations, aliases and frees.
+is not being re-used enough \emph{precisely because} of its explicit
+allocations, aliases and frees.

### 2018-10-26 write-up/paper/intro.tex
Add points about finding bugs

@@
 and recursion into a small, easily understandable, yet expressive set of core
-constructs. In addition to this, \lang's implementation supports several
-syntactic conveniences as well as a usable integration with real-world OCaml
-code.
+constructs. \lang\ allows a novice to understand and work with complicated
+linear algebra library APIs; as well as point out subtle aliasing bugs and
+reduce memory usage in existing programs. In addition to this, \lang's
+implementation supports several syntactic conveniences as well as a
+\emph{usable} integration with real OCaml libraries.
 
 \subsection{Contributions}
+
+In this paper
 \begin{itemize}
-    \item We have designed the \lang\ programming language
-    \item We illustrate that the design is sensible with many matrix-y examples
-    \item We give a soundness proof for \lang, using a step-indexed logical relation
-    \item Incredibly simple type inference algorihtm for polymorphic fractional permissions
-        \begin{itemize}
-            \item Compare to Bierhof et al's \emph{Fraction Polymorphic Permission Inference},
-                which uses a fancy dataflow analysis
-            \item We use exactly the same unification algorithm type polymorphism does
+    \item we describe \lang, a linearly typed language for linear algebra programs
+    \item we illustrate that \lang's design and features are well-suited to its
+        intended domain with progressively sophisticated examples
+    \item we prove \lang's soundness, using a step-indexed logical relation
+    \item we describe a very simple, unification based type-inference algorithm
+        for polymorphic fractional permissions (similar to ones used for
+        parametric polymorphism), demonstrating an alternative approach to the
+        dataflow analysis of Bierhof et al's \emph{Fraction Polymorphic
+        Permission Inference}
+    \item we describe an implementation that is both compatible with and usable
+        from existing code.
 \end{itemize}
-    \item We have an implementation - compatible with and usable from existing code!
-\end{itemize}

### 2018-10-28 write-up/paper/implementation.tex
Finish desugaring and matrix expression figures

--- 2018-10-21 write-up/paper/implementation.tex
@@
 \lang\ transpiles to OCaml and its implementation follows the structure of a
-typical domain-specific language compiler. Although \lang's current
-implementation is not as embedded DSL, its the general design is simple enough
+typical domain-specific language (DSL) compiler. Although \lang's current
+implementation is not as an embedded DSL, its the general design is simple enough
 to adapt to being so and also to target other languages.
@@
 A very pleasant way to use \lang\ is to have the build system generate code at
-\emph{compile-time} and then have it be used by other modules like normal OCaml
-functions. This makes it possible and even easy to use \lang\ alongside
-existing OCaml libraries; in fact, this is exactly how the benchmarking program
-and test-suite use code written in \lang.
+\emph{compile-time} and then have the generated code be used by other modules
+like normal OCaml functions. This makes it possible and even easy to use \lang\
+alongside existing OCaml libraries; in fact, this is exactly how the
+benchmarking program and test-suite use code written in \lang.
+
+\subsubsection{Desugaring, Matrix Expressions and Type Checking}\label{subsubsec:initial}
+
+Desugaring is conventional, outlined in Figure~\ref{fig:lang_desugar}.  Matrix
+expression are translated into BLAS/ LAPACK calls via pattern-matching,
+outlined in Figure~\ref{fig:lang_matexp}. Type checking is mostly standard for
+a linearly typed language, with the exception of fractional permission
+inference.  Because all functions must have their argument types explicitly
+annotated, inferring the correct fraction at a call-site is simply a matter of
+unification.
+
+\textsc{What more can/should we say here?}
+
+\begin{figure}[t]
+\begin{center}
+\[\def\arraystretch{1.3}
+    \begin{array}{rcl}
+    x[e] &
+    \Rightarrow &
+    \mathbf{get}\ \_\ x\ (e) \;\qquad \textrm{(similarly for matrices)}
+\\
+    x[e_1] := e_2 &
+    \Rightarrow &
+    \mathbf{set}\ x\ (e_1)\ (e_2) \quad \textrm{(similarly for matrices)}
+\\
+\\
+    pat & ::= & ()\ \mid\ x\ \mid\ !x\ \mid\ \mathbf{Many\ } pat\ \mid\ (pat, pat)
+\\
+    \mathbf{let}\ !x = e_1\ \mathbf{in}\ e_2 &
+    \Rightarrow &
+    \specialcell[t]{l}{\mathbf{let\ Many}\ x = e_1\ \mathbf{in\ } \\
+    \mathbf{let\ Many}\ x = \mathbf{Many}\ (\mathbf{Many}\ x)\ \mathbf{in}\ e_2}
+\\
+    \mathbf{let\ Many} \langle pat_x \rangle\ = e_1\ \mathbf{in}\ e_2 &
+    \Rightarrow &
+    \specialcell[t]{l}{%
+        \mathbf{let\ Many}\ x = x\ \mathbf{in\ } \\
+        \mathbf{let\ }\ \langle pat_x \rangle\ = x\ \mathbf{in\ } e_2}
+\\
+    \mathbf{let}\ (\langle pat_a \rangle, \langle pat_b \rangle)\  = e_1\ \mathbf{in}\ e_2 &
+    \Rightarrow &
+    \specialcell[t]{l}{%
+        \mathbf{let\ }\ (a,b)\ = a\_b\ \mathbf{in\ }
+        \ \mathbf{let\ }\ \langle pat_a \rangle\ = a\ \mathbf{in\ } \\
+        \mathbf{let\ }\ \langle pat_b \rangle\ = b\ \mathbf{in\ } e_2}
+\\
+    \mathbf{fun}\ (\langle pat_x \rangle\ : t) \rightarrow e &
+    \Rightarrow &
+    \mathbf{fun}\ (x : t) \rightarrow \mathbf{let\ }\ \langle pat_x \rangle = x\ \mathbf{in\ } e
+\\
+\\
+    arg & ::= & \langle pat \rangle : t\ \mid\ {'x} \textrm{ (fractional permission variable)}
+\\
+    \mathbf{fun}\ \langle arg_{1 .. n} \rangle \rightarrow e &
+    \Rightarrow &
+    \mathbf{fun}\ \langle arg_1 \rangle \rightarrow ..
+    \ \mathbf{fun}\ \langle arg_n \rangle \rightarrow e
+\\
+    \mathbf{let}\ f\ {\langle arg_{1 .. n} \rangle} = e_1\ \mathbf{in}\ e_2 &
+    \Rightarrow &
+    \mathbf{let}\ f = \mathbf{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1\
+    \mathbf{in}\ e_2
+\\
+    \mathbf{let}\ f\ {\langle arg_{1 .. n} \rangle} = e_1\ \mathbf{in}\ e_2 &
+    \Rightarrow &
+    \mathbf{let}\ f = \mathbf{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1\
+    \mathbf{in}\ e_2
+\\
+    \mathbf{let}\ !f\ {\langle arg_{1 .. n} \rangle} = e_1\ \mathbf{in}\ e_2 &
+    \Rightarrow &
+    \mathbf{let\ Many}\ f = \mathbf{Many}\ (\mathbf{fun}\ {\langle arg_{1 .. n} \rangle}
+    \rightarrow e_1)\ \mathbf{in}\ e_2
+\\
+    \mathbf{let\ rec}\ f\ (x : t)\ {\langle arg_{1 .. n} \rangle} : {t'} = e_1\ \mathbf{in}\ e_2 &
+    \Rightarrow &
+    \specialcell[t]{l}{\mathrm{fixpoint} \equiv \mathbf{fix}\ (f, x : t,
+        \mathbf{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1 : {t'} ) \\
+    \mathbf{let\ Many}\ f = \mathbf{Many}\ \mathrm{fixpoint}\ \mathbf{in}\ e_2}
+    \end{array}
+\]
+\end{center}
+\caption{Desugaring from \lang\ concrete syntax to core constructs.}\label{fig:lang_desugar}
+\end{figure}
+
+\begin{figure}[t]
+\begin{center}
+\[\def\arraystretch{1.3}
+    \begin{array}{rcl}
+    \mathbf{let\ } v \leftarrow x[e]\ \mathbf{in\ } e &
+    \Rightarrow &
+    \mathbf{let\ } (x, !v)\ = x[e]\ \mathbf{in\ } e \qquad \textrm{(similarly for matrices)}
+\\
+    \mathbf{let\ } x_2 \leftarrow \mathbf{new}\ [|\ x_1\ |]\ \mathbf{in\ } e &
+    \Rightarrow &
+    \mathbf{let\ } (x_1, x_2)\ = \mathbf{copyM\ } \_\ x_1\ \mathbf{in\ } e
+\\
+    \mathbf{let\ } x_2 \leftarrow [|\ x_1\ |]\ \mathbf{in\ } e &
+    \Rightarrow &
+    \mathbf{let\ } (x_1, x_2)\ = \mathbf{copyM\_to\ } \_\  x_1\ x_2\ \mathbf{in\ } e
+    \end{array}
+\]
+\[
+    M ::= X\ \mid\ X^T\ \mid\ \textrm{sym}( X )
+\]
+\end{center}
+\begin{align*}
+% new
+    \mathbf{let\ } &Y \leftarrow \mathbf{new}\ (n, k)\ [|\ \alpha M_1 M_2 \ |]\ \mathbf{in\ } e
+    \Rightarrow \\
+    & \mathbf{let\ } Y\ = \mathbf{matrix\ } n\ k\ \mathbf{in}
+    \ \mathbf{let\ } Y \leftarrow [|\ \alpha M_1 M_2 + 0Y\ |]\ \mathbf{in\ } e \\
+\\
+%syrk
+    \mathbf{let\ } &Y \leftarrow [|\ \alpha X X^T + \beta Y\ |]\ \mathbf{in\ } e
+    \Rightarrow \\
+    & \mathbf{let\ } (X,Y)\ = \mathbf{syrk\ } \mathbf{false}
+    \ \alpha\ \_\ X\ \beta\ Y\ \mathbf{in\ } e \\
+\\
+    \mathbf{let\ } &Y \leftarrow [|\ \alpha X^T X + \beta Y\ |]\ \mathbf{in\ } e
+    \Rightarrow \\
+    & \mathbf{let\ } (X,Y)\ = \mathbf{syrk\ } \mathbf{true}
+    \ \alpha\ \_\ X\ \beta\ Y\ \mathbf{in\ } e \\
+%symm
+\\
+    \mathbf{let\ } &Y \leftarrow\ [|\ \alpha\,\textrm{sym}(X_1)\,X_2 + \beta Y\ |]\ \mathbf{in\ } e
+    \Rightarrow \\
+    & \mathbf{let\ } ((X_1,X_2),Y)\ = \mathbf{symm\ } \mathbf{false}
+        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \mathbf{in\ }e \\
+\\
+    \mathbf{let\ } &Y \leftarrow\ [|\ \alpha\,X_2\,\textrm{sym}(X_1) + \beta Y\ |]\ \mathbf{in\ } e
+    \Rightarrow \\
+    & \mathbf{let\ } ((X_1,X_2),Y)\ = \mathbf{symm\ } \mathbf{false}
+        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \mathbf{in\ } e \\
+\\
+%gemm
+    \mathbf{let\ } &Y \leftarrow [|\ \alpha X_1^{T?} X_2^{T?} + \beta Y\ |]\ \mathbf{in\ } e
+    \Rightarrow \\
+    & \mathbf{let\ } ((X_1,X_2),Y)\ =\ \mathbf{gemm\ } \alpha
+        \ \_\ \left(X_1, \substack{\mathbf{true}\\\mathbf{false}}\right)
+        \ \_\ \left(X_2, \substack{\mathbf{true}\\\mathbf{false}}\right)
+        \ \beta\ Y\ \mathbf{in\ } e
+\end{align*}
+
+\caption{Pattern-matching on matrix expressions.}\label{fig:lang_matexp}
+\end{figure}
+
+\subsubsection{Code Generation}
 

### 2018-10-28 write-up/paper/intro.tex
Finish desugaring and matrix expression figures

@@
         for polymorphic fractional permissions (similar to ones used for
-        parametric polymorphism), demonstrating an alternative approach to the
-        dataflow analysis of Bierhof et al's \emph{Fraction Polymorphic
-        Permission Inference}
+        parametric polymorphism), demonstrating an alternative approach to
+        dataflow analysis \cite{bierhoff}
     \item we describe an implementation that is both compatible with and usable

### 2018-10-28 write-up/paper/lang_and_examples.tex
Finish desugaring and matrix expression figures

@@
 syntactic convenience for declaring the value to used intuitionistically, its
-full and precise meaning is described in Section~\ref{subsubsec:desugaring}.
+full and precise meaning is described in Section~\ref{subsubsec:initial}.
 
@@
 
+L1-Norm minimisation is often used in optimisation problems, as a
+\emph{regularisation} term for reducing the influence of outliers.  Although
+the below formuation\cite{bronstein} is intended to be used with \emph{sparse}
+computations, \lang's current implementation only implements dense ones.
+However, it still serves as a useful example of explaining \lang's features.
+
 \begin{figure}[t]
@@
 \subsubsection{Kalman Filter}
+
+A \emph{Kalman Filter}\cite{kalman} is a an algorithm for combining prior
+knowledge of a state, a statistical model and measurements from (noisy) sensors
+to produce an estimate a more reliable estimated of the current state.  It has
+various applications (navigation, signal-processing, econometrics) and is
+relevant here because it is usually presented as a series of complex matrix
+equations.
 

### 2018-10-28 write-up/paper/implementation.tex
Start code generation subsubsection

@@
 Desugaring is conventional, outlined in Figure~\ref{fig:lang_desugar}.  Matrix
-expression are translated into BLAS/ LAPACK calls via pattern-matching,
-outlined in Figure~\ref{fig:lang_matexp}. Type checking is mostly standard for
-a linearly typed language, with the exception of fractional permission
-inference.  Because all functions must have their argument types explicitly
-annotated, inferring the correct fraction at a call-site is simply a matter of
-unification.
+expression are translated into BLAS/ LAPACK calls via purely syntactic
+pattern-matching, outlined in Figure~\ref{fig:lang_matexp}. Type checking is
+mostly standard for a linearly typed language, with the exception of fractional
+permission inference.  Because all functions must have their argument types
+explicitly annotated, inferring the correct fraction at a call-site is simply a
+matter of unification.
 
@@
 
-\caption{Pattern-matching on matrix expressions.}\label{fig:lang_matexp}
+\caption{Purely syntactic pattern-matching translations of
+    matrix expressions.}\label{fig:lang_matexp}
 \end{figure}
 
-\subsubsection{Code Generation}
+\subsubsection{Code Generation} is a straightforward mapping from \lang's core
+constructs to high-level OCaml ones. We embed \lang's type- and term-
+constructors into OCaml as a sanity check on the output
+(Figure~\ref{fig:type_grammar}).
 
-\begin{figure}[tp]
+\begin{figure}[t]
     \centering
@@
 \end{figure}
+
+This is also useful when using \lang\ from within OCaml; for example, we can
+use existing tools to inspect the type of the function we are using
+(Figure~\ref{fig:build}). It is worth reiterating that only the type- and term-
+constructors are translated into OCaml, \lang's precise control over linearity
+and aliasing are not brought over.
+
+We actually use this fact to our advantage to clean up the output code by
+removing what would otherwise be redundant re-bindings in the output OCaml.
+Combined with a code-formatter, the resulting code is not obviously correct and
+exactly what a programmer would intend to write by hand, but now with the
+guarantees and safety of \lang\ behind it.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=\textwidth]{impl_build}
+    \caption{Using \lang\ functions from OCaml.}\label{fig:build}
+\end{figure}
+
 A few examples?

### 2018-10-28 write-up/paper/intro.tex
Start code generation subsubsection

@@
 and recursion into a small, easily understandable, yet expressive set of core
-constructs. \lang\ allows a novice to understand and work with complicated
-linear algebra library APIs; as well as point out subtle aliasing bugs and
-reduce memory usage in existing programs. In addition to this, \lang's
-implementation supports several syntactic conveniences as well as a
+constructs.
+
+\lang\ allows a novice to understand and work with complicated linear algebra
+library APIs, as well as point out subtle aliasing bugs and reduce memory usage
+in existing programs. In fact, we were able to use \lang\ to find linearity and
+aliasing bugs in a linear algebra program that was \emph{generated} by another
+program \emph{specifically designed to translate matrix expressions into an
+efficient sequence of calls linear algebra routines}. We were also able to
+reduce the number of temporaries used by the same program, using \lang's type
+system to guide us.
+
+\lang's implementation supports several syntactic conveniences as well as a
 \emph{usable} integration with real OCaml libraries.
@@
     \item we describe an implementation that is both compatible with and usable
-        from existing code.
+        from existing code
+    \item we show an example of how using \lang\ helped highlight linearity
+        and aliasing bugs, and reduce the memory usage of a \emph{generated}
+        linear algebra program
+    \item we show that using \lang, we can achieve parity with C for linear
+        algebra routines, whilst having much better static guarantees about the
+        linearity and aliasing behaviour of our programs.
 \end{itemize}

### 2018-10-28 write-up/paper/appendix.tex
Add Appendix and more figures

\section{Appendix}

\begin{figure}[t]
\begin{center}
\[\def\arraystretch{1.3}
    \begin{array}{rcl}
    x[e] &
    \Rightarrow &
    \mathbf{get}\ \_\ x\ (e) \;\qquad \textrm{(similarly for matrices)}
\\
    x[e_1] := e_2 &
    \Rightarrow &
    \mathbf{set}\ x\ (e_1)\ (e_2) \quad \textrm{(similarly for matrices)}
\\
\\
    pat & ::= & ()\ \mid\ x\ \mid\ !x\ \mid\ \mathbf{Many\ } pat\ \mid\ (pat, pat)
\\
    \mathbf{let}\ !x = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \specialcell[t]{l}{\mathbf{let\ Many}\ x = e_1\ \mathbf{in\ } \\
    \mathbf{let\ Many}\ x = \mathbf{Many}\ (\mathbf{Many}\ x)\ \mathbf{in}\ e_2}
\\
    \mathbf{let\ Many} \langle pat_x \rangle\ = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \specialcell[t]{l}{%
        \mathbf{let\ Many}\ x = x\ \mathbf{in\ } \\
        \mathbf{let\ }\ \langle pat_x \rangle\ = x\ \mathbf{in\ } e_2}
\\
    \mathbf{let}\ (\langle pat_a \rangle, \langle pat_b \rangle)\  = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \specialcell[t]{l}{%
        \mathbf{let\ }\ (a,b)\ = a\_b\ \mathbf{in\ }
        \ \mathbf{let\ }\ \langle pat_a \rangle\ = a\ \mathbf{in\ } \\
        \mathbf{let\ }\ \langle pat_b \rangle\ = b\ \mathbf{in\ } e_2}
\\
    \mathbf{fun}\ (\langle pat_x \rangle\ : t) \rightarrow e &
    \Rightarrow &
    \mathbf{fun}\ (x : t) \rightarrow \mathbf{let\ }\ \langle pat_x \rangle = x\ \mathbf{in\ } e
\\
\\
    arg & ::= & \langle pat \rangle : t\ \mid\ {'x} \textrm{ (fractional permission variable)}
\\
    \mathbf{fun}\ \langle arg_{1 .. n} \rangle \rightarrow e &
    \Rightarrow &
    \mathbf{fun}\ \langle arg_1 \rangle \rightarrow ..
    \ \mathbf{fun}\ \langle arg_n \rangle \rightarrow e
\\
    \mathbf{let}\ f\ {\langle arg_{1 .. n} \rangle} = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \mathbf{let}\ f = \mathbf{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1\
    \mathbf{in}\ e_2
\\
    \mathbf{let}\ f\ {\langle arg_{1 .. n} \rangle} = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \mathbf{let}\ f = \mathbf{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1\
    \mathbf{in}\ e_2
\\
    \mathbf{let}\ !f\ {\langle arg_{1 .. n} \rangle} = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \mathbf{let\ Many}\ f = \mathbf{Many}\ (\mathbf{fun}\ {\langle arg_{1 .. n} \rangle}
    \rightarrow e_1)\ \mathbf{in}\ e_2
\\
    \mathbf{let\ rec}\ f\ (x : t)\ {\langle arg_{1 .. n} \rangle} : {t'} = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \specialcell[t]{l}{\mathrm{fixpoint} \equiv \mathbf{fix}\ (f, x : t,
        \mathbf{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1 : {t'} ) \\
    \mathbf{let\ Many}\ f = \mathbf{Many}\ \mathrm{fixpoint}\ \mathbf{in}\ e_2}
    \end{array}
\]
\end{center}
\caption{Desugaring from \lang\ concrete syntax to core constructs.}\label{fig:lang_desugar}
\end{figure}

\begin{figure}[p]
    \centering
    \begin{minted}[fontsize=\footnotesize]{ocaml}
let kalman sigma h mu r_1 data_1 =
  let h, _p_k_n_p_ = Prim.size_mat h in
  let k, n = _p_k_n_p_ in
  let sigma_h = Prim.matrix k n in
  let (sigma, h), sigma_h =
    Prim.symm (Many true) (Many 1.) sigma h (Many 0.) sigma_h
  in
  let (sigma_h, h), r_2 =
    Prim.gemm (Many 1.) (sigma_h, Many false) (h, Many true) (Many 1.) r_1
  in
  let (h, mu), data_2 =
    Prim.gemm (Many 1.) (h, Many false) (mu, Many false) (Many (-1.)) data_1
  in
  let h, new_h = Prim.copy_mat_to h sigma_h in
  let r_2, new_r = Prim.copy_mat r_2 in
  let chol_r, sol_h = Prim.posv new_r new_h in
  let chol_r, sol_data = Prim.potrs chol_r data_2 in
  let () = Prim.free_mat chol_r in
  let h_sol_h = Prim.matrix n n in
  let (h, sol_h), h_sol_h =
    Prim.gemm (Many 1.) (h, Many true) (sol_h, Many false) (Many 0.) h_sol_h
  in
  let () = Prim.free_mat sol_h in
  let h_sol_data = Prim.matrix n (Many 1) in
  let (h, sol_data), h_sol_data =
    Prim.gemm (Many 1.) (h, Many true) (sol_data, Many false) (Many 0.) h_sol_data
  in
  let mu, mu_copy = Prim.copy_mat mu in
  let (sigma, h_sol_data), new_mu =
    Prim.symm (Many false) (Many 1.) sigma h_sol_data (Many 1.) mu_copy
  in
  let () = Prim.free_mat h_sol_data in
  let h_sol_h_sigma = Prim.matrix n n in
  let (sigma, h_sol_h), h_sol_h_sigma =
    Prim.symm (Many true) (Many 1.) sigma h_sol_h (Many 0.) h_sol_h_sigma
  in
  let sigma, sigma_copy = Prim.copy_mat_to sigma h_sol_h in
  let (sigma, h_sol_h_sigma), new_sigma =
    Prim.symm (Many false) (Many (-1.)) sigma h_sol_h_sigma (Many 1.) sigma_copy
  in
  let () = Prim.free_mat h_sol_h_sigma in
  ((sigma, (h, (mu, (r_2, sol_data)))), (new_mu, new_sigma)) )
in
kalman
    \end{minted}
    \caption{OCaml code for a Kalman filter, generated (at \emph{compile time})
        from the code in Figure~\ref{fig:lang_kalman}, passed through
        \texttt{ocamlformat} for presentation.}\label{fig:ocaml_kalman}

\end{figure}

\begin{landscape}
\begin{figure}[p]
    \centering
    \begin{minted}[fontsize=\footnotesize]{c}
static void kalman( const int n,               const int k,                const double *sigma, /* n,n */
                    const double *h, /* k,n */ const double *mu, /* n,1 */ double *r,           /* k,k */
                    double *data,    /* k,1 */ double **ret_mu,  /* k,1 */ double **ret_sigma   /* n,n */ ) {
        double* k_by_n = (double *) malloc(k * n * sizeof(double));
/*16*/  cblas_dsymm(CblasRowMajor, CblasRight, CblasUpper, k, n, 1., sigma, n, h, n, 0., k_by_n, n);
/*17*/  cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasTrans, k, k, n, 1., k_by_n, n, h, n, 1., r, k);
/*18*/  cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans, k, 1, n, 1., h, n, mu, 1, -1., data, 1);
/*19*/  cblas_dcopy(k * n, h, 1, k_by_n, 1);
        double* k_by_k = (double *) malloc(k * k * sizeof(double));
/*20*/  cblas_dcopy(k * k, r, 1, k_by_k, 1);
/*21*/  LAPACKE_dposv(LAPACK_ROW_MAJOR, 'U', k, n, k_by_k, k, k_by_n, n);
/*23*/  LAPACKE_dpotrs(LAPACK_ROW_MAJOR, 'U', k, 1, k_by_k, k, data, 1);
        free(k_by_k);
        double* n_by_n = (double *) malloc(n * n * sizeof(double));
/*24*/  cblas_dgemm(CblasRowMajor, CblasTrans, CblasNoTrans, n, n, k, 1., h, n, k_by_n, n, 0., n_by_n, n);
        free(k_by_n);
        double* n_by_1 = (double *) malloc(n * sizeof(double));
/*25*/  cblas_dgemm(CblasRowMajor, CblasTrans, CblasNoTrans, n, 1, k, 1., h, n, data, 1, 0., n_by_1, 1);
        double* new_mu = (double *) malloc(n * sizeof(double));
/*26*/  cblas_dcopy(n, mu, 1, new_mu, 1);
/*27*/  cblas_dsymm(CblasRowMajor, CblasLeft, CblasUpper, n, 1, 1., sigma, n, n_by_1, 1, 1., new_mu, 1);
        free(n_by_1);
        double* n_by_n2 = (double *) malloc(n * n * sizeof(double));
/*28*/  cblas_dsymm(CblasRowMajor, CblasRight, CblasUpper, n, n, 1., sigma, n, n_by_n, n, 0., n_by_n2, n);
/*29*/  cblas_dcopy(n*n, sigma, 1, n_by_n, 1);
/*30*/  cblas_dsymm(CblasRowMajor, CblasLeft, CblasUpper, n, n, -1., sigma, n, n_by_n2, n, 1., n_by_n, n);
        free(n_by_n2);
        *ret_sigma = n_by_n;
        *ret_mu = new_mu; }
    \end{minted}
    \caption{CBLAS/LAPACKE implementation of a Kalman filter. I used C instead
        of Fortran because it is what Owl uses under the hood and OCaml FFI
        support for C is better and easier to use than that for Fortran. A distinct
        `measure\_kalman' function that sandwiches a call to this function with
        \texttt{getrusage} is omitted for brevity.}\label{fig:cblas_kalman}

\end{figure}
\end{landscape}

### 2018-10-28 write-up/paper/implementation.tex
Add Appendix and more figures

@@
     \begin{array}{rcl}
-    x[e] &
-    \Rightarrow &
-    \mathbf{get}\ \_\ x\ (e) \;\qquad \textrm{(similarly for matrices)}
-\\
-    x[e_1] := e_2 &
-    \Rightarrow &
-    \mathbf{set}\ x\ (e_1)\ (e_2) \quad \textrm{(similarly for matrices)}
-\\
-\\
-    pat & ::= & ()\ \mid\ x\ \mid\ !x\ \mid\ \mathbf{Many\ } pat\ \mid\ (pat, pat)
-\\
-    \mathbf{let}\ !x = e_1\ \mathbf{in}\ e_2 &
-    \Rightarrow &
-    \specialcell[t]{l}{\mathbf{let\ Many}\ x = e_1\ \mathbf{in\ } \\
-    \mathbf{let\ Many}\ x = \mathbf{Many}\ (\mathbf{Many}\ x)\ \mathbf{in}\ e_2}
-\\
-    \mathbf{let\ Many} \langle pat_x \rangle\ = e_1\ \mathbf{in}\ e_2 &
-    \Rightarrow &
-    \specialcell[t]{l}{%
-        \mathbf{let\ Many}\ x = x\ \mathbf{in\ } \\
-        \mathbf{let\ }\ \langle pat_x \rangle\ = x\ \mathbf{in\ } e_2}
-\\
-    \mathbf{let}\ (\langle pat_a \rangle, \langle pat_b \rangle)\  = e_1\ \mathbf{in}\ e_2 &
-    \Rightarrow &
-    \specialcell[t]{l}{%
-        \mathbf{let\ }\ (a,b)\ = a\_b\ \mathbf{in\ }
-        \ \mathbf{let\ }\ \langle pat_a \rangle\ = a\ \mathbf{in\ } \\
-        \mathbf{let\ }\ \langle pat_b \rangle\ = b\ \mathbf{in\ } e_2}
-\\
-    \mathbf{fun}\ (\langle pat_x \rangle\ : t) \rightarrow e &
-    \Rightarrow &
-    \mathbf{fun}\ (x : t) \rightarrow \mathbf{let\ }\ \langle pat_x \rangle = x\ \mathbf{in\ } e
-\\
-\\
-    arg & ::= & \langle pat \rangle : t\ \mid\ {'x} \textrm{ (fractional permission variable)}
-\\
-    \mathbf{fun}\ \langle arg_{1 .. n} \rangle \rightarrow e &
-    \Rightarrow &
-    \mathbf{fun}\ \langle arg_1 \rangle \rightarrow ..
-    \ \mathbf{fun}\ \langle arg_n \rangle \rightarrow e
-\\
-    \mathbf{let}\ f\ {\langle arg_{1 .. n} \rangle} = e_1\ \mathbf{in}\ e_2 &
-    \Rightarrow &
-    \mathbf{let}\ f = \mathbf{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1\
-    \mathbf{in}\ e_2
-\\
-    \mathbf{let}\ f\ {\langle arg_{1 .. n} \rangle} = e_1\ \mathbf{in}\ e_2 &
-    \Rightarrow &
-    \mathbf{let}\ f = \mathbf{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1\
-    \mathbf{in}\ e_2
-\\
-    \mathbf{let}\ !f\ {\langle arg_{1 .. n} \rangle} = e_1\ \mathbf{in}\ e_2 &
-    \Rightarrow &
-    \mathbf{let\ Many}\ f = \mathbf{Many}\ (\mathbf{fun}\ {\langle arg_{1 .. n} \rangle}
-    \rightarrow e_1)\ \mathbf{in}\ e_2
-\\
-    \mathbf{let\ rec}\ f\ (x : t)\ {\langle arg_{1 .. n} \rangle} : {t'} = e_1\ \mathbf{in}\ e_2 &
-    \Rightarrow &
-    \specialcell[t]{l}{\mathrm{fixpoint} \equiv \mathbf{fix}\ (f, x : t,
-        \mathbf{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1 : {t'} ) \\
-    \mathbf{let\ Many}\ f = \mathbf{Many}\ \mathrm{fixpoint}\ \mathbf{in}\ e_2}
-    \end{array}
-\]
-\end{center}
-\caption{Desugaring from \lang\ concrete syntax to core constructs.}\label{fig:lang_desugar}
-\end{figure}
-
-\begin{figure}[t]
-\begin{center}
-\[\def\arraystretch{1.3}
-    \begin{array}{rcl}
     \mathbf{let\ } v \leftarrow x[e]\ \mathbf{in\ } e &
@@
     & \mathbf{let\ } Y\ = \mathbf{matrix\ } n\ k\ \mathbf{in}
-    \ \mathbf{let\ } Y \leftarrow [|\ \alpha M_1 M_2 + 0Y\ |]\ \mathbf{in\ } e \\
-\\
+    \ \mathbf{let\ } Y \leftarrow [|\ \alpha M_1 M_2 + 0Y\ |]\ \mathbf{in\ } e
+\\[0.1\baselineskip]
 %syrk
@@
     & \mathbf{let\ } (X,Y)\ = \mathbf{syrk\ } \mathbf{false}
-    \ \alpha\ \_\ X\ \beta\ Y\ \mathbf{in\ } e \\
-\\
+    \ \alpha\ \_\ X\ \beta\ Y\ \mathbf{in\ } e
+\\[0.1\baselineskip]
     \mathbf{let\ } &Y \leftarrow [|\ \alpha X^T X + \beta Y\ |]\ \mathbf{in\ } e
@@
     & \mathbf{let\ } (X,Y)\ = \mathbf{syrk\ } \mathbf{true}
-    \ \alpha\ \_\ X\ \beta\ Y\ \mathbf{in\ } e \\
+    \ \alpha\ \_\ X\ \beta\ Y\ \mathbf{in\ } e
 %symm
-\\
+\\[0.1\baselineskip]
     \mathbf{let\ } &Y \leftarrow\ [|\ \alpha\,\textrm{sym}(X_1)\,X_2 + \beta Y\ |]\ \mathbf{in\ } e
@@
     & \mathbf{let\ } ((X_1,X_2),Y)\ = \mathbf{symm\ } \mathbf{false}
-        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \mathbf{in\ }e \\
-\\
+        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \mathbf{in\ } e
+\\[0.1\baselineskip]
     \mathbf{let\ } &Y \leftarrow\ [|\ \alpha\,X_2\,\textrm{sym}(X_1) + \beta Y\ |]\ \mathbf{in\ } e
@@
     & \mathbf{let\ } ((X_1,X_2),Y)\ = \mathbf{symm\ } \mathbf{false}
-        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \mathbf{in\ } e \\
-\\
+        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \mathbf{in\ } e
+\\[0.1\baselineskip]
 %gemm
@@
 
-We actually use this fact to our advantage to clean up the output code by
-removing what would otherwise be redundant re-bindings in the output OCaml.
-Combined with a code-formatter, the resulting code is not obviously correct and
-exactly what a programmer would intend to write by hand, but now with the
-guarantees and safety of \lang\ behind it.
-
 \begin{figure}[t]
@@
 
-A few examples?
+We actually use this fact to our advantage to clean up the output OCaml by
+removing what would otherwise be redundant re-bindings.  Combined with a
+code-formatter, the resulting code is not obviously correct and exactly what an
+expert would intend to write by hand, but now with the guarantees and safety of
+\lang\ behind it. A small example is shown in Figure~\ref{fig:ocaml_sumarray},
+a larger one in Figure~\ref{fig:ocaml_kalman}.
+
+\begin{figure}[tp]
+    \centering
+    \begin{minted}[fontsize=\small]{ocaml}
+let rec f i n x0 row =
+  if Prim.extract @@ Prim.eqI i n then (row, x0)
+  else
+    let row, x1 = Prim.get row i in
+    f (Prim.addI i (Many 1)) n (Prim.addE x0 x1) row
+in
+f
+    \end{minted}
+    \caption{Recursive OCaml function for a summing over an array, generated (at
+        \emph{compile time}) from the code in Figure~\ref{fig:lang_sumarray},
+        passed through \texttt{ocamlformat} for presentation.}\label{fig:ocaml_sumarray}
+\end{figure}
 
@@
 the straightforward math transcribed into (Matlab/R/Numpy?).
+
+\begin{figure}[t]
+    \centering
+    \input{timings.tex}
+    \caption{Comparison of execution times (error bars are present but quite
+        small). Small matrices and timings $n \le 5^3$ were micro-benchmarked
+        with the Core\_bench library. Larger ones used Unix's
+        \texttt{getrusage} functionality, sandwiched between calls to
+        \highl{Gc.full_major} for the OCaml implementations.}\label{fig:timings}
+\end{figure}

### 2018-10-28 write-up/paper/lang_and_examples.tex
Add Appendix and more figures

@@
         equations this code implements. Line numbers in comments refer to
-        equivalent lines in a C implementation.}\label{fig:lang_kalman}
+        equivalent lines in a C implementation
+        (Figure~\ref{fig:cblas_kalman}).}\label{fig:lang_kalman}
 \end{figure}
@@
 \begin{figure}[t]
-    \begin{align*}
+    % align* uses too much vertical space
+    {\centering
+    $  \displaystyle
+    \begin{aligned}
         \mu' &= \mu + \Sigma H^T (R + H \Sigma H^T)^{-1} (H \mu - \textrm{data})\\
         \Sigma' &= \Sigma ( I - H^T (R + H \Sigma H^T)^{-1} H \Sigma )
-    \end{align*}
+    \end{aligned}
+    $ \par}
     \caption{Kalman filter equations (credit:
@@
 
-With expressions as complex as those in Figure~\ref{fig:kalman_eqns}, it can
-get very difficult to ensure memory is not being re-used incorrectly/enough
-whilst also enusring the aliasing assumptions for the BLAS/LAPACK routine calls
-are guaranteed. However, using \lang\ makes incorrect aliasing and memory
-re-use impossible. Additionally, \lang\ makes it possible to spot where memory
-is not being re-used enough \emph{precisely because} of its explicit
-allocations, aliases and frees.
+The \lang\ implementation is much longer than the mathematical equations for
+two reasons. First, the \lang\ implementation is a let-normalised form of the
+Kalman equations: since there a large number of unary/binary (and occasionally
+ternary) sub-expressions in the equations, naming each one line at a time makes
+the implementation much longer.  Second, \lang\ has the additional task of
+handling explicit allocations, aliasing and frees of matrices. However, it is
+exactly this which makes it possible (and often, easy) to spot additional
+opportunities for memory re-use. Furthermore, a programmer can explore those
+opportunities easily because \lang's type system statically enforces correct
+memory management and the aliasing assumptions of BLAS/LAPACK routines.

### 2018-10-28 write-up/paper/implementation.tex
Add section on OCaml 'optimisation'

@@
 We actually use this fact to our advantage to clean up the output OCaml by
-removing what would otherwise be redundant re-bindings.  Combined with a
-code-formatter, the resulting code is not obviously correct and exactly what an
-expert would intend to write by hand, but now with the guarantees and safety of
-\lang\ behind it. A small example is shown in Figure~\ref{fig:ocaml_sumarray},
-a larger one in Figure~\ref{fig:ocaml_kalman}.
+removing what would otherwise be redundant re-bindings
+(Figure~\ref{fig:lang_optimise}).  Combined with a code-formatter, the
+resulting code is not obviously correct and exactly what an expert would intend
+to write by hand, but now with the guarantees and safety of \lang\ behind it. A
+small example is shown in Figure~\ref{fig:ocaml_sumarray}, a larger one in
+Figure~\ref{fig:ocaml_kalman}.
+
+\begin{figure}[t]
+\begin{center}
+    \begin{tabular}{rcl}
+        \alsocell[b]{l}{\highl{let Many x = x in} \\
+        \highl{let Many x = Many (Many x) in <exp>} } &
+        $\Rightarrow$ &
+        \highl{<exp>}
+    \\ \\
+        \alsocell[c]{l}{\highl{let Many x = <exp> in} \\
+            \highl{let Many x = Many (Many x) in <body>} \\
+            \textbf{OR} \highl{let Many x = Many <exp> in <body>} \\
+            \textbf{OR} \highl{(fun x : t -> <body>) <exp>}} &
+        $\Rightarrow$ &
+        \highl{let x = <exp> in <body>}
+    \\ \\
+        \alsocell[c]{l}{\highl{(* fixp = fix (f, x:t, <exp> : t') *)} \\
+            \highl{let Many f = Many fixp in <body>} \\
+            \textbf{OR} \highl{let f = fixp in <body>} } &
+        $\Rightarrow$ &
+        \highl{let rec f x = <exp> in <body>}
+    \end{tabular}
+\end{center}
+\caption{Removing redundant re-bindings during tranlation to OCaml.}\label{fig:lang_optimise}
+\end{figure}
 

### 2018-10-29 write-up/paper/intro.tex
Add frac. perm. sub. lemma

@@
 program \emph{specifically designed to translate matrix expressions into an
-efficient sequence of calls linear algebra routines}. We were also able to
+efficient sequence of calls to linear algebra routines}. We were also able to
 reduce the number of temporaries used by the same program, using \lang's type

### 2018-10-29 write-up/semantics/lemmas.tex
Add points about finding bugs
Finish desugaring and matrix expression figures
Add frac. perm. sub. lemma

@@
 
-$\theta$ is irrelevant because it only maps fractional capability variables to
-fractional capabilities. By construction, $\delta$ and $\gamma$ only map
+$\theta$ is irrelevant because it only maps fractional permission variables to
+fractional permissions. By construction, $\delta$ and $\gamma$ only map
 variables to values, and values are closed under substitution.
@@
 $}\label{stepInC}
+
+(Here, $\theta$ is the empty substitution for $\varsigma = \den{H}{}{\sigma}\theta$.)
 
@@
     \delta \in \den{I}{k}{\Delta}\theta \wedge \gamma \in \pi_2[\den{L}{k}{\Gamma}\theta]
-    \Rightarrow \dom(\Delta) = \dom(\delta)$ and $\dom(\Gamma) = \dom(\gamma)$}\label{samedom}
+    \Rightarrow \\ \dom(\Delta) = \dom(\delta)$ and $\dom(\Gamma) = \dom(\gamma)$}\label{samedom}
 
@@
 \pf~By induction on $\Gamma'$.
+
+\subsection{$\forall e, \sigma, e', \sigma', \theta.\
+    \langle e, \sigma \rangle \rightarrow \langle e' , \sigma' \rangle
+    \Rightarrow \langle \theta(e), \theta(\sigma)
+    \rangle \rightarrow \langle \theta(e') , \theta(\sigma') \rangle$}
+
+\pf~By induction on $\rightarrow$.
+\begin{proof}
+
+    \step{}{\assume{Arbitrary $e, \sigma, e', \sigma', \theta$ such that 
+        $\langle e, \sigma \rangle \rightarrow \langle e' , \sigma' \rangle$.}}
+
+    \step{}{\suffices{To consider only the following rules which mention
+        fractional permission variables.}
+        \textsc{Op\_Frac\_Perm}, \textsc{Op\_Share}, \textsc{Op\_Unshare\_(N)Eq}
+        and \textsc{Op\_Gemm\_(Mis)Match}.}
+
+    \step{}{\case{\textsc{Op\_Frac\_Perm.}}{Because substitution avoids capture,
+        \\ $\langle \theta((\ottkw{fun} fc \rightarrow v) \, [f]),
+        \theta(\sigma) \rangle \rightarrow \langle \theta(v\, [fc/f]),
+        \theta(\sigma'\, [fc/f]) \rangle$.}}
+
+    \step{}{The rest of the cases are parametric in their use of fractional
+        permission variables and so will take the same step ater any substitution.}
+
+    \step{}{\textsc{Corollary:}
+        If $\langle e\, [fc/f_1], \sigma \, [fc/f_1] \rangle \rightarrow^n
+        \langle v_2, \sigma_2 \rangle$ and
+        $\langle e\, [fc/f_2], \sigma \, [fc/f_2] \rangle \rightarrow^n
+        \langle v_2, \sigma_2 \rangle$, then
+        $\exists\,v, \sigma.\ v_1 = v\,[fc/f_1] \wedge v_2 = v\,[fc/f_2]
+        \sigma_1 = \sigma\,[fc/f_1] \wedge \sigma_2 = \sigma\,[fc/f_2]$.}
+
+\end{proof}

### 2018-10-29 write-up/semantics/lemmas.tex
Use frac. perm. sub. lemma
Thread through theta-substitution in heaps

--- 2018-10-23 write-up/semantics/lemmas.tex
@@
 $}\label{stepInC}
+
+In the lemma, and for the rest of its proof, $\varsigma = \den{H}{}{\sigma}$.
 
@@
 
+For the rest of this proof, $\varsigma = \den{H}{}{\sigma}$.\\
 Lemma~\ref{stepInC} is the inductive step for this lemma for the $\C{}{}$ case.\\
@@
 \pf~By induction on $\Gamma'$.
+
+\subsection{$\forall e, \sigma, e', \sigma', \theta.\
+    \langle \sigma, e \rangle \rightarrow \langle \sigma',  e' \rangle
+    \Rightarrow \langle \theta(\sigma), \theta(e) \rangle \rightarrow
+    \langle \theta(\sigma') , \theta(e') \rangle$}\label{fracPermSub}
+
+\pf~By induction on $\rightarrow$.
+\begin{proof}
+
+    \step{}{\assume{Arbitrary $e, \sigma, e', \sigma', \theta$ such that 
+        $\langle \sigma, e \rangle \rightarrow \langle \sigma', e' \rangle$.}}
+
+    \step{}{\suffices{To consider only the following rules which mention
+        fractional permission variables.}
+        \textsc{Op\_Frac\_Perm}, \textsc{Op\_Share}, \textsc{Op\_Unshare\_(N)Eq}
+        and \textsc{Op\_Gemm\_(Mis)Match}.}
+
+    \step{}{\case{\textsc{Op\_Frac\_Perm.}}{Because substitution avoids capture,
+        \\ $\langle \theta(\sigma), \theta((\ottkw{fun} fc \rightarrow v) \, [f])
+         \rangle \rightarrow \langle \theta(\sigma'\, [fc/f]),
+        \theta(v\, [fc/f]) \rangle$.}}
+
+    \step{}{The rest of the cases are parametric in their use of fractional
+        permission variables and so will take the same step ater any substitution.}
+
+    \step{}{\textsc{Corollary:}
+        If $\langle \sigma \, [fc/f_1], e\, [fc/f_1] \rangle \rightarrow^n
+        \langle \sigma_2, e'_2 \rangle$ and
+        $\langle \sigma\, [fc/f_2], e \, [fc/f_2] \rangle \rightarrow^n
+        \langle \sigma_2, e'_2 \rangle$, then
+        $\exists\,\sigma, e'.\ \sigma_1 = \sigma\,[fc/f_1] \wedge
+        \sigma_2 = \sigma\,[fc/f_2] \wedge
+        e'_1 = e'\,[fc/f_1] \wedge e'_2 = e'\,[fc/f_2]$.}
+
+\end{proof}

### 2018-10-30 write-up/paper/discussion.tex
Add frac. perm. sub. lemma
Start Discussion

@@
 \section{Discussion and Related Work}\label{sec:discussion_related_work}
+
+\subsection{Finding Bugs in SymPy's Output}
+
+Prior to this project, we had little experience with linear algebra libraries
+or the problem of matrix expression compilation. As such, we based our initial
+\lang\ implementation of a Kalman filter using BLAS and LAPACK, on a popular
+GitHub gist of a Fortran implementation, one that was \emph{automatically
+generated} from SymPy's matrix expression compiler~\cite{rocklin_thesis}.
+
+Once we translated the implementation from Fortran to \lang, we attempted to
+compile it and found that (to our surprise) it did not type-check. This was
+because the original implementation contained incorrect aliasing, unused and
+unnecessary temporaries, and did not adhere to Fortran's read/write permissions
+(with respect to \texttt{intent} annotations \texttt{in}, \texttt{out} and
+\texttt{inout}) all of which were now highlighted by \lang's type system.
+
+The original implementation used 6 temporaries, one of which was immediately
+spotted as never being used due to linearity. It also contained two variables
+which were marked as \texttt{intent(in)} but would have been written over by
+calls to `gemm', spotted by the fractional-capabilities feature. Furthermore,
+it used a matrix \emph{twice} in a call to `symm', once with a read permission
+but once with a \emph{write} permission.  Fortran assumes that any parameter
+being written to is not aliased and so this call was not only incorrect, but
+illegal according to the standard, both aspects of which were captured by
+linearity and fractional-capabilities.
+
+Lastly, it contained another unnecessary temporary, however one that was not
+obvious without linear types. To spot it, we first performed live-range
+splitting (checked by linearity) by hoisting calls to \highl{freeM} and then
+annotated the freed matrices with their dimensions.  After doing so and
+spotting two disjoint live-ranges of the same size, we replaced a call to
+\highl{freeM} followed by allocating call to \highl{copy} with one, in-place
+call to \highl{copyM_to}. We believe the ability to boldly refactor code which
+manages memory is good evidence of the usefulness of linearity as a tool for
+programming.
+
+\subsection{Related Work}
 
@@
 \item Boyland fractional permissions
+\item Disadvantage of graph-based approaches to matrix expression generation
+\item multi-stage: once dimensions known, they are fixed?
 \end{itemize}
+
+\subsection{Simplicity and Further Work}
+
+We are pleasantly surprised at how simple the overall design and implementation
+of \lang\ is, given its expressive power and usability. Indeed, the focus on
+getting a working prototype early on (so that we could test it with real
+BLAS/LAPACK routines as soon as possible) meant that we only added features to
+the type system when it was clear that they were absolutely necessary: these
+features were !-types and value-restriction for the \highl{Many} constructor.
+Although we considered parametric polymorphism, size-types and tracking pointer
+identities at the type-level\cite{morrisett} (all of which would certainly be
+useful and are potential avenues for further work) it turns out these features
+are not necessary.
+
+We believe we have stumbled across a particularly nice problem-solution fit:
+the idea of linear types for linear algebra -- one that used to just be
+commonly believed folklore -- actually has some merit and is worth exploring.
+We also think that this concept (and the general design of its implementation)
+need not be limited to linear algebra: we could conceivably `backport' this
+idea to other contexts that need linearity (concurrency, single-use
+continuations, zero-copy buffer, streaming I/O) or combine it with dependent
+types to achieve even more expressive power to split up a single block of
+memory into multiple regions in an arbitrary manner\cite{space_monads}.

### 2018-10-31 write-up/paper/formal_system.tex
Start on Formal System

@@
 
-% TODO value restriction
+\subsection{Core Type Theory}
 
-\subsection{The Core Type Theory and Dynamic Semantics}
+The full typing rules are in Appendix \label{subsec:static_sem}, but the key
+ideas are as follow: \textsf{Perm}, \textsf{Type}, $\Theta, \Delta, \Gamma$,
+value restriction, lambda, Gen/Spc, and Fix.
 
-Describe the typing rules and operational semantics here
+\subsection{Dynamic Semantics}\label{subsec:semantics}
+
+and operational semantics here
 

### 2018-10-31 write-up/paper/implementation.tex
Add frac. perm. sub. lemma
Start on Formal System

@@
 expression are translated into BLAS/ LAPACK calls via purely syntactic
-pattern-matching, outlined in Figure~\ref{fig:lang_matexp}. Type checking is
-mostly standard for a linearly typed language, with the exception of fractional
-permission inference.  Because all functions must have their argument types
-explicitly annotated, inferring the correct fraction at a call-site is simply a
-matter of unification.
+pattern-matching, outlined in Figure~\ref{fig:lang_matexp}.
+
+\subsubsection{Type checking} is mostly standard for a linearly typed language,
+with the exception of fractional permission inference. Because all functions
+must have their argument types explicitly annotated, inferring the correct
+fraction at a call-site is simply a matter of unification.
+
+There are a few differences between the type system as presented in
+\ref{subsec:semantics} and how we implemented it: the environment
+\emph{changes} as a result of type checking an expression (the standard
+transformation to avoid a non-deterministic split of the environment for
+checking pairs); variables are \emph{marked as used} rather than removed for
+better error messages; variables are \emph{tagged} as linear or intuitionistic
+in \emph{one} environment as opposed to being stored in \emph{two} separate
+ones (this allows scoping/variable look-up to be handled uniformly).
 
@@
     \Rightarrow \\
-    & \mathbf{let\ } ((X_1,X_2),Y)\ = \mathbf{symm\ } \mathbf{false}
+    & \mathbf{let\ } ((X_1,X_2),Y)\ = \mathbf{symm\ } \mathbf{true}
         \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \mathbf{in\ } e

### 2018-11-01 write-up/paper/formal_system.tex
Add interpretation and proof

@@
 
-The full typing rules are in Appendix \label{subsec:static_sem}, but the key
-ideas are as follow: \textsf{Perm}, \textsf{Type}, $\Theta, \Delta, \Gamma$,
+The full typing rules are in Appendix \ref{subsec:static_sem}, but the key
+ideas are: \textsf{Perm}, \textsf{Type}, $\Theta, \Delta, \Gamma$,
 value restriction, lambda, Gen/Spc, and Fix.
@@
 
-\subsection{The logical relation}
+\subsection{Logical Relation}
 

### 2018-11-01 write-up/paper/paper.tex
Sketch out Implementation section
Add points about finding bugs
Finish desugaring and matrix expression figures
Add Appendix and more figures
Add section on OCaml 'optimisation'
Add frac. perm. sub. lemma
Add Ott Semantics
Add interpretation and proof

@@
-% This is samplepaper.tex, a sample chapter demonstrating the
-% LLNCS macro package for Springer Computer Science proceedings;
-% Version 2.20 of 2017/10/04
-%
-\documentclass[runningheads]{llncs}
-%
-\usepackage{graphicx}
-% Used for displaying a sample figure. If possible, figure files should
-% be included in EPS format.
-%
-% If you use the hyperref package, please uncomment the following line
-% to display URLs in blue roman font according to Springer's eBook style:
+\documentclass[runningheads,a4paper,twoside]{llncs}
 \usepackage{hyperref}
@@
     linkcolor={black},
-    citecolor={green}
+    citecolor={black},
+    urlcolor={blue}
 }
-\renewcommand\UrlFont{\color{blue}\rmfamily}
 
@@
 \RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
-
+% Convenient inline syntax highlighting
 \newmintinline[highl]{ocaml}{breaklines}
 
-\usepackage{amsmath}
+\usepackage{amsmath,amssymb}
+
+% Ott Rules
+\usepackage{supertabular}%
+\usepackage[margin=1in]{geometry}%
+\usepackage{ifthen}%
+\usepackage{alltt}%hack%
+
+\input{../semantics/semantics_def}%
+\usepackage{ottlayout}%
+\renewcommand{\ottpremise}[1]{\premiseSTY{#1}}%
+\renewcommand{\ottusedrule}[1]{\usedruleSTY{#1}}%
+\renewcommand{\ottdrule}[4][]{\druleSTY[#1]{#2}{#3}{#4}}%
+\renewenvironment{ottdefnblock}[3][]{\defnblockSTY[#1]{#2}{#3}}{\enddefnblockSTY}%
+
+% Proof macros
+\newcommand{\den}[3]{ \mathcal{#1}_{#2} [\![ #3 ]\!] }%
+\newcommand{\V}[2]{ \den{V}{#1}{#2} }%
+\newcommand{\C}[2]{ \den{C}{#1}{#2} }%
+
+\newcommand{\Unit}{\ottkw{unit}}%
+\newcommand{\Bang}{\ottkw{!}}
+\newcommand{\Bool}{\ottkw{bool}}%
+\newcommand{\Int}{\ottkw{int}}%
+\newcommand{\Elt}{\ottkw{elt}}%
+\newcommand{\Mat}{\ottkw{mat}}%
+\newcommand{\Zf}{\ottkw{Z}}%
+\newcommand{\Sf}{\ottkw{S}}%
+\newcommand{\Many}{\ottkw{Many}}%
+\newcommand{\dom}{\mathrm{dom}}%
+\newcommand{\empH}{\emptyset}%
+
+\usepackage{pf2}
+\beforePfSpace{15pt, 10pt, 10pt, 10pt, 5pt, 2pt}
+\afterPfSpace{15pt, 10pt, 10pt, 10pt, 5pt, 2pt}
+\interStepSpace{15pt, 10pt, 10pt, 10pt, 5pt, 2pt}
+\pflongindent%
+
+% Multi-line table cells
+% tex.stackexchange.com/questions/2441/how-to-add-a-forced-line-break-inside-a-table-cell#19678
+\newcommand{\specialcell}[3][c]{%
+  \begin{array}[#1]{@{}#2@{}}#3\end{array}}
+
+\newcommand{\alsocell}[3][c]{%
+  \begin{tabular}[#1]{@{}#2@{}}#3\end{tabular}}
+
+% Figures
+\usepackage{graphicx}
+\usepackage[dvipsnames]{xcolor}
+\usepackage{lscape}
+\usepackage{pgfplots}
+
+% PL Stuff
+\usepackage[nounderscore]{syntax}
+\renewcommand{\syntleft}{\normalfont\itshape}
+\renewcommand{\syntright}{}
+\renewcommand{\ulitleft}{\normalfont\bf}%\syn@ttspace\frenchspacing}
+\renewcommand{\ulitright}{}
+\renewcommand{\litleft}{\bgroup\ulitleft}
+\renewcommand{\litright}{\ulitright\egroup}
 
@@
 
-\title{NumLin: Linear Types for Numerical Libraries}
+\title{NumLin: Linear Types for Linear Algebra}
 
-% \titlerunning{NumLin: Linear Types for Numerical Libraries}
+% \titlerunning{NumLin: Linear Types for Linear Algebra}
 % If the paper title is too long for the running head, you can set
@@
 %
-\maketitle              % typeset the header of the contribution
+\maketitle
 %
@@
 Briefly summarize the contents of the paper in 150--250 words.
-
+%
 \keywords{numerical, linear, algebra, types, permissions, OCaml}
 \end{abstract}
-%
-%
-%
+
 \input{intro}
@@
 
+\appendix
+\input{appendix}
+
 \end{document}

### 2018-11-04 bin/kalman.h
Improve LT4LA and CBLAS implementations of Kalman

@@ -14,7 +14,7 @@
 
 double *new_sigma;
 
-double sigma[N][N] = {
+double const sigma[N][N] = {
     { 1.682490, 0.621964, 0.959947, 1.228820, 1.029410, } ,
     { 0.621964, 0.631446, 0.551902, 0.723342, 0.756674, } ,
     { 0.959947, 0.551902, 1.100060, 0.908402, 1.032840, } ,
@@ -22,13 +22,11 @@ double sigma[N][N] = {
     { 1.029410, 0.756674, 1.032840, 1.011350, 1.302410, } ,
 };
 
-double h[K][N] = {
+double const h[K][N] = {
     { 0.4621110, 0.833041, 0.0395867, 0.529315, 0.241678, },
     { 0.0507828, 0.340120, 0.8726660, 0.836114, 0.571528, },
     { 0.7779080, 0.541655, 0.8691540, 0.286846, 0.265820, },
 };
-
-double* new_mu;
 
 double mu[N][1] = {
     { 0.8015420 },
@@ -55,38 +53,25 @@ static void kalman(
     const int k,
     const double *sigma, /* n,n */
     const double *h,     /* k,n */
-    const double *mu,    /* n,1 */
+    double *mu,          /* n,1 */
     double *r,           /* k,k */
     double *data,        /* k,1 */
-    double **ret_mu,     /* k,1 */
     double **ret_sigma   /* n,n */
 ) {
-        double* k_by_n = (double *) malloc(k * n * sizeof(double));
-/*20*/  cblas_dsymm(CblasRowMajor, CblasRight, CblasUpper, k, n, 1., sigma, n, h, n, 0., k_by_n, n);
-/*21*/  cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasTrans, k, k, n, 1., k_by_n, n, h, n, 1., r, k);
-/*22*/  cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans, k, 1, n, 1., h, n, mu, 1, -1., data, 1);
-/*23*/  cblas_dcopy(k * n, h, 1, k_by_n, 1);
-        double* k_by_k = (double *) malloc(k * k * sizeof(double));
-/*24*/  cblas_dcopy(k * k, r, 1, k_by_k, 1);
[…]

### 2018-11-05 write-up/paper/appendix.tex
Add section on OCaml 'optimisation'
Add frac. perm. sub. lemma
Start on Formal System
Add Ott Semantics
Add interpretation and proof
Start on Soundness Theorem

@@
-\section{Appendix}
+\section{\lang\ Specification}
 
-\begin{figure}[t]
+\ottstyledefaults{premiselayout=justify}%
+\subsection{Static Semantics}\label{subsec:static_sem}
+\ottdefnsTypes%
+
+\subsection{Dynamic Semantics}\label{subsec:dyn_sem}
+\ottdefnsOpXXSem%
+
+\input{../semantics/interpretation}
+
+\input{../semantics/lemmas}
+
+\input{../semantics/soundness}
+
+\subsection{Well-formed types}
+\ottdefnsWellXXFormed%
+
+\section{\lang\ Grammar}\label{sec:grammar_def}
+\ottgrammar%
+
+\section{Desugaring \lang}\label{sec:lang_desugar}
+
 \begin{center}
@@
         \mathbf{let\ Many}\ x = x\ \mathbf{in\ } \\
-        \mathbf{let\ }\ \langle pat_x \rangle\ = x\ \mathbf{in\ } e_2}
+        \mathbf{let\ } \langle pat_x \rangle\ = x\ \mathbf{in\ } e_2}
 \\
@@
     \specialcell[t]{l}{%
-        \mathbf{let\ }\ (a,b)\ = a\_b\ \mathbf{in\ }
-        \ \mathbf{let\ }\ \langle pat_a \rangle\ = a\ \mathbf{in\ } \\
-        \mathbf{let\ }\ \langle pat_b \rangle\ = b\ \mathbf{in\ } e_2}
+        \mathbf{let\ } (a,b)\ = a\_b\ \mathbf{in\ }
+        \ \mathbf{let\ } \langle pat_a \rangle\ = a\ \mathbf{in\ } \\
+        \mathbf{let\ } \langle pat_b \rangle\ = b\ \mathbf{in\ } e_2}
 \\
-    \mathbf{fun}\ (\langle pat_x \rangle\ : t) \rightarrow e &
+    \mathbf{fun}\ (\langle pat_x \rangle : t) \rightarrow e &
     \Rightarrow &
-    \mathbf{fun}\ (x : t) \rightarrow \mathbf{let\ }\ \langle pat_x \rangle = x\ \mathbf{in\ } e
+    \mathbf{fun}\ (x : t) \rightarrow \mathbf{let\ } \langle pat_x \rangle = x\ \mathbf{in\ } e
 \\
@@
 \\
-    \mathbf{let}\ f\ {\langle arg_{1 .. n} \rangle} = e_1\ \mathbf{in}\ e_2 &
-    \Rightarrow &
-    \mathbf{let}\ f = \mathbf{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1\
-    \mathbf{in}\ e_2
-\\
     \mathbf{let}\ !f\ {\langle arg_{1 .. n} \rangle} = e_1\ \mathbf{in}\ e_2 &
@@
 \\
+    \mathrm{fixpoint} & \equiv & \mathbf{fix}\ (f, x : t, \mathbf{fun}
+    \ {\langle arg_{1 .. n} \rangle} \rightarrow e_1 : {t'} )
+\\
     \mathbf{let\ rec}\ f\ (x : t)\ {\langle arg_{1 .. n} \rangle} : {t'} = e_1\ \mathbf{in}\ e_2 &
     \Rightarrow &
-    \specialcell[t]{l}{\mathrm{fixpoint} \equiv \mathbf{fix}\ (f, x : t,
-        \mathbf{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1 : {t'} ) \\
-    \mathbf{let\ Many}\ f = \mathbf{Many}\ \mathrm{fixpoint}\ \mathbf{in}\ e_2}
+    \mathbf{let}\ f = \mathrm{fixpoint}\ \mathbf{in}\ e_2
+\\
+    \mathbf{let\ rec}\ !f\ (x : t)\ {\langle arg_{1 .. n} \rangle} : {t'} = e_1\ \mathbf{in}\ e_2 &
+    \Rightarrow &
+    \mathbf{let\ Many}\ f = \mathbf{Many}\ \mathrm{fixpoint}\ \mathbf{in}\ e_2
     \end{array}
@@
 \end{center}
-\caption{Desugaring from \lang\ concrete syntax to core constructs.}\label{fig:lang_desugar}
-\end{figure}
 
-\begin{figure}[p]
-    \centering
+\clearpage
+\section{Primitives}\label{subsec:primitives}
+
+\vspace*{\fill}
+\begin{center}
+\begin{minted}[fontsize=\small]{ocaml}
+module Arr = Owl.Dense.Ndarray.D
+type z = Z
+type 'a s = Succ
+type 'a arr = A of Arr.arr [@@unboxed]
+type 'a mat = M of Arr.arr [@@unboxed]
+type 'a bang = Many of 'a [@@unboxed]
+module Prim :
+sig
+  val extract : 'a bang -> 'a
+  (** Boolean *)
+  val not_ : bool bang -> bool bang
+  (** Arithmetic, many omitted for brevity *)
+  val addI : int bang -> int bang -> int bang
+  val eqI : int bang -> int bang -> bool bang
+  (** Arrays *)
+  val set : z arr -> int bang -> float bang -> z arr
+  val get : 'a arr -> int bang -> 'a arr * float bang
+  val share : 'a arr -> 'a s arr * 'a s arr
+  val unshare : 'a s arr -> 'a s arr -> 'a arr
+  val free : z arr -> unit
+  (** Owl *)
+  val array : int bang -> z arr
+  val copy : 'a arr -> 'a arr * z arr
+  val sin : z arr -> z arr
+  val hypot : z arr -> 'a arr -> 'a arr * z arr
+  (** Level 1 BLAS *)
+  val asum : 'a arr -> 'a arr * float bang
+  val axpy : float bang -> 'a arr -> z arr -> 'a arr * z arr
+  val dot : 'a arr -> 'b arr -> ('a arr * 'b arr) * float bang
+  val scal : float bang -> z arr -> z arr
+  val amax : 'a arr -> 'a arr * int bang
+  (* Matrix, some omitted for brevity *)
+  val matrix : int bang -> int bang -> z mat
+  val eye : int bang -> z mat
+  val copy_mat : 'a mat -> 'a mat * z mat
+  val copy_mat_to : 'a mat -> z mat -> 'a mat * z mat
+  val size_mat : 'a mat -> 'a mat * (int bang * int bang)
+  val transpose : 'a mat -> 'a mat * z mat
+  (* Level 3 BLAS/LAPACK *)
+  val gemm : float bang -> ('a mat * bool bang) -> ('b mat * bool bang) ->
+             float bang -> z mat -> ('a mat * 'b mat) * z mat
+  val symm : bool bang -> float bang -> 'a mat -> 'b mat -> float bang ->
+             z mat -> ('a mat * 'b mat) * z mat
+  val gesv : z mat -> z mat -> z mat * z mat
+  val posv : z mat -> z mat -> z mat * z mat
+  val potrs : 'a mat -> z mat -> 'a mat * z mat
+  val syrk : bool bang -> float bang -> 'a mat -> float bang -> z mat ->
+           'a mat * z mat
+end
+\end{minted}
+\end{center}
+\vfill
+
+\clearpage
+\section{Kalman Filters from \lang\ and C}
+
+\vspace*{\fill}
+\begin{figure}[h]
+\begin{center}
     \begin{minted}[fontsize=\footnotesize]{ocaml}
@@
   let k, n = _p_k_n_p_ in
-  let sigma_h = Prim.matrix k n in
-  let (sigma, h), sigma_h =
-    Prim.symm (Many true) (Many 1.) sigma h (Many 0.) sigma_h
+  let h_sigma = Prim.matrix k n in
+  let (sigma, h), h_sigma =
+    Prim.symm (Many true) (Many 1.) sigma h (Many 0.) h_sigma
   in
-  let (sigma_h, h), r_2 =
-    Prim.gemm (Many 1.) (sigma_h, Many false) (h, Many true) (Many 1.) r_1
+  let (h_sigma, h), r_2 =
+    Prim.gemm (Many 1.) (h_sigma, Many false) (h, Many true) (Many 1.) r_1
   in
@@
   in
-  let h, new_h = Prim.copy_mat_to h sigma_h in
+  let h, new_h = Prim.copy_mat_to h h_sigma in
   let r_2, new_r = Prim.copy_mat r_2 in
@@
         \texttt{ocamlformat} for presentation.}\label{fig:ocaml_kalman}
-
+\end{center}
 \end{figure}
+\vfill
 

### 2018-11-05 write-up/paper/discussion.tex
Start on Soundness Theorem

@@
 compile it and found that (to our surprise) it did not type-check. This was
-because the original implementation contained incorrect aliasing, unused and
-unnecessary temporaries, and did not adhere to Fortran's read/write permissions
-(with respect to \texttt{intent} annotations \texttt{in}, \texttt{out} and
-\texttt{inout}) all of which were now highlighted by \lang's type system.
+because the original implementation contained incorrect aliasing, unused
+variables and unnecessary temporaries, and did not adhere to Fortran's
+read/write permissions (with respect to \texttt{intent} annotations
+\texttt{in}, \texttt{out} and \texttt{inout}) all of which were now highlighted
+by \lang's type system.
 

### 2018-11-05 write-up/paper/formal_system.tex
Start on Soundness Theorem

@@
 The full typing rules are in Appendix \ref{subsec:static_sem}, but the key
-ideas are: \textsf{Perm}, \textsf{Type}, $\Theta, \Delta, \Gamma$,
-value restriction, lambda, Gen/Spc, and Fix.
+ideas are as follow.
+
+A typing judgement consists of $ \Theta; \Delta; \Gamma
+\vdash e : t$.
+
+$\Theta$ is the environment that tracks which fractional permission variables
+in scope. Fractional permissions (the \textsf{Perm} judgement) and types (the
+\textsf{Type} judgement) are \emph{well-formed} if all of their free fractional
+variables are in $\Theta$.
+
+$\Delta$ is the environment storing non-linearly or \emph{inuitionistically}
+typed variables.
+
+$\Gamma$ is the environment storing linearly typed variables.  Note that rules
+for typing $()$, booleans, integers and elements are typed with respect to an
+\emph{empty} linear environment: this means no linear values are needed to
+produce a value of those types.
+
+\[
+    \ottdruleTyXXUnitXXIntro{}
+\]
+
+Conversely, whenever two or more subexpressions need to be typed, they must
+consume a disjoint set of linear values (pairs, let-expressions).  In the case
+of if-expressions, both branches must consume the same set of linear values
+(disjoint to the ones used to evaluate the condition).
+
+\[
+    \ottdruleTyXXBoolXXElim{}
+\]
+
+The \highl{Many} introduction and elimination rules are very important.
+Producing !-type values may only be done if the expression inside is a
+syntactic value which is not a location. This allows all safely duplicable
+resources, including functions which capture non-linear resources from their
+environments, but prevents producing aliases of (pointers to) arrays and
+matrices. This is exactly the same as value-restriction from the world of
+parametric polymorphism.
+
+\[
+    \ottdruleTyXXBangXXIntro{}
+\]
+
+Consuming a !-type value \emph{moves it} from the linear environment $\Gamma$
+and \emph{into} the intuitionistic environment $\Delta$. This is exactly why
+$\mathbf{let}\ !x = e_1\ \mathbf{in}\ e_2$ desugars to $\mathbf{let\ Many}\ x =
+e_1\ \mathbf{in\ } \mathbf{let\ Many}\ x = \mathbf{Many}\ (\mathbf{Many}\ x)\
+\mathbf{in}\ e_2$.
+
+\[
+    \ottdruleTyXXBangXXElim{}
+\]
+
+Rules \textsc{Ty\_Gen} and \textsc{Ty\_Spc} are for fractional permission
+generalisation and specialisation respectively. They allow the definition and
+use of functions that are polymorphic in the fractional permission index of
+their results and one or more of their arguments.
+
+\[
+    \ottdruleTyXXGen{} \qquad\qquad \ottdruleTyXXSpc{}
+\]
+
+Rule \textsc{Ty\_Fix} shows how recursive functions are typed. Even though
+recursive functions are fully annotated, type checking them is interesting for
+two reasons: to type check the body of the fixpoint, the type of the recusive
+function is in the \emph{intuitionistic} environment $\Delta$ (without this,
+you would not be able to write a base case) whilst the argument and its type
+are the \emph{only things in the linear environment} $\Gamma$. The latter means
+that recursive functions can be type checked in an empty environment (thus be
+wrapped in \highl{Many} and used zero or multiple times).
+
+\[
+    \ottdruleTyXXFix{}
+\]
+
+Lastly, types of almost all \lang\ primitives, as embedded in OCaml's type
+system, are shown in Appendix \ref{subsec:primitives}, with some similar ones
+(like those for binary arithmetic operators) omitted for brevity. The
+main difference between the OCaml type of a primitive like \highl{gemm} and its
+\lang\ counterpart is the inclusion of explicit `$\forall$'s.  So,
+\highl{float bang -> ('a mat * bool bang) -> ('b mat * bool bang) -> float
+bang -> z mat -> ('a mat * 'b mat) * z mat}
+will correspond to \\
+$!\ottkw{elt} \multimap \forall\, x.\ x \ \ottkw{mat} \ \otimes \ !\ottkw{bool}
+\multimap \forall\, y.\  y \ \ottkw{mat} \ \otimes \ !\ottkw{bool} \multimap
+\ !\ottkw{elt} \multimap z\ \ottkw{mat} \multimap ( x \ \ottkw{mat} \ \otimes y
+\ \ottkw{mat} ) \ \otimes z\ \ottkw{mat}$
 
@@
 
-and operational semantics here
+The full, small-step transition relation is in Appendix \ref{subsec:dyn_sem},
+but the key ideas are as follow.
+
+Heaps ($\sigma$) are multisets containing triples of an abstract location $l$,
+a fractional permission $f$ and sized matrices $m_{n,k}$. The notation $l
+\mapsto_f m_{k_1, k_2}$ should be read as ``location $l$ represents $f$
+ownership over matrix $m$ (of size $k_1 \times k_2$)''.  Each heap-and-expression
+either steps to another heap-and-expression or a runtime error $\mathbf{err}$.
+In the full grammar definition we see a definition of values and contexts in
+the language.
+
+We draw the reader's attention to the definitions relating to fractional
+permissions. Specifically, unlike a lambda, the body of a $\ottkw{fun}\, f\!c
+\rightarrow \_$ must be a syntactic value. The context $\ottkw{fun}\, f\!c
+\rightarrow [-]$ means expressions can be reduced inside a fractional
+permission generalisation. This is to emphasize that fractions are merely
+\emph{compile-time constructs} and do not affect runtime behaviour. Correct
+usage of fractions is enforced by the type system, so programs do not get
+stuck. Fractional permissions are specialised using substitution over both the
+heap and an expression (\textsc{Op\_Frac\_Perm}).
+\[
+    \ottdruleOpXXFracXXPerm{}
+\]
+
+Like with the static semantics, the interesting rules in the dynamic semantics
+are those relating to primitives. Creating a matrix ($\ottkw{matrix}\ k_1\
+k_2$) successfully (\textsc{Op\_Matrix}) requires non-negative dimensions and
+returns a (fresh) location of a matrix of those dimensions, extending the heap
+to reflect that $l$ represents a complete ownership over the new matrix.
+\[
+    \ottdruleOpXXMatrix{}
+\]
+
+Dually, \textsc{Op\_Free}, requires a location represent complete ownership
+before removing it and the matrix it points to from the heap.
+\[
+    \ottdruleOpXXFree{}
+\]
+
+Choosing a multiset representation as opposed to a set allows for two
+convenient invariants: multiplicity of a triple $l \mapsto_f m_{k_1, k_2}$ in
+the heap corresponds to the number of aliases of $l$ in the expression with
+type $f\ \ottkw{mat}$ and the sum of all the fractions for $l$ will always be
+$1$ (for a closed, well-typed expression). With this in mind, the rules
+\textsc{Op\_Share} and \textsc{Op\_Unshare\_Eq} are fairly natural.
+\[
+    \ottdruleOpXXShare{} \\
+\]
+\[
+    \ottdruleOpXXUnshareXXEq{}
+\]
+
+Combining all of these features, we see that \textsc{Op\_Gemm\_Match} requires
+that the location being updated ($l_3$) has complete ownership of over matrix
+$m_3$ and can thus change what value it stores to $m_1 m_2 + m_3$. In
+particular, this places no restriction on $l_2$ and $l_3$: they could be
+$\ottkw{share}$d aliases of the same matrix. Transition rules for other
+primitives (omitted) follow the same structure: $\mapsto_1$ for any locations
+that are written to and $\mapsto_{f\!c}$ for anything else.
+\[
+    \ottdruleOpXXGemmXXMatch{}
+\]
 
@@
 
-Describe the step-indexed logical relation and its main properties
+First, we define an interpretation of heaps with fractional permissions in the
+style of Bornat et. al~\cite{bornat} (interpreting the multiset as a partial
+map from locations to the sum of all its associated fractions and a matrix) as
+well as the n-fold iteration of $\rightarrow$.
+\[
+    \den{H}{}{\sigma} = \bigstar_{(l,f,m) \in \sigma} [ l \mapsto_f m ]
+\]
+where
+\[
+    (\varsigma_1 \star \varsigma_2)(l) \equiv
+    \begin{cases}
+        \varsigma_1(l) & \textrm{if } l \in \dom(\varsigma_1) \wedge l \notin \dom(\varsigma_2) \\
+        \varsigma_2(l) & \textrm{if } l \in \dom(\varsigma_2) \wedge l \notin \dom(\varsigma_1) \\
+        (f_1 + f_2, m) & \textrm{if } (f_1, m) = \varsigma_1(l) \wedge (f_2, m) = \varsigma_2(l) \wedge f_1 + f_2 \leq 1 \\
+        \textrm{undefined} & \textrm{otherwise}
+    \end{cases}
+\]
+
+We then define a step-indexed logical relation in the style of Morrisett et.
+al~\cite{morrisett}. $(\varsigma, v) \in \V{k}{t}$ means it takes a heap with
+exactly $\varsigma$ resources to produce a value $v$ of type $t$ in at most $k$
+steps. So, something like a $\ottkw{unit}$ or a $!t$ need no resources, whereas
+a $f\, \ottkw{mat}$ needs exactly $f$ ownership of a some matrix and a pair
+needs a $\star$ combination of the heaps required for each component.
+\begin{align*}
+  \V{k}{ \Unit } &= \{ (\empH, \ast) \} \\
+  \V{k}{ f \, \Mat } &= \{ (\{ l \mapsto_{2^{-f}} \_ \} , l) \} \\
+  \V{k}{ \Bang t } &= \{ (\empH, \Many\, v) \mid (\empH, v) \in \V{k}{t} \} \\
+  \V{k}{ t_1 \otimes t_2 } &= \{ (\varsigma_1 \star \varsigma_2, ( v_1, v_2 )) \mid (\varsigma_1, v_1) \in \V{k}{t_1} \wedge (\varsigma_2, v_2) \in \V{k}{t_2} \}
+\end{align*}
+
+The definition of $\V{k}{\forall f\!c.\ t}$ says a value and heap
+must be the same regardless of what fraction is substituted into both; the
+$k-1$ is to take into account fraction specialisation takes ones step
+(\textsc{Op\_Spc}).
+\[
+    \V{k}{ \forall f\!c.\  t } = \{ (\varsigma, \ottkw{fun}\, f\!c \rightarrow \, v) \mid \forall f.\ (\varsigma [ f\!c / f ], v [ f\!c / f ]) \in \V{k-1}{ t [ f\!c / f ] } \}
+\]
+
+To understand the definition of $\V{k}{t' \multimap t}$, we must first look at
+$\C{k}{t}$, the computational interpretation of types. Intuitively, it is a
+combination of a frame rule on heaps (no interference), type-preservation and
+termination (in $j < k$ steps) to either an error or a heap-and-expression,
+with the further condition that if the expression is a syntactic value then it
+is also one semantically.
+\begin{align*}
+    \C{k}{ t } &= \{ (\varsigma_s, e_s) \mid \forall \, j < k, \sigma_r.\ \varsigma_s \star \varsigma_r \textrm{ defined } \Rightarrow \langle \sigma_s + \sigma_r, e_s \rangle \rightarrow^j \ottkw{err}\ \vee \exists \sigma_f, e_f.\\
+               & \qquad \qquad \langle \sigma_s + \sigma_r, e_s \rangle \rightarrow^j \langle \sigma_f + \sigma_r, e_f \rangle \wedge ( e_f \textrm { is a value } \Rightarrow ( \varsigma_f \star \varsigma_r, e_f ) \in \V{k-j}{t} ) \}
+\end{align*}
+
+In this light, $\V{k}{t' \multimap t}$ simply says
+that $v$ is a function and that evaluating the application of it to any
+argument (of the correct type, requiring its own set of resources, bounded by
+$k$ steps) satisfies all the aforementioned properties.
+\begin{align*}
+    \V{k}{ t' \multimap t } &= \{ (\varsigma_v, v ) \mid ( v \equiv \ottkw{fun}\, x : t' \rightarrow e \vee v \equiv \ottkw{fix} (g, x : t' , e : t) ) \, \wedge\\
+                            & \qquad \qquad \forall j \leq k, (\varsigma_{v'}, v') \in \V{j}{ t' }.\ \varsigma_v \star \varsigma_v' \textrm{ defined } \Rightarrow (\varsigma_v \star \varsigma_v', v\, v') \in \C{j}{t} \}
+\end{align*}
+
+The interpretation of typing environments $\Delta$ and $\Gamma$ are with
+respect to an arbitrary substitution of fractional permissions $\theta$. Note
+that only the interpretation of $\Gamma$ involves a non-empty heap.
+\begin{align*}
+    \den{I}{k}{ \Delta, x : t } \theta &= \{ \delta[x \mapsto v_x ] \mid \delta \in \den{I}{k}{\Delta}\theta \wedge (\empH, v_x) \in \V{k}{\theta(t)} \} \\
+    \den{L}{k}{ \Gamma, x : t } \theta &= \{ (\varsigma \star \varsigma_x, \gamma[x \mapsto v_x ]) \mid (\varsigma, \gamma) \in \den{L}{k}{\Gamma}\theta \wedge (\varsigma_x, v_x) \in \V{k}{\theta(t)} \}
+\end{align*}
+
+And so the final semantic interpretation of a typing judgement simply
+quantifies over all possible fractional permission substitutions $\theta$,
+linear value substitutions $\gamma$, intuitionistic value substitutions
+$\delta$ and heaps $\sigma$.  Note that, $\varsigma \equiv \den{H}{}{\theta(\sigma)}$.
+\begin{align*}
+\den{}{k}{ \Theta; \Delta ; \Gamma \vdash e : t } &= \forall \theta, \delta, \gamma, \sigma.\ \Theta = \dom(\theta) \wedge (\varsigma, \gamma) \in \den{L}{k}{ \Gamma }\theta \wedge \delta \in \den{I}{k}{ \Delta }\theta \Rightarrow \\
+                                                     & \qquad \qquad (\varsigma, \theta(\delta(\gamma(e)))) \in \C{k}{ \theta(t) }
+\end{align*}
 
@@
 
-State the fundamental lemma, and sketch the proof a little
+\[
+    \forall \Theta, \Delta, \Gamma, e, t.\ \Theta; \Delta ; \Gamma \vdash e : t \Rightarrow
+    \forall k.\ \den{}{k}{ \Theta; \Delta ; \Gamma \vdash e : t }
+\]
+
+To prove the above theorem, we need several lemmas; the interesting ones are:
+\ref{frame} (a proof of the frame rule), \ref{subsetKJ} (monotonicity for the
+step-index), \ref{restriction} (splitting up environments corresponds to
+splitting up heaps) and \ref{fracPermSub} (a heap-and-expression take the same
+steps of evaluation under any substitution of their free fractional
+permissions).
+
+The proof proceeds by induction on the typing judgement.

### 2018-11-05 write-up/paper/implementation.tex
Add interpretation and proof
Start on Soundness Theorem

@@
     \item \textbf{Matrix Expressions} are also desugared into the abstract
-        syntax tree through some simple pattern-matching.
+        syntax tree through pattern-matching.
 
@@
         process is type-preserving: \lang's type system is embedded into
-        OCaml's (Figure~\ref{fig:type_grammar}), and so the OCaml type checker
+        OCaml's (Figure~\ref{fig:type_grammar}), so the OCaml type checker
         acts as a sanity check on the generated code.
@@
 
-Desugaring is conventional, outlined in Figure~\ref{fig:lang_desugar}.  Matrix
+Desugaring is conventional, outlined in Appendix~\ref{sec:lang_desugar}.  Matrix
 expression are translated into BLAS/ LAPACK calls via purely syntactic
@@
 \subsubsection{Type checking} is mostly standard for a linearly typed language,
-with the exception of fractional permission inference. Because all functions
-must have their argument types explicitly annotated, inferring the correct
-fraction at a call-site is simply a matter of unification.
+with the exception of fractional permission inference. By restricting fractions
+to be non-positive integer powers of two, we only need to keep track of the
+logarithm of the fractions used. Explicit sharing and unsharing removes the
+need for performing dataflow analysis. As a result, all fractional arithmetic
+can be solved with unification, and in doing so, fractions become directly
+usable in \lang's type-system as opposed to a convenient theoretical tool.
+
+Because all functions must have their argument types explicitly annotated,
+inferring the correct fraction at a call-site is simply a matter of
+unification. We strongly suspect that \emph{full-inference of fractional
+permissions is similarly just matter of unification} (thanks to an
+experimental implementation of just this feature), even though the formal
+system we present here is for an explicitly-typed language.
 
@@
 ones (this allows scoping/variable look-up to be handled uniformly).
-
-\textsc{What more can/should we say here?}
 
@@
 \end{center}
-\caption{Removing redundant re-bindings during tranlation to OCaml.}\label{fig:lang_optimise}
+\caption{Removing redundant re-bindings during translation to OCaml.}\label{fig:lang_optimise}
 \end{figure}

### 2018-11-05 write-up/paper/formal_system.tex
Finish Formal System

@@
 respect to an arbitrary substitution of fractional permissions $\theta$. Note
-that only the interpretation of $\Gamma$ involves a non-empty heap.
+that only the interpretation of $\Gamma$ involves a (potentially) non-empty heap.
 \begin{align*}
@@
 To prove the above theorem, we need several lemmas; the interesting ones are:
-\ref{frame} (a proof of the frame rule), \ref{subsetKJ} (monotonicity for the
-step-index), \ref{restriction} (splitting up environments corresponds to
-splitting up heaps) and \ref{fracPermSub} (a heap-and-expression take the same
-steps of evaluation under any substitution of their free fractional
-permissions).
+the moral equivalent of the frame rule (\ref{frame}), monotonicity for the
+step-index (\ref{subsetKJ}), splitting up environments corresponds to splitting
+up heaps (\ref{restriction}) and heap-and-expressions take the same steps of
+evaluation under any substitution of their free fractional permissions
+(\ref{fracPermSub}).
 
-The proof proceeds by induction on the typing judgement.
+The proof proceeds by induction on the typing judgement.  The case for
+\textsc{Ty\_Fix} is the reason we quantify over the step-index $k$ in the
+\emph{conclusion} of the soundness theorem. It allows us to then induct over
+the step-index and assume exactly the thing we need to prove at a smaller index.
+
+The case for \textsc{Ty\_Gen} follows a similar pattern, but has the extra
+complication of reducing an expression with an arbitrary fractional permission
+variable in it, and then instantiating it at the last momemnt to conclude,
+which is where \ref{fracPermSub} (heap-and-expressions take the same steps of
+evaluation under any substitution of their free fractional permissions) is
+used.
+
+The rest of the cases are either very simple base cases (variables, unit,
+boolean, integer or element literals) or follow very similar patterns; for
+these, only \textsc{Ty\_Let} is presented in full and other similar cases
+simply highlight exactly what would be different.  The general idea is to split
+up the linear substitution and heap along the same split of $\Gamma/\Gamma'$,
+then (by induction) use $\C{k}{-}$ and one `half' of the  linear substitution
+and heap to conclude the `first' sub-expression either takes $j< k$ steps to
+$\ottkw{err}$ or another heap-and-expression.
+
+In the first case, you use \textsc{Op\_Context\_Err} to conclude the whole
+let-expression does the same. Similarly we use \textsc{Op\_Context} $j$ times
+in the second case. However, a small book-keeping wrinkle needs to be taken
+care of in the case that the heap-and-expression turns into a value in $i \leq
+j$ steps: \textsc{Op\_Context} is not functorial for the n-fold iteration of
+$\rightarrow$.  Basically, the following is not quite true:
+\[
+\ottdrule{%
+    \ottpremise{\langle  \sigma  \ottsym{,}  \ottnt{e}  \rangle  \rightarrow^j  \langle  \sigma'  \ottsym{,}  \ottnt{e'}  \rangle}%
+    }{
+    \langle  \sigma  \ottsym{,}  \ottnt{C}  \ottsym{[}  \ottnt{e}  \ottsym{]}  \rangle  \rightarrow^j  \langle  \sigma'  \ottsym{,}  \ottnt{C}  \ottsym{[}  \ottnt{e'}  \ottsym{]}  \rangle}{%
+    {\ottdrulename{Op\_Context}}{}%
+}
+\]
+because after the $i$ steps, we need to invoke \textsc{Op\_Let\_Var} to proceed
+evalution for any remaining $j-i$ steps. After that, it suffices to use the
+induction hypothesis on the second sub-expression to finish the proof.  To do
+so, we need to construct a valid linear substitution and heap (i.e., one in
+$\den{L}{k}{\Gamma', x : t}\theta$). We take the other `half' of the linear
+substitution and heap (from the inital split at the start) and extend it with
+$[x \mapsto v]$, (where $x$ is the variable bound in the let-expression and $v$
+is the value we assume the first sub-expression evaluated to in $i$ steps).

### 2018-11-12 write-up/paper/implementation.tex
Start Performance Metrics

@@
 inferring the correct fraction at a call-site is simply a matter of
-unification. We strongly suspect that \emph{full-inference of fractional
-permissions is similarly just matter of unification} (thanks to an
-experimental implementation of just this feature), even though the formal
-system we present here is for an explicitly-typed language.
+unification. We believe \emph{full-inference of fractional permissions is
+similarly just matter of unification} (thanks to an experimental implementation
+of just this feature), even though the formal system we present here is for an
+explicitly-typed language.
 
@@
 
-Here, evaluate the performance of the examples from the second
-section.  Compare with your C implementations, and perhaps as well as
-the straightforward math transcribed into (Matlab/R/Numpy?).
+% Here, evaluate the performance of the examples from the second
+% section. Compare with your C implementations, and perhaps as well as
+% the straightforward math transcribed into (Matlab/R/Numpy?).
+
+We think that using \lang has two primary benefits: safety and performance. We
+discuss safety in \ref{subsec:finding_bugs}, where we describe how we used
+\lang\ to find linearity and aliasing bugs in a linear algebra algorithm that
+was \emph{generated} by another program.
+
+For performance, we measured the execution four of three equivalent
+implementations of a Kalman filter: in C (using \textsc{Cblas}), \lang\ (using
+\textsc{Owl}'s low-level \textsc{Cblas} bindings), OCaml (using \textsc{Owl}'s
+intended, safe/copying-by-default interface), and Python (using
+\textsc{NumPy}).  Measurements were made in in micro-seconds, against an
+exponentially (powers of 5) increasing scaling factor for matrix size
+parameters $n=5$ and $k=3$. The results in Figure~\ref{fig:timings} are as we
+expected: C is the fastest, followed by \lang, with OCaml and Python last.
+
+We expected the C implementation to be faster than the \lang\ one because the
+latter has the additional (but relatively low) overhead of dimension checks and
+crossing the OCaml/C FFI for each call to a \textsc{Cblas} routine, even though
+the calls and their order are exactly the same. We expected the OCaml and
+Python implementations to be slower because of the unnecessary temporaries they
+allocate.
+
+% Need to re-run timings
+Differences is timings are quite pronounced at small matrix sizes, but are
+still significant at larger ones: for $n=625$, \textsc{Cblas} took $112 \pm
+35\, ms$, \lang\ took $105 \pm 25 \, ms$, \textsc{Owl} took $124 \pm 38\, ms$
+and \textsc{NumPy} took $112 \pm 12\, ms$;  for $n=3125$, \textsc{Cblas} took
+$10.8 \pm 0.7\, s$, \lang\ took $12.0 \pm 1.2 \, s$, \textsc{Owl} took $13.3
+\pm 0.2\, s$ and \textsc{NumPy} took $12.7 \pm 0.6\, s$.
 
@@
 \end{figure}
+
+% Talk about memory and temporaries here
+
+% Caveats - set-up problems

### 2018-11-12 write-up/paper/intro.tex
Start Performance Metrics

--- 2018-10-28 write-up/paper/intro.tex
@@
 in existing programs. In fact, we were able to use \lang\ to find linearity and
-aliasing bugs in a linear algebra program that was \emph{generated} by another
-program \emph{specifically designed to translate matrix expressions into an
-efficient sequence of calls linear algebra routines}. We were also able to
-reduce the number of temporaries used by the same program, using \lang's type
-system to guide us.
+aliasing bugs in a linear algebra algorithm that was \emph{generated} by
+another program \emph{specifically designed to translate matrix expressions
+into an efficient sequence of calls to linear algebra routines}. We were also
+able to reduce the number of temporaries used by the same algorithm, using
+\lang's type system to guide us.
 

### 2018-11-12 write-up/paper/appendix.tex
Update Appendix code

@@
   let k, n = _p_k_n_p_ in
-  let h_sigma = Prim.matrix k n in
-  let (sigma, h), h_sigma =
-    Prim.symm (Many true) (Many 1.) sigma h (Many 0.) h_sigma
-  in
-  let (h_sigma, h), r_2 =
-    Prim.gemm (Many 1.) (h_sigma, Many false) (h, Many true) (Many 1.) r_1
-  in
+  let sigma_hT = Prim.matrix n k in
+  let (sigma, h), sigma_hT =
+    Prim.gemm (Many 1.) (sigma, Many false) (h, Many true) (Many 0.) sigma_hT in
+  let (h, sigma_hT), r_2 =
+    Prim.gemm (Many 1.) (h, Many false) (sigma_hT, Many false) (Many 1.) r_1 in
+  let k_by_k, x = Prim.posv_flip r_2 sigma_hT in
   let (h, mu), data_2 =
-    Prim.gemm (Many 1.) (h, Many false) (mu, Many false) (Many (-1.)) data_1
-  in
-  let h, new_h = Prim.copy_mat_to h h_sigma in
-  let r_2, new_r = Prim.copy_mat r_2 in
-  let chol_r, sol_h = Prim.posv new_r new_h in
-  let chol_r, sol_data = Prim.potrs chol_r data_2 in
-  let () = Prim.free_mat chol_r in
-  let h_sol_h = Prim.matrix n n in
-  let (h, sol_h), h_sol_h =
-    Prim.gemm (Many 1.) (h, Many true) (sol_h, Many false) (Many 0.) h_sol_h
-  in
-  let () = Prim.free_mat sol_h in
-  let h_sol_data = Prim.matrix n (Many 1) in
-  let (h, sol_data), h_sol_data =
-    Prim.gemm (Many 1.) (h, Many true) (sol_data, Many false) (Many 0.) h_sol_data
-  in
-  let mu, mu_copy = Prim.copy_mat mu in
-  let (sigma, h_sol_data), new_mu =
-    Prim.symm (Many false) (Many 1.) sigma h_sol_data (Many 1.) mu_copy
-  in
-  let () = Prim.free_mat h_sol_data in
-  let h_sol_h_sigma = Prim.matrix n n in
-  let (sigma, h_sol_h), h_sol_h_sigma =
-    Prim.symm (Many true) (Many 1.) sigma h_sol_h (Many 0.) h_sol_h_sigma
-  in
-  let sigma, sigma_copy = Prim.copy_mat_to sigma h_sol_h in
-  let (sigma, h_sol_h_sigma), new_sigma =
-    Prim.symm (Many false) (Many (-1.)) sigma h_sol_h_sigma (Many 1.) sigma_copy
-  in
-  let () = Prim.free_mat h_sol_h_sigma in
-  ((sigma, (h, (mu, (r_2, sol_data)))), (new_mu, new_sigma)) )
-in
-kalman
+    Prim.gemm (Many 1.) (h, Many false) (mu, Many false) (Many (-1.)) data_1 in
+  let (x, data_2), new_mu =
+    Prim.gemm (Many 1.) (x, Many false) (data_2, Many false) (Many 1.) mu in
+  let x_h = Prim.matrix n n in
+  let (x, h), x_h =
+    Prim.gemm (Many 1.) (x, Many false) (h, Many false) (Many 0.) x_h in
+  let () = Prim.free_mat x in
+  let sigma, sigma2 = Prim.copy_mat sigma in
+  let (sigma, x_h), new_sigma =
+    Prim.symm (Many true) (Many (-1.)) sigma x_h (Many 1.) sigma2 in
+  let () = Prim.free_mat x_h in
+  ((sigma, h), (new_sigma, (new_mu, (k_by_k, data_2))))
     \end{minted}
@@
                     const double *h, /* k,n */ const double *mu,  /* n,1 */ double *r,           /* k,k */
-                    double *data,    /* k,1 */ double **ret_mu,  /* k,1 */ double **ret_sigma   /* n,n */ ) {
-        double* k_by_n = (double *) malloc(k * n * sizeof(double));
-/*16*/  cblas_dsymm(CblasRowMajor, CblasRight, CblasUpper, k, n, 1., sigma, n, h, n, 0., k_by_n, n);
-/*17*/  cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasTrans, k, k, n, 1., k_by_n, n, h, n, 1., r, k);
-/*18*/  cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans, k, 1, n, 1., h, n, mu, 1, -1., data, 1);
-/*19*/  cblas_dcopy(k * n, h, 1, k_by_n, 1);
-        double* k_by_k = (double *) malloc(k * k * sizeof(double));
-/*20*/  cblas_dcopy(k * k, r, 1, k_by_k, 1);
-/*21*/  LAPACKE_dposv(LAPACK_ROW_MAJOR, 'U', k, n, k_by_k, k, k_by_n, n);
-/*23*/  LAPACKE_dpotrs(LAPACK_ROW_MAJOR, 'U', k, 1, k_by_k, k, data, 1);
-        free(k_by_k);
+                    double *data,    /* k,1 */ double **ret_sigma /* n,n */ ) {
+    double* n_by_k = (double *) malloc(n * k * sizeof(double));
+    cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasTrans, n, k, n, 1., sigma, n, h, n, 0., n_by_k, k);
+    cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans, k, k, n, 1., h, n, n_by_k, k, 1., r, k);
+    LAPACKE_dposv(LAPACK_COL_MAJOR, 'U', k, n, r, k, n_by_k, k);
+    cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans, k, 1, n, 1., h, n, mu, 1, -1., data, 1);
+    cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans, n, 1, k, 1., n_by_k, k, data, 1, 1., mu, 1);
     double* n_by_n = (double *) malloc(n * n * sizeof(double));
-/*24*/  cblas_dgemm(CblasRowMajor, CblasTrans, CblasNoTrans, n, n, k, 1., h, n, k_by_n, n, 0., n_by_n, n);
-        free(k_by_n);
-        double* n_by_1 = (double *) malloc(n * sizeof(double));
-/*25*/  cblas_dgemm(CblasRowMajor, CblasTrans, CblasNoTrans, n, 1, k, 1., h, n, data, 1, 0., n_by_1, 1);
-        double* new_mu = (double *) malloc(n * sizeof(double));
-/*26*/  cblas_dcopy(n, mu, 1, new_mu, 1);
-/*27*/  cblas_dsymm(CblasRowMajor, CblasLeft, CblasUpper, n, 1, 1., sigma, n, n_by_1, 1, 1., new_mu, 1);
-        free(n_by_1);
+    cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans, n, n, k, 1., n_by_k, k, h, n, 0., n_by_n, n);
+    free(n_by_k);
     double* n_by_n2 = (double *) malloc(n * n * sizeof(double));
-/*28*/  cblas_dsymm(CblasRowMajor, CblasRight, CblasUpper, n, n, 1., sigma, n, n_by_n, n, 0., n_by_n2, n);
-/*29*/  cblas_dcopy(n*n, sigma, 1, n_by_n, 1);
-/*30*/  cblas_dsymm(CblasRowMajor, CblasLeft, CblasUpper, n, n, -1., sigma, n, n_by_n2, n, 1., n_by_n, n);
-        free(n_by_n2);
-        *ret_sigma = n_by_n;
-        *ret_mu = new_mu; }
+    cblas_dcopy(n*n, sigma, 1, n_by_n2, 1);
+    cblas_dsymm(CblasRowMajor, CblasRight, CblasUpper, n, n, -1., sigma, n, n_by_n, n, 1., n_by_n2, n);
+    free(n_by_n);
+    *ret_sigma = n_by_n2; }
     \end{minted}

### 2018-11-15 write-up/paper/implementation.tex
Finish Performance Metrics (for Kalman, for now)

--- 2018-11-05 write-up/paper/implementation.tex
@@
 
-Here, evaluate the performance of the examples from the second
-section.  Compare with your C implementations, and perhaps as well as
-the straightforward math transcribed into (Matlab/R/Numpy?).
+We think that using \lang\ has two primary benefits: safety and performance. We
+discuss safety in \ref{subsec:finding_bugs}, where we describe how we used
+\lang\ to find linearity and aliasing bugs in a linear algebra algorithm that
+was \emph{generated} by another program.
+
+\subsubsection{Setup}
+
+For performance, we measured the execution of four equivalent
+implementations of a Kalman filter: in C (using \textsc{Cblas}), \lang\ (using
+\textsc{Owl}'s low-level \textsc{Cblas} bindings), OCaml (using \textsc{Owl}'s
+intended, safe/copying-by-default interface), and Python (using \textsc{NumPy},
+with the interpreter started and functions interpreted). We measured execution
+time in micro-seconds, against an exponentially (powers of 5) increasing
+scaling factor for matrix size parameters $n=5$ and $k=3$.
+
+For large scaling factors ($n = 5^4, 5^5$), we triggered a full
+garbage-collection before measuring the execution time of a single call of a
+function. However, due to the limitations of the micro-benchmarking library we
+used, for smaller scaling factors ($n = 5^1, 5^2, 5^3$), we measured the
+execution time of \emph{multiple} calls to a function in a loop, thus including
+potential garbage-collection effects.
+
+\subsubsection{Hypothesis}
+
+We expected the C implementation to be faster than the \lang\ one because the
+latter has the additional (but relatively low) overhead of dimension checks and
+crossing the OCaml/C FFI for each call to a \textsc{Cblas} routine, even though
+the calls and their order are exactly the same. We expected the OCaml and
+Python implementations to be slower because they allocate more temporaries (so
+possibly less cache-friendly) and carry out more floating-point operations --
+the \textsc{Cblas} and \lang\ implementations use ternary kernels (coalescing
+steps), a Cholesky decomposition (of a symmetric matrix, which is more
+efficient than the LU decomposition used for inverting a matrix in \textsc{Owl}
+and \textsc{NumPy}) and \texttt{symm} (symmetric matrix multiplication, halving
+the number of floating-point multiplications required).
+
+\subsubsection{Results}
+
+% Need to re-run timings
+
+The results in Figure~\ref{fig:timings} are as we expected: C is the fastest,
+followed by \lang, with OCaml and Python last.  Differences in timings are
+quite pronounced at small matrix sizes, but are still significant at larger
+ones: for $n=625$, \textsc{Cblas} took $112 \pm 35\, ms$, \lang\ took $105 \pm
+25 \, ms$, \textsc{Owl} took $124 \pm 38\, ms$ and \textsc{NumPy} took $112
+\pm 12\, ms$;  for $n=3125$, \textsc{Cblas} took $10.8 \pm 0.7\, s$, \lang\
+took $12.0 \pm 1.2 \, s$, \textsc{Owl} took $13.3 \pm 0.2\, s$ and
+\textsc{NumPy} took $12.7 \pm 0.6\, s$.
+
+Worth highlighting here is the other major advantage of using \lang\ is reduced
+memory usage.  Whilst the \textsc{Owl} and \textsc{NumPy} use 11 temporary
+matrices (\emph{excluding} the 2 matrices which store the results), using $n +
+n^2 + 4nk + 3k^2 + 2k \approx 4n^2$ (for $k = 3n/5$) words of memory,
+\textsc{Cblas} and \lang\ use only \emph{2} temporary m`atrices (excluding the
+\emph{one} matrix which stores one of the results), using only $n^2 + nk \leq
+2n^2$ words of memory.
 
@@
 \end{figure}
+
+\subsubsection{Analysis}
+
+As matrix sizes increase, assuming sufficient memory, the difference in the
+number of floating-point operations ($O(n^3))$ dominates execution times.
+However for small matrix sizes, since $n$ is small and the measurements were
+over multiple calls to a function in a loop, the large number of temporaries
+show the adverse effect of not re-using memory at even quite small matrix
+sizes: creating pressure on the garbage collector.
+
+% Other examples impl
+% Sampled linear reg example

### 2018-11-15 write-up/paper/discussion.tex
Start Performance Metrics
Start Related Work

@@
 
-\subsection{Finding Bugs in SymPy's Output}
+\subsection{Finding Bugs in SymPy's Output}\label{subsec:finding_bugs}
 
@@
 
-The main point we want to make is that using linear types for BLAS
-is an ``obvious'' idea, but is surprisingly under-explored.
+% The main point we want to make is that using linear types for BLAS
+% is an ``obvious'' idea, but is surprisingly under-explored.
+
+Using linear types for BLAS routines is a particularly good domain fit (given
+the implicit restrictions on aliasing arguments), but is under-explored in the
+academic literature. Linear algebra libraries written in Rust take advantage of
+the distinction that Rust's type system offers between mutable views/references
+to arrays, offering a \emph{seemingly} different approach to the
+fractional-permissions based one we took in this paper.  However, recent
+research\cite{weiss} suggests that Rust's borrow-checker could be explained in
+simpler terms using \emph{fractional-permissions}; in this light, our paper
+neatly bridges the gap by demonstrating the theoretical and practical
+simplicity of this way of thinking.
+
+ATS\cite{cui2005ats} is a language which combines dependent and linear types;
+although it provides BLAS bindings, the main focus in the API provided there is
+for statically enforcing dimensions related constraints of the arguments,
+rather than aliasing restrictions as shown here.
 
 \begin{itemize}
-\item Rust
-\item ATS
-\item Single-assignment C 
-\item Linear Haskell
-\item Bernardy and Sveningsson
-\item L3
+\item Single-assignment C: for hardware, doesn't allow higher-order fun.
+\item Linear Haskell: linearity on the arrows.
+\item Bernardy and Sveningsson: very theoretical, no implementation.
+\item L3: can use existential types to statically track pointer identities to
+    prevent that source of runtime errors.
 \item Boyland fractional permissions
-\item Disadvantage of graph-based approaches to matrix expression generation
-\item multi-stage: once dimensions known, they are fixed?
+\item Disadvantage of graph-based approaches to matrix expression generation:
+    buggy, like SymPy.
+\item multi-stage: once dimensions known, they are fixed, so we can take
+    dimension checking one-step further and implement a matrix chaining alg.
 \end{itemize}
+
+% Dimensions: easy enough to integrate, even w/o ATS/dep types
+% generative phantom types
+% \cite{abe2015simple}
 

### 2018-11-16 write-up/paper/discussion.tex
Finish Related Work

@@
 
-% The main point we want to make is that using linear types for BLAS
-% is an ``obvious'' idea, but is surprisingly under-explored.
-
 Using linear types for BLAS routines is a particularly good domain fit (given
@@
 
-ATS\cite{cui2005ats} is a language which combines dependent and linear types;
-although it provides BLAS bindings, the main focus in the API provided there is
-for statically enforcing dimensions related constraints of the arguments,
-rather than aliasing restrictions as shown here.
+Linear Haskell\cite{bernardy2017linear}, takes a slightly different definition
+of linearity, that is one on \emph{arrows} as opposed to \emph{kinds}: for $f :
+a \multimap b$, if $f u$ is used exactly once \emph{then} $u$ is used exactly
+once. Whilst this has the advantage of being backwards-compatible, \lang's
+approach of compile-time code-generation enables using linearity \emph{as a
+library}, whilst the use of fractions gives extra control over aliasing not
+available in Linear Haskell.
 
-\begin{itemize}
-\item Single-assignment C: for hardware, doesn't allow higher-order fun.
-\item Linear Haskell: linearity on the arrows.
-\item Bernardy and Sveningsson: very theoretical, no implementation.
-\item L3: can use existential types to statically track pointer identities to
-    prevent that source of runtime errors.
-\item Boyland fractional permissions
-\item Disadvantage of graph-based approaches to matrix expression generation:
-    buggy, like SymPy.
-\item multi-stage: once dimensions known, they are fixed, so we can take
-    dimension checking one-step further and implement a matrix chaining alg.
-\end{itemize}
-
-% Dimensions: easy enough to integrate, even w/o ATS/dep types
-% generative phantom types
-% \cite{abe2015simple}
+In general, using substructural types to express array computations is not
+particularly new\cite{scholz,henriksen,bernardy2016}, but these approaches
+limit recursion and higher-order functions, and have special constructs to
+enable \emph{implementation} of efficient BLAS-like libraries rather than
+enforcing correct \emph{usage}.
 
@@
 We are pleasantly surprised at how simple the overall design and implementation
-of \lang\ is, given its expressive power and usability. Indeed, the focus on
-getting a working prototype early on (so that we could test it with real
-BLAS/LAPACK routines as soon as possible) meant that we only added features to
-the type system when it was clear that they were absolutely necessary: these
-features were !-types and value-restriction for the \highl{Many} constructor.
-Although we considered parametric polymorphism, size-types and tracking pointer
-identities at the type-level\cite{morrisett} (all of which would certainly be
-useful and are potential avenues for further work) it turns out these features
-are not necessary.
+of \lang\ is, given its expressive power and usability.  So simple in fact,
+that fractions, a convenient theoretical abstraction until this point, could be
+implemented by restricting division and multiplication to be by 2 only
+\cite{boyland2003}, thus turning any required arithmetic into unification.
 
-We believe we have stumbled across a particularly nice problem-solution fit:
-the idea of linear types for linear algebra -- one that used to just be
-commonly believed folklore -- actually has some merit and is worth exploring.
+Indeed, the focus on getting a working prototype early on (so that we could
+test it with real BLAS/LAPACK routines as soon as possible) meant that we only
+added features to the type system when it was clear that they were absolutely
+necessary: these features were !-types and value-restriction for the
+\highl{Many} constructor.
+
+Going forwards, one may wish to eliminate even more runtime errors from \lang,
+by extending its type system. For example, we could have used existential types
+to statically track pointer identities\cite{morrisett}, or parametric
+polymorphism.
+
+We could also attempt to catch mismatched dimensions at compile time as well.
+While this could be done with generative phantom types\cite{abe2015simple},
+using dependent types may offer more flexibility in \emph{partitioning}
+regions\cite{space_monads} or statically enforcing dimensions related
+constraints of the arguments at compile-time.  ATS\cite{cui2005ats} is just
+such a language which combines dependent and linear types; although it provides
+BLAS bindings, it does not aim to provide aliasing restrictions as demonstrated
+in this paper.
+
+Taking this idea one step even further, since matrix dimensions are typically
+fixed at runtime, we could \emph{stage} \lang\ programs and compile matrix
+expressions using more sophisticated algorithms\cite{barthels}. However, it is
+worth noting that without care, such algorithms\cite{rocklin_thesis}, usually
+based on graph-based, ad-hoc dataflow analysis, can produce erroneous output
+which would not get past a linear type system with fractions.
+
 We also think that this concept (and the general design of its implementation)

### 2018-11-16 write-up/paper/lang_and_examples.tex
Start Performance Metrics
Finish Related Work

@@
     \centering
+%   \begin{minipage}{.4\textwidth}
+%   \centering
     \inputminted[fontsize=\small]{ocaml}{../../test/examples/kalman.lt}
+%   \end{minipage}
+%   \begin{minipage}{.4\textwidth}
+%   \centering
+%   \begin{minted}[linenos,fontsize=\small]{ocaml}
+%   let owl_kalman
+%       ~sigma
+%       ~h
+%       ~mu
+%       ~r
+%       ~data =
+%     let open Owl.Mat in
+%     let ( * ) = dot in
+%     let h' = transpose h in
+%     let sigma_h' = sigma * h' in
+%     let x = sigma_h' * (inv @@ r + h * sigma_h') in
+%     let new_mu = mu + x * (h * mu - data) in
+%     let new_sigma = sigma - x * h * sigma in
+%     new_sigma, new_mu
+%   ;;
+%   \end{minted}
+%   \end{minipage}
     \caption{Kalman filter: see Figure~\ref{fig:kalman_eqns} for the
@@
         (Figure~\ref{fig:cblas_kalman}).}\label{fig:lang_kalman}
+\end{figure}
+
+\begin{figure}[t]
+    \centering
+    \begin{minipage}{.3\textwidth}
+        \centering
+        \begin{grammar}
+            <f> ::= `'
+            \alt <fc>
+            \alt `Z'
+            \alt `S' <f>
+
+            <t> ::= `'
+            \alt `unit'
+            \alt `bool'
+            \alt `int'
+            \alt `elt'
+            \alt <f> `arr'
+            \alt <f> `mat'
+            \alt `!' <t>
+            \alt \synt{$\forall$} <fc.> <t>
+            \alt <t> \lit{$\otimes$} \synt{$t'$}
+            \alt <t> \lit{$\multimap$} \synt{$t'$}
+        \end{grammar}
+    \end{minipage}
+    \begin{minipage}{.3\textwidth}
+        \centering
+        \begin{minted}[fontsize=\small]{ocaml}
+module Arr =
+  Owl.Dense.Ndarray.D
+
+type z = Z
+type 'a s = Succ
+
+type 'a arr =
+  A of Arr.arr
+  [@@unboxed]
+
+type 'a mat =
+  M of Arr.arr
+  [@@unboxed]
+
+type 'a bang =
+  Many of 'a
+  [@@unboxed]
+        \end{minted}
+    \end{minipage}
+    \begin{minipage}{.3\textwidth}
+        \begin{align*}
+            [\![ f\!c ]\!] &= \texttt{'fc} \\
+            [\![ \textbf{Z} ]\!] &= \texttt{z}\\
+            [\![ \textbf{S} \, f ]\!] &= [\![ f ]\!]\, \texttt{s}\\
+            [\![ \textbf{unit} ]\!] &= \texttt{unit}\\
+            [\![ \textbf{bool} ]\!] &= \texttt{bool}\\
+            [\![ \textbf{int} ]\!] &= \texttt{int}\\
+            [\![ \textbf{elt} ]\!] &= \texttt{float}\\
+            [\![ f\, \textbf{arr} ]\!] &= [\![ f ]\!]\, \texttt{arr}\\
+            [\![ f\, \textbf{mat} ]\!] &= [\![ f ]\!]\, \texttt{mat}\\
+            [\![ \textbf{!} \, t ]\!] &= [\![ t ]\!]\, \texttt{bang}\\
+            [\![ \forall f\!c.\, t ]\!] &= [\![ t ]\!]\\
+            [\![ t \otimes t' ]\!] &= [\![ t ]\!] \texttt{*} [\![ t' ]\!]\\
+            [\![ t \multimap t' ]\!] &= [\![ t ]\!] \rightarrow [\![ t' ]\!]
+        \end{align*}
+    \end{minipage}
+    \caption{\lang's type grammar (left) and its embedding into OCaml
+        (right).}\label{fig:type_grammar}
 \end{figure}
@@
 
-    \item \texttt{potrs \_ chol\_r data\_2} uses a pre-computed Cholesky decomposition
-        to multiply \texttt{data\_2} $ = H \mu - \textrm{data}$ by
-        \texttt{r\_2} $= (R + H \Sigma H^T)^{-1}$
+    \item \texttt{posvFlip} is like \texttt{posv} except for solving $XA = B$
 
@@
         \begin{itemize}
-            \item \texttt{r\_1} and \texttt{r\_2}
-            \item \texttt{data\_1}, \texttt{data\_2} and \texttt{sol\_data}
-            \item \texttt{new\_h} and \texttt{sol\_h}
-            \item \texttt{h\_sol\_h}, \texttt{sigma\_copy} and \texttt{new\_sigma}
-            \item \texttt{mu\_copy} and \texttt{new\_mu}.
+            \item \texttt{r\_1}, \texttt{r\_2} and \texttt{k\_by\_k}
+            \item \texttt{data\_1} and \texttt{data\_2}
+            \item \texttt{mu} and \texttt{new\_mu}
+            \item \texttt{sigma\_hT} and \texttt{x}
         \end{itemize}

### 2018-11-16 write-up/paper/implementation.tex
Declare Draft

@@
 \section{Implementation}\label{sec:implementation}
-
-Talk about how you implemented \lang\ and the general
-architecture. Talk about how simple everything is, and also
-about how implementing inference for fractions is.
 
@@
     \\ \\
-        \alsocell[c]{l}{\highl{let Many x = <exp> in} \\
-            \highl{let Many x = Many (Many x) in <body>} \\
-            \textbf{OR} \highl{let Many x = Many <exp> in <body>} \\
-            \textbf{OR} \highl{(fun x : t -> <body>) <exp>}} &
+        \alsocell[c]{l}{\highl{(* fixp = fix (f, x:t, <exp> : t') *)} \\
+            \highl{(*1*) let Many f = Many fixp in <body>} \\
+            \highl{(*2*) let f = fixp in <body>} } &
+        $\Rightarrow$ &
+        \highl{let rec f x = <exp> in <body>}
+    \\ \\
+    \alsocell[c]{l}{\highl{(*1*) let Many x = <exp> in} \\
+            \highl{(*-*) let Many x = Many (Many x) in <body>} \\
+            \highl{(*2*) let Many x = Many <exp> in <body>} \\
+            \highl{(*3*) (fun x : t -> <body>) <exp>}} &
         $\Rightarrow$ &
         \highl{let x = <exp> in <body>}
-    \\ \\
-        \alsocell[c]{l}{\highl{(* fixp = fix (f, x:t, <exp> : t') *)} \\
-            \highl{let Many f = Many fixp in <body>} \\
-            \textbf{OR} \highl{let f = fixp in <body>} } &
-        $\Rightarrow$ &
-        \highl{let rec f x = <exp> in <body>}
     \end{tabular}

### 2018-11-16 write-up/paper/paper.tex
Start on Soundness Theorem
Add Abstract

@@
 \begin{abstract}
-Briefly summarize the contents of the paper in 150--250 words.
-%
+    We present \lang, a a functional programming language designed to express
+    the APIs of low-level linear algebra libraries (such as BLAS/LAPACK) safely
+    and explicitly, through a brief description of its key features and several
+    illustrative examples. We show that the \lang's type system is sound and that
+    its implementation improves upon na{\"i}ve implementations of linear
+    algebra programs, almost towards C-levels of performance. Lastly, we contrast
+    it to other recent developments in linear types and show that using linear
+    types and fractional permissions to express the APIs of low-level linear
+    algebra libraries is a simple and effective idea.
 \keywords{numerical, linear, algebra, types, permissions, OCaml}
@@
 
+\clearpage
 \appendix

### 2018-11-29 write-up/dissertation/background.tex
Organise write-ups
Add more evaluation examples

--- 2018-05-29 write-up/background.tex
@@
     \centering
-    \inputminted[fontsize=\small]{ocaml}{../test/examples/factorial.lt}
+    \inputminted[fontsize=\small]{ocaml}{../../examples/factorial.lt}
     \caption{Factorial function in LT4LA.}\label{fig:ltfla_factorial}
@@
     \centering
-    \inputminted[fontsize=\small]{ocaml}{../test/examples/sum_array.lt}
+    \inputminted[fontsize=\small]{ocaml}{../../examples/sum_array.lt}
     \caption{Summing over an array in LT4LA.}\label{fig:ltfla_sumarray}
@@
 (* Are you trying to write to/free/unshare an array you don't own? *)
-(* In test/examples/sum_array.lt, at line: 7 and column: 19        *)
+(* In examples/sum_array.lt, at line: 7 and column: 19        *)
     \end{minted}

### 2018-11-29 write-up/paper/appendix.tex
Finish Performance Metrics (for Kalman, for now)
Add more evaluation examples

@@
 
-\vspace*{\fill}
 \begin{figure}[h]
@@
         \texttt{ocamlformat} for presentation.}\label{fig:ocaml_kalman}
-\end{center}
-\end{figure}
-\vfill
 
-\begin{landscape}
-\begin{figure}[p]
-    \centering
+\vspace{\baselineskip}
+
     \begin{minted}[fontsize=\footnotesize]{c}
-static void kalman( const int n,               const int k,                const double *sigma, /* n,n */
-                    const double *h, /* k,n */ const double *mu,  /* n,1 */ double *r,           /* k,k */
+static void kalman( const int n,                   const int k,                
+                    const double *sigma, /* n,n */ const double *h,   /* k,n */
+                    const double *mu,    /* n,1 */ double *r,         /* k,k */
                     double *data,        /* k,1 */ double **ret_sigma /* n,n */ ) {
     double* n_by_k = (double *) malloc(n * k * sizeof(double));
-    cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasTrans, n, k, n, 1., sigma, n, h, n, 0., n_by_k, k);
-    cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans, k, k, n, 1., h, n, n_by_k, k, 1., r, k);
+    cblas_dgemm(RowMajor, NoTrans, Trans, n, k, n, 1., sigma, n, h, n, 0., n_by_k, k);
+    cblas_dgemm(RowMajor, NoTrans, NoTrans, k, k, n, 1., h, n, n_by_k, k, 1., r, k);
     LAPACKE_dposv(LAPACK_COL_MAJOR, 'U', k, n, r, k, n_by_k, k);
-    cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans, k, 1, n, 1., h, n, mu, 1, -1., data, 1);
-    cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans, n, 1, k, 1., n_by_k, k, data, 1, 1., mu, 1);
+    cblas_dgemm(RowMajor, NoTrans, NoTrans, k, 1, n, 1., h, n, mu, 1, -1., data, 1);
+    cblas_dgemm(RowMajor, NoTrans, NoTrans, n, 1, k, 1., n_by_k, k, data, 1, 1., mu, 1);
     double* n_by_n = (double *) malloc(n * n * sizeof(double));
-    cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans, n, n, k, 1., n_by_k, k, h, n, 0., n_by_n, n);
+    cblas_dgemm(RowMajor, NoTrans, NoTrans, n, n, k, 1., n_by_k, k, h, n, 0., n_by_n, n);
     free(n_by_k);
@@
     cblas_dcopy(n*n, sigma, 1, n_by_n2, 1);
-    cblas_dsymm(CblasRowMajor, CblasRight, CblasUpper, n, n, -1., sigma, n, n_by_n, n, 1., n_by_n2, n);
+    cblas_dsymm(RowMajor, Right, Upper, n, n, -1., sigma, n, n_by_n, n, 1., n_by_n2, n);
     free(n_by_n);
@@
 
+\end{center}
 \end{figure}
-\end{landscape}

### 2018-11-29 write-up/paper/implementation.tex
Add more evaluation examples

@@
 
-For performance, we measured the execution of four equivalent
+For performance, we measured the execution times of four equivalent
 implementations of a Kalman filter: in C (using \textsc{Cblas}), \lang\ (using
@@
 potential garbage-collection effects.
+
+We also measured the execution times of L1-norm minimisation and the
+``linear-regression'' ($\mathbf{(X^T X)^{-1} X^T y}$) similary, but without a C
+implementation.
 
@@
 
-The results in Figure~\ref{fig:timings} are as we expected: C is the fastest,
+The results in Figures~\ref{fig:timings} are as we expected: C is the fastest,
 followed by \lang, with OCaml and Python last.  Differences in timings are
 quite pronounced at small matrix sizes, but are still significant at larger
-ones: for $n=625$, \textsc{Cblas} took $112 \pm 35\, ms$, \lang\ took $105 \pm
-25 \, ms$, \textsc{Owl} took $124 \pm 38\, ms$ and \textsc{NumPy} took $112
-\pm 12\, ms$;  for $n=3125$, \textsc{Cblas} took $10.8 \pm 0.7\, s$, \lang\
-took $12.0 \pm 1.2 \, s$, \textsc{Owl} took $13.3 \pm 0.2\, s$ and
-\textsc{NumPy} took $12.7 \pm 0.6\, s$.
+ones. Specifically for the Kalman filter, for $n=625$, \textsc{Cblas} took $112
+\pm 35\, ms$, \lang\ took $105 \pm 25 \, ms$, \textsc{Owl} took $124 \pm 38\,
+ms$ and \textsc{NumPy} took $112 \pm 12\, ms$;  for $n=3125$, \textsc{Cblas}
+took $10.8 \pm 0.7\, s$, \lang\ took $12.0 \pm 1.2 \, s$, \textsc{Owl} took
+$13.3 \pm 0.2\, s$ and \textsc{NumPy} took $12.7 \pm 0.6\, s$.
 
@@
 memory usage.  Whilst the \textsc{Owl} and \textsc{NumPy} use 11 temporary
-matrices (\emph{excluding} the 2 matrices which store the results), using $n +
-n^2 + 4nk + 3k^2 + 2k \approx 4n^2$ (for $k = 3n/5$) words of memory,
-\textsc{Cblas} and \lang\ use only \emph{2} temporary m`atrices (excluding the
-\emph{one} matrix which stores one of the results), using only $n^2 + nk \leq
-2n^2$ words of memory.
+matrices for the Kalman filter, (\emph{excluding} the 2 matrices which store
+the results), using $n + n^2 + 4nk + 3k^2 + 2k \approx 4n^2$ (for $k = 3n/5$)
+words of memory, \textsc{Cblas} and \lang\ use only \emph{2} temporary
+matrices (excluding the \emph{one} matrix which stores one of the results),
+using only $n^2 + nk \leq 2n^2$ words of memory.
 
@@
     \centering
-    \input{timings.tex}
+    \input{kalman_timings.tex}
+    \begin{minipage}{.49\textwidth}
+        \centering
+        \input{l1_norm_min_timings.tex}
+    \end{minipage}
+    \begin{minipage}{.49\textwidth}
+        \centering
+        \input{lin_reg_timings.tex}
+    \end{minipage}
     \caption{Comparison of execution times (error bars are present but quite
@@
 sizes: creating pressure on the garbage collector.
-
-% Other examples impl
-% Sampled linear reg example

### 2018-11-29 write-up/paper/lang_and_examples.tex
Add more evaluation examples

--- 2018-10-28 write-up/paper/lang_and_examples.tex
@@
     \centering
-    \inputminted[fontsize=\small, linenos]{ocaml}{../../test/examples/factorial.lt}
+    \inputminted[fontsize=\small, linenos]{ocaml}{../../examples/factorial.lt}
     \caption{Factorial function in \lang.}\label{fig:lang_factorial}
@@
     \centering
-    \inputminted[fontsize=\small]{ocaml}{../../test/examples/sum_array.lt}
+    \inputminted[fontsize=\small]{ocaml}{../../examples/sum_array.lt}
     \caption{Summing over an array in \lang.}\label{fig:lang_sumarray}
@@
     \centering
-    \inputminted[fontsize=\small]{ocaml}{../../test/examples/weighted_avg_infer.lt}
+    \inputminted[fontsize=\small]{ocaml}{../../examples/weighted_avg_infer.lt}
     \caption{\emph{Simplified} one-dimensional convolution.}\label{fig:lang_oned_conv}
@@
     \centering
-    \inputminted[fontsize=\small]{ocaml}{../../test/examples/square.lt}
+    \inputminted[fontsize=\small]{ocaml}{../../examples/square.lt}
     \caption{Linear regression (OLS): $\hat\beta =
@@
     \centering
-    \inputminted[fontsize=\small]{ocaml}{../../test/examples/lin_reg.lt}
+    \inputminted[fontsize=\small]{ocaml}{../../examples/lin_reg.lt}
     \caption{Linear regression (OLS): $\hat\beta =
@@
     \centering
-    \inputminted[fontsize=\small]{ocaml}{../../test/examples/l1_norm_min.lt}
+    \inputminted[fontsize=\small]{ocaml}{../../examples/l1_norm_min.lt}
     \caption{L1-norm minimisation on manifolds:
@@
     \centering
-    \inputminted[fontsize=\small]{ocaml}{../../test/examples/kalman.lt}
+%   \begin{minipage}{.4\textwidth}
+%   \centering
+    \inputminted[fontsize=\small]{ocaml}{../../examples/kalman.lt}
+%   \end{minipage}
+%   \begin{minipage}{.4\textwidth}
+%   \centering
+%   \begin{minted}[linenos,fontsize=\small]{ocaml}
+%   let owl_kalman
+%       ~sigma
+%       ~h
+%       ~mu
+%       ~r
+%       ~data =
+%     let open Owl.Mat in
+%     let ( * ) = dot in
+%     let h' = transpose h in
+%     let sigma_h' = sigma * h' in
+%     let x = sigma_h' * (inv @@ r + h * sigma_h') in
+%     let new_mu = mu + x * (h * mu - data) in
+%     let new_sigma = sigma - x * h * sigma in
+%     new_sigma, new_mu
+%   ;;
+%   \end{minted}
+%   \end{minipage}
     \caption{Kalman filter: see Figure~\ref{fig:kalman_eqns} for the

### 2019-01-03 write-up/paper/paper.tex
Fix abstract

--- 2018-11-01 write-up/paper/paper.tex
@@
 \begin{abstract}
-Briefly summarize the contents of the paper in 150--250 words.
-%
+    We present \lang, a functional programming language designed to express
+    the APIs of low-level linear algebra libraries (such as BLAS/LAPACK) safely
+    and explicitly, through a brief description of its key features and several
+    illustrative examples. We show that \lang's type system is sound and that
+    its implementation improves upon na{\"i}ve implementations of linear
+    algebra programs, almost towards C-levels of performance. Lastly, we contrast
+    it to other recent developments in linear types and show that using linear
+    types and fractional permissions to express the APIs of low-level linear
+    algebra libraries is a simple and effective idea.
 \keywords{numerical, linear, algebra, types, permissions, OCaml}

### 2019-01-09 write-up/paper/discussion.tex
Beefed up the related work

@@
 
-Using linear types for BLAS routines is a particularly good domain fit (given
-the implicit restrictions on aliasing arguments), but is under-explored in the
-academic literature. Linear algebra libraries written in Rust take advantage of
-the distinction that Rust's type system offers between mutable views/references
-to arrays, offering a \emph{seemingly} different approach to the
-fractional-permissions based one we took in this paper.  However, recent
-research\cite{weiss} suggests that Rust's borrow-checker could be explained in
-simpler terms using \emph{fractional-permissions}; in this light, our paper
-neatly bridges the gap by demonstrating the theoretical and practical
-simplicity of this way of thinking.
+Using linear types for BLAS routines is a particularly good domain fit
+(given the implicit restrictions on aliasing arguments), and as a
+result the idea of using substructural types to express array
+computations is not a particularly new
+one\cite{scholz,henriksen,bernardy2016}.  However, many of these
+designs have been focused on building languages to \emph{implement}
+the kernel linear algebra functions, and as a result, they tend to add
+additional limitations on the language design. Both
+Futhark~\cite{henriksen} and Single Assignment C~\cite{scholz} omit
+higher-order functions facilitating compilation to GPUs. The work of
+\citet{bernardy2016} forbids term-level recursion, in order to ensure
+that all higher-order computations can be statically normalized away
+and thereby maximize opportunities for array fusion.
 
-Linear Haskell\cite{bernardy2017linear}, takes a slightly different definition
-of linearity, that is one on \emph{arrows} as opposed to \emph{kinds}: for $f :
-a \multimap b$, if $f u$ is used exactly once \emph{then} $u$ is used exactly
-once. Whilst this has the advantage of being backwards-compatible, \lang's
-approach of compile-time code-generation enables using linearity \emph{as a
-library}, whilst the use of fractions gives extra control over aliasing not
-available in Linear Haskell.
+In contrast, our approach is to begin with the assumption that we can
+take existing efficient BLAS-like libraries, and then enforce their
+correct \emph{usage} using a linear type discipline with fractional
+permissions.
 
-In general, using substructural types to express array computations is not
-particularly new\cite{scholz,henriksen,bernardy2016}, but these approaches
-limit recursion and higher-order functions, and have special constructs to
-enable \emph{implementation} of efficient BLAS-like libraries rather than
-enforcing correct \emph{usage}.
+This approach is similar to the one taken in linear algebra libraries
+for Rust -- these libraries typically take advantage of the
+distinction that Rust's type system offers between mutable
+views/references to arrays.  The work of \citet{weiss} and
+\citet{rustbelt} suggest that Rust's borrow-checker can be expressed
+in simpler terms using \emph{fractional-permissions}, though to our
+knowledge the programmer-visible lifetime analysis in Rust has never
+been formalized.
+
+Using fractional permissions directly, as we did, rather than relying
+on a complex has two main benefits. First, our type system
+demonstrates that type systems for fractional permissions can be
+\emph{dramatically} simpler than existing state-of-the-art approaches,
+including both industrial languages like Rust, as well as the academic
+languages developed by \citet{bierhoff}.  Their type system, much like
+Rust, build a complex dataflow analysis into the typing rules to infer
+when variables can be shared or not. This allows for very
+natural-looking user programs, but creates the impression that using
+fractional permissions requires a heavy theoretical and engineering
+effort going well beyond that needed for supporting basic linear
+types.
+
+Instead, our approach of requiring sharing to be made explicit means
+that the existing unification machinery already in place for ordinary
+ML-style type inference, can be reused to support fractions. Basically,
+we can view sharing a value as dividing a fraction by two, and after
+taking logarithms all fractions are Peano numbers, whose equality can
+be established with ordinary unification.
+
+This fact is important because there are major upcoming
+implementations of linear types such as Linear
+Haskell~\cite{bernardy2017linear}, which do not have built-in support
+for fractional permissions. Instead, Linear Haskell takes a slightly
+different definition of linearity, one based on \emph{arrows} as
+opposed to \emph{kinds}: for $f : a \multimap b$, if $f u$ is used
+exactly once \emph{then} $u$ is used exactly once. Whilst this has the
+advantage of being backwards-compatible, it also means that the type
+system has no built-in support for the concurrent reader, exclusive
+writer pattern that fractional permissions enable.
+
+However, since our type system shows demonstrates unification is ``all
+one needs'' for fractions, it should be possible to \emph{encode}
+\lang's approach to fractional permissions in Linear Haskell by adding
+a GADT-style natural number index to array types tracking the
+fraction, which should enable supporting high-performance BLAS
+bindings in Linear Haskell. Actually implementing this is something we
+leave for future work, as there remains an issue which we do not see a
+good encoding for. Namely, only having support for linear functions
+makes it a bit inconvenient to manipulate linear values directly --
+programs end up taking on a CPS-like structure. This seems to remain
+an advantage of a direct implementation of linear types.
 

### 2019-01-10 write-up/paper/discussion.tex
Update to LIPIcs style

@@
 higher-order functions facilitating compilation to GPUs. The work of
-\citet{bernardy2016} forbids term-level recursion, in order to ensure
+\cite{bernardy2016} forbids term-level recursion, in order to ensure
 that all higher-order computations can be statically normalized away
@@
 distinction that Rust's type system offers between mutable
-views/references to arrays.  The work of \citet{weiss} and
-\citet{rustbelt} suggest that Rust's borrow-checker can be expressed
+views/references to arrays.  The work of \cite{weiss} and
+\cite{rustbelt} suggest that Rust's borrow-checker can be expressed
 in simpler terms using \emph{fractional-permissions}, though to our
@@
 including both industrial languages like Rust, as well as the academic
-languages developed by \citet{bierhoff}.  Their type system, much like
+languages developed by \cite{bierhoff}.  Their type system, much like
 Rust, build a complex dataflow analysis into the typing rules to infer

### 2019-01-10 write-up/paper/paper.tex
Update to LIPIcs style

@@
-\documentclass[runningheads,a4paper,twoside]{llncs}
-\usepackage{hyperref}
-\hypersetup{%
-    colorlinks=true,
-    linkcolor={black},
-    citecolor={black},
-    urlcolor={blue}
-}
+\documentclass[a4paper,UKenglish]{lipics-v2019}
+%This is a template for producing LIPIcs articles. 
+%See lipics-manual.pdf for further information.
+%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
+%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
+%for section-numbered lemmas etc., use "numberwithinsect"
+%for enabling cleveref support, use "cleveref"
+%for enabling cleveref support, use "autoref"
+
+\bibliographystyle{plainurl}
+
+\title{NumLin: Linear Types for Linear Algebra}
+
+\author{Dhruv C.~Makwana}{Unaffiliated \url{dhruvmakwana.com} }{dcm41@cam.ac.uk}{[orcid]}{}
+
+\author{Neelakantan R.~Krishnaswami}{Department of Computer Science and Technology, University of Cambridge, United Kingdom}{nk480@cl.cam.ac.uk}{[orcid]}{}
+
+\authorrunning{D.\,C. Makwana and N.\,R. Krishnaswami}
+
+\Copyright{Dhruv C. Makwana and Neelakantan R. Krishnaswami}% LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+
+\ccsdesc[300]{Theory of computation~Program specifications}
+
+\keywords{numerical, linear, algebra, types, permissions, OCaml}
+
+\supplement{\url{www.github.com/dc-mak/lt4la}}
+
+%\nolinenumbers %uncomment to disable line numbering
+
+%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
+
+%Editor-only macros:: begin (do not touch as author)%
+\EventEditors{John Q. Open and Joan R. Access}
+\EventNoEds{2}
+\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
+\EventShortTitle{CVIT 2016}
+\EventAcronym{CVIT}
+\EventYear{2016}
+\EventDate{December 24--27, 2016}
+\EventLocation{Little Whinging, United Kingdom}
+\EventLogo{}
+\SeriesVolume{42}
+\ArticleNo{23}
 
@@
 \usepackage{supertabular}%
-\usepackage[margin=1in]{geometry}%
+% \usepackage{geometry}%
 \usepackage{ifthen}%
@@
 
+% NumLin
+\newcommand{\lang}{\textsc{NumLin}}
+
 \begin{document}
 
-\title{NumLin: Linear Types for Linear Algebra}
-
-% \titlerunning{NumLin: Linear Types for Linear Algebra}
-% If the paper title is too long for the running head, you can set
-% an abbreviated paper title here
-
-\newcommand{\lang}{\textsc{NumLin}}
-
-\author{Dhruv C.~Makwana\inst{1}\orcidID{orcidID} \and
-Neelakantan R.~Krishnaswami\inst{2}\orcidID{orcidID}}
-%
-\authorrunning{D.C. Makwana and N.R. Krishnaswami}
-
-\institute{\email{dcm41@cam.ac.uk}
-\url{dhruvmakwana.com} \and
-Department of Computer Science, University of Cambridge\\
-\email{nk480@cl.cam.ac.uk}}
-%
-\maketitle
-%
 \begin{abstract}
@@
     algebra libraries is a simple and effective idea.
-\keywords{numerical, linear, algebra, types, permissions, OCaml}
 \end{abstract}
@@
 
-\bibliographystyle{splncs04}
-
+\clearpage
 \bibliography{ourbib}

### 2019-01-11 write-up/paper/paper.tex
Add ORCiDs

@@
 
-\author{Dhruv C.~Makwana}{Unaffiliated \url{dhruvmakwana.com} }{dcm41@cam.ac.uk}{[orcid]}{}
+\author{Dhruv C.~Makwana}{Unaffiliated \url{dhruvmakwana.com} }{dcm41@cam.ac.uk}{https://orcid.org/0000-0001-7220-4991}{}
 
-\author{Neelakantan R.~Krishnaswami}{Department of Computer Science and Technology, University of Cambridge, United Kingdom}{nk480@cl.cam.ac.uk}{[orcid]}{}
+\author{Neelakantan R.~Krishnaswami}{Department of Computer Science and Technology, University of Cambridge, United Kingdom}{nk480@cl.cam.ac.uk}{https://orcid.org/0000-0003-2838-5865}{}
 

### 2019-01-11 write-up/dissertation/background.tex
Add title

@@
 The syntax is intended to resemble OCaml's, apart from the spurious
-\ltfla{`!'}s found here and there (they are annotations to show that we can
+\highl{`!'}s found here and there (they are annotations to show that we can
 use a value more than once). In Figure~\ref{fig:ltfla_sumarray}, we see
-\ltfla{row} has type (\ltfla{'x arr}). More detailed explanations of
+\highl{row} has type (\highl{'x arr}). More detailed explanations of
 \emph{what and why} will be given in Chapter~\ref{chap:impl}, but for now, it
-is enough to know it means we can only \emph{read} from \ltfla{row}, and not
-write to it. If we did try to write to or free \ltfla{row}, we would get a
+is enough to know it means we can only \emph{read} from \highl{row}, and not
+write to it. If we did try to write to or free \highl{row}, we would get a
 helpful error message, as shown in Figure~\ref{fig:ltfla_univ}.
@@
 routine (called `simple\_dgemm'\footnote{For the purposes of this example, I
-assume `simple\_dgemm' has type \ltfla{'x . 'x mat --o 'y .  'y mat --o z mat
+assume `simple\_dgemm' has type \highl{'x . 'x mat --o 'y .  'y mat --o z mat
 --o ('x mat * 'y mat) * z mat}.} in Figures~\ref{fig:ltfla_square}
@@
 
-To solve this, we use a special primitive called \ltfla{share} to produce
+To solve this, we use a special primitive called \highl{share} to produce
 (more) read-only \emph{aliases} of any matrix. We then pass these into the
@@
 we may not want the original anymore, and thus decide to free it. Before we do
-so, we must first \ltfla{unshare} any read-only aliases that exist, to get back
+so, we must first \highl{unshare} any read-only aliases that exist, to get back
 a single, read-write handle.
@@
     \end{minted}
-    \caption{Squaring a matrix in LT4LA (assuming \ltfla{c} is initially
+    \caption{Squaring a matrix in LT4LA (assuming \highl{c} is initially
         0).}\label{fig:ltfla_square}
@@
 If we tried to free one of the read-only aliases before or instead of
-\ltfla{unshare}-ing them, then we would get the error shown in
-Figure~\ref{fig:ltfla_free}. Briefly, a \ltfla{z arr} is a read-write array,
-everything else (\ltfla{'x arr} or \ltfla{_ s arr}) is read-only. The types of
-\ltfla{free}, \ltfla{share} and \ltfla{unshare} (see
+\highl{unshare}-ing them, then we would get the error shown in
+Figure~\ref{fig:ltfla_free}. Briefly, a \highl{z arr} is a read-write array,
+everything else (\highl{'x arr} or \highl{_ s arr}) is read-only. The types of
+\highl{free}, \highl{share} and \highl{unshare} (see
 Appendix~\ref{chap:primitives}) are set up so that you can only free something
@@
 name in scope with this capability (aliases can only be read-only). Conversely,
-if we \emph{forgot} to \ltfla{free a}, we would also get \texttt{Variable a not
+if we \emph{forgot} to \highl{free a}, we would also get \texttt{Variable a not
 used} error.
@@
 concluding that it computes $A := A^2 + A$ in-place. However, the type of
-\ltfla{share} prevents this as well -- \ltfla{let (a11, a12) = share _ a1 in
+\highl{share} prevents this as well -- \highl{let (a11, a12) = share _ a1 in
 simple_dgemm _ a11 _ a12 a2} would result in an error similar to the one shown

### 2019-01-11 write-up/dissertation/evaluation.tex
Add Appendix and more figures
Add title

@@
 library, for larger factors (125 and greater), I used the \texttt{getrusage}
-system call (called \ltfla{Unix.times} in OCaml), sandwiched between calls to
-\ltfla{Gc.full_major} to minimise the effects of garbage-collection.
+system call (called \highl{Unix.times} in OCaml), sandwiched between calls to
+\highl{Gc.full_major} to minimise the effects of garbage-collection.
 Core\_bench performs a linear regression (here, time against batch-size) so
@@
         \texttt{getrusage} functionality, sandwiched between calls to
-        \ltfla{Gc.full_major} for the OCaml implementations.}\label{fig:timings}
+        \highl{Gc.full_major} for the OCaml implementations.}\label{fig:timings}
 \end{figure}
@@
 that memory usage for temporaries in LT4LA is on par with that of CBLAS. One
-caveat is that the \ltfla{freeM} primitive is a no-op in LT4LA, so deallocation
+caveat is that the \highl{freeM} primitive is a no-op in LT4LA, so deallocation
 still relies on OCaml's garbage-collector.
@@
         \texttt{getrusage} functionality, sandwiched between calls to
-        \ltfla{Gc.full_major} for the OCaml implementations.}\label{fig:timings_all}
+        \highl{Gc.full_major} for the OCaml implementations.}\label{fig:timings_all}
 \end{figure}
@@
 The trend here is that for $n=625$ and $n=3125$ (sizes for which
-\ltfla{Gc.full_major} was called before each measurement), mean execution times
+\highl{Gc.full_major} was called before each measurement), mean execution times
 start to split into two groups: Chol and Owl as one group (similar,
@@
 first performed live-range splitting (checked by linearity) by hoisting calls
-to \ltfla{freeM} and then annotated the freed matrices with their dimensions.
+to \highl{freeM} and then annotated the freed matrices with their dimensions.
 After doing so and spotting two disjoint live-ranges of the same size, I
-replaced a call to \ltfla{freeM} followed by allocating call to \ltfla{copy}
-with one, in-place call to \ltfla{copyM_to}. I believe the ability to boldly
+replaced a call to \highl{freeM} followed by allocating call to \highl{copy}
+with one, in-place call to \highl{copyM_to}. I believe the ability to boldly
 refactor code which manages memory is good evidence of the usefulness of

### 2019-01-11 write-up/dissertation/implementation.tex
Organise write-ups
Add more evaluation examples
Add title

--- 2018-05-29 write-up/implementation.tex
@@
 at the type-level of the DSL and its corresponding introduction (the
-\ltfla{Many}-constructor) and elimination (the \ltfla{let Many <id> = .. in ..
+\highl{Many}-constructor) and elimination (the \highl{let Many <id> = .. in ..
 } destructor) forms at the term-level of the DSL.  The idea behind the
@@
 anything which can be passed around by copying, will have a \mbox{!-type}. This
-includes integers, elements and booleans. So \ltfla{3 : !int} and \ltfla{3. +.
+includes integers, elements and booleans. So \highl{3 : !int} and \highl{3. +.
 4. : !elt}. However, all bindings are still linear by default, so to emulate
-intuitionism, I desugar \ltfla{let !x = <exp> in <body>} to \ltfla{let Many x =
+intuitionism, I desugar \highl{let !x = <exp> in <body>} to \highl{let Many x =
 <exp> in let Many x = Many (Many x) in <body>} (similarly for function argument
 bindings). The reader can check (using the rules in
-Appendix~\ref{chap:ott_spec}) that this has the effect of moving \ltfla{x : !t}
-from the linear to the intuitionistic environments, only if \ltfla{<exp> : !t}.
+Appendix~\ref{chap:ott_spec}) that this has the effect of moving \highl{x : !t}
+from the linear to the intuitionistic environments, only if \highl{<exp> : !t}.
 
@@
 an array or matrix and moving it into the intuitionistic environments. Why?
-There are certain situations in which we \emph{should not} use the \ltfla{Many}
-constructor.  Consider the following code: \ltfla{let Many x = Many (array 5)
-in <body>}; the expression \ltfla{array 5} uses no linearly-typed variables
+There are certain situations in which we \emph{should not} use the \highl{Many}
+constructor.  Consider the following code: \highl{let Many x = Many (array 5)
+in <body>}; the expression \highl{array 5} uses no linearly-typed variables
 from the environment. Although we could just reject types of the form
-\ltfla{!(_ arr)} to fix this simple example, what about pairs \ltfla{let Many
+\highl{!(_ arr)} to fix this simple example, what about pairs \highl{let Many
 xy = Many (3, array 5) in <body>}?  Ad-hoc pattern matching on the type cannot
-account for all possible situations. With the last case, we can use \ltfla{xy}
+account for all possible situations. With the last case, we can use \highl{xy}
 as many times as we would like, destruct the pair to get the second component
@@
 Not quite, but to understand how we can fix this problem, we need to question
-an assumption left implicit up until this point: what does \ltfla{Many} even
+an assumption left implicit up until this point: what does \highl{Many} even
 mean in the DSL? How is it translated to OCaml? What is the OCaml runtime
-behaviour of the DSL's \ltfla{Many}-constructor? One option is to go down the
-C++ route and make \ltfla{Many} behave like a \ltfla{shared_ptr} and act as a
+behaviour of the DSL's \highl{Many}-constructor? One option is to go down the
+C++ route and make \highl{Many} behave like a \highl{shared_ptr} and act as a
 runtime reference-count for arrays. I chose to not go for this option because
@@
 \emph{library}. In that spirit, the simplest thing that the DSL's
-\ltfla{Many}-constructor can mean in the OCaml runtime is \emph{nothing}.
-LT4LA's !/\ltfla{Many} constructors are translated into OCaml
-\ltfla{bang}/\ltfla{Many} constructors declared as \ltfla{type 'a bang = Many
+\highl{Many}-constructor can mean in the OCaml runtime is \emph{nothing}.
+LT4LA's !/\highl{Many} constructors are translated into OCaml
+\highl{bang}/\highl{Many} constructors declared as \highl{type 'a bang = Many
 of 'a [@@unboxed]}.  The \emph{unboxed} annotation means that the type and its
 constructor only exist for the purpose of type checking in OCaml; \emph{the
-OCaml runtime representation of values of type \ltfla{'a} and \ltfla{'a bang}
+OCaml runtime representation of values of type \highl{'a} and \highl{'a bang}
 is exactly the same}.
@@
 types applied to linearity instead. We then have the rule that in LT4LA, we can
-only use \ltfla{Many} on expressions that are defined to be \emph{values} and
+only use \highl{Many} on expressions that are defined to be \emph{values} and
 \emph{use no linear variables}.
@@
 represents complete ownership of a value. Creating an array gives you ownership
-of it; the function \ltfla{array : !int --o z arr} (where \ltfla{z} represents
-`0').  Once you have ownership of an array, you can free it: \ltfla{free : z
+of it; the function \highl{array : !int --o z arr} (where \highl{z} represents
+`0').  Once you have ownership of an array, you can free it: \highl{free : z
 arr --o unit}.  Importantly, because a linear value may only be used once, the
@@
 \textbf{preventing use-after-free}.  Ownership also enables you to write to the
-array: \ltfla{set : z arr --o !int --o !elt --o z arr} (the syntax \ltfla{w[i]
-:= j} is just sugar for \ltfla{set w i j}). Here, linearity prevents accessing
+array: \highl{set : z arr --o !int --o !elt --o z arr} (the syntax \highl{w[i]
+:= j} is just sugar for \highl{set w i j}). Here, linearity prevents accessing
 aliases which represented the array \emph{before} the mutation.
@@
 Any fraction less than 1 (for simplicity, limited to $2^{-k}$ in this system,
-for a positive integer $k$) represents read-only access. So, the \ltfla{'x}
-represents a natural number (either a zero \ltfla{z}, variable \ltfla{'x} or a
+for a positive integer $k$) represents read-only access. So, the \highl{'x}
+represents a natural number (either a zero \highl{z}, variable \highl{'x} or a
 successor (+1) of a natural number). Hence, you can read from (index) any array
-\ltfla{get : 'x .  'x arr --o !int --o 'x arr * !elt} (the syntax \ltfla{let !v
-<- w[i]} is just sugar for \ltfla{let (w, !v) = get _ w i}). In general, a
-left-arrow \ltfla{<-} signifies transparent rebinding with returned values: it
+\highl{get : 'x .  'x arr --o !int --o 'x arr * !elt} (the syntax \highl{let !v
+<- w[i]} is just sugar for \highl{let (w, !v) = get _ w i}). In general, a
+left-arrow \highl{<-} signifies transparent rebinding with returned values: it
 means a program can \emph{appear} to use a variable multiple times, important
@@
 Fractions exist to provide both safe aliasing and read/write permissions, via
-the primitives \ltfla{share : 'x . 'x arr --o ('x s) arr * ('x s) arr} and
-\ltfla{unshare: 'x .  ('x s) arr --o ('x s) arr --o 'x arr}.  For the former,
+the primitives \highl{share : 'x . 'x arr --o ('x s) arr * ('x s) arr} and
+\highl{unshare: 'x .  ('x s) arr --o ('x s) arr --o 'x arr}.  For the former,
 the two arrays returned (which are just aliases of the given array) can now
@@
 must use the latter to combine other read-only aliases until you are left with
-only one value of type \ltfla{z arr}, guaranteeing no other aliases exists.
+only one value of type \highl{z arr}, guaranteeing no other aliases exists.
 
@@
 linear variables from the surrounding scope. So, we can desugar something like
-\ltfla{let x = 3 in let !f (y : !int) = x + y in <body>} to \ltfla{let x = 3 in
+\highl{let x = 3 in let !f (y : !int) = x + y in <body>} to \highl{let x = 3 in
 let Many f =  Many (fun y : !int -> x + y) in  <body>}.  Recursion is slightly
 more complicated: we can desugar the factorial function
-(Figure~\ref{fig:ltfla_factorial}) to \ltfla{let Many f = fix (f, x : !int, if
-(*..*) : !int) in <body> }. However, \ltfla{fix}, like \ltfla{Many}, must also
+(Figure~\ref{fig:ltfla_factorial}) to \highl{let Many f = fix (f, x : !int, if
+(*..*) : !int) in <body> }. However, \highl{fix}, like \highl{Many}, must also
 not use any linear variables from its surrounding scope, because a
@@
     \centering
-    \inputminted[fontsize=\small]{ocaml}{../test/examples/kalman.lt}
+    \inputminted[fontsize=\small]{ocaml}{../../examples/kalman.lt}
     \caption{Kalman filter in LT4LA. In contrast with the Fortran
@@
         matrix functions. Further annotations on the matrix variables
-        (\ltfla{sym} or \ltfla{^T}) determine which BLAS routine is called and
+        (\highl{sym} or \highl{^T}) determine which BLAS routine is called and
         what parameters it is passed.}\label{fig:mat_patterns}
@@
 Code generation is a straightforward mapping from core LT4LA constructs to
-OCaml constructs, with the addition of \ltfla{Many} constructors to wrap
+OCaml constructs, with the addition of \highl{Many} constructors to wrap
 integer, element and boolean literals.
@@
 
-    \item \ltfla{let xy, z = <exp> in let x,y = xy in <body>} with
-        \ltfla{let (x,y), z = <exp> in <body>},
+    \item \highl{let xy, z = <exp> in let x,y = xy in <body>} with
+        \highl{let (x,y), z = <exp> in <body>},
 
-    \item \ltfla{let Many x = x in let Many x = Many (Many x) in <body>} with
-        \ltfla{<body>},
+    \item \highl{let Many x = x in let Many x = Many (Many x) in <body>} with
+        \highl{<body>},
 
-    \item \ltfla{let Many x = <exp> in let Many x = Many (Many x) in <body>} with
-        \ltfla{let x = <exp> in <body>},
+    \item \highl{let Many x = <exp> in let Many x = Many (Many x) in <body>} with
+        \highl{let x = <exp> in <body>},
 
-    \item \ltfla{let Many x = Many <exp> in <body>} with \ltfla{let x = <exp> in <body>},
+    \item \highl{let Many x = Many <exp> in <body>} with \highl{let x = <exp> in <body>},
 
-    \item \ltfla{let Many f = fix (f, x, .., <exp>, ..) in <body>} with
-        \ltfla{let rec f x = <exp> in <body>}.
+    \item \highl{let Many f = fix (f, x, .., <exp>, ..) in <body>} with
+        \highl{let rec f x = <exp> in <body>}.
 
@@
 
-I explained how a few core features -- linearity, the \ltfla{Many}
+I explained how a few core features -- linearity, the \highl{Many}
 constructor, value-restriction, fractional-capabilities with inference,

### 2019-01-11 write-up/paper/intro.tex
Beefed up intro; added a stronger opening to language overview

--- 2018-10-28 write-up/paper/intro.tex
@@
 
-\lang\ is a functional programming language designed to express the APIs of
-low-level linear algebra libraries (such as BLAS/LAPACK) safely and explicitly.
-It does so by combining linear types, fractional permissions, runtime errors
-and recursion into a small, easily understandable, yet expressive set of core
-constructs.
+Programmers writing numerical software often find themselves caught on
+the horns of a dilemma. The foundational, low-level linear algebra
+libraries such as BLAS and LAPACK offer programmers very precise
+control over the memory lifetime and usage of vector and matrix
+values. However, this power comes paired with the responsibility to
+manually manage the memory associated with each array object, and in
+addition to bringing in the familiar difficulties of reasoning about
+lifetimes, aliasing and sharing that plague low-level systems
+programming, this also moves the APIs away from the linear-algebraic,
+mathematical style of thinking that numerical programmers want to use.
 
-\lang\ allows a novice to understand and work with complicated linear algebra
-library APIs, as well as point out subtle aliasing bugs and reduce memory usage
-in existing programs. In fact, we were able to use \lang\ to find linearity and
-aliasing bugs in a linear algebra program that was \emph{generated} by another
-program \emph{specifically designed to translate matrix expressions into an
-efficient sequence of calls linear algebra routines}. We were also able to
-reduce the number of temporaries used by the same program, using \lang's type
-system to guide us.
+As a result, programmers often turn to higher-level languages such as
+Matlab, R and Numpy, which offer very high level array abstractions
+that can be viewed as ordinary mathematical values. This makes
+programming safer, as well as making prototyping and verification much
+easier, since it lets programmers write programs which bear a closer
+resemblance to the formulas that the mathematicians and statisticians
+designing these algorithms prefer to work with, and ensures that
+program bugs will reflect incorrectly-computed values rather than heap
+corruption.
+
+The intention is that these languages can use libraries BLAS and
+LAPACK, without having to expose programmers to explicit memory
+management. However, this benefit comes at a price: because user
+programs do not worry about aliasing, the language implementations
+cannot in general exploit the underlying features of the low-level
+libraries that let them explicitly manage and reuse memory. As a
+result, programs written in high-level statistical languages can
+be much less memory-efficient than programs that make full use
+of the powers the low-level APIs offer.
+
+So in practice, programmers face a tradeoff: they can eschew safety
+and exploit the full power of the underlying linear algebra libraries,
+or they can obtain safety at the price of unneeded copies and worse
+memory efficiency. In this work, show that this tradeoff is not a
+fundamental one.
+
+\lang\ is a functional programming language designed to express the
+APIs of low-level linear algebra libraries (such as BLAS/LAPACK)
+safely and explicitly.  It does so by combining linear types,
+fractional permissions, runtime errors and recursion into a small,
+easily understandable, yet expressive set of core constructs.
+
+\lang\ allows a novice to understand and work with complicated linear
+algebra library APIs, as well as point out subtle aliasing bugs and
+reduce memory usage in existing programs. In fact, we were able to use
+\lang\ to find linearity and aliasing bugs in a linear algebra
+algorithm that was \emph{generated} by another program
+\emph{specifically designed to translate matrix expressions into an
+  efficient sequence of calls to linear algebra routines}. We were
+also able to reduce the number of temporaries used by the same
+algorithm, using \lang's type system to guide us.
 

### 2019-01-11 write-up/paper/lang_and_examples.tex
Beefed up intro; added a stronger opening to language overview

@@
 
-Linearity is at the heart of \lang. Linearity allows us to express a
-pure-functional API for numerical library routines that mutate arrays and
-matrices. Linearity also restricts aliasing of (values which represent)
-pointers.
+The core type theory of \lang\ is a nearly off-the-shelf linear type
+theory, supporting familiar features such as linear function spaces
+$A \multimap B$ and tensor products $A \otimes B$. We adopt linearity
+-- the restriction that each program variable be used at most once --
+since it allows us to express purely functional APIs for numerical
+library routines that mutate arrays and matrices~\cite{Wadler90}. Due
+to linearity, values cannot alias and are only used once, which means
+that linearly-typed updates result in no \emph{observable} mutation.
+
+As a result, programmers can reason about \lang\ expressions as if they
+were ordinary mathematical expressions -- as indeed they are! We are
+merely adopting a stricter type discipline than usual to make managing
+memory safe.
 

### 2019-01-11 write-up/paper/discussion.tex
Typo fix

@@
 Futhark~\cite{henriksen} and Single Assignment C~\cite{scholz} omit
-higher-order functions facilitating compilation to GPUs. The work of
+higher-order functions to facilitate compilation to GPUs. The work of
 \cite{bernardy2016} forbids term-level recursion, in order to ensure
@@
 
-Using fractional permissions directly, as we did, rather than relying
-on a complex has two main benefits. First, our type system
-demonstrates that type systems for fractional permissions can be
-\emph{dramatically} simpler than existing state-of-the-art approaches,
-including both industrial languages like Rust, as well as the academic
-languages developed by \cite{bierhoff}.  Their type system, much like
-Rust, build a complex dataflow analysis into the typing rules to infer
-when variables can be shared or not. This allows for very
-natural-looking user programs, but creates the impression that using
-fractional permissions requires a heavy theoretical and engineering
-effort going well beyond that needed for supporting basic linear
-types.
+Working explicitly with fractional permissions has two main
+benefits. First, our type system demonstrates that type systems for
+fractional permissions can be dramatically simpler than existing
+state-of-the-art approaches, including both industrial languages like
+Rust, as well as academic (such as those developed by
+\cite{bierhoff}).  Bierhoff \emph{et al}'s type system, much like
+Rust's, builds a complex dataflow analysis into the typing rules to
+infer when variables can be shared or not. This allows for more
+natural-looking user programs, but can create the impression that
+using fractional permissions requires a heavy theoretical and
+engineering effort going well beyond that needed for supporting basic
+linear types.
 
-Instead, our approach of requiring sharing to be made explicit means
-that the existing unification machinery already in place for ordinary
-ML-style type inference, can be reused to support fractions. Basically,
-we can view sharing a value as dividing a fraction by two, and after
-taking logarithms all fractions are Peano numbers, whose equality can
-be established with ordinary unification.
+Instead, our approach, of requiring sharing to be made explicit, lets
+us demonstrate that the existing unification machinery already in
+place for ordinary ML-style type inference can be reused to support
+fractions. Basically, we can view sharing a value as dividing a
+fraction by two, and after taking logarithms all fractions are Peano
+numbers, whose equality can be established with ordinary unification.
 
@@
 bindings in Linear Haskell. Actually implementing this is something we
-leave for future work, as there remains an issue which we do not see a
-good encoding for. Namely, only having support for linear functions
+leave for future work, as there remains one issue which we do not see
+a good encoding for. Namely, only having support for linear functions
 makes it a bit inconvenient to manipulate linear values directly --
 programs end up taking on a CPS-like structure. This seems to remain
-an advantage of a direct implementation of linear types.
+an advantage of a direct implementation of linear types over the
+Linear Haskell style.
 
@@
 
-We could also attempt to catch mismatched dimensions at compile time as well.
-While this could be done with generative phantom types\cite{abe2015simple},
-using dependent types may offer more flexibility in \emph{partitioning}
-regions\cite{space_monads} or statically enforcing dimensions related
-constraints of the arguments at compile-time.  ATS\cite{cui2005ats} is just
-such a language which combines dependent and linear types; although it provides
-BLAS bindings, it does not aim to provide aliasing restrictions as demonstrated
-in this paper.
+We could also attempt to catch mismatched dimensions at compile time
+as well.  While this could be done with generative phantom
+types\cite{abe2015simple}, using dependent types may offer more
+flexibility in \emph{partitioning} regions\cite{space_monads} or
+statically enforcing dimensions related constraints of the arguments
+at compile-time.  ATS\cite{cui2005ats} is an example of a language
+which combines linear types with a sophisticated proof layer. But
+although it provides BLAS bindings, it does not aim to provide
+aliasing restrictions as demonstrated in this paper.
 

### 2019-01-11 write-up/paper/intro.tex
Tidy up punctuation and figures

@@
 lifetimes, aliasing and sharing that plague low-level systems
-programming, this also moves the APIs away from the linear-algebraic,
+programming; this also moves the APIs away from the linear-algebraic,
 mathematical style of thinking that numerical programmers want to use.
@@
 As a result, programmers often turn to higher-level languages such as
-Matlab, R and Numpy, which offer very high level array abstractions
+Matlab, R and Numpy, which offer very high-level array abstractions
 that can be viewed as ordinary mathematical values. This makes

### 2019-01-11 write-up/paper/formal_system.tex
Beefed up intro; added a stronger opening to language overview
Use theorem environment

@@
 
-A typing judgement consists of $ \Theta; \Delta; \Gamma
-\vdash e : t$.
+\begin{itemize}
 
-$\Theta$ is the environment that tracks which fractional permission variables
+    \item A typing judgement consists of $ \Theta; \Delta; \Gamma \vdash e : t$.
+
+    \item $\Theta$ is the environment that tracks which fractional permission variables
         in scope. Fractional permissions (the \textsf{Perm} judgement) and types (the
@@
 
-$\Delta$ is the environment storing non-linearly or \emph{inuitionistically}
+    \item $\Delta$ is the environment storing non-linearly or \emph{inuitionistically}
         typed variables.
 
-$\Gamma$ is the environment storing linearly typed variables.  Note that rules
-for typing $()$, booleans, integers and elements are typed with respect to an
-\emph{empty} linear environment: this means no linear values are needed to
-produce a value of those types.
+    \item $\Gamma$ is the environment storing linearly typed variables.
+
+\end{itemize}
+
+Note that rules for typing $()$, booleans, integers and elements are typed
+with respect to an \emph{empty} linear environment: this means no linear
+values are needed to produce a value of those types.
 
@@
 well as the n-fold iteration of $\rightarrow$.
+
 \[
@@
 
+\begin{theorem}
+(The Fundamental Lemma of Logical Relations)
 \[
@@
 \]
+\end{theorem}
 

### 2019-01-12 write-up/paper/lang_and_examples.tex
Update figure captions

@@
     \centering
-%   \begin{minipage}{.4\textwidth}
-%   \centering
     \inputminted[fontsize=\small]{ocaml}{../../examples/kalman.lt}
-%   \end{minipage}
-%   \begin{minipage}{.4\textwidth}
-%   \centering
-%   \begin{minted}[linenos,fontsize=\small]{ocaml}
-%   let owl_kalman
-%       ~sigma
-%       ~h
-%       ~mu
-%       ~r
-%       ~data =
-%     let open Owl.Mat in
-%     let ( * ) = dot in
-%     let h' = transpose h in
-%     let sigma_h' = sigma * h' in
-%     let x = sigma_h' * (inv @@ r + h * sigma_h') in
-%     let new_mu = mu + x * (h * mu - data) in
-%     let new_sigma = sigma - x * h * sigma in
-%     new_sigma, new_mu
-%   ;;
-%   \end{minted}
-%   \end{minipage}
     \caption{Kalman filter: see Figure~\ref{fig:kalman_eqns} for the
-        equations this code implements. Line numbers in comments refer to
-        equivalent lines in a C implementation
-        (Figure~\ref{fig:cblas_kalman}).}\label{fig:lang_kalman}
+        equations this code implements and Figure~\ref{fig:cblas_kalman}
+        for an equivalent \textsc{Cblas/Lapacke} implementation.}\label{fig:lang_kalman}
 \end{figure}

### 2019-03-13 write-up/response.txt
Initial outline of response

# Questions:

## Questions common to all the reviews, and big questions

## Reviewer-specific questions

### Review A

* The exact research contribution of this paper is, however, unclear
  to me. Linear types as well as fractional permissions have been used
  before.  Is the contibution that these two type systems are applied
  together, or that they are applied to a new domain? If it is the
  latter, it would be helpful if the paper could point out differences
  to previous type systems, or challenges, if there are any.

* I have an additional question: how were the types of the BLAS/LAPACK
  routines obtained? Did you just annotate the APIs based on
  assumptions documented somewhere, or where these APIs also
  type-checked?

* Finally, the related work focuses almost exclusively on type
  systems. Since ultimately the goal is to provide (memory) safe and
  fast linear algebra, I think it would be nice to also mention
  non-type based efforts, e.g. Spiral in Scala which compiles to
  efficient code [1], even though it does not use BLAS/LAPACK.  [1]
  Spiral in Scala: Towards the Systematic Construction of Generators
  for Performance Libraries. GPCE ’13

### Review B

* One question I have is why OCaml code is generated? If the program
  only uses higher-order functions in certain ways, would it be
  possible to generate C code?

* Also, could the aliasing requirements be checked dynamically? I.e.,
  provide the OCaml interface in Appendix G, but with dynamic aliasing
  checks? What overhead would that have?

* I was confused by the desugaring "let !x = e1 in e2" => `let Many x
  = e1 in let Many x = Many (Many x) in e2`. Can you explain this
  more?

* The statement to line 302 talks about moving values from the linear
  environment to the intuitionistic one. But only variables live in
  environments?

* I was initially confused by the presence (in Figure 11) of the
  seemling useless wrappers around banged values. Is this to make the
  API of the generated OCaml code easier to use correctly?

* I was confused by the comment on Linear Haskell starting on line
  610. What relevance does a system based on kinds have to NumLin
  (which isn't based on kinds either)?

### Review C

* How exactly could they incorporate this knowledge of what is
  legal/illegal wrt Fortran in the Numlin framework? Or was it spotted
  incidentally when they examined the permissions under which an
  external routine was attempted to be invoked?

* Compared to the learning curve of Rust (which the authors cite) what
  is the learning curve of Numlin, in the authors' estimation?

* Could the authors sketch the evolution of linear types from its
  initial classical introduction (say in Wadler's "Linear types can
  change the world" paper) to its current sophisticated form? Is this
  really an "off the shelf" linear type system?

* Can they explain their soundness theorem once in English?

* My biggest comment is concerning concurrency. These sort of routines
  will be used in a setting where multicore concurrency will be
  employed to perform concurrent updates on arrays. In that setting,
  one might employ separation logic to deal with safe concurrent
  accesses. Now, separation logic also employs fractional
  permissions. Would Numlin be usable in that setting? Would the use
  of Separation Logic and Numlin (Linear Types) be orthogonal? Or do
  the authors feel the need to work out a concurrency extension of
  Numlin?

----

ECOOP '19 Paper #35 Reviews and Comments
===========================================================================
Paper #35 NumLin: Linear Types for Linear Algebra

Review #35A
===========================================================================

Overall merit
-------------
C. Weak reject: I think this paper should be rejected but I will not argue
   strongly against it

Reviewer expertise
------------------
Y. Knowledgeable

Paper summary
-------------
The paper presents a language and type system called NumLin for writing linear algebra programs which use BLAS/LAPACK routines.  NumLin's type system is based on linear and fractional permission types and ensures correct memory usage and aliasing. The performance achieved is comparable to an unsafe C implementation.

Comments for author
-------------------
The paper targets an important domain; high-performance linear algebra routines are used in many different applications.  However, common libraries such as BLAS or LAPACK are written in low-level languages and require careful memory management. Providing more (memory) safety without sacrificing performance is thus of practical importance.  
The paper demonstrates how to provide this safety using a type system, and provides a prototype solution in the form of a  domain-specific language which compiles down to OCaml. Thus, it provides a practical implementation which is of potential interest to a wide audience.

The exact research contribution of this paper is, however, unclear to me. Linear types as well as fractional permissions have been used before.  Is the contibution that these two type systems are applied together, or that they are applied to a new domain? If it is the latter, it would be helpful if the paper could point out differences to previous type systems, or challenges, if there are any.

My lack of understanding of the contribution is probably also due to the fact that I found the paper hard to read:
- For instance, the exact usage of the language was unclear to me initiatially. The beginning of the paper talks about expressing the APIs of BLAS/LAPACK libraries, but much what follows is about writing programs (using those APIs), but the APIs are hidden. 
- Similarly, the overview of NumLin in section 2 read like a background section, but did not really explain to me how this fits together in NumLin.
- The examples in section 2.2 hinted at different features, but left too many details for later to help understanding. I found the examples were trying to introduce too much at the same time: syntax of the language, syntactic sugar, different BLAS routines which are supported, as well as the type system. For me, it would have been clearer to discuss these (more) individually, and I believe this would make the paper more accessible to a wider audience.

I have an additional question: how were the types of the BLAS/LAPACK routines obtained? Did you just annotate the APIs based on assumptions documented somewhere, or where these APIs also type-checked?

Finally, the related work focuses almost exclusively on type systems. Since ultimately the goal is to provide (memory) safe and fast linear algebra, I think it would be nice to also mention non-type based efforts, e.g. Spiral in Scala which compiles to efficient code [1], even though it does not use BLAS/LAPACK.
[1] Spiral in Scala: Towards the Systematic Construction of Generators for Performance Libraries. GPCE ’13

Review #35B
===========================================================================

Overall merit
-------------
A. Accept: I will champion this paper

Reviewer expertise
------------------
X. Expert

Paper summary
-------------
The APIs of low level linear algebra libraries, like LAPACK and BLAS go to great lengths to allow users full flexibility in how memory to store matricies is allocated, freed, and reused. However, this comes at the cost of potential memory safety bugs as the API does not automatically enforce the (non-)aliasing assumed by the implementations of the routines. High-level wrappers around these APIs ensure memory safety, but at the cost of additional temporary storage and defensive copying.

This paper proposes a "middle way": a high level programming language that directly uses the low-level routines, and uses a linear type system to ensure that the (non-)aliasing constraints required by the low-level routines are satisfied. The system is proved sound using a step-indexed logical relation. An implementation has been constructed that generates OCaml code, and it has been benchmarked against high-level code in Owl (an OCaml library) and NumPy (a Python library). The benchmark results show a convincing speedup in cases where the execution time is not dominated by the low level routines themselves.

Comments for author
-------------------
This paper introduces a non-trivial system to solve a real problem, provides both a proof of soundness and an implementation, and uses that implementation to run a convincing evaluation. I recommend this paper for acceptance.

One question I have is why OCaml code is generated? If the program only uses higher-order functions in certain ways, would it be possible to generate C code?

Also, could the aliasing requirements be checked dynamically? I.e., provide the OCaml interface in Appendix G, but with dynamic aliasing checks? What overhead would that have?

Points in favour of this paper:

  - The type system is introduced in a clear and easy to follow way, with the linear aspects of the type system introduced in a logical order. Several examples are given, progressing from fibonacci to relatively involved examples, showing that the library is actually usable for real work.

  - There is a soundness proof, carried out using step-indexed logical relations. While this is an adaptation of previous work by Morrisett et al., it is non-trivial to apply it to this system, and it includes the novel feature (for a type system like this) of using fractional permissions.

  - The inclusion of fractional permissions into a unification based type inference system appears to be novel, and while the paper does take pains to not claim that this feature has been formally proved complete, it does make the implementation feasible to use for real examples.

  - The existence of an implementation means that a good empirical evaluation can be carried out, and this shows that the NumLin system does indeed have the claimed benefits of reducing the overhead of generating temporary arrays and low-level copying.

Things that could be improved / suggestions:

  - The type system appears to be based on Andrew Barber's DILL (Dual Intuitionistic Linear Logic), but this is not referenced.

  - I was confused by the desugaring "let !x = e1 in e2" => `let Many x = e1 in let Many x = Many (Many x) in e2`. Can you explain this more?

  - The statement to line 302 talks about moving values from the linear environment to the intuitionistic one. But only variables live in environments?

  - The "value" restriction in the !-introduction rule should be explained more, with an example that shows a "bad" use. I guess the restriction arises from the existence of allocating primtivies in the language, such as `array`, and allowing these in the !-introduction rule would permit unrestricted duplication of references to newly allocated elements (another way round this would be to have a token passed around encapsulating the ability to allocate). Ahmed et al. (2007) [1] also have a similar value restriction for the same reason (I also think this would be a better reference than the Morrisett et al. TLCA2005 paper).

  - The types of some of the primitive should be given in the paper's main body, so that it is possible to see how to type the aliasing restrictions required for a given routine. As given, the reader has to reverse engineer the NumLin type from the OCaml type given in the appendix.

  - I was initially confused by the presence (in Figure 11) of the seemling useless wrappers around banged values. Is this to make the API of the generated OCaml code easier to use correctly?

  - I was confused by the comment on Linear Haskell starting on line 610. What relevance does a system based on kinds have to NumLin (which isn't based on kinds either)?

Typos:

  - Line 327: "follow" => "follows"

  - Line 494: Missing words at the start of the line.

[1] Amal Ahmed, Matthew Fluet, Greg Morrisett: L3: A Linear Language with Locations. Fundam. Inform. 77(4): 397-449 (2007)

Review #35C
===========================================================================

Overall merit
-------------
A. Accept: I will champion this paper

Reviewer expertise
------------------
Z. Outsider

Paper summary
-------------
This paper is on the use of linear types in a new programming language NUMLIN to express a large array of important algorithms involving aggregate data types (arrays primarily). Existing array based programs end up introducing many errors that are hard to detect including incorrect aliases. There also are sources of inefficiency including useless temporaries. The use of linear types that these authors propose elegantly characterizes the usage disciplines intended and also helps catch these bugs.

Comments for author
-------------------
While I am an outsider, I really find the use of linear types to be quite fascinating in terms of how it has advanced since the 80s and now applies to practical routines. The authors do a fantastic job covering a whole gamut of challenging case studies.

NUMLIN appears to impose some modest burden on part of the programmer to "correctly" annotate a program as to the intended uses of the underlying array types. Given this, the typing system can infer permissions and check for correct usage of aggregate data structures. For instance, "free" or "set" need full permissions. Multiple read uses of an array are supported via fractional permissions.

In terms of results, the slowdown is not that pronounced even compared to a C implementation. They also detected an incorrect usage in the setting of the Kalman filter wrt Fortran conventions.

Here are my suggestions and questions:

1) How exactly could they incorporate this knowledge of what is legal/illegal wrt Fortran in the Numlin framework? Or was it spotted incidentally when they examined the permissions under which an external routine was attempted to be invoked?

2) Page 2: achieving parity with C -> explain this as parity wrt runtime efficiency (no other parity with C, I suppose)

3) Compared to the learning curve of Rust (which the authors cite) what is the learning curve of Numlin, in the authors' estimation?

4) Could the authors sketch the evolution of linear types from its initial classical introduction (say in Wadler's "Linear types can change the world" paper) to its current sophisticated form? Is this really an "off the shelf" linear type system? Giving a history of the evolution of linear types to the kind used in Numlin would be very valuable

5) Can they explain their soundness theorem once in English? This can help novices read the paper better

6) My biggest comment is concerning concurrency. These sort of routines will be used in a setting where multicore concurrency will be employed to perform concurrent updates on arrays. In that setting, one might employ separation logic to deal with safe concurrent accesses. Now, separation logic also employs fractional permissions. Would Numlin be usable in that setting? Would the use of Separation Logic and Numlin (Linear Types) be orthogonal? Or do the authors feel the need to work out a concurrency extension of Numlin? This would be highly relevant to discuss, as otherwise it is unclear how we are going to employ multicores on these class of routines (run on a single core, the performance would be terrible).

### 2019-03-13 write-up/response.txt
Initial stuff on the response

@@
 
-## Questions common to all the reviews, and big questions
+## Questions common to multiple reviews
+
+1. Reviews A and C both ask how we got the Numlin types for the
+   BLAS/LAPACK routines.
+
+The documentation records the expected aliasing behaviour of each of
+these routines, and whether or not the routine modifies or consumes
+its argument in any way. We used that to derive the types we ascribed
+to the low-level routines. Since most of these low-level routines are
+very careful not to do any allocation themselves, it was generally
+very easy to give them a Numlin type -- every argument that can
+modify/consume its argument needs a full permission, and all others
+can be fraction-polymorphic.
 
@@
 
-* I have an additional question: how were the types of the BLAS/LAPACK
-  routines obtained? Did you just annotate the APIs based on
-  assumptions documented somewhere, or where these APIs also
-  type-checked?
+We certainly agree that both linear types and fractional permissions
+have been studied before. To our surprise, it did not appear that
+linear types had been much applied to checking linear algebra
+libraries. It was only when we tried to use it, that we realized that
+we needed fractional permissions.
+
+In our view, this explains why it hadn't been used before: many prior
+type systems for fractional permissions (eg, from Boyland's original
+work to Aldrich and his students' work on Plural) are extremely
+complex. This is because these type systems typically encode a complex
+analysis to automatically infer how fractional permissions should be
+split and rejoined.
+
+By requiring sharing and merging to be made explicit, we were able to
+drastically simplify the type system. Our formal system is very close
+to standard presentations of linear logic, and the implementation
+complexity is no worse than HM type inference.
 
@@
 
+We're happy to discuss this in the related work. However, NumLin is
+more likely to be a consumer of efforts like Spiral than a competitor:
+we want to check whether client code conforms to the aliasing rules
+that low-level libraries (whether from platform libraries like BLAS or
+generated via systems like Spiral) want clients to adhere to.
+
 ### Review B
@@
 
+There's no technical obstacle to generating C code, but as we had
+local access to the developer of the Owl numerical programming library
+for Ocaml, it made sense to use Ocaml. Eventually we'd like to extend
+Owl with something like Numlin to offer safe access to the low-level
+platform numerics libraries.
+
 * Also, could the aliasing requirements be checked dynamically? I.e.,
@@
   checks? What overhead would that have?
+
+It might be possible, but we're not sure how. Our system enforces full
+linearity (to support explicit deallocation), and to our knowledge
+contract systems for linearity checking (eg, Tov and Pucella's ESOP
+2010 "Stateful Contracts for Affine Types") only enforce affine usage
+(ie, a dynamic check can enforce at-most-one, but not exactly-once).
 
@@
 
-* The statement to line 302 talks about moving values from the linear
-  environment to the intuitionistic one. But only variables live in
-  environments?
+[TODO]
 
@@
 
+Yes, this is correct. Ocaml's underlying value representation does not
+allocate anything for unary datatypes, so there is no performance
+penalty for using them like Haskell-style newtypes.
+
+* The statement to line 302 talks about moving values from the linear
+  environment to the intuitionistic one. But only variables live in
+  environments?
+
+[TODO]
+
 * I was confused by the comment on Linear Haskell starting on line
@@
 
+We were trying to explain that (a) it seems like we can use GADTs with
+natural-number indices to encode fractions a la Numlin in Linear
+Haskell, but (b) we don't see how to support linear values (since
+Linear Haskell only supports a linear arrow). We're happy to tighten
+up the discussion in the final version.
+
 ### Review C
-
-* How exactly could they incorporate this knowledge of what is
-  legal/illegal wrt Fortran in the Numlin framework? Or was it spotted
-  incidentally when they examined the permissions under which an
-  external routine was attempted to be invoked?
 
@@
   is the learning curve of Numlin, in the authors' estimation?
+
+The basic intuitions about ownership are similar, so we expect an
+experienced Rust programmer would pick up Numlin pretty
+easily. However, Rust is a much bigger language, supporting a lot more
+features than linear types, so the other direction would be more work!
 
@@
 
+We'll add this historical discussion to the final paper. Our system is
+indeed fairly close to off-the-shelf, and as we mentioned to reviewer
+A, it was one of the surprising discoveries of this paper that this
+works.
+
 * Can they explain their soundness theorem once in English?
+
+[TODO]
 
@@
   Numlin?
+
+Numlin's type system enforces a concurrent-reader/exclusive-writer
+pattern through the use of fractions, and our heap model was
+deliberately chosen to be quite similar to the work on fractions from
+separation logic. As a result, we would expect that proving that it
+works in the presence of concurrency would be relatively
+straightforward -- though without doing the work we don't want to say
+that it would definitely work.
+
+That said, developing a variant of separation logic suitable for
+Numlin would be interesting, because it would let us prove things
+about (eg) array bounds without having to throw them into the type
+system.
 

### 2019-03-14 write-up/response.md
Finish TODOs in response

--- 2019-03-13 write-up/response.txt
@@
 
-1. Reviews A and C both ask how we got the Numlin types for the
+* Reviews A and C both ask how we got the NumLin types for the
   BLAS/LAPACK routines.
@@
 very careful not to do any allocation themselves, it was generally
-very easy to give them a Numlin type -- every argument that can
+very easy to give them a NumLin type -- every argument that can
 modify/consume its argument needs a full permission, and all others
-can be fraction-polymorphic.
+can be fraction-polymorphic. Taking Fortran as an example, it has
+a notion of `in`, `out` and `inout` parameters. The latter two would
+need full permissions, the first would be fraction-polymorphic.
 
@@
   Spiral in Scala: Towards the Systematic Construction of Generators
-  for Performance Libraries. GPCE ’13
+  for Performance Libraries. GPCE '13
 
@@
 that low-level libraries (whether from platform libraries like BLAS or
-generated via systems like Spiral) want clients to adhere to.
+generated via systems like Spiral, Cl1ck, FLAME) want clients to adhere to.
 
@@
 local access to the developer of the Owl numerical programming library
-for Ocaml, it made sense to use Ocaml. Eventually we'd like to extend
-Owl with something like Numlin to offer safe access to the low-level
+for OCaml, it made sense to use OCaml. Eventually we'd like to extend
+Owl with something like NumLin to offer safe access to the low-level
 platform numerics libraries.
@@
 
-* I was confused by the desugaring "let !x = e1 in e2" => `let Many x
+* I was confused by the desugaring `let !x = e1 in e2` => `let Many x
   = e1 in let Many x = Many (Many x) in e2`. Can you explain this
@@
 
-[TODO]
+There are two aspects to answering this, one is the type-based justification
+and the second is *why* that's even necessary. Starting with why, the primary
+motivation for having this was for binary arithmetic operations (of type
+`!int * !int --o !int`). For the sake of ergonomics, we'd like to be able to write
+code like "let x = 5 * 5 in x * x"; this is morally right because integers and
+their operations are !-types but this is non-linear in x. Intuitively, we wish
+for a rule like the following to "just do the right thing" when
+destructuring/evaluating expressions of a !-type; which we can derive as a
+meta-rule using the desugaring in question.
+
+```
+Int; Lin |- e : !t 
+Int, x : !t; Lin' |- e' : t'
+ ---------------------------------------
+Int; Lin, Lin' |- let !x = e in e' : t'
+```
 
@@
 
-Yes, this is correct. Ocaml's underlying value representation does not
-allocate anything for unary datatypes, so there is no performance
-penalty for using them like Haskell-style newtypes.
+Yes, this is correct, we essentially embed the NumLin type system into the
+OCaml one for a sanity check on the output code, ease of interop with OCaml and
+tooling such as Merlin support.  Annotating with "[@@unboxed]" means OCaml's
+underlying value representation does not allocate anything for unary datatypes,
+so there is no performance penalty for using them like Haskell-style newtypes.
 
@@
 
-[TODO]
+Yes, that's correct, this is a typo, we meant to write "moves the variable
+referring to that value from the linear environment to the intuitionistic one".
 
@@
 We were trying to explain that (a) it seems like we can use GADTs with
-natural-number indices to encode fractions a la Numlin in Linear
+natural-number indices to encode fractions a la NumLin in Linear
 Haskell, but (b) we don't see how to support linear values (since
@@
 * Compared to the learning curve of Rust (which the authors cite) what
-  is the learning curve of Numlin, in the authors' estimation?
+  is the learning curve of NumLin, in the authors' estimation?
 
-The basic intuitions about ownership are similar, so we expect an
-experienced Rust programmer would pick up Numlin pretty
-easily. However, Rust is a much bigger language, supporting a lot more
-features than linear types, so the other direction would be more work!
+We conjecture that the learning curve of NumLin is less steep than that of
+Rust, for two reaons: Rust supports more than just linear types and is smarter
+about inferring the appropriate ownership. We suspect the explicit nature of
+NumLin makes it easier to pick up and sets up an easier transition to Rust's
+more complex system. We will add a mention of Weiss et. al. (2018/19) which
+proposes a fraction-based framework to understand/model Rust's borrow checker.
 
@@
 
-[TODO]
+We'll add an explanation of the following terse explanation to the final paper.
+If an expression is syntactically type-checked, then for an arbitrary number of
+steps (k), under *any* substitution of free fractional-permission variables,
+linear variables (with a suitable heap) and intuitionistic variables, that
+suitable heap and substituted expression are in the computational
+interpretation (explained in English in the paper) of the (substituted) type of
+that expression.
 
@@
   accesses. Now, separation logic also employs fractional
-  permissions. Would Numlin be usable in that setting? Would the use
-  of Separation Logic and Numlin (Linear Types) be orthogonal? Or do
+  permissions. Would NumLin be usable in that setting? Would the use
+  of Separation Logic and NumLin (Linear Types) be orthogonal? Or do
   the authors feel the need to work out a concurrency extension of
-  Numlin?
+  NumLin?
 
-Numlin's type system enforces a concurrent-reader/exclusive-writer
+NumLin's type system enforces a concurrent-reader/exclusive-writer
 pattern through the use of fractions, and our heap model was
@@
 That said, developing a variant of separation logic suitable for
-Numlin would be interesting, because it would let us prove things
+NumLin would be interesting, because it would let us prove things
 about (eg) array bounds without having to throw them into the type
 system.
-
-----
-
-ECOOP '19 Paper #35 Reviews and Comments
-===========================================================================
-Paper #35 NumLin: Linear Types for Linear Algebra
-
-Review #35A
-===========================================================================
-
-Overall merit
--------------
-C. Weak reject: I think this paper should be rejected but I will not argue
-   strongly against it
-
-Reviewer expertise
-------------------
-Y. Knowledgeable
-
-Paper summary
--------------
-The paper presents a language and type system called NumLin for writing linear algebra programs which use BLAS/LAPACK routines.
-NumLin's type system is based on linear and fractional permission types and ensures correct memory usage and aliasing.
-The performance achieved is comparable to an unsafe C implementation.
-
-Comments for author
--------------------
-The paper targets an important domain; high-performance linear algebra routines are used in many different applications.
-However, common libraries such as BLAS or LAPACK are written in low-level languages and require careful memory management.
-Providing more (memory) safety without sacrificing performance is thus of practical importance.
-The paper demonstrates how to provide this safety using a type system, and provides a prototype solution in the form of a  domain-specific language which compiles down to OCaml.
-Thus, it provides a practical implementation which is of potential interest to a wide audience.
-
-The exact research contribution of this paper is, however, unclear to me.
-Linear types as well as fractional permissions have been used before.
-Is the contibution that these two type systems are applied together, or that they are applied to a new domain?
-If it is the latter, it would be helpful if the paper could point out differences to previous type systems, or challenges, if there are any.
-
-My lack of understanding of the contribution is probably also due to the fact that I found the paper hard to read:
-- For instance, the exact usage of the language was unclear to me initiatially.
-The beginning of the paper talks about expressing the APIs of BLAS/LAPACK libraries, but much what follows is about writing programs (using those APIs), but the APIs are hidden.
-- Similarly, the overview of NumLin in section 2 read like a background section, but did not really explain to me how this fits together in NumLin.
-- The examples in section 2.2 hinted at different features, but left too many details for later to help understanding.
-I found the examples were trying to introduce too much at the same time: syntax of the language, syntactic sugar, different BLAS routines which are supported, as well as the type system.
-For me, it would have been clearer to discuss these (more) individually, and I believe this would make the paper more accessible to a wider audience.
-
-I have an additional question: how were the types of the BLAS/LAPACK routines obtained? Did you just annotate the APIs based on assumptions documented somewhere, or where these APIs also type-checked?
-
-Finally, the related work focuses almost exclusively on type systems.
-Since ultimately the goal is to provide (memory) safe and fast linear algebra, I think it would be nice to also mention non-type based efforts, e.g.
-Spiral in Scala which compiles to efficient code [1], even though it does not use BLAS/LAPACK.
-[1] Spiral in Scala: Towards the Systematic Construction of Generators for Performance Libraries. GPCE ’13
-
-Review #35B
-===========================================================================
-
-Overall merit
--------------
-A. Accept: I will champion this paper
-
-Reviewer expertise
-------------------
-X. Expert
-
-Paper summary
--------------
-The APIs of low level linear algebra libraries, like LAPACK and BLAS go to great lengths to allow users full flexibility in how memory to store matricies is allocated, freed, and reused.
-However, this comes at the cost of potential memory safety bugs as the API does not automatically enforce the (non-)aliasing assumed by the implementations of the routines.
-High-level wrappers around these APIs ensure memory safety, but at the cost of additional temporary storage and defensive copying.
-
-This paper proposes a "middle way": a high level programming language that directly uses the low-level routines, and uses a linear type system to ensure that the (non-)aliasing constraints required by the low-level routines are satisfied.
-The system is proved sound using a step-indexed logical relation.
-An implementation has been constructed that generates OCaml code, and it has been benchmarked against high-level code in Owl (an OCaml library) and NumPy (a Python library).
-The benchmark results show a convincing speedup in cases where the execution time is not dominated by the low level routines themselves.
-
-Comments for author
--------------------
-This paper introduces a non-trivial system to solve a real problem, provides both a proof of soundness and an implementation, and uses that implementation to run a convincing evaluation.
-I recommend this paper for acceptance.
-
-One question I have is why OCaml code is generated? If the program only uses higher-order functions in certain ways, would it be possible to generate C code?
-
-Also, could the aliasing requirements be checked dynamically? I.e., provide the OCaml interface in Appendix G, but with dynamic aliasing checks? What overhead would that have?
-
-Points in favour of this paper:
-
-  - The type system is introduced in a clear and easy to follow way, with the linear aspects of the type system introduced in a logical order.
-Several examples are given, progressing from fibonacci to relatively involved examples, showing that the library is actually usable for real work.
-
-  - There is a soundness proof, carried out using step-indexed logical relations.
-While this is an adaptation of previous work by Morrisett et al., it is non-trivial to apply it to this system, and it includes the novel feature (for a type system like this) of using fractional permissions.
-
-  - The inclusion of fractional permissions into a unification based type inference system appears to be novel, and while the paper does take pains to not claim that this feature has been formally proved complete, it does make the implementation feasible to use for real examples.
-
-  - The existence of an implementation means that a good empirical evaluation can be carried out, and this shows that the NumLin system does indeed have the claimed benefits of reducing the overhead of generating temporary arrays and low-level copying.
-
-Things that could be improved / suggestions:
-
-  - The type system appears to be based on Andrew Barber's DILL (Dual Intuitionistic Linear Logic), but this is not referenced.
-
-  - I was confused by the desugaring "let !x = e1 in e2" => `let Many x = e1 in let Many x = Many (Many x) in e2`. Can you explain this more?
-
-  - The statement to line 302 talks about moving values from the linear environment to the intuitionistic one. But only variables live in environments?
-
-  - The "value" restriction in the !-introduction rule should be explained more, with an example that shows a "bad" use.
-I guess the restriction arises from the existence of allocating primtivies in the language, such as `array`, and allowing these in the !-introduction rule would permit unrestricted duplication of references to newly allocated elements (another way round this would be to have a token passed around encapsulating the ability to allocate).
-Ahmed et al.
-(2007) [1] also have a similar value restriction for the same reason (I also think this would be a better reference than the Morrisett et al.
-TLCA2005 paper).
-
-  - The types of some of the primitive should be given in the paper's main body, so that it is possible to see how to type the aliasing restrictions required for a given routine.
-As given, the reader has to reverse engineer the NumLin type from the OCaml type given in the appendix.
-
-  - I was initially confused by the presence (in Figure 11) of the seemling useless wrappers around banged values. Is this to make the API of the generated OCaml code easier to use correctly?
-
-  - I was confused by the comment on Linear Haskell starting on line 610. What relevance does a system based on kinds have to NumLin (which isn't based on kinds either)?
-
-Typos:
-
-  - Line 327: "follow" => "follows"
-
-  - Line 494: Missing words at the start of the line.
-
-[1] Amal Ahmed, Matthew Fluet, Greg Morrisett: L3: A Linear Language with Locations. Fundam. Inform. 77(4): 397-449 (2007)
-
-Review #35C
-===========================================================================
-
-Overall merit
--------------
-A. Accept: I will champion this paper
-
-Reviewer expertise
-------------------
-Z. Outsider
-
-Paper summary
--------------
-This paper is on the use of linear types in a new programming language NUMLIN to express a large array of important algorithms involving aggregate data types (arrays primarily).
-Existing array based programs end up introducing many errors that are hard to detect including incorrect aliases.
-There also are sources of inefficiency including useless temporaries.
-The use of linear types that these authors propose elegantly characterizes the usage disciplines intended and also helps catch these bugs.
-
-Comments for author
--------------------
-While I am an outsider, I really find the use of linear types to be quite fascinating in terms of how it has advanced since the 80s and now applies to practical routines.
-The authors do a fantastic job covering a whole gamut of challenging case studies.
-
-NUMLIN appears to impose some modest burden on part of the programmer to "correctly" annotate a program as to the intended uses of the underlying array types.
-Given this, the typing system can infer permissions and check for correct usage of aggregate data structures.
-For instance, "free" or "set" need full permissions.
-Multiple read uses of an array are supported via fractional permissions.
-
-In terms of results, the slowdown is not that pronounced even compared to a C implementation. They also detected an incorrect usage in the setting of the Kalman filter wrt Fortran conventions.
-
-Here are my suggestions and questions:
-
-1) How exactly could they incorporate this knowledge of what is legal/illegal wrt Fortran in the Numlin framework?
-Or was it spotted incidentally when they examined the permissions under which an external routine was attempted to be invoked?
-
-2) Page 2: achieving parity with C -> explain this as parity wrt runtime efficiency (no other parity with C, I suppose)
-
-3) Compared to the learning curve of Rust (which the authors cite) what is the learning curve of Numlin, in the authors' estimation?
-
-4) Could the authors sketch the evolution of linear types from its initial classical introduction (say in Wadler's "Linear types can change the world" paper) to its current sophisticated form?
-Is this really an "off the shelf" linear type system?
-Giving a history of the evolution of linear types to the kind used in Numlin would be very valuable
-
-5) Can they explain their soundness theorem once in English? This can help novices read the paper better
-
-6) My biggest comment is concerning concurrency.
-These sort of routines will be used in a setting where multicore concurrency will be employed to perform concurrent updates on arrays.
-In that setting, one might employ separation logic to deal with safe concurrent accesses.
-Now, separation logic also employs fractional permissions.
-Would Numlin be usable in that setting?
-Would the use of Separation Logic and Numlin (Linear Types) be orthogonal?
-Or do the authors feel the need to work out a concurrency extension of Numlin?
-This would be highly relevant to discuss, as otherwise it is unclear how we are going to employ multicores on these class of routines (run on a single core, the performance would be terrible).

### 2019-03-14 write-up/response.md
Updates to response

@@
 
-## Questions common to multiple reviews
+## Major Questions
 
@@
 
+* Reviewer A asks:
+
+    > The exact research contribution of this paper is, however, unclear
+    > to me. Linear types as well as fractional permissions have been used
+    > before.  Is the contibution that these two type systems are applied
+    > together, or that they are applied to a new domain? If it is the
+    > latter, it would be helpful if the paper could point out differences
+    > to previous type systems, or challenges, if there are any.
+
+    We certainly agree that both linear types and fractional
+permissions have been studied before. To our surprise, it did not
+appear that linear types had been much applied to checking linear
+algebra libraries. It was only when we tried to use it, that we
+realized that we needed fractional permissions.
+
+    In our view, this explains why it hadn't been used before: many prior
+type systems for fractional permissions (eg, from Boyland's original
+work to Aldrich and his collaborators' work on Plural) are extremely
+complex. This is because these type systems typically encode a complex
+analysis to automatically infer how fractional permissions should be
+split and rejoined.
+
+    By requiring sharing and merging to be made explicit, we were able
+to drastically simplify the type system. Our formal system is very
+close to standard presentations of linear logic, and the
+implementation complexity is no worse than HM type inference.
+
+* Reviewer C asks:
+
+    > My biggest comment is concerning concurrency. These sort of routines
+    > will be used in a setting where multicore concurrency will be
+    > employed to perform concurrent updates on arrays. In that setting,
+    > one might employ separation logic to deal with safe concurrent
+    > accesses. Now, separation logic also employs fractional
+    > permissions. Would NumLin be usable in that setting? Would the use
+    > of Separation Logic and NumLin (Linear Types) be orthogonal? Or do
+    > the authors feel the need to work out a concurrency extension of
+    > NumLin?
+
+    NumLin's type system enforces a concurrent-reader/exclusive-writer
+pattern through the use of fractions (a pattern widely used in
+concurrent programming), and our heap model was deliberately chosen to
+be quite similar to the work on fractions from separation logic. As a
+result, we would expect that the language would work unchanged in the
+concurrent setting, and that extending the proof to show that it works
+in the presence of concurrency would be relatively straightforward --
+though without doing the work we don't want to say for sure that it
+would definitely work.
+
+    That said, developing a variant of separation logic suitable for
+NumLin would be interesting, because it would let us prove things
+about (eg) array bounds without having to throw them into the type
+system.
+
+----
+
 ## Reviewer-specific questions
@@
 
-* The exact research contribution of this paper is, however, unclear
-  to me. Linear types as well as fractional permissions have been used
-  before.  Is the contibution that these two type systems are applied
-  together, or that they are applied to a new domain? If it is the
-  latter, it would be helpful if the paper could point out differences
-  to previous type systems, or challenges, if there are any.
-
-We certainly agree that both linear types and fractional permissions
-have been studied before. To our surprise, it did not appear that
-linear types had been much applied to checking linear algebra
-libraries. It was only when we tried to use it, that we realized that
-we needed fractional permissions.
-
-In our view, this explains why it hadn't been used before: many prior
-type systems for fractional permissions (eg, from Boyland's original
-work to Aldrich and his students' work on Plural) are extremely
-complex. This is because these type systems typically encode a complex
-analysis to automatically infer how fractional permissions should be
-split and rejoined.
-
-By requiring sharing and merging to be made explicit, we were able to
-drastically simplify the type system. Our formal system is very close
-to standard presentations of linear logic, and the implementation
-complexity is no worse than HM type inference.
-
-* Finally, the related work focuses almost exclusively on type
-  systems. Since ultimately the goal is to provide (memory) safe and
-  fast linear algebra, I think it would be nice to also mention
-  non-type based efforts, e.g. Spiral in Scala which compiles to
-  efficient code [1], even though it does not use BLAS/LAPACK.  [1]
-  Spiral in Scala: Towards the Systematic Construction of Generators
-  for Performance Libraries. GPCE '13
+* > Finally, the related work focuses almost exclusively on type
+  > systems. Since ultimately the goal is to provide (memory) safe and
+  > fast linear algebra, I think it would be nice to also mention
+  > non-type based efforts, e.g. Spiral in Scala which compiles to
+  > efficient code [1], even though it does not use BLAS/LAPACK.  [1]
+  > Spiral in Scala: Towards the Systematic Construction of Generators
+  > for Performance Libraries. GPCE '13
 
@@
 
-* One question I have is why OCaml code is generated? If the program
+* > One question I have is why OCaml code is generated? If the program
   only uses higher-order functions in certain ways, would it be
@@
 
-* Also, could the aliasing requirements be checked dynamically? I.e.,
+* > Also, could the aliasing requirements be checked dynamically? I.e.,
   provide the OCaml interface in Appendix G, but with dynamic aliasing
@@
 
-* I was confused by the desugaring `let !x = e1 in e2` => `let Many x
+* > I was confused by the desugaring `let !x = e1 in e2` => `let Many x
   = e1 in let Many x = Many (Many x) in e2`. Can you explain this
@@
 `!int * !int --o !int`). For the sake of ergonomics, we'd like to be able to write
-code like "let x = 5 * 5 in x * x"; this is morally right because integers and
+code like "let x = 5 + 5 in x - x"; this is morally right because integers and
 their operations are !-types but this is non-linear in x. Intuitively, we wish
@@
 
-```
+  > ```
 Int; Lin |- e : !t 
@@
 
-* I was initially confused by the presence (in Figure 11) of the
+   It should be possible to handle this using the Benton/Wadler
+LNL-style presentation of linear types, which uses adjoint modalities
+to distinguish between intrinsically linear and intrinsically
+nonlinear types, but that is a pretty big digression from our main
+goal.
+
+* > I was initially confused by the presence (in Figure 11) of the
   seemling useless wrappers around banged values. Is this to make the
@@
 
-* The statement to line 302 talks about moving values from the linear
+* > The statement to line 302 talks about moving values from the linear
   environment to the intuitionistic one. But only variables live in
@@
 
-* I was confused by the comment on Linear Haskell starting on line
-  610. What relevance does a system based on kinds have to NumLin
+* > I was confused by the comment on Linear Haskell starting on line 610. 
+  What relevance does a system based on kinds have to NumLin
   (which isn't based on kinds either)?
@@
 
-* Compared to the learning curve of Rust (which the authors cite) what
+* > Compared to the learning curve of Rust (which the authors cite) what
   is the learning curve of NumLin, in the authors' estimation?
@@
 
-* Could the authors sketch the evolution of linear types from its
+* > Could the authors sketch the evolution of linear types from its
   initial classical introduction (say in Wadler's "Linear types can
@@
 
-* Can they explain their soundness theorem once in English?
+* > Can they explain their soundness theorem once in English?
 
-We'll add an explanation of the following terse explanation to the final paper.
-If an expression is syntactically type-checked, then for an arbitrary number of
-steps (k), under *any* substitution of free fractional-permission variables,
-linear variables (with a suitable heap) and intuitionistic variables, that
-suitable heap and substituted expression are in the computational
-interpretation (explained in English in the paper) of the (substituted) type of
-that expression.
+    We'll add an expansion of the following terse explanation to the
+final paper. If an expression is syntactically type-checked, then for
+an arbitrary number of steps (k), under *any* substitution of free
+fractional-permission variables, linear variables (with a suitable
+heap) and intuitionistic variables, that suitable heap and substituted
+expression are in the computational interpretation (explained in
+English in the paper) of the (substituted) type of that expression.
 
-* My biggest comment is concerning concurrency. These sort of routines
-  will be used in a setting where multicore concurrency will be
-  employed to perform concurrent updates on arrays. In that setting,
-  one might employ separation logic to deal with safe concurrent
-  accesses. Now, separation logic also employs fractional
-  permissions. Would NumLin be usable in that setting? Would the use
-  of Separation Logic and NumLin (Linear Types) be orthogonal? Or do
-  the authors feel the need to work out a concurrency extension of
-  NumLin?
-
-NumLin's type system enforces a concurrent-reader/exclusive-writer
-pattern through the use of fractions, and our heap model was
-deliberately chosen to be quite similar to the work on fractions from
-separation logic. As a result, we would expect that proving that it
-works in the presence of concurrency would be relatively
-straightforward -- though without doing the work we don't want to say
-that it would definitely work.
-
-That said, developing a variant of separation logic suitable for
-NumLin would be interesting, because it would let us prove things
-about (eg) array bounds without having to throw them into the type
-system.
+    The "computational interpretation" just identifies executions that do
+no un- or ill-defined behaviours (eg, adding a boolean or an integer).
+Since our operational semantics explicitly models deallocation, we now
+know no well-typed program will ever try to access deallocated memory,
+establishing the correctness of our memory management checking.

### 2019-03-14 write-up/response.md
Update response.md

@@
 `!int * !int --o !int`). For the sake of ergonomics, we'd like to be able to write
-code like "let x = 5 + 5 in x - x"; this is morally right because integers and
+code like `let x = 5 + 5 in x - x`; this is morally right because integers and
 their operations are !-types but this is non-linear in x. Intuitively, we wish
@@
 
-  > ```
+    ```
     Int; Lin |- e : !t
@@
 fractional-permission variables, linear variables (with a suitable
-heap) and intuitionistic variables, that suitable heap and substituted
-expression are in the computational interpretation (explained in
-English in the paper) of the (substituted) type of that expression.
+heap) and intuitionistic variables, that suitable heap and expression
+are in the computational interpretation of the type of that expression.
 

### 2019-03-14 write-up/response.md
Remove separator line
Heading suffices to demarcate sections.
Update response.md

@@
     In our view, this explains why it hadn't been used before: many prior
-type systems for fractional permissions (eg, from Boyland's original
+type systems for fractional permissions (e.g. from Boyland's original
 work to Aldrich and his collaborators' work on Plural) are extremely
@@
 NumLin would be interesting, because it would let us prove things
-about (eg) array bounds without having to throw them into the type
+about (for example) array bounds without having to throw them into the type
 system.
-
-----
 
@@
 linearity (to support explicit deallocation), and to our knowledge
-contract systems for linearity checking (eg, Tov and Pucella's ESOP
+contract systems for linearity checking (e.g. Tov and Pucella's ESOP
 2010 "Stateful Contracts for Affine Types") only enforce affine usage
@@
 OCaml one for a sanity check on the output code, ease of interop with OCaml and
-tooling such as Merlin support.  Annotating with "[@@unboxed]" means OCaml's
+tooling support (eg Merlin).  Annotating with `[@@unboxed]` means OCaml's
 underlying value representation does not allocate anything for unary datatypes,
@@
     The "computational interpretation" just identifies executions that do
-no un- or ill-defined behaviours (eg, adding a boolean or an integer).
+no un- or ill-defined behaviours (e.g. adding a boolean or an integer).
 Since our operational semantics explicitly models deallocation, we now

### 2019-04-09 write-up/paper/intro.tex
Fix for shepherding: improve introduction
to a) emphasise that the main contribution of this paper is the
definition of the NumLin language and its application as a safe language
for invoking BLAS routines, and b) to clarify the relationship to
previous work on linear types and fractional permissions (as explained
in the response).

@@
 
-\lang\ is a functional programming language designed to express the
-APIs of low-level linear algebra libraries (such as BLAS/LAPACK)
-safely and explicitly.  It does so by combining linear types,
-fractional permissions, runtime errors and recursion into a small,
-easily understandable, yet expressive set of core constructs.
+\lang\ is a functional programming language whose type system is
+designed to enforce the safe usage of the APIs of low-level linear algebra
+libraries (such as BLAS/LAPACK).  It does so by combining linear types,
+fractional permissions, runtime errors and recursion into a small, easily
+understandable, yet expressive set of core constructs.
 

### 2019-04-09 write-up/paper/paper.tex
Add title
Beefed up intro; added a stronger opening to language overview
Fix for shepherding: improve introduction
to a) emphasise that the main contribution of this paper is the
definition of the NumLin language and its application as a safe language
for invoking BLAS routines, and b) to clarify the relationship to
previous work on linear types and fractional permissions (as explained
in the response).

@@
 
+\maketitle%
+
 \begin{abstract}
-    We present \lang, a functional programming language designed to express
-    the APIs of low-level linear algebra libraries (such as BLAS/LAPACK) safely
-    and explicitly, through a brief description of its key features and several
-    illustrative examples. We show that \lang's type system is sound and that
-    its implementation improves upon na{\"i}ve implementations of linear
-    algebra programs, almost towards C-levels of performance. Lastly, we contrast
-    it to other recent developments in linear types and show that using linear
-    types and fractional permissions to express the APIs of low-level linear
-    algebra libraries is a simple and effective idea.
+    We present \lang, a functional programming language whose type system is
+    designed to enforce the safe usage of the APIs of low-level linear algebra
+    libraries (such as BLAS/LAPACK). We do so through a brief description of
+    its key features and several illustrative examples. We show that \lang's
+    type system is sound and that its implementation improves upon na{\"i}ve
+    implementations of linear algebra programs, almost towards C-levels of
+    performance. By doing so, we demonstrate (a) that linear types are
+    well-suited to expressing the APIs of low-level linear algebra libraries
+    accurately and concisely and (b) that, despite the complexity of prior work
+    on it, fractional permissions can actually be implemented using simple,
+    well-known techniques and be used practically in real programs.
 \end{abstract}

### 2019-04-09 write-up/paper/formal_system.tex
Fix for shepherding: accessible soundness result

@@
 \]
+
 \end{theorem}
+
+\subsubsection{Explanation}
+
+If an expression is syntactically type-checked, then for an arbitrary number of
+steps $k$, under any substitution $\theta$ of free fractional-permission
+variables, linear variables $\gamma$ (with a suitable heap $\varsigma$) and
+intuitionistic variables $\delta$, that suitable heap and expression
+$(\varsigma, \theta(\delta(\gamma(e))))$ are in the computational
+interpretation of the type of that expression $\C{k}{ \theta(t) }$.
+
+The \emph{computational interpretation} is as defined before: a combination of
+a frame rule on heaps (no interference), type-preservation and termination (in
+$j < k$ steps) to either an error or a heap-and-expression, with the further
+condition that if the expression is a syntactic value then it is also one
+semantically.
+
+It identifies executions that do no un- or ill-defined behaviours (e.g.\ adding
+a boolean and an integer). Since our operational semantics explicitly models
+deallocation, we now know no well-typed program will ever try to access
+deallocated memory, establishing the correctness of our memory management
+checking.
+
+\subsubsection{Proof Sketch}
 

### 2019-04-12 write-up/paper/implementation.tex
Tidy up punctuation and figures
Fix formatting

@@
 
-\subsubsection{Desugaring, Matrix Expressions and Type Checking}\label{subsubsec:initial}
+\subsubsection{Desugaring, Matrix Expressions and Type Checking}
 
@@
 
-\subsubsection{Type checking} is mostly standard for a linearly typed language,
+\subsubsection{Type checking}
+
+Type checking is mostly standard for a linearly typed language,
 with the exception of fractional permission inference. By restricting fractions

### 2019-04-12 write-up/paper/lang_and_examples.tex
Fix formatting

@@
 
-\subsubsection{Factorial}
+\subsubsection{Factorial}\label{subsubsec:factorial}
 
@@
 syntactic convenience for declaring the value to used intuitionistically, its
-full and precise meaning is described in Section~\ref{subsubsec:initial}.
+full and precise meaning is described in Section~\ref{subsec:core_tt}.
 

### 2019-04-12 write-up/paper/formal_system.tex
Fix formatting
Fix for shepherding: improve let !x = e in e'

@@
 
-\subsection{Core Type Theory}
+\subsection{Core Type Theory}\label{subsec:core_tt}
 
@@
 
+\vspace{-\baselineskip}
 \[
@@
 
-Consuming a !-type value \emph{moves it} from the linear environment $\Gamma$
-and \emph{into} the intuitionistic environment $\Delta$. This is exactly why
-$\mathbf{let}\ !x = e_1\ \mathbf{in}\ e_2$ desugars to $\mathbf{let\ Many}\ x =
-e_1\ \mathbf{in\ } \mathbf{let\ Many}\ x = \mathbf{Many}\ (\mathbf{Many}\ x)\
-\mathbf{in}\ e_2$.
+Consuming a variable that refers to a !-type value \emph{moves it} from the
+linear environment $\Gamma$ and \emph{into} the intuitionistic environment
+$\Delta$.
 
+\vspace{-\baselineskip}
 \[
     \ottdruleTyXXBangXXElim{}
+\]
+
+Using this, we can explain how the ! annotation on variables -- first
+introduced in the factorial example in \ref{subsubsec:factorial} -- works. That
+is, we can explain why the meaning of $\mathbf{let}\ !x = e\ \mathbf{in}\ e'$
+can be expressed using only the rules presented thus far, as $\mathbf{let\
+Many}\ x = e\ \mathbf{in\ } \mathbf{let\ Many}\ x = \mathbf{Many}\
+(\mathbf{Many}\ x)\ \mathbf{in}\ e'$.
+\footnote{\emph{Why} we have this at all is for the sake of ergonomics when
+    using binary arithmetic operations (for example, of type
+    $!\ottkw{int} \, \otimes \, !\ottkw{int} \multimap \; !\ottkw{int}$): a
+    programmer should be able to write \highl{let x = 5 + 5 in x - x},
+    which, although non-linear in \highl{x}, is morally right because
+    integers and operations on them rarely need to be linear. Though it
+    should be possible to handle this using a LNL-style presentation of
+    linear types~\cite{benton1995} (using adjoint modalities to distinguish
+    between intrinsically linear and intrinsically intuitionistic types) that
+    is a pretty big digression from the stated goals of this paper.}
+The reader is invited to quickly convince themselves that the following
+meta-rule is provable using \textsc{Ty\_Bang\_Intro} (twice),
+\textsc{Ty\_Bang\_Elim} (twice) and weakening the intuitionistic environment
+$\Delta$ (once).
+
+\vspace{-\baselineskip}
+\[
+\ottdrule{%
+\ottpremise{\Theta  \ottsym{;}  \Delta  \ottsym{;}  \Gamma  \vdash  \ottnt{e}  \ottsym{:}  \: !  \ottnt{t}}%
+\ottpremise{\Theta  \ottsym{;}  \Delta  \ottsym{,}  \mathit{x}  \ottsym{:}  \: !  \ottnt{t}  \ottsym{;}  \Gamma'  \vdash  \ottnt{e'}  \ottsym{:}  \ottnt{t'}}%
+}{
+\Theta  \ottsym{;}  \Delta  \ottsym{;}  \Gamma  \ottsym{,}  \Gamma'  \vdash  \ottkw{let} \, ! \mathit{x} \ottsym{=} \ottnt{e} \, \ottkw{in} \, \ottnt{e'}  \ottsym{:}  \ottnt{t'}}{%
+{\ottdrulename{Meta\_Ty\_Bang\_Var\_Elim}}{}%
+}
 \]
@@
 heap and an expression (\textsc{Op\_Frac\_Perm}).
+
+\vspace{-\baselineskip}
 \[
@@
 to reflect that $l$ represents a complete ownership over the new matrix.
+
+\vspace{-\baselineskip}
 \[
@@
 before removing it and the matrix it points to from the heap.
+
+\vspace{-\baselineskip}
 \[
@@
 \textsc{Op\_Share} and \textsc{Op\_Unshare\_Eq} are fairly natural.
+
+\vspace{-\baselineskip}
 \[
@@
 \]
+
+\vspace{-\baselineskip}
 \[
@@
 that are written to and $\mapsto_{f\!c}$ for anything else.
+
+\vspace{-\baselineskip}
 \[
@@
 \]
+
 where
+
+\vspace{-\baselineskip}
 \[
@@
 (\textsc{Op\_Spc}).
+
+\vspace{-\baselineskip}
 \[
@@
 \]
-
 \end{theorem}
@@
 $\rightarrow$.  Basically, the following is not quite true:
+
+\vspace{-\baselineskip}
 \[
@@
 \]
+
 because after the $i$ steps, we need to invoke \textsc{Op\_Let\_Var} to proceed

### 2019-04-12 write-up/paper/discussion.tex
Fix formatting

@@
 which were marked as \texttt{intent(in)} but would have been written over by
-calls to `gemm', spotted by the fractional-capabilities feature. Furthermore,
+calls to `gemm', spotted by the fractional capabilities feature. Furthermore,
 it used a matrix \emph{twice} in a call to `symm', once with a read permission
@@
 illegal according to the standard, both aspects of which were captured by
-linearity and fractional-capabilities.
+linearity and fractional capabilities.
 
@@
 
-Using linear types for BLAS routines is a particularly good domain fit
-(given the implicit restrictions on aliasing arguments), and as a
-result the idea of using substructural types to express array
-computations is not a particularly new
-one\cite{scholz,henriksen,bernardy2016}.  However, many of these
-designs have been focused on building languages to \emph{implement}
-the kernel linear algebra functions, and as a result, they tend to add
-additional limitations on the language design. Both
-Futhark~\cite{henriksen} and Single Assignment C~\cite{scholz} omit
-higher-order functions to facilitate compilation to GPUs. The work of
-\cite{bernardy2016} forbids term-level recursion, in order to ensure
-that all higher-order computations can be statically normalized away
-and thereby maximize opportunities for array fusion.
+Using linear types for BLAS routines is a particularly good domain fit (given
+the implicit restrictions on aliasing arguments), and as a result the idea of
+using substructural types to express array computations is not a particularly
+new one~\cite{scholz,henriksen,bernardy2016}.  However, many of these designs
+have been focused on building languages to \emph{implement} the kernel linear
+algebra functions, and as a result, they tend to add additional limitations on
+the language design. Both Futhark~\cite{henriksen} and Single Assignment
+C~\cite{scholz} omit higher-order functions to facilitate compilation to GPUs.
+The work of \cite{bernardy2016} forbids term-level recursion, in order to
+ensure that all higher-order computations can be statically normalized away and
+thereby maximize opportunities for array fusion.
 
-In contrast, our approach is to begin with the assumption that we can
-take existing efficient BLAS-like libraries, and then enforce their
-correct \emph{usage} using a linear type discipline with fractional
-permissions.
+In contrast, our approach is to begin with the assumption that we can take
+existing efficient BLAS-like libraries, and then enforce their correct
+\emph{usage} using a linear type discipline with fractional permissions.
 
-This approach is similar to the one taken in linear algebra libraries
-for Rust -- these libraries typically take advantage of the
-distinction that Rust's type system offers between mutable
-views/references to arrays.  The work of \cite{weiss} and
-\cite{rustbelt} suggest that Rust's borrow-checker can be expressed
-in simpler terms using \emph{fractional-permissions}, though to our
-knowledge the programmer-visible lifetime analysis in Rust has never
-been formalized.
+This approach is similar to the one taken in linear algebra libraries for Rust
+-- these libraries typically take advantage of the distinction that Rust's type
+system offers between mutable views/references to arrays.  The work of
+\cite{weiss} and \cite{rustbelt} suggest that Rust's borrow-checker \emph{can
+be expressed in simpler terms} using fractional permissions, though to our
+knowledge the programmer-visible lifetime analysis in Rust has never been
+formalized.
 
-Working explicitly with fractional permissions has two main
-benefits. First, our type system demonstrates that type systems for
-fractional permissions can be dramatically simpler than existing
-state-of-the-art approaches, including both industrial languages like
-Rust, as well as academic (such as those developed by
-\cite{bierhoff}).  Bierhoff \emph{et al}'s type system, much like
-Rust's, builds a complex dataflow analysis into the typing rules to
-infer when variables can be shared or not. This allows for more
-natural-looking user programs, but can create the impression that
-using fractional permissions requires a heavy theoretical and
-engineering effort going well beyond that needed for supporting basic
-linear types.
+Working explicitly with fractional permissions has two main benefits. First,
+our type system demonstrates that type systems for fractional permissions can
+be dramatically simpler than existing state-of-the-art approaches, including
+both industrial languages like Rust, as well as academic (such as those
+developed by \cite{bierhoff}).  Bierhoff \emph{et al}'s type system, much like
+Rust's, builds a complex dataflow analysis into the typing rules to infer when
+variables can be shared or not. This allows for more natural-looking user
+programs, but can create the impression that using fractional permissions
+requires a heavy theoretical and engineering effort going well beyond that
+needed for supporting basic linear types.
 
-Instead, our approach, of requiring sharing to be made explicit, lets
-us demonstrate that the existing unification machinery already in
-place for ordinary ML-style type inference can be reused to support
-fractions. Basically, we can view sharing a value as dividing a
-fraction by two, and after taking logarithms all fractions are Peano
-numbers, whose equality can be established with ordinary unification.
+Instead, our approach, of requiring sharing to be made explicit, lets us
+demonstrate that the existing unification machinery already in place for
+ordinary ML-style type inference can be reused to support fractions. Basically,
+we can view sharing a value as dividing a fraction by two, and after taking
+logarithms all fractions are Peano numbers, whose equality can be established
+with ordinary unification.
 
-This fact is important because there are major upcoming
-implementations of linear types such as Linear
-Haskell~\cite{bernardy2017linear}, which do not have built-in support
-for fractional permissions. Instead, Linear Haskell takes a slightly
-different definition of linearity, one based on \emph{arrows} as
-opposed to \emph{kinds}: for $f : a \multimap b$, if $f u$ is used
-exactly once \emph{then} $u$ is used exactly once. Whilst this has the
-advantage of being backwards-compatible, it also means that the type
-system has no built-in support for the concurrent reader, exclusive
-writer pattern that fractional permissions enable.
+This fact is important because there are major upcoming implementations of
+linear types such as Linear Haskell~\cite{bernardy2017linear}, which do not
+have built-in support for fractional permissions. Instead, Linear Haskell takes
+a slightly different definition of linearity, one based on \emph{arrows} as
+opposed to \emph{kinds}: for $f : a \multimap b$, if $f u$ is used exactly once
+\emph{then} $u$ is used exactly once. Whilst this has the advantage of being
+backwards-compatible, it also means that the type system has no built-in
+support for the concurrent reader, exclusive writer pattern that fractional
+permissions enable.
 
-However, since our type system shows demonstrates unification is ``all
-one needs'' for fractions, it should be possible to \emph{encode}
-\lang's approach to fractional permissions in Linear Haskell by adding
-a GADT-style natural number index to array types tracking the
-fraction, which should enable supporting high-performance BLAS
-bindings in Linear Haskell. Actually implementing this is something we
-leave for future work, as there remains one issue which we do not see
-a good encoding for. Namely, only having support for linear functions
-makes it a bit inconvenient to manipulate linear values directly --
-programs end up taking on a CPS-like structure. This seems to remain
-an advantage of a direct implementation of linear types over the
-Linear Haskell style.
+However, since our type system shows demonstrates unification is ``all one
+needs'' for fractions, it should be possible to \emph{encode} \lang's approach
+to fractional permissions in Linear Haskell by adding a GADT-style natural
+number index to array types tracking the fraction, which should enable
+supporting high-performance BLAS bindings in Linear Haskell. Actually
+implementing this is something we leave for future work, as there remains one
+issue which we do not see a good encoding for. Namely, only having support for
+linear functions makes it a bit inconvenient to manipulate linear values
+directly -- programs end up taking on a CPS-like structure. This seems to
+remain an advantage of a direct implementation of linear types over the Linear
+Haskell style.
 
@@
 by extending its type system. For example, we could have used existential types
-to statically track pointer identities\cite{morrisett}, or parametric
+to statically track pointer identities~\cite{morrisett}, or parametric
 polymorphism.
 
-We could also attempt to catch mismatched dimensions at compile time
-as well.  While this could be done with generative phantom
-types\cite{abe2015simple}, using dependent types may offer more
-flexibility in \emph{partitioning} regions\cite{space_monads} or
-statically enforcing dimensions related constraints of the arguments
-at compile-time.  ATS\cite{cui2005ats} is an example of a language
-which combines linear types with a sophisticated proof layer. But
-although it provides BLAS bindings, it does not aim to provide
+We could also attempt to catch mismatched dimensions at compile time as well.
+While this could be done with generative phantom types~\cite{abe2015simple},
+using dependent types may offer more flexibility in \emph{partitioning}
+regions~\cite{space_monads} or statically enforcing dimensions related
+constraints of the arguments at compile-time.  ATS~\cite{cui2005ats} is an
+example of a language which combines linear types with a sophisticated proof
+layer. But although it provides BLAS bindings, it does not aim to provide
 aliasing restrictions as demonstrated in this paper.
@@
 fixed at runtime, we could \emph{stage} \lang\ programs and compile matrix
-expressions using more sophisticated algorithms\cite{barthels}. However, it is
-worth noting that without care, such algorithms\cite{rocklin_thesis}, usually
+expressions using more sophisticated algorithms~\cite{barthels}. However, it is
+worth noting that without care, such algorithms~\cite{rocklin_thesis}, usually
 based on graph-based, ad-hoc dataflow analysis, can produce erroneous output
@@
 types to achieve even more expressive power to split up a single block of
-memory into multiple regions in an arbitrary manner\cite{space_monads}.
+memory into multiple regions in an arbitrary manner~\cite{space_monads}.

### 2019-04-12 write-up/paper/formal_system.tex
Fix formatting

@@
 Many}\ x = e\ \mathbf{in\ } \mathbf{let\ Many}\ x = \mathbf{Many}\
-(\mathbf{Many}\ x)\ \mathbf{in}\ e'$.
-\footnote{\emph{Why} we have this at all is for the sake of ergonomics when
-    using binary arithmetic operations (for example, of type
-    $!\ottkw{int} \, \otimes \, !\ottkw{int} \multimap \; !\ottkw{int}$): a
-    programmer should be able to write \highl{let x = 5 + 5 in x - x},
-    which, although non-linear in \highl{x}, is morally right because
-    integers and operations on them rarely need to be linear. Though it
-    should be possible to handle this using a LNL-style presentation of
-    linear types~\cite{benton1995} (using adjoint modalities to distinguish
-    between intrinsically linear and intrinsically intuitionistic types) that
-    is a pretty big digression from the stated goals of this paper.}
+(\mathbf{Many}\ x)\ \mathbf{in}\ e'$.\footnote{\emph{Why} we have this at all
+    is for the sake of ergonomics when using binary arithmetic operations
+    (e.g.\ of type $!\ottkw{int} \, \otimes \, !\ottkw{int} \multimap \;
+    !\ottkw{int}$): a programmer should be able to write \highl{let x = 5 + 5
+    in x - x}, which, although non-linear in \highl{x}, is morally right
+    because integers and operations on them rarely need to be linear. Though it
+    should be possible to handle this using a LNL-style presentation of linear
+    types~\cite{benton1995} (using adjoint modalities to distinguish between
+    intrinsically linear and intrinsically intuitionistic types) that is a
+    pretty big digression from the stated goals of this paper.}
 The reader is invited to quickly convince themselves that the following
@@
 \Theta  \ottsym{;}  \Delta  \ottsym{;}  \Gamma  \ottsym{,}  \Gamma'  \vdash  \ottkw{let} \, ! \mathit{x} \ottsym{=} \ottnt{e} \, \ottkw{in} \, \ottnt{e'}  \ottsym{:}  \ottnt{t'}}{%
-{\ottdrulename{Meta\_Ty\_Bang\_Var\_Elim}}{}%
+{\ottdrulename{Meta\_Ty\_Let\_Bang}}{}%
 }
@@
 induction hypothesis on the second sub-expression to finish the proof.  To do
-so, we need to construct a valid linear substitution and heap (i.e., one in
+so, we need to construct a valid linear substitution and heap (one in
 $\den{L}{k}{\Gamma', x : t}\theta$). We take the other `half' of the linear

### 2019-04-12 write-up/paper/implementation.tex
Fix formatting

@@
 
-Desugaring is conventional, outlined in Appendix~\ref{sec:lang_desugar}.  Matrix
-expression are translated into BLAS/ LAPACK calls via purely syntactic
-pattern-matching, outlined in Figure~\ref{fig:lang_matexp}.
+As seen earlier (Figure~\ref{fig:lang_desugar}), desugaring is conventional.
+Matrix expressions are translated into BLAS/LAPACK calls via purely syntactic
+pattern-matching (also seen earlier in Figure~\ref{fig:lang_matexp}).
 
@@
 
-\begin{figure}[t]
-\begin{center}
-\[\def\arraystretch{1.3}
-    \begin{array}{rcl}
-    \mathbf{let\ } v \leftarrow x[e]\ \mathbf{in\ } e &
-    \Rightarrow &
-    \mathbf{let\ } (x, !v)\ = x[e]\ \mathbf{in\ } e \qquad \textrm{(similarly for matrices)}
-\\
-    \mathbf{let\ } x_2 \leftarrow \mathbf{new}\ [|\ x_1\ |]\ \mathbf{in\ } e &
-    \Rightarrow &
-    \mathbf{let\ } (x_1, x_2)\ = \mathbf{copyM\ } \_\ x_1\ \mathbf{in\ } e
-\\
-    \mathbf{let\ } x_2 \leftarrow [|\ x_1\ |]\ \mathbf{in\ } e &
-    \Rightarrow &
-    \mathbf{let\ } (x_1, x_2)\ = \mathbf{copyM\_to\ } \_\  x_1\ x_2\ \mathbf{in\ } e
-    \end{array}
-\]
-\[
-    M ::= X\ \mid\ X^T\ \mid\ \textrm{sym}( X )
-\]
-\end{center}
-\begin{align*}
-% new
-    \mathbf{let\ } &Y \leftarrow \mathbf{new}\ (n, k)\ [|\ \alpha M_1 M_2 \ |]\ \mathbf{in\ } e
-    \Rightarrow \\
-    & \mathbf{let\ } Y\ = \mathbf{matrix\ } n\ k\ \mathbf{in}
-    \ \mathbf{let\ } Y \leftarrow [|\ \alpha M_1 M_2 + 0Y\ |]\ \mathbf{in\ } e
-\\[0.1\baselineskip]
-%syrk
-    \mathbf{let\ } &Y \leftarrow [|\ \alpha X X^T + \beta Y\ |]\ \mathbf{in\ } e
-    \Rightarrow \\
-    & \mathbf{let\ } (X,Y)\ = \mathbf{syrk\ } \mathbf{false}
-    \ \alpha\ \_\ X\ \beta\ Y\ \mathbf{in\ } e
-\\[0.1\baselineskip]
-    \mathbf{let\ } &Y \leftarrow [|\ \alpha X^T X + \beta Y\ |]\ \mathbf{in\ } e
-    \Rightarrow \\
-    & \mathbf{let\ } (X,Y)\ = \mathbf{syrk\ } \mathbf{true}
-    \ \alpha\ \_\ X\ \beta\ Y\ \mathbf{in\ } e
-%symm
-\\[0.1\baselineskip]
-    \mathbf{let\ } &Y \leftarrow\ [|\ \alpha\,\textrm{sym}(X_1)\,X_2 + \beta Y\ |]\ \mathbf{in\ } e
-    \Rightarrow \\
-    & \mathbf{let\ } ((X_1,X_2),Y)\ = \mathbf{symm\ } \mathbf{false}
-        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \mathbf{in\ } e
-\\[0.1\baselineskip]
-    \mathbf{let\ } &Y \leftarrow\ [|\ \alpha\,X_2\,\textrm{sym}(X_1) + \beta Y\ |]\ \mathbf{in\ } e
-    \Rightarrow \\
-    & \mathbf{let\ } ((X_1,X_2),Y)\ = \mathbf{symm\ } \mathbf{true}
-        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \mathbf{in\ } e
-\\[0.1\baselineskip]
-%gemm
-    \mathbf{let\ } &Y \leftarrow [|\ \alpha X_1^{T?} X_2^{T?} + \beta Y\ |]\ \mathbf{in\ } e
-    \Rightarrow \\
-    & \mathbf{let\ } ((X_1,X_2),Y)\ =\ \mathbf{gemm\ } \alpha
-        \ \_\ \left(X_1, \substack{\mathbf{true}\\\mathbf{false}}\right)
-        \ \_\ \left(X_2, \substack{\mathbf{true}\\\mathbf{false}}\right)
-        \ \beta\ Y\ \mathbf{in\ } e
-\end{align*}
+\subsubsection{Code Generation}
 
-\caption{Purely syntactic pattern-matching translations of
-    matrix expressions.}\label{fig:lang_matexp}
-\end{figure}
-
-\subsubsection{Code Generation} is a straightforward mapping from \lang's core
-constructs to high-level OCaml ones. We embed \lang's type- and term-
-constructors into OCaml as a sanity check on the output
-(Figure~\ref{fig:type_grammar}).
+Code generation is a straightforward mapping from \lang's core constructs to
+high-level OCaml ones. We embed \lang's type- and term- constructors into OCaml
+as a sanity check on the output (Figure~\ref{fig:type_grammar}).
 
@@
             <f> ::= `'
-            \alt <fc>
-            \alt `Z'
-            \alt `S' <f>
+            \alt <'fc>
+            \alt `z'
+            \alt <f> `s'
 
@@
             \alt `!' <t>
-            \alt \synt{$\forall$} <fc.> <t>
+            \alt <'fc.> <t>
             \alt <t> \lit{$\otimes$} \synt{$t'$}
@@
         \begin{align*}
-            [\![ f\!c ]\!] &= \texttt{'fc} \\
-            [\![ \textbf{Z} ]\!] &= \texttt{z}\\
-            [\![ \textbf{S} \, f ]\!] &= [\![ f ]\!]\, \texttt{s}\\
+            [\![ '\!f\!c ]\!] &= \texttt{'fc} \\
+            [\![ \textbf{z} ]\!] &= \texttt{z}\\
+            [\![ f \, \textbf{s} ]\!] &= [\![ f ]\!]\, \texttt{s}\\
             [\![ \textbf{unit} ]\!] &= \texttt{unit}\\
@@
             [\![ \textbf{!} \, t ]\!] &= [\![ t ]\!]\, \texttt{bang}\\
-            [\![ \forall f\!c.\, t ]\!] &= [\![ t ]\!]\\
+            [\![  '\!f\!c.\, t ]\!] &= [\![ t ]\!]\\
             [\![ t \otimes t' ]\!] &= [\![ t ]\!] \texttt{*} [\![ t' ]\!]\\

### 2019-04-13 write-up/paper/lang_and_examples.tex
Fix for shepherding: frontload syntax/primitives

@@
 
-\subsection{Overview}
+\subsection{Syntax}
+
+\lang's concrete syntax is inspired by that of OCaml. It desugars
+(Figure~\ref{fig:lang_desugar}) into a smaller, core type and expression
+grammar (Figure~\ref{fig:core_grammar}).
+
+\begin{figure}[p]
+    \begin{center}
+    \[\def\arraystretch{1.3}
+    \begin{array}{rcl}
+        f & ::= & {'\!f\!c}
+                \mid \texttt{z}
+                \mid f\ \texttt{s} \\
+                \\
+        t & ::= & \Unit
+                \mid \Bool
+                \mid \Int
+                \mid \Elt
+                \mid f\:\ottkw{arr}
+                \mid f\:\Mat
+                \mid \Bang t
+                \mid {'\!f\!c}. \: t
+                \mid t \otimes t'
+                \mid t \multimap t' \\
+                \\
+        e & ::= & p \textrm{ (primitives)}
+                \mid x \textrm{ (variable)}
+                \mid \Let\ x = e\ \In\ e'
+                \mid ()
+                \mid \Let\ () = e\ \In\ e'
+                \mid \ottkw{true}
+                \mid \ottkw{false}
+                \\ && \ottkw{if}\ e\ \ottkw{then}\ e_1\ \ottkw{else}\ e_2
+                \mid k \textrm{ (integer)}
+                \mid l \textrm{ (heap location)}
+                \mid el \textrm{ (array element)}
+                \\ && \Many\ v
+                \mid \Let\ \Many\ x = e\ \In\ e'
+                \mid \ottkw{fun}\ '\!f\!c \rightarrow e
+                    \textrm{ (frac. perm. abstraction)}
+                \\ && e\ [f] \textrm{ (frac. perm. specialisiation)}
+                \mid (e, e')
+                \mid \Let\ (a,b) = e\ \In\ e'
+                \\ && \ottkw{fun}\ x : t \rightarrow e
+                \mid e\ e'
+                \mid \ottkw{fix} \: (g, x : t, e : t')
+    \end{array} \]
+    \end{center}
+    \caption{Core fraction $f$, type $t$ and expression $e$ grammar of
+        \lang. Values $v$ are a subset of the expressions, their full
+        definition and a list of all primitives $p$ is in
+        Appendix~\ref{sec:grammar_def}.}\label{fig:core_grammar}
+\end{figure}
+
+\begin{figure}[p]
+\begin{center}
+\[\def\arraystretch{1.3}
+    \begin{array}{rcl}
+    x[e] &
+    \Rightarrow &
+    \mathbf{get}\ \_\ x\ (e) \;\qquad \textrm{(similarly for matrices)}
+\\
+    x[e_1] := e_2 &
+    \Rightarrow &
+    \mathbf{set}\ x\ (e_1)\ (e_2) \quad \textrm{(similarly for matrices)}
+\\
+\\
+    pat & ::= & ()\ \mid\ x\ \mid\ !x\ \mid\ \mathbf{Many\ } pat\ \mid\ (pat, pat)
+\\
+    \mathbf{let}\ !x = e_1\ \mathbf{in}\ e_2 &
+    \Rightarrow &
+    \specialcell[t]{l}{\mathbf{let\ Many}\ x = e_1\ \mathbf{in\ } \\
+    \mathbf{let\ Many}\ x = \mathbf{Many}\ (\mathbf{Many}\ x)\ \mathbf{in}\ e_2}
+\\
+    \mathbf{let\ Many}\ \langle pat_x \rangle\ = e_1\ \mathbf{in}\ e_2 &
+    \Rightarrow &
+    \specialcell[t]{l}{%
+        \mathbf{let\ Many}\ x = x\ \mathbf{in\ } \\
+        \mathbf{let\ } \langle pat_x \rangle\ = x\ \mathbf{in\ } e_2}
+\\
+    \mathbf{let}\ (\langle pat_a \rangle, \langle pat_b \rangle)\  = e_1\ \mathbf{in}\ e_2 &
+    \Rightarrow &
+    \specialcell[t]{l}{%
+        \mathbf{let\ } (a,b)\ = a\_b\ \mathbf{in\ }
+        \ \mathbf{let\ } \langle pat_a \rangle\ = a\ \mathbf{in\ } \\
+        \mathbf{let\ } \langle pat_b \rangle\ = b\ \mathbf{in\ } e_2}
+\\
+    \mathbf{fun}\ (\langle pat_x \rangle : t) \rightarrow e &
+    \Rightarrow &
+    \mathbf{fun}\ (x : t) \rightarrow \mathbf{let\ } \langle pat_x \rangle = x\ \mathbf{in\ } e
+\\
+\\
+    arg & ::= & \langle pat \rangle : t\ \mid\ {'x} \textrm{ (fractional permission variable)}
+\\
+    \mathbf{fun}\ \langle arg_{1 .. n} \rangle \rightarrow e &
+    \Rightarrow &
+    \mathbf{fun}\ \langle arg_1 \rangle \rightarrow ..
+    \ \mathbf{fun}\ \langle arg_n \rangle \rightarrow e
+\\
+    \mathbf{let}\ f\ {\langle arg_{1 .. n} \rangle} = e_1\ \mathbf{in}\ e_2 &
+    \Rightarrow &
+    \mathbf{let}\ f = \mathbf{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1\
+    \mathbf{in}\ e_2
+\\
+    \mathbf{let}\ !f\ {\langle arg_{1 .. n} \rangle} = e_1\ \mathbf{in}\ e_2 &
+    \Rightarrow &
+    \mathbf{let\ Many}\ f = \mathbf{Many}\ (\mathbf{fun}\ {\langle arg_{1 .. n} \rangle}
+    \rightarrow e_1)\ \mathbf{in}\ e_2
+\\
+    \mathrm{fixpoint} & \equiv & \mathbf{fix}\ (f, x : t, \mathbf{fun}
+    \ {\langle arg_{1 .. n} \rangle} \rightarrow e_1 : {t'} )
+\\
+    \mathbf{let\ rec}\ f\ (x : t)\ {\langle arg_{1 .. n} \rangle} : {t'} = e_1\ \mathbf{in}\ e_2 &
+    \Rightarrow &
+    \mathbf{let}\ f = \mathrm{fixpoint}\ \mathbf{in}\ e_2
+\\
+    \mathbf{let\ rec}\ !f\ (x : t)\ {\langle arg_{1 .. n} \rangle} : {t'} = e_1\ \mathbf{in}\ e_2 &
+    \Rightarrow &
+    \mathbf{let\ Many}\ f = \mathbf{Many}\ \mathrm{fixpoint}\ \mathbf{in}\ e_2
+    \end{array}
+\]
+\end{center}
+\caption{Desugaring \lang.}\label{fig:lang_desugar}
+\end{figure}
+
+\subsection{Type System and Other Features}
 
@@
 
+\subsubsection{Digression: Types of Primitives}
+
+\emph{The most pertinent aspect of \lang\ is the types of its primitives}
+(Figure~\ref{fig:primitive_types}).  While the types of operations such as
+\texttt{get} and \texttt{set} might be borderline obvious, the types of
+BLAS/LAPACK routines become an \emph{incredibly useful, automated check for
+using the API correctly.}
+
+\begin{figure}
+    \begin{center}
+    \[\def\arraystretch{1.3}
+    \begin{array}{rcl}
+%
+        \ottkw{symm} &:& \; !\Bool \multimap \; !\Elt \multimap \; {'x}. \:
+            {'x}\ \Mat \multimap \; {'y}. \:  {'y}\ \Mat \multimap \; !\Elt
+            \multimap \texttt{z}\ \Mat \multimap
+            \\ && ( {'x}\ \Mat \otimes {'y}\ \Mat ) \otimes \texttt{z}\ \Mat \\
+%
+        \ottkw{gemm} &:& \; !\Elt \multimap \; {'x}. \:  {'x}\ \Mat \: \otimes \;
+            !\Bool \multimap \; {'y}. \:  {'y}\ \Mat \: \otimes \; !\Bool
+            \multimap \; !\Elt \multimap
+            \\ && \texttt{z}\ \Mat \multimap ( {'x}\ \Mat \otimes {'y}\ \Mat )
+            \otimes \texttt{z}\ \Mat \\
+%
+        \ottkw{gesv} &:& \texttt{z}\ \Mat \multimap \texttt{z}\ \Mat \multimap
+            \texttt{z}\ \Mat \otimes \texttt{z}\ \Mat \\
+%
+        \ottkw{posv} &:& \texttt{z}\ \Mat \multimap \texttt{z}\ \Mat \multimap
+            \texttt{z}\ \Mat \otimes \texttt{z}\ \Mat \\
+%
+        \ottkw{posv\_flip} &:& \texttt{z}\ \Mat \multimap \texttt{z}\ \Mat
+            \multimap \texttt{z}\ \Mat \otimes \texttt{z}\ \Mat \\
+%
+        \ottkw{potrs} &:& {'x}. \:  {'x}\ \Mat \multimap \texttt{z}\ \Mat
+            \multimap {'x}\ \Mat \otimes \texttt{z}\ \Mat \\
+%
+        \ottkw{syrk} &:& \; !\Bool \multimap \; !\Elt \multimap \; {'x}. \:
+            {'x}\ \Mat \multimap \; !\Elt \multimap \texttt{z}\ \Mat \multimap \;
+            {'x}\ \Mat \otimes \texttt{z}\ \Mat \\
+%
+    \end{array} \]
+    \end{center}
+    \caption{Types of some \lang\ primitives.}\label{fig:primitive_types}
+\end{figure}
+
+We determine the types for these routines by consulting their documentation.
+Each routine has a record of the expected aliasing behaviour and whether or not
+it modifies or consumes its argument in any way. We use that to derive the
+types in Figure~\ref{fig:primitive_types}. Since most of these low-level
+routines are very careful not to do any allocation themselves, it is generally
+very easy to give each a \lang\ type -- every argument that can modify/consume
+its argument needs a full permission, and all others can be
+fraction-polymorphic.  Taking Fortran as an example, it has a notion of
+\texttt{in}, \texttt{out} and \texttt{inout} parameters. The latter two would
+need full \texttt{z} permissions; the first would be fraction-polymorphic.
+
 \subsubsection{Squaring a Matrix}
@@
 \end{figure}
-
-\emph{The most pertinent aspect of \lang\ is the types of its primitives.}
-While the types of operations such as \texttt{get} and \texttt{set} might be
-borderline obvious, the types of BLAS/LAPACK routines become an
-\emph{incredibly useful, automated check for using the API correctly.}
 
@@
 
-By using some simple pattern-matching and syntactic sugar, we can:
+By using some simple pattern-matching and syntactic sugar
+(Figure~\ref{fig:lang_matexp}), we can:
 \begin{itemize}
@@
 \end{itemize}
+
+\begin{figure}[t]
+\begin{center}
+\[\def\arraystretch{1.3}
+    \begin{array}{rcl}
+    \mathbf{let\ } v \leftarrow x[e]\ \mathbf{in\ } e &
+    \Rightarrow &
+    \mathbf{let\ } (x, !v)\ = x[e]\ \mathbf{in\ } e \qquad \textrm{(similarly for matrices)}
+\\
+    \mathbf{let\ } x_2 \leftarrow \mathbf{new}\ [|\ x_1\ |]\ \mathbf{in\ } e &
+    \Rightarrow &
+    \mathbf{let\ } (x_1, x_2)\ = \mathbf{copyM\ } \_\ x_1\ \mathbf{in\ } e
+\\
+    \mathbf{let\ } x_2 \leftarrow [|\ x_1\ |]\ \mathbf{in\ } e &
+    \Rightarrow &
+    \mathbf{let\ } (x_1, x_2)\ = \mathbf{copyM\_to\ } \_\  x_1\ x_2\ \mathbf{in\ } e
+    \end{array}
+\]
+\[
+    M ::= X\ \mid\ X^T\ \mid\ \textrm{sym}( X )
+\]
+\end{center}
+\begin{align*}
+% new
+    \mathbf{let\ } &Y \leftarrow \mathbf{new}\ (n, k)\ [|\ \alpha M_1 M_2 \ |]\ \mathbf{in\ } e
+    \Rightarrow \\
+    & \mathbf{let\ } Y\ = \mathbf{matrix\ } n\ k\ \mathbf{in}
+    \ \mathbf{let\ } Y \leftarrow [|\ \alpha M_1 M_2 + 0Y\ |]\ \mathbf{in\ } e
+\\[0.1\baselineskip]
+%syrk
+    \mathbf{let\ } &Y \leftarrow [|\ \alpha X X^T + \beta Y\ |]\ \mathbf{in\ } e
+    \Rightarrow \\
+    & \mathbf{let\ } (X,Y)\ = \mathbf{syrk\ } \mathbf{false}
+    \ \alpha\ \_\ X\ \beta\ Y\ \mathbf{in\ } e
+\\[0.1\baselineskip]
+    \mathbf{let\ } &Y \leftarrow [|\ \alpha X^T X + \beta Y\ |]\ \mathbf{in\ } e
+    \Rightarrow \\
+    & \mathbf{let\ } (X,Y)\ = \mathbf{syrk\ } \mathbf{true}
+    \ \alpha\ \_\ X\ \beta\ Y\ \mathbf{in\ } e
+%symm
+\\[0.1\baselineskip]
+    \mathbf{let\ } &Y \leftarrow\ [|\ \alpha\,\textrm{sym}(X_1)\,X_2 + \beta Y\ |]\ \mathbf{in\ } e
+    \Rightarrow \\
+    & \mathbf{let\ } ((X_1,X_2),Y)\ = \mathbf{symm\ } \mathbf{false}
+        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \mathbf{in\ } e
+\\[0.1\baselineskip]
+    \mathbf{let\ } &Y \leftarrow\ [|\ \alpha\,X_2\,\textrm{sym}(X_1) + \beta Y\ |]\ \mathbf{in\ } e
+    \Rightarrow \\
+    & \mathbf{let\ } ((X_1,X_2),Y)\ = \mathbf{symm\ } \mathbf{true}
+        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \mathbf{in\ } e
+\\[0.1\baselineskip]
+%gemm
+    \mathbf{let\ } &Y \leftarrow [|\ \alpha X_1^{T?} X_2^{T?} + \beta Y\ |]\ \mathbf{in\ } e
+    \Rightarrow \\
+    & \mathbf{let\ } ((X_1,X_2),Y)\ =\ \mathbf{gemm\ } \alpha
+        \ \_\ \left(X_1, \substack{\mathbf{true}\\\mathbf{false}}\right)
+        \ \_\ \left(X_2, \substack{\mathbf{true}\\\mathbf{false}}\right)
+        \ \beta\ Y\ \mathbf{in\ } e
+\end{align*}
+
+\caption{Purely syntactic pattern-matching translations of
+    matrix expressions.}\label{fig:lang_matexp}
+\end{figure}
 
@@
 
-\subsubsection{L1-Norm Minimisation on Manifolds}
+\subsubsection{L1-norm Minimisation on Manifolds}
 
-L1-Norm minimisation is often used in optimisation problems, as a
+L1-norm minimisation is often used in optimisation problems, as a
 \emph{regularisation} term for reducing the influence of outliers.  Although
-the below formuation\cite{bronstein} is intended to be used with \emph{sparse}
+the below formulation~\cite{bronstein} is intended to be used with \emph{sparse}
 computations, \lang's current implementation only implements dense ones.
@@
 
-A \emph{Kalman Filter}\cite{kalman} is a an algorithm for combining prior
+A \emph{Kalman Filter}~\cite{kalman} is a an algorithm for combining prior
 knowledge of a state, a statistical model and measurements from (noisy) sensors

### 2019-04-13 write-up/paper/formal_system.tex
Make stuff consistent
Syntax & spacing, with a few small edits from Reviewer C's review.

@@
 matrices. This is exactly the same as value-restriction from the world of
-parametric polymorphism.
+parametric polymorphism; without it, the expression $\Let\ \Many\ x = \Many\
+(\ottkw{array}\ 5)\ \In\ \Let\ () = \ottkw{free}\ x\ \In\ x[0]$ would
+type-check but error at runtime.
 
@@
 Lastly, types of almost all \lang\ primitives, as embedded in OCaml's type
-system, are shown in Appendix \ref{subsec:primitives}, with some similar ones
-(like those for binary arithmetic operators) omitted for brevity. The
-main difference between the OCaml type of a primitive like \highl{gemm} and its
-\lang\ counterpart is the inclusion of explicit `$\forall$'s.  So,
-\highl{float bang -> ('a mat * bool bang) -> ('b mat * bool bang) -> float
-bang -> z mat -> ('a mat * 'b mat) * z mat}
-will correspond to \\
-$!\ottkw{elt} \multimap \forall\, x.\ x \ \ottkw{mat} \ \otimes \ !\ottkw{bool}
-\multimap \forall\, y.\  y \ \ottkw{mat} \ \otimes \ !\ottkw{bool} \multimap
-\ !\ottkw{elt} \multimap z\ \ottkw{mat} \multimap ( x \ \ottkw{mat} \ \otimes y
-\ \ottkw{mat} ) \ \otimes z\ \ottkw{mat}$
+system, are shown in Appendix~\ref{subsec:primitives}, with some similar ones
+(like those for binary arithmetic operators) omitted for brevity. The main
+difference between the OCaml type of a primitive like \highl{gemm} and its
+\lang\ counterpart (Figure~\ref{fig:primitive_types}) is the inclusion of
+explicit universal-quantification of fractional permission variables in the
+latter.
 
@@
 We draw the reader's attention to the definitions relating to fractional
-permissions. Specifically, unlike a lambda, the body of a $\ottkw{fun}\, f\!c
-\rightarrow \_$ must be a syntactic value. The context $\ottkw{fun}\, f\!c
+permissions. Specifically, unlike a lambda, the body of a $\ottkw{fun}\ f\!c
+\rightarrow v$ must be a syntactic value. The context $\ottkw{fun}\ f\!c
 \rightarrow [-]$ means expressions can be reduced inside a fractional
@@
 \[
-    \den{H}{}{\sigma} = \bigstar_{(l,f,m) \in \sigma} [ l \mapsto_f m ]
+    \den{H}{}{\sigma} = \bigstar_{(l,f,m) \in \sigma} [l \mapsto_f m]
 \]
@@
 We then define a step-indexed logical relation in the style of Morrisett et.
-al~\cite{morrisett}. $(\varsigma, v) \in \V{k}{t}$ means it takes a heap with
+al~\cite{ahmed20073}. $(\varsigma, v) \in \V{k}{t}$ means it takes a heap with
 exactly $\varsigma$ resources to produce a value $v$ of type $t$ in at most $k$
@@
 
-The definition of $\V{k}{\forall f\!c.\ t}$ says a value and heap
+The definition of $\V{k}{ '\!f\!c.\ t}$ says a value and heap
 must be the same regardless of what fraction is substituted into both; the
@@
 \[
-    \V{k}{ \forall f\!c.\  t } = \{ (\varsigma, \ottkw{fun}\, f\!c \rightarrow \, v) \mid \forall f.\ (\varsigma [ f\!c / f ], v [ f\!c / f ]) \in \V{k-1}{ t [ f\!c / f ] } \}
+    \V{k}{ '\!f\!c.\  t } = \{ (\varsigma, \ottkw{fun}\ '\!f\!c \rightarrow \, v) \mid \forall f.\ (\varsigma [f / f\!c], v [f / f\!c]) \in \V{k-1}{ t [f / f\!c] } \}
 \]
@@
 \begin{align*}
-    \V{k}{ t' \multimap t } &= \{ (\varsigma_v, v ) \mid ( v \equiv \ottkw{fun}\, x : t' \rightarrow e \vee v \equiv \ottkw{fix} (g, x : t' , e : t) ) \, \wedge\\
+    \V{k}{ t' \multimap t } &= \{ (\varsigma_v, v ) \mid ( v \equiv \ottkw{fun}\ x : t' \rightarrow e \vee v \equiv \ottkw{fix} (g, x : t' , e : t) ) \, \wedge\\
                             & \qquad \qquad \forall j \leq k, (\varsigma_{v'}, v') \in \V{j}{ t' }.\ \varsigma_v \star \varsigma_v' \textrm{ defined } \Rightarrow (\varsigma_v \star \varsigma_v', v\, v') \in \C{j}{t} \}
@@
 \begin{align*}
-    \den{I}{k}{ \Delta, x : t } \theta &= \{ \delta[x \mapsto v_x ] \mid \delta \in \den{I}{k}{\Delta}\theta \wedge (\empH, v_x) \in \V{k}{\theta(t)} \} \\
-    \den{L}{k}{ \Gamma, x : t } \theta &= \{ (\varsigma \star \varsigma_x, \gamma[x \mapsto v_x ]) \mid (\varsigma, \gamma) \in \den{L}{k}{\Gamma}\theta \wedge (\varsigma_x, v_x) \in \V{k}{\theta(t)} \}
+    \den{I}{k}{ \Delta, x : t } \theta &= \{ \delta[x \mapsto v_x] \mid \delta \in \den{I}{k}{\Delta}\theta \wedge (\empH, v_x) \in \V{k}{\theta(t)} \} \\
+    \den{L}{k}{ \Gamma, x : t } \theta &= \{ (\varsigma \star \varsigma_x, \gamma[x \mapsto v_x]) \mid (\varsigma, \gamma) \in \den{L}{k}{\Gamma}\theta \wedge (\varsigma_x, v_x) \in \V{k}{\theta(t)} \}
 \end{align*}

### 2019-04-13 write-up/paper/intro.tex
Make stuff consistent
Syntax & spacing, with a few small edits from Reviewer C's review.

@@
 As a result, programmers often turn to higher-level languages such as
-Matlab, R and Numpy, which offer very high-level array abstractions
+Matlab, R and NumPy, which offer very high-level array abstractions
 that can be viewed as ordinary mathematical values. This makes
@@
 or they can obtain safety at the price of unneeded copies and worse
-memory efficiency. In this work, show that this tradeoff is not a
+memory efficiency. In this work, we show that this tradeoff is not a
 fundamental one.

### 2019-04-13 write-up/paper/lang_and_examples.tex
Make stuff consistent
Syntax & spacing, with a few small edits from Reviewer C's review.

--- 2019-04-12 write-up/paper/lang_and_examples.tex
@@
 
-\subsection{Overview}
+\subsection{Syntax}
+
+\lang's concrete syntax is inspired by that of OCaml. It desugars
+(Figure~\ref{fig:lang_desugar}) into a smaller, core type and expression
+grammar (Figure~\ref{fig:core_grammar}).
+
+\begin{figure}[p]
+    \begin{center}
+    \[\def\arraystretch{1.3}
+    \begin{array}{rcl}
+        f & ::= & {'\!f\!c}
+                \mid \Zf
+                \mid f\ \Sf \\
+                \\
+        t & ::= & \Unit
+                \mid \Bool
+                \mid \Int
+                \mid \Elt
+                \mid f\ \Arr
+                \mid f\ \Mat
+                \mid \Bang t
+                \mid {'\!f\!c}.\ t
+                \mid t \otimes t'
+                \mid t \multimap t' \\
+                \\
+        e & ::= & p \textrm{ (primitives)}
+                \mid x \textrm{ (variable)}
+                \mid \Let\ x = e\ \In\ e'
+                \mid ()
+                \mid \Let\ () = e\ \In\ e'
+                \mid \ottkw{true}
+                \mid \ottkw{false}
+                \\ && \ottkw{if}\ e\ \ottkw{then}\ e_1\ \ottkw{else}\ e_2
+                \mid k \textrm{ (integer)}
+                \mid l \textrm{ (heap location)}
+                \mid el \textrm{ (array element)}
+                \\ && \Many\ v
+                \mid \Let\ \Many\ x = e\ \In\ e'
+                \mid \ottkw{fun}\ '\!f\!c \rightarrow e
+                    \textrm{ (frac. perm. abstraction)}
+                \\ && e\ [f] \textrm{ (frac. perm. specialisiation)}
+                \mid (e, e')
+                \mid \Let\ (a,b) = e\ \In\ e'
+                \\ && \ottkw{fun}\ x : t \rightarrow e
+                \mid e\ e'
+                \mid \ottkw{fix} \: (g, x : t, e : t')
+    \end{array} \]
+    \end{center}
+    \caption{Core fraction $f$, type $t$ and expression $e$ grammar of
+        \lang. Values $v$ are a subset of the expressions, their full
+        definition and a list of all primitives $p$ is in
+        Appendix~\ref{sec:grammar_def}.}\label{fig:core_grammar}
+\end{figure}
+
+\begin{figure}[p]
+\begin{center}
+\[\def\arraystretch{1.3}
+    \begin{array}{rcl}
+    x[e] &
+    \Rightarrow &
+    \ottkw{get}\ \_\ x\ (e) \:\qquad \textrm{(similarly for matrices)}
+\\
+    x[e_1] := e_2 &
+    \Rightarrow &
+    \ottkw{set}\ x\ (e_1)\ (e_2) \quad \textrm{(similarly for matrices)}
+\\
+\\
+    pat & ::= & () \mid x \mid {!x} \mid \Many\ pat \mid (pat, pat)
+\\
+    \Let\ !x = e_1\ \In\ e_2 &
+    \Rightarrow &
+    \specialcell[t]{l}{%
+        \Let\ \Many\ x = e_1\ \In \\
+        \Let\ \Many\ x = \Many\ (\Many\ x)\ \In\ e_2}
+\\
+    \Let\ \Many\ \langle pat_x \rangle\ = e_1\ \In\ e_2 &
+    \Rightarrow &
+    \specialcell[t]{l}{%
+        \Let\ \Many\ x = x\ \In \\
+        \Let\ \langle pat_x \rangle\ = x\ \In\ e_2}
+\\
+    \Let\ (\langle pat_a \rangle, \langle pat_b \rangle)\ = e_1\ \In\ e_2 &
+    \Rightarrow &
+    \specialcell[t]{l}{%
+        \Let\ (a,b)\ = a\_b\ \In\
+        \Let\ \langle pat_a \rangle\ = a\ \In \\
+        \Let\ \langle pat_b \rangle\ = b\ \In\ e_2}
+\\
+    \ottkw{fun}\ (\langle pat_x \rangle : t) \rightarrow e &
+    \Rightarrow &
+    \ottkw{fun}\ (x : t) \rightarrow \Let\ \langle pat_x \rangle = x\ \In\ e
+\\
+\\
+    arg & ::= & \langle pat \rangle : t\ \mid\ {'x} \textrm{ (fractional permission variable)}
+\\
+    \ottkw{fun}\ \langle arg_{1 .. n} \rangle \rightarrow e &
+    \Rightarrow &
+    \ottkw{fun}\ \langle arg_1 \rangle \rightarrow ..
+    \ \ottkw{fun}\ \langle arg_n \rangle \rightarrow e
+\\
+    \Let\ f\ {\langle arg_{1 .. n} \rangle} = e_1\ \In\ e_2 &
+    \Rightarrow &
+    \Let\ f = \ottkw{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1\
+    \In\ e_2
+\\
+    \Let\ !f\ {\langle arg_{1 .. n} \rangle} = e_1\ \In\ e_2 &
+    \Rightarrow &
+    \Let\ \Many\ f = \Many\ (\ottkw{fun}\ {\langle arg_{1 .. n} \rangle}
+    \rightarrow e_1)\ \In\ e_2
+\\
+    \mathrm{fixpoint} & \equiv & \ottkw{fix}\ (f, x : t, \ottkw{fun}
+    \ {\langle arg_{0 .. n} \rangle} \rightarrow e_1 : {t'} )
+\\
+    \Let\ \ottkw{rec}\ f\ (x : t)\ {\langle arg_{0 .. n} \rangle} : {t'} = e_1\ \In\ e_2 &
+    \Rightarrow &
+    \Let\ f = \mathrm{fixpoint}\ \In\ e_2
+\\
+    \Let\ \ottkw{rec}\ !f\ (x : t)\ {\langle arg_{0 .. n} \rangle} : {t'} = e_1\ \In\ e_2 &
+    \Rightarrow &
+    \Let\ \Many\ f = \Many\ \mathrm{fixpoint}\ \In\ e_2
+    \end{array}
+\]
+\end{center}
+\caption{Desugaring \lang.}\label{fig:lang_desugar}
+\end{figure}
+
+\subsection{Type System and Other Features}
 
 The core type theory of \lang\ is a nearly off-the-shelf linear type
-theory, supporting familiar features such as linear function spaces
-$A \multimap B$ and tensor products $A \otimes B$. We adopt linearity
--- the restriction that each program variable be used at most once --
-since it allows us to express purely functional APIs for numerical
-library routines that mutate arrays and matrices~\cite{Wadler90}. Due
-to linearity, values cannot alias and are only used once, which means
-that linearly-typed updates result in no \emph{observable} mutation.
+theory~\cite{barber1996}, supporting familiar features such as linear function
+spaces $A \multimap B$ and tensor products $A \otimes B$. We adopt linearity --
+the restriction that each program variable be used at most once -- since it
+allows us to express purely functional APIs for numerical library routines that
+mutate arrays and matrices~\cite{Wadler90}. Due to linearity, values cannot
+alias and are only used once, which means that linearly-typed updates result in
+no \emph{observable} mutation.
 
@@
 An example of this consuming and re-binding is in \highl{let (row, !x1) =
-row[i]}. Indexing is implemented as a primitive \highl{get: 'x. 'x arr --o
-!int --o 'x arr * !elt}. Although fractional permissions can be passed around
-explicitly  (as done in the recursive call), they can also be
-\emph{automatically inferred at call sites}: \highl{row[i] == get _ row i}
-takes advantage of this convenience.
+row[i]}. Indexing is implemented as a primitive $\ottkw{get}:\ {'x}.\ {{'x}\
+\Arr} \multimap {!\Int} \multimap {'x\ \Arr} \otimes {!\Elt}$.
+Although fractional permissions can be passed around explicitly  (as done in
+the recursive call), they can also be \emph{automatically inferred at call
+sites}: \highl{row[i] == get _ row i} takes advantage of this convenience.
 
@@
 
-Mutating an array is implemented similarly to indexing one: a primitive
-\highl{set: z arr --o !int --o !elt --o z arr}.  It consumes the original
-array and returns a new array with the updated value.  \highl{let written =
-write[i] := <exp> } is just syntactic sugar for \highl{let written = set write i <exp>}.
+Mutating an array is implemented similarly to indexing one -- a primitive
+$\ottkw{set}:\ {\Zf\ \Arr} \multimap {!\Int} \multimap {!\Elt} \multimap
+{\Zf\ \Arr}$. It consumes the original array and returns a new array with
+the updated value.  \highl{let written = write[i] := <exp> } is just syntactic
+sugar for \highl{let written = set write i <exp>}.
 
-Since \highl{write: z arr} (where \texttt{z} stands for $k=0$, representing a
+Since \highl{write: z arr} (where \Zf stands for $k=0$, representing a
 fractional permission of $2^{-k} = 2^{-0} = 1$), we may mutate it, but since we
@@
 permission based on the given arguments.
+
+\subsubsection{Digression: Types of Primitives}
+
+\emph{The most pertinent aspect of \lang\ is the types of its primitives}
+(Figure~\ref{fig:primitive_types}).  While the types of operations such as
+\textbf{get} and \textbf{set} might be borderline obvious, the types of
+BLAS/LAPACK routines become an \emph{incredibly useful, automated check for
+using the API correctly.}
+
+\begin{figure}
+    \begin{center}
+    \[\def\arraystretch{1.3}
+    \begin{array}{rcl}
+%
+        \ottkw{symm} &:& {!\Bool} \multimap {!\Elt} \multimap {'x}.\
+            {{'x}\ \Mat} \multimap {'y}.\ {{'y}\ \Mat} \multimap {!\Elt}
+            \multimap {\Zf\ \Mat} \multimap
+            \\ && ( {{'x}\ \Mat} \otimes {{'y}\ \Mat} ) \otimes {\Zf\ \Mat} \\
+%
+        \ottkw{gemm} &:& {!\Elt} \multimap {'x}.\ {{'x}\ \Mat} \otimes
+            {!\Bool} \multimap {'y}.\ {{'y}\ \Mat} \otimes {!\Bool}
+            \multimap {!\Elt} \multimap
+            \\ && {\Zf\ \Mat} \multimap ( {{'x}\ \Mat} \otimes {{'y}\ \Mat} )
+            \otimes {\Zf\ \Mat} \\
+%
+        \ottkw{gesv} &:& {\Zf\ \Mat} \multimap {\Zf\ \Mat} \multimap
+            {\Zf\ \Mat} \otimes {\Zf\ \Mat} \\
+%
+        \ottkw{posv} &:& {\Zf\ \Mat} \multimap {\Zf\ \Mat} \multimap
+            {\Zf\ \Mat} \otimes {\Zf\ \Mat} \\
+%
+        \ottkw{potrs} &:& {'x}.\ {{'x}\ \Mat} \multimap {\Zf\ \Mat}
+            \multimap {{'x}\ \Mat} \otimes {\Zf\ \Mat} \\
+%
+        \ottkw{syrk} &:& {!\Bool} \multimap {!\Elt} \multimap {'x}.\
+            {{'x}\ \Mat} \multimap {!\Elt} \multimap {\Zf\ \Mat} \multimap
+            {{'x}\ \Mat} \otimes {\Zf\ \Mat} \\
+%
+    \end{array} \]
+    \end{center}
+    \caption{Types of some \lang\ primitives.}\label{fig:primitive_types}
+\end{figure}
+
+We determine the types for these routines by consulting their documentation.
+Each routine has a record of the expected aliasing behaviour and whether or not
+it modifies or consumes its argument in any way. We use that to derive the
+types in Figure~\ref{fig:primitive_types}. Since most of these low-level
+routines are very careful not to do any allocation themselves, it is generally
+very easy to give each a \lang\ type -- every argument that can modify/consume
+its argument needs a full permission, and all others can be
+fraction-polymorphic.  Taking Fortran as an example, it has a notion of
+\texttt{in}, \texttt{out} and \texttt{inout} parameters. The latter two would
+need full \Zf permissions; the first would be fraction-polymorphic.
 
@@
 
-\emph{The most pertinent aspect of \lang\ is the types of its primitives.}
-While the types of operations such as \texttt{get} and \texttt{set} might be
-borderline obvious, the types of BLAS/LAPACK routines become an
-\emph{incredibly useful, automated check for using the API correctly.}
-
 Figure~\ref{fig:lang_square} shows how a linearly-typed matrix squaring
@@
 \texttt{x}. Then, because we need to use \texttt{x} twice (so that we can
-multiply it by itself) but linearity only allows one use, we use \highl{shareM:
-'x. 'x mat --o 'x s mat * 'x s mat} to split the permission \texttt{'x} (which
-represents $2^{-x}$) into two halves (\texttt{'x s}, which represents $2^{-(x+1)}$).
+multiply it by itself) but linearity only allows one use, we use
+$\ottkw{shareM}:\ {'x}.\ {{'x}\ \Mat} \multimap {{'x}\ \Sf\ \Mat} \otimes
+{{'x}\ \Sf\ \Mat}$ to split the permission \texttt{'x} (which represents
+$2^{-x}$) into two halves (\texttt{'x s}, which represents $2^{-(x+1)}$).
 
@@
 \end{itemize}
+
+\begin{figure}[t]
+\begin{center}
+\[\def\arraystretch{1.3}
+    \begin{array}{rcl}
+    \Let\ v \leftarrow x[e]\ \In\ e &
+    \Rightarrow &
+    \Let\ (x, !v)\ = x[e]\ \In\ e \qquad \textrm{(similarly for matrices)}
+\\
+    \Let\ x_2 \leftarrow \ottkw{new}\ [|\ x_1\ |]\ \In\ e &
+    \Rightarrow &
+    \Let\ (x_1, x_2)\ = \ottkw{copyM}\ \_\ x_1\ \In\ e
+\\
+    \Let\ x_2 \leftarrow [|\ x_1\ |]\ \In\ e &
+    \Rightarrow &
+    \Let\ (x_1, x_2)\ = \ottkw{copyM\_to}\ \_\ x_1\ x_2\ \In\ e
+    \end{array}
+\]
+\[
+    M ::= X\ \mid\ X^T\ \mid\ \textrm{sym}( X )
+\]
+\end{center}
+\begin{align*}
+% new
+    \Let\ &Y \leftarrow \ottkw{new}\ (n, k)\ [|\ \alpha M_1 M_2 \ |]\ \In\ e
+    \Rightarrow \\
+    & \Let\ Y\ = \ottkw{matrix}\ n\ k\ \In
+    \ \Let\ Y \leftarrow [|\ \alpha M_1 M_2 + 0Y\ |]\ \In\ e
+\\[0.1\baselineskip]
+%syrk
+    \Let\ &Y \leftarrow [|\ \alpha X X^T + \beta Y\ |]\ \In\ e
+    \Rightarrow \\
+    & \Let\ (X,Y)\ = \ottkw{syrk}\ \ottkw{false}
+    \ \alpha\ \_\ X\ \beta\ Y\ \In\ e
+\\[0.1\baselineskip]
+    \Let\ &Y \leftarrow [|\ \alpha X^T X + \beta Y\ |]\ \In\ e
+    \Rightarrow \\
+    & \Let\ (X,Y)\ = \ottkw{syrk}\ \ottkw{true}
+    \ \alpha\ \_\ X\ \beta\ Y\ \In\ e
+%symm
+\\[0.1\baselineskip]
+    \Let\ &Y \leftarrow\ [|\ \alpha\,\textrm{sym}(X_1)\,X_2 + \beta Y\ |]\ \In\ e
+    \Rightarrow \\
+    & \Let\ ((X_1,X_2),Y)\ = \ottkw{symm}\ \ottkw{false}
+        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \In\ e
+\\[0.1\baselineskip]
+    \Let\ &Y \leftarrow\ [|\ \alpha\,X_2\,\textrm{sym}(X_1) + \beta Y\ |]\ \In\ e
+    \Rightarrow \\
+    & \Let\ ((X_1,X_2),Y)\ = \ottkw{symm}\ \ottkw{true}
+        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \In\ e
+\\[0.1\baselineskip]
+%gemm
+    \Let\ &Y \leftarrow [|\ \alpha X_1^{T?} X_2^{T?} + \beta Y\ |]\ \In\ e
+    \Rightarrow \\
+    & \Let\ ((X_1,X_2),Y)\ =\ \ottkw{gemm}\ \alpha
+        \ \_\ \left(X_1, \substack{\ottkw{true}\\\ottkw{false}}\right)
+        \ \_\ \left(X_2, \substack{\ottkw{true}\\\ottkw{false}}\right)
+        \ \beta\ Y\ \In\ e
+\end{align*}
+
+\caption{Purely syntactic pattern-matching translations of
+    matrix expressions.}\label{fig:lang_matexp}
+\end{figure}
 
@@
 After computing \texttt{x\_T\_x}, we need to invert it and then multiply it by
-\texttt{xy}. The BLAS routine \highl{posv: z mat --o z mat --o z mat * z mat}
-does exactly that: assuming the first argument is symmetric, \texttt{posv}
-mutates its second argument to contain the desired value. Its first argument is
-also mutated to contain the (upper triangular) Cholesky decomposition factor of
-the original matrix. Since we do not need that matrix (or its memory) again, we
+\texttt{xy}. The BLAS routine $\ottkw{posv}:\ {\Zf\ \Mat} \multimap
+{\Zf\ \Mat} \multimap {\Zf\ \Mat} \otimes {\Zf\ \Mat}$ does
+exactly that: assuming the first argument is symmetric, \texttt{posv} mutates
+its second argument to contain the desired value. Its first argument is also
+mutated to contain the (upper triangular) Cholesky decomposition factor of the
+original matrix. Since we do not need that matrix (or its memory) again, we
 \texttt{free} it. If we forgot to, we would get a \texttt{Variable to\_del not
@@
 Figure~\ref{fig:lang_l1_norm_min} shows even more pattern-matching. Patterns of
-the form \highl{let <id> <- [| beta * c + alpha * a * b |]} are also desugared to
-\texttt{gemm} calls. Primitives like \highl{transpose: 'x. 'x mat --o 'x mat *
-z mat} and \highl{eye: !int --o z mat} allocate new matrices;
-\texttt{transpose} returns the transpose of a given matrix and \texttt{eye k}
-evaluates to a $k \times k$ identity matrix.
+the form \highl{let <id> <- [| beta * c + alpha * a * b |]} are also desugared
+to \texttt{gemm} calls. Primitives like $\ottkw{transpose}:\ {'x}.\ {{'x}\ \Mat}
+\multimap {{'x}\ \Mat} \otimes {\Zf\ \Mat}$ and $\ottkw{eye}:\ {!\Int}
+\multimap {\Zf\ \Mat}$ allocate new matrices; \texttt{transpose} returns
+the transpose of a given matrix and \texttt{eye k} evaluates to a $k \times k$
+identity matrix.
 
@@
 
-A \emph{Kalman Filter}\cite{kalman} is a an algorithm for combining prior
+A \emph{Kalman Filter}~\cite{kalman} is an algorithm for combining prior
 knowledge of a state, a statistical model and measurements from (noisy) sensors

### 2019-04-13 write-up/paper/formal_system.tex
Fix spacing

@@
 
-Heaps ($\sigma$) are multisets containing triples of an abstract location $l$,
+Heaps $\sigma$ are multisets containing triples of an abstract location $l$,
 a fractional permission $f$ and sized matrices $m_{n,k}$. The notation $l
@@
 We draw the reader's attention to the definitions relating to fractional
-permissions. Specifically, unlike a lambda, the body of a $\ottkw{fun}\ f\!c
-\rightarrow v$ must be a syntactic value. The context $\ottkw{fun}\ f\!c
+permissions. Specifically, unlike a lambda, the body of a $\ottkw{fun}\ '\!f\!c
+\rightarrow v$ must be a syntactic value. The context $\ottkw{fun}\ '\!f\!c
 \rightarrow [-]$ means expressions can be reduced inside a fractional

### 2019-05-26 write-up/paper/discussion.tex
Make stuff consistent
Syntax & spacing, with a few small edits from Reviewer C's review.
Incorporate final review feedback
Also reduce warnings.

@@
 which were marked as \texttt{intent(in)} but would have been written over by
-calls to `gemm', spotted by the fractional capabilities feature. Furthermore,
+calls to `gemm', spotted by the fractional permissions feature. Furthermore,
 it used a matrix \emph{twice} in a call to `symm', once with a read permission
@@
 illegal according to the standard, both aspects of which were captured by
-linearity and fractional capabilities.
+linearity and fractional permissions.
 
@@
 
+\subsubsection{Linear types for \emph{implementing} linear algebra routines}
+
 Using linear types for BLAS routines is a particularly good domain fit (given
@@
 
+\subsubsection{Our contribution: linear types for enforcing correct
+    \emph{usage} of linear algebra routines}
+
 In contrast, our approach is to begin with the assumption that we can take
@@
 
-This approach is similar to the one taken in linear algebra libraries for Rust
+\subsubsection{Traditionally complex approaches to sharing}
+
+Our approach is similar to the one taken in linear algebra libraries for Rust
 -- these libraries typically take advantage of the distinction that Rust's type
@@
 
+\subsubsection{Our contribution: a simpler approach to sharing}
+
 Instead, our approach, of requiring sharing to be made explicit, lets us
@@
 with ordinary unification.
+
+\subsubsection{Implications}
 
@@
 
-However, since our type system shows demonstrates unification is ``all one
+However, since our type system shows unification is ``all one
 needs'' for fractions, it should be possible to \emph{encode} \lang's approach
@@
 by extending its type system. For example, we could have used existential types
-to statically track pointer identities~\cite{morrisett}, or parametric
+to statically track pointer identities~\cite{ahmed20073}, or parametric
 polymorphism.

### 2019-05-26 write-up/paper/formal_system.tex
Fix reference
Incorporate final review feedback
Also reduce warnings.

@@
     is for the sake of ergonomics when using binary arithmetic operations
-    (e.g.\ of type $!\ottkw{int} \, \otimes \, !\ottkw{int} \multimap \;
+    (e.g.\ of type $!\ottkw{int} \multimap \; !\ottkw{int} \multimap \;
     !\ottkw{int}$): a programmer should be able to write \highl{let x = 5 + 5
@@
 
-\subsection{Logical Relation}
+\subsection{Logical Relation}\label{subsec:logical_rel}
 
@@
 
-We then define a step-indexed logical relation in the style of Morrisett et.
+We then define a step-indexed logical relation in the style of Ahmed et.
 al~\cite{ahmed20073}. $(\varsigma, v) \in \V{k}{t}$ means it takes a heap with
@@
 must be the same regardless of what fraction is substituted into both; the
-$k-1$ is to take into account fraction specialisation takes ones step
+$k-1$ is to take into account fraction specialisation takes one step
 (\textsc{Op\_Spc}).
@@
 combination of a frame rule on heaps (no interference), type-preservation and
-termination (in $j < k$ steps) to either an error or a heap-and-expression,
-with the further condition that if the expression is a syntactic value then it
-is also one semantically.
+termination (in $j < k$ steps) to either an error or a heap-and-expression.
+For the case of termination to a heap-and-expression, there is a further
+condition: if the expression is a value syntactically then it is also one
+semantically.
 \begin{align*}
@@
 
-If an expression is syntactically type-checked, then for an arbitrary number of
-steps $k$, under any substitution $\theta$ of free fractional-permission
-variables, linear variables $\gamma$ (with a suitable heap $\varsigma$) and
-intuitionistic variables $\delta$, that suitable heap and expression
-$(\varsigma, \theta(\delta(\gamma(e))))$ are in the computational
-interpretation of the type of that expression $\C{k}{ \theta(t) }$.
-
-The \emph{computational interpretation} is as defined before: a combination of
-a frame rule on heaps (no interference), type-preservation and termination (in
-$j < k$ steps) to either an error or a heap-and-expression, with the further
-condition that if the expression is a syntactic value then it is also one
-semantically.
-
-It identifies executions that do no un- or ill-defined behaviours (e.g.\ adding
-a boolean and an integer). Since our operational semantics explicitly models
-deallocation, we now know no well-typed program will ever try to access
-deallocated memory, establishing the correctness of our memory management
-checking.
+If an expression $e$ is syntactically type-checked (against a type $t$), then
+\begin{itemize}
+    \item for an arbitrary number of steps $k$,
+    \item under any substitution of
+        \begin{itemize}
+            \item free fractional permission variables $\theta$,
+            \item linear variables with a suitable heap $(\gamma, \varsigma)$ and
+            \item intuitionistic variables $\delta$,
+        \end{itemize}
+    \item the aforementioned suitable heap and expression $(\varsigma, \theta(\delta(\gamma(e))))$
+    \item are in the computational interpretation $\C{k}{ \theta(t) }$ of the type $t$.
+\end{itemize}
+The \emph{computational interpretation} is as defined before
+(Section~\ref{subsec:logical_rel}); it identifies executions that do no un- or
+ill-defined behaviours (e.g.\ adding a boolean and an integer). Since our
+operational semantics explicitly models deallocation, we now know no well-typed
+program will ever try to access deallocated memory, establishing the
+correctness of our memory management checking.
 
@@
 complication of reducing an expression with an arbitrary fractional permission
-variable in it, and then instantiating it at the last momemnt to conclude,
+variable in it, and then instantiating it at the last moment to conclude,
 which is where \ref{fracPermSub} (heap-and-expressions take the same steps of

### 2019-05-26 write-up/paper/intro.tex
Incorporate final review feedback
Also reduce warnings.

@@
 
-So in practice, programmers face a tradeoff: they can eschew safety
+So in practice, programmers face a trade-off: they can eschew safety
 and exploit the full power of the underlying linear algebra libraries,
 or they can obtain safety at the price of unneeded copies and worse
-memory efficiency. In this work, we show that this tradeoff is not a
+memory efficiency. In this work, we show that this trade-off is not a
 fundamental one.
@@
 
+Our contribution is the idea applying of linear types with fractional
+permissions to enforce the correct \emph{usage} (as opposed to
+\emph{implementation}) of linear algebra libraries. We explain the idea in
+detail and provide evidence for its efficacy. Prior type systems for fractional
+permissions \cite{boyland2003, bierhoff, bierhoff2008plural} are quite complex.
+This is because these type systems typically encode a sophisticated analysis to
+automatically infer how fractional permissions should be split and rejoined.
+
+In contrast, in \lang, we made sharing and merging explicit. As a result, we
+were able to drastically simplify the type system. Thefore, our formal system
+is very close to standard presentations of linear logic, and the implementation
+complexity is no worse than that for parametric polymorphism.
+
 In this paper
@@
         dataflow analysis \cite{bierhoff}
-    \item we describe an implementation that is both compatible with and usable
+    \item we describe an implementation that is compatible with and usable
         from existing code

### 2019-05-26 write-up/paper/lang_and_examples.tex
Fix spacing
Fix punctuation
Incorporate final review feedback
Also reduce warnings.

@@
 \section{\lang\ Overview and Examples}\label{sec:lang_and_examples}
-
-\subsection{Syntax}
-
-\lang's concrete syntax is inspired by that of OCaml. It desugars
-(Figure~\ref{fig:lang_desugar}) into a smaller, core type and expression
-grammar (Figure~\ref{fig:core_grammar}).
-
-\begin{figure}[p]
-    \begin{center}
-    \[\def\arraystretch{1.3}
-    \begin{array}{rcl}
-        f & ::= & {'\!f\!c}
-                \mid \Zf
-                \mid f\ \Sf \\
-                \\
-        t & ::= & \Unit
-                \mid \Bool
-                \mid \Int
-                \mid \Elt
-                \mid f\ \Arr
-                \mid f\ \Mat
-                \mid \Bang t
-                \mid {'\!f\!c}.\ t
-                \mid t \otimes t'
-                \mid t \multimap t' \\
-                \\
-        e & ::= & p \textrm{ (primitives)}
-                \mid x \textrm{ (variable)}
-                \mid \Let\ x = e\ \In\ e'
-                \mid ()
-                \mid \Let\ () = e\ \In\ e'
-                \mid \ottkw{true}
-                \mid \ottkw{false}
-                \\ && \ottkw{if}\ e\ \ottkw{then}\ e_1\ \ottkw{else}\ e_2
-                \mid k \textrm{ (integer)}
-                \mid l \textrm{ (heap location)}
-                \mid el \textrm{ (array element)}
-                \\ && \Many\ v
-                \mid \Let\ \Many\ x = e\ \In\ e'
-                \mid \ottkw{fun}\ '\!f\!c \rightarrow e
-                    \textrm{ (frac. perm. abstraction)}
-                \\ && e\ [f] \textrm{ (frac. perm. specialisiation)}
-                \mid (e, e')
-                \mid \Let\ (a,b) = e\ \In\ e'
-                \\ && \ottkw{fun}\ x : t \rightarrow e
-                \mid e\ e'
-                \mid \ottkw{fix} \: (g, x : t, e : t')
-    \end{array} \]
-    \end{center}
-    \caption{Core fraction $f$, type $t$ and expression $e$ grammar of
-        \lang. Values $v$ are a subset of the expressions, their full
-        definition and a list of all primitives $p$ is in
-        Appendix~\ref{sec:grammar_def}.}\label{fig:core_grammar}
-\end{figure}
-
-\begin{figure}[p]
-\begin{center}
-\[\def\arraystretch{1.3}
-    \begin{array}{rcl}
-    x[e] &
-    \Rightarrow &
-    \ottkw{get}\ \_\ x\ (e) \:\qquad \textrm{(similarly for matrices)}
-\\
-    x[e_1] := e_2 &
-    \Rightarrow &
-    \ottkw{set}\ x\ (e_1)\ (e_2) \quad \textrm{(similarly for matrices)}
-\\
-\\
-    pat & ::= & () \mid x \mid {!x} \mid \Many\ pat \mid (pat, pat)
-\\
-    \Let\ !x = e_1\ \In\ e_2 &
-    \Rightarrow &
-    \specialcell[t]{l}{%
-        \Let\ \Many\ x = e_1\ \In \\
-        \Let\ \Many\ x = \Many\ (\Many\ x)\ \In\ e_2}
-\\
-    \Let\ \Many\ \langle pat_x \rangle\ = e_1\ \In\ e_2 &
-    \Rightarrow &
-    \specialcell[t]{l}{%
-        \Let\ \Many\ x = x\ \In \\
-        \Let\ \langle pat_x \rangle\ = x\ \In\ e_2}
-\\
-    \Let\ (\langle pat_a \rangle, \langle pat_b \rangle)\ = e_1\ \In\ e_2 &
-    \Rightarrow &
-    \specialcell[t]{l}{%
-        \Let\ (a,b)\ = a\_b\ \In\
-        \Let\ \langle pat_a \rangle\ = a\ \In \\
-        \Let\ \langle pat_b \rangle\ = b\ \In\ e_2}
-\\
-    \ottkw{fun}\ (\langle pat_x \rangle : t) \rightarrow e &
-    \Rightarrow &
-    \ottkw{fun}\ (x : t) \rightarrow \Let\ \langle pat_x \rangle = x\ \In\ e
-\\
-\\
-    arg & ::= & \langle pat \rangle : t\ \mid\ {'x} \textrm{ (fractional permission variable)}
-\\
-    \ottkw{fun}\ \langle arg_{1 .. n} \rangle \rightarrow e &
-    \Rightarrow &
-    \ottkw{fun}\ \langle arg_1 \rangle \rightarrow ..
-    \ \ottkw{fun}\ \langle arg_n \rangle \rightarrow e
-\\
-    \Let\ f\ {\langle arg_{1 .. n} \rangle} = e_1\ \In\ e_2 &
-    \Rightarrow &
-    \Let\ f = \ottkw{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1\
-    \In\ e_2
-\\
-    \Let\ !f\ {\langle arg_{1 .. n} \rangle} = e_1\ \In\ e_2 &
-    \Rightarrow &
-    \Let\ \Many\ f = \Many\ (\ottkw{fun}\ {\langle arg_{1 .. n} \rangle}
-    \rightarrow e_1)\ \In\ e_2
-\\
-    \mathrm{fixpoint} & \equiv & \ottkw{fix}\ (f, x : t, \ottkw{fun}
-    \ {\langle arg_{0 .. n} \rangle} \rightarrow e_1 : {t'} )
-\\
-    \Let\ \ottkw{rec}\ f\ (x : t)\ {\langle arg_{0 .. n} \rangle} : {t'} = e_1\ \In\ e_2 &
-    \Rightarrow &
-    \Let\ f = \mathrm{fixpoint}\ \In\ e_2
-\\
-    \Let\ \ottkw{rec}\ !f\ (x : t)\ {\langle arg_{0 .. n} \rangle} : {t'} = e_1\ \In\ e_2 &
-    \Rightarrow &
-    \Let\ \Many\ f = \Many\ \mathrm{fixpoint}\ \In\ e_2
-    \end{array}
-\]
-\end{center}
-\caption{Desugaring \lang.}\label{fig:lang_desugar}
-\end{figure}
 
@@
 spaces $A \multimap B$ and tensor products $A \otimes B$. We adopt linearity --
-the restriction that each program variable be used at most once -- since it
+the restriction that each program variable be used exactly once -- since it
 allows us to express purely functional APIs for numerical library routines that
@@
 
-\subsubsection{Intuitionism: ! and \highl{Many}}
+\subsubsection{Intuitionism: ! and \texorpdfstring{\highl{Many}}{Many}}
 
@@
 
-\subsubsection{Fractional Permissions}
+\subsubsection{Fractional Permissions}\label{subsubsec:frac_perm}
 
@@
 
+\subsection{Syntax}\label{subsec:syntax}
+
+\lang's concrete syntax is inspired by that of OCaml. It desugars
+(Figure~\ref{fig:lang_desugar}) into a smaller, core type and expression
+grammar (Figure~\ref{fig:core_grammar}).
+
+\begin{figure}[p]
+    \begin{center}
+    \[\def\arraystretch{1.3}
+    \begin{array}{rcl}
+        f & ::= & {'\!f\!c}
+                \mid \Zf
+                \mid f\ \Sf \\
+                \\
+        t & ::= & \Unit
+                \mid \Bool
+                \mid \Int
+                \mid \Elt
+                \mid f\ \Arr
+                \mid f\ \Mat
+                \mid \Bang t
+                \mid {'\!f\!c}.\ t
+                \mid t \otimes t'
+                \mid t \multimap t' \\
+                \\
+        e & ::= & p \textrm{ (primitives)}
+                \mid x \textrm{ (variable)}
+                \mid \Let\ x = e\ \In\ e'
+                \mid ()
+                \mid \Let\ () = e\ \In\ e'
+                \mid \ottkw{true}
+                \mid \ottkw{false}
+                \\ && \ottkw{if}\ e\ \ottkw{then}\ e_1\ \ottkw{else}\ e_2
+                \mid k \textrm{ (integer)}
+                \mid l \textrm{ (heap location)}
+                \mid el \textrm{ (array element)}
+                \\ && \Many\ v
+                \mid \Let\ \Many\ x = e\ \In\ e'
+                \mid \ottkw{fun}\ '\!f\!c \rightarrow e
+                    \textrm{ (frac. perm. abstraction)}
+                \\ && e\ [f] \textrm{ (frac. perm. specialisation)}
+                \mid (e, e')
+                \mid \Let\ (a,b) = e\ \In\ e'
+                \\ && \ottkw{fun}\ x : t \rightarrow e
+                \mid e\ e'
+                \mid \ottkw{fix} \: (g, x : t, e : t')
+    \end{array} \]
+    \end{center}
+    \caption{Core fraction $f$, type $t$ and expression $e$ grammar of
+        \lang. Values $v$ are a subset of the expressions, their full
+        definition and a list of all primitives $p$ is in
+        Appendix~\ref{sec:grammar_def}.}\label{fig:core_grammar}
+\end{figure}
+
+\begin{figure}[p]
+\begin{center}
+\[\def\arraystretch{1.3}
+    \begin{array}{rcl}
+    x[e] &
+    \Rightarrow &
+    \ottkw{get}\ \_\ x\ (e) \:\qquad \textrm{(similarly for matrices)}
+\\
+    x[e_1] := e_2 &
+    \Rightarrow &
+    \ottkw{set}\ x\ (e_1)\ (e_2) \quad \textrm{(similarly for matrices)}
+\\
+\\
+    pat & ::= & () \mid x \mid {!x} \mid \Many\ pat \mid (pat, pat)
+\\
+    \Let\ !x = e_1\ \In\ e_2 &
+    \Rightarrow &
+    \specialcell[t]{l}{%
+        \Let\ \Many\ x = e_1\ \In \\
+        \Let\ \Many\ x = \Many\ (\Many\ x)\ \In\ e_2}
+\\
+    \Let\ \Many\ \langle pat_x \rangle\ = e_1\ \In\ e_2 &
+    \Rightarrow &
+    \specialcell[t]{l}{%
+        \Let\ \Many\ x = x\ \In \\
+        \Let\ \langle pat_x \rangle\ = x\ \In\ e_2}
+\\
+    \Let\ (\langle pat_a \rangle, \langle pat_b \rangle)\ = e_1\ \In\ e_2 &
+    \Rightarrow &
+    \specialcell[t]{l}{%
+        \Let\ (a,b)\ = a\_b\ \In\
+        \Let\ \langle pat_a \rangle\ = a\ \In \\
+        \Let\ \langle pat_b \rangle\ = b\ \In\ e_2}
+\\
+    \ottkw{fun}\ (\langle pat_x \rangle : t) \rightarrow e &
+    \Rightarrow &
+    \ottkw{fun}\ (x : t) \rightarrow \Let\ \langle pat_x \rangle = x\ \In\ e
+\\
+\\
+    arg & ::= & (\langle pat \rangle : t)\ \mid\ ({'x}) \textrm{ (fractional permission variable)}
+\\
+    \ottkw{fun}\ \langle arg_{1 .. n} \rangle \rightarrow e &
+    \Rightarrow &
+    \ottkw{fun}\ \langle arg_1 \rangle \rightarrow ..
+    \ \ottkw{fun}\ \langle arg_n \rangle \rightarrow e
+\\
+    \Let\ f\ {\langle arg_{1 .. n} \rangle} = e_1\ \In\ e_2 &
+    \Rightarrow &
+    \Let\ f = \ottkw{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1\
+    \In\ e_2
+\\
+    \Let\ !f\ {\langle arg_{1 .. n} \rangle} = e_1\ \In\ e_2 &
+    \Rightarrow &
+    \Let\ \Many\ f = \Many\ (\ottkw{fun}\ {\langle arg_{1 .. n} \rangle}
+    \rightarrow e_1)\ \In\ e_2
+\\
+    \mathrm{fixpoint} & \equiv & \ottkw{fix}\ (f, x : t, \ottkw{fun}
+    \ {\langle arg_{0 .. n} \rangle} \rightarrow e_1 : {t'} )
+\\
+    \Let\ \ottkw{rec}\ f\ (x : t)\ {\langle arg_{0 .. n} \rangle} : {t'} = e_1\ \In\ e_2 &
+    \Rightarrow &
+    \Let\ f = \mathrm{fixpoint}\ \In\ e_2
+\\
+    \Let\ \ottkw{rec}\ !f\ (x : t)\ {\langle arg_{0 .. n} \rangle} : {t'} = e_1\ \In\ e_2 &
+    \Rightarrow &
+    \Let\ \Many\ f = \Many\ \mathrm{fixpoint}\ \In\ e_2
+    \end{array}
+\]
+\end{center}
+\caption{Desugaring \lang.}\label{fig:lang_desugar}
+\end{figure}
+
+\begin{figure}[t]
+\begin{center}
+\[\def\arraystretch{1.3}
+    \begin{array}{rcl}
+    \Let\ v \leftarrow x[e]\ \In\ e' &
+    \Rightarrow &
+    \Let\ (x, !v)\ = x[e]\ \In\ e' \qquad \textrm{(similarly for matrices)}
+\\
+    \Let\ x_2 \leftarrow \ottkw{new}\ [|\ x_1\ |]\ \In\ e &
+    \Rightarrow &
+    \Let\ (x_1, x_2)\ = \ottkw{copyM}\ \_\ x_1\ \In\ e
+\\
+    \Let\ x_2 \leftarrow [|\ x_1\ |]\ \In\ e &
+    \Rightarrow &
+    \Let\ (x_1, x_2)\ = \ottkw{copyM\_to}\ \_\ x_1\ x_2\ \In\ e
+    \end{array}
+\]
+\[
+    M ::= X\ \mid\ X^T\ \mid\ \textrm{sym}( X )
+\]
+\end{center}
+\begin{align*}
+% new
+    \Let\ &Y \leftarrow \ottkw{new}\ (n, k)\ [|\ \alpha M_1 M_2 \ |]\ \In\ e
+    \Rightarrow \\
+    & \Let\ Y\ = \ottkw{matrix}\ n\ k\ \In
+    \ \Let\ Y \leftarrow [|\ \alpha M_1 M_2 + 0Y\ |]\ \In\ e
+\\[0.1\baselineskip]
+%syrk
+    \Let\ &Y \leftarrow [|\ \alpha X X^T + \beta Y\ |]\ \In\ e
+    \Rightarrow \\
+    & \Let\ (X,Y)\ = \ottkw{syrk}\ \ottkw{false}
+    \ \alpha\ \_\ X\ \beta\ Y\ \In\ e
+\\[0.1\baselineskip]
+    \Let\ &Y \leftarrow [|\ \alpha X^T X + \beta Y\ |]\ \In\ e
+    \Rightarrow \\
+    & \Let\ (X,Y)\ = \ottkw{syrk}\ \ottkw{true}
+    \ \alpha\ \_\ X\ \beta\ Y\ \In\ e
+%symm
+\\[0.1\baselineskip]
+    \Let\ &Y \leftarrow\ [|\ \alpha\,\textrm{sym}(X_1)\,X_2 + \beta Y\ |]\ \In\ e
+    \Rightarrow \\
+    & \Let\ ((X_1,X_2),Y)\ = \ottkw{symm}\ \ottkw{false}
+        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \In\ e
+\\[0.1\baselineskip]
+    \Let\ &Y \leftarrow\ [|\ \alpha\,X_2\,\textrm{sym}(X_1) + \beta Y\ |]\ \In\ e
+    \Rightarrow \\
+    & \Let\ ((X_1,X_2),Y)\ = \ottkw{symm}\ \ottkw{true}
+        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \In\ e
+\\[0.1\baselineskip]
+%gemm
+    \Let\ &Y \leftarrow [|\ \alpha X_1^{T?} X_2^{T?} + \beta Y\ |]\ \In\ e
+    \Rightarrow \\
+    & \Let\ ((X_1,X_2),Y)\ =\ \ottkw{gemm}\ \alpha
+        \ \_\ \left(X_1, \substack{\ottkw{true}\\\ottkw{false}}\right)
+        \ \_\ \left(X_2, \substack{\ottkw{true}\\\ottkw{false}}\right)
+        \ \beta\ Y\ \In\ e
+\end{align*}
+
+\caption{Purely syntactic pattern-matching translations of
+    matrix expressions.}\label{fig:lang_matexp}
+\end{figure}
+
+As described in Section~\ref{subsubsec:frac_perm}, fractional permissions $f$
+are either variables $'\!f\!c$, $\Zf$ ($2^0$) or $f\ \Sf$ ($2^{-(f + 1)}$).
+
+Types are either simple ($\Unit$, $\Bool$, $\Int$, $\Elt$), indexed
+by fractional permission ($f\ \Arr$, $f\ \Mat$) or compound (!
+constructor for intuitionism, universally-quantifying over a
+fractional-permission $'\!f\!c$ in $t$ for fraction-polymorphic types, pairs,
+linear functions).
+
+Expressions are standard with the exception of $'\!f\!c.\ t$ introduction
+(fractional permission abstraction) and elimination (fractional permission
+specialisation) forms.
+
+The ! annotation on variables is a syntactic convenience for declaring the
+variable to used intuitionistically; why it desugars the way it does is
+explained in Section~\ref{subsec:core_tt}.
+
+Array/matrix indexing and duplication (non-destructive and destructive) also
+have special syntax to lessen the syntactic overhead of re-binding identifiers.
+Furthermore, to aid readability, there is support for using BLAS methods via
+conventional-looking matrix expressions.
+
+In particular, the syntax \highl{let y <- new (m,n) [| alpha * x1 * x2 |]} is
+syntactic sugar for first creating a new $m \times n$ matrix (\highl{let y =
+matrix m n}) and then storing the result of the multiplication in it
+(\highl{let ((x1, x2), y) = .. in ..}).
+
+Note that, the pattern \highl{let y <- [| x^T * x + beta * y |]} translates to
+(\highl{syrk true 1. x beta y}), which uses \texttt{x} once only.
+
 \subsection{Examples}
@@
 Although a factorial function (Figure~\ref{fig:lang_factorial}) may seem like
-an aggresively pedestrian first example, in a linearly typed language such as
+an aggressively pedestrian first example, in a linearly typed language such as
 \lang\ it represents the culmination of many features.
 
-To simplifiy the design and implementation of \lang's type system, recursive
+To simplify the design and implementation of \lang's type system, recursive
 functions must have full type annotations (non-recursive functions need only
-their argument types annotated). Its body is a closed expression (with respect
-to the function's arguments), so it type-checks (since it does not capture any
-linear values from its environment).
+their argument types annotated). The body of the factorial function is a closed
+expression (with respect to the function's arguments), so it type-checks (since
+it does not capture any linear values from its environment).
 
-The only argument is \highl{!x : !int}. The ! annotation on \texttt{x} is a
-syntactic convenience for declaring the value to used intuitionistically, its
-full and precise meaning is described in Section~\ref{subsec:core_tt}.
+The only argument is \highl{!x : !int}. As explained before
+(Section~\ref{subsec:syntax}), this declares \texttt{x} to be used
+intuitionistically.
 
@@
 \highl{!elt} (types of elements of arrays/matrices, typically 64-bit
-floating-point numbers). The short-circuting \texttt{||} behaves in exactly the
+floating-point numbers). The short-circuiting \texttt{||} behaves in exactly the
 same way as a boolean-valued \highl{if}-expression.
@@
 the recursive call), they can also be \emph{automatically inferred at call
-sites}: \highl{row[i] == get _ row i} takes advantage of this convenience.
+sites}: \highl{row[i]} $\Rightarrow$ \highl{get _ row i} takes advantage of
+this convenience.
 
@@
 {\Zf\ \Arr}$. It consumes the original array and returns a new array with
-the updated value.  \highl{let written = write[i] := <exp> } is just syntactic
-sugar for \highl{let written = set write i <exp>}.
+the updated value.
 
-Since \highl{write: z arr} (where \Zf stands for $k=0$, representing a
+Since \highl{write: z arr} (where \Zf\ stands for $k=0$, representing a
 fractional permission of $2^{-k} = 2^{-0} = 1$), we may mutate it, but since we
@@
 \texttt{in}, \texttt{out} and \texttt{inout} parameters. The latter two would
-need full \Zf permissions; the first would be fraction-polymorphic.
+need full \Zf\ permissions; the first would be fraction-polymorphic.
 
@@
     \inputminted[fontsize=\small]{ocaml}{../../examples/square.lt}
-    \caption{Linear regression (OLS): $\hat\beta =
-        \mathbf{(X^T X)^{-1} X^T y}$}\label{fig:lang_square}
+    \caption{Squaring a matrix}\label{fig:lang_square}
 \end{figure}
@@
 
-The line \highl{let answer <- new (m,n) [| x1 * x2 |]} is syntactic sugar for
-first creating a new $m \times n$ matrix (\highl{let answer = matrix m n}) and
-then storing the result of the multiplication in it (\highl{let ((x1, x2),
-answer) = gemm 1. _ (x1, false) _ (x2, false) 0. answer}). \highl{false} means
-the matrix should not be accessed with indices transposed.
-
 By using some simple pattern-matching and syntactic sugar
@@
 \end{itemize}
-
-\begin{figure}[t]
-\begin{center}
-\[\def\arraystretch{1.3}
-    \begin{array}{rcl}
-    \Let\ v \leftarrow x[e]\ \In\ e &
-    \Rightarrow &
-    \Let\ (x, !v)\ = x[e]\ \In\ e \qquad \textrm{(similarly for matrices)}
-\\
-    \Let\ x_2 \leftarrow \ottkw{new}\ [|\ x_1\ |]\ \In\ e &
-    \Rightarrow &
-    \Let\ (x_1, x_2)\ = \ottkw{copyM}\ \_\ x_1\ \In\ e
-\\
-    \Let\ x_2 \leftarrow [|\ x_1\ |]\ \In\ e &
-    \Rightarrow &
-    \Let\ (x_1, x_2)\ = \ottkw{copyM\_to}\ \_\ x_1\ x_2\ \In\ e
-    \end{array}
-\]
-\[
-    M ::= X\ \mid\ X^T\ \mid\ \textrm{sym}( X )
-\]
-\end{center}
-\begin{align*}
-% new
-    \Let\ &Y \leftarrow \ottkw{new}\ (n, k)\ [|\ \alpha M_1 M_2 \ |]\ \In\ e
-    \Rightarrow \\
-    & \Let\ Y\ = \ottkw{matrix}\ n\ k\ \In
-    \ \Let\ Y \leftarrow [|\ \alpha M_1 M_2 + 0Y\ |]\ \In\ e
-\\[0.1\baselineskip]
-%syrk
-    \Let\ &Y \leftarrow [|\ \alpha X X^T + \beta Y\ |]\ \In\ e
-    \Rightarrow \\
-    & \Let\ (X,Y)\ = \ottkw{syrk}\ \ottkw{false}
-    \ \alpha\ \_\ X\ \beta\ Y\ \In\ e
-\\[0.1\baselineskip]
-    \Let\ &Y \leftarrow [|\ \alpha X^T X + \beta Y\ |]\ \In\ e
-    \Rightarrow \\
-    & \Let\ (X,Y)\ = \ottkw{syrk}\ \ottkw{true}
-    \ \alpha\ \_\ X\ \beta\ Y\ \In\ e
-%symm
-\\[0.1\baselineskip]
-    \Let\ &Y \leftarrow\ [|\ \alpha\,\textrm{sym}(X_1)\,X_2 + \beta Y\ |]\ \In\ e
-    \Rightarrow \\
-    & \Let\ ((X_1,X_2),Y)\ = \ottkw{symm}\ \ottkw{false}
-        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \In\ e
-\\[0.1\baselineskip]
-    \Let\ &Y \leftarrow\ [|\ \alpha\,X_2\,\textrm{sym}(X_1) + \beta Y\ |]\ \In\ e
-    \Rightarrow \\
-    & \Let\ ((X_1,X_2),Y)\ = \ottkw{symm}\ \ottkw{true}
-        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \In\ e
-\\[0.1\baselineskip]
-%gemm
-    \Let\ &Y \leftarrow [|\ \alpha X_1^{T?} X_2^{T?} + \beta Y\ |]\ \In\ e
-    \Rightarrow \\
-    & \Let\ ((X_1,X_2),Y)\ =\ \ottkw{gemm}\ \alpha
-        \ \_\ \left(X_1, \substack{\ottkw{true}\\\ottkw{false}}\right)
-        \ \_\ \left(X_2, \substack{\ottkw{true}\\\ottkw{false}}\right)
-        \ \beta\ Y\ \In\ e
-\end{align*}
-
-\caption{Purely syntactic pattern-matching translations of
-    matrix expressions.}\label{fig:lang_matexp}
-\end{figure}
 
@@
 
-However, the line \highl{let x_T_x <- new (m,m) [| x^T * x |]}, works for a
-slightly different reason: that pattern is matched to a BLAS call to
-(\highl{syrk true 1. x 0. x_T_x}), which only uses \texttt{x} once. Hence
-\texttt{x} can appear \emph{twice} in the \emph{pattern} without any calls to
-\texttt{share}.
+Note that \texttt{x} can appear \emph{twice} in the \emph{pattern} without any
+calls to \texttt{share} because the pattern is matched to a BLAS call to
+\highl{syrk true 1. x 0. x_T_x}, which uses \texttt{x} once only.
 
@@
 
-Figure~\ref{fig:lang_l1_norm_min} shows even more pattern-matching. Patterns of
-the form \highl{let <id> <- [| beta * c + alpha * a * b |]} are also desugared
-to \texttt{gemm} calls. Primitives like $\ottkw{transpose}:\ {'x}.\ {{'x}\ \Mat}
+Primitives like $\ottkw{transpose}:\ {'x}.\ {{'x}\ \Mat}
 \multimap {{'x}\ \Mat} \otimes {\Zf\ \Mat}$ and $\ottkw{eye}:\ {!\Int}
@@
 with \texttt{to\_del} and \texttt{posv} in the previous example, we do not need
-the value stored in \texttt{tmp\_5\_5} after the call to \texttt{gesv} (a
+the value stored in \texttt{tmp\_n\_n} after the call to \texttt{gesv} (a
 primitive similar to \texttt{posv} but for a non-symmetric first argument).
 However, we can re-use its memory much later to store \texttt{answer} with
-\highl{let answer <- [| 0. * tmp_5_5 + q_inv_u * inv_u_T |]}. Again, thanks to
-linearity, the identifiers \texttt{q} and \texttt{tmp\_5\_5} are out of scope
+\highl{let answer <- [| 0. * tmp_n_n + q_inv_u * inv_u_T |]}. Again, thanks to
+linearity, the identifiers \texttt{q} and \texttt{tmp\_n\_n} are out of scope
 by the time \texttt{answer} is bound. Although during execution, all three
@@
 
-    \item \highl{let new_r <- new [| r_2 |]} creates a copy of \texttt{r\_2}
-
     \item \texttt{posvFlip} is like \texttt{posv} except for solving $XA = B$
@@
             \item \texttt{mu} and \texttt{new\_mu}
-            \item \texttt{sigma\_hT} and \texttt{x}
+            \item \texttt{sigma\_hT} and \texttt{x}.
         \end{itemize}

### 2019-05-26 write-up/paper/paper.tex
Fix formatting
Make stuff consistent
Syntax & spacing, with a few small edits from Reviewer C's review.
Incorporate final review feedback
Also reduce warnings.

@@
 
-\supplement{\url{www.github.com/dc-mak/lt4la}}
+\supplement{\url{www.github.com/dc-mak/NumLin}}
 
-%\nolinenumbers %uncomment to disable line numbering
+\acknowledgements{We would like to thank Stephen Dolan for his advice and
+support with the implementation and evaluation of NumLin. We would also like to
+thank the (anonymous) reviewers for their feedback and suggestions.}
+
+\nolinenumbers %uncomment to disable line numbering
 
@@
 %Editor-only macros:: begin (do not touch as author)%
-\EventEditors{John Q. Open and Joan R. Access}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
+\EventEditors{Alastair F. Donaldson}
+\EventNoEds{1}
+\EventLongTitle{33rd European Conference on Object-Oriented Programming (ECOOP 2019)}
+\EventShortTitle{ECOOP 2019}
+\EventAcronym{ECOOP}
+\EventYear{2019}
+\EventDate{July 15--19, 2019}
+\EventLocation{London, United Kingdom}
 \EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
+\SeriesVolume{134}
+\ArticleNo{9}
 
@@
 \newcommand{\Elt}{\ottkw{elt}}%
+\newcommand{\Arr}{\ottkw{arr}}%
 \newcommand{\Mat}{\ottkw{mat}}%
-\newcommand{\Zf}{\ottkw{Z}}%
-\newcommand{\Sf}{\ottkw{S}}%
+\newcommand{\Zf}{\texttt{z}}%
+\newcommand{\Sf}{\texttt{s}}%
+\newcommand{\Let}{\ottkw{let}}%
+\newcommand{\In}{\ottkw{in}}%
 \newcommand{\Many}{\ottkw{Many}}%

### 2019-05-26 write-up/semantics/lemmas.tex
Start on Soundness Theorem
Update to LIPIcs style
Make stuff consistent
Syntax & spacing, with a few small edits from Reviewer C's review.
Incorporate final review feedback
Also reduce warnings.

--- 2018-10-23 write-up/semantics/lemmas.tex
@@
 
-\subsection{$
+\subsection{Moral equivalent of frame-rule
+\texorpdfstring{$
     \forall \sigma_s, \sigma_r, e.\ \varsigma_s \star \varsigma_r \textrm{ defined }
@@
     \langle \sigma_s + \sigma_r , e \rangle \rightarrow^n
-$}\label{frame}
+$}{}}\label{frame}
 
@@
 
-    \step{}{\case{\textsc{Op\_Free}, $\sigma_s \equiv \sigma' + \{ l
+    \step{<1>1}{\case{\textsc{Op\_Free}, $\sigma_s \equiv \sigma' + \{ l
         \mapsto_1 m \}$, $\sigma_f = \sigma'$.}{\pf~Instantiate
@@
 
-    \step{}{\case{\textsc{Op\_Matrix}}{
+    \step{<1>2}{\case{\textsc{Op\_Matrix}}{
         \pf~Rule has no requirements on $\sigma_s$ so will also work with
@@
 
-    \step{}{\case{\textsc{Op\_Share}, $\sigma_s \equiv \sigma' + \{ l \mapsto_f
+    \step{<1>3}{\case{\textsc{Op\_Share}, $\sigma_s \equiv \sigma' + \{ l \mapsto_f
         m \}$, $\sigma_f = \sigma' + \{ l \mapsto_{\frac{1}{2} \cdot f} m \} +
@@
 
-    \step{}{\case{\textsc{Op\_Unshare\_Eq}, $\sigma_s \equiv \sigma' + \{ l
+    \step{<1>4}{\case{\textsc{Op\_Unshare\_Eq}, $\sigma_s \equiv \sigma' + \{ l
         \mapsto_{\frac{1}{2} \cdot f} m \} + \{ l \mapsto_{\frac{1}{2} \cdot f}
@@
 
-        \step{}{Union-ing $\sigma_r$ does not remove $l \mapsto_{\frac{1}{2}
+        \step{<2>1}{Union-ing $\sigma_r$ does not remove $l \mapsto_{\frac{1}{2}
             \cdot f} m$, so that can still be split out of $ \sigma_s + \sigma_r$.}
 
-        \step{}{There may also be other valid splits introduced by $\sigma_r$.}
+        \step{<2>2}{There may also be other valid splits introduced by $\sigma_r$.}
 
-        \step{}{However, by assumption of $\varsigma_s \star \varsigma_r$
+        \step{<2>3}{However, by assumption of $\varsigma_s \star \varsigma_r$
             defined, any splitting of $\sigma_s + \sigma_r$ will
@@
 
-    \step{}{\case{\textsc{Op\_Gemm\_Match}}{}}
+    \step{<1>5}{\case{\textsc{Op\_Gemm\_Match}}{}}
 
@@
 
-        \step{}{By assumption of $\varsigma_s \star \varsigma_r$ defined,
+        \step{<2>1}{By assumption of $\varsigma_s \star \varsigma_r$ defined,
             either $l_1$ (or $l_2$, or both) are not in $\sigma_r$, or they are
@@
 
-        \step{}{The permissions (of $l_1$ and/or $l_2$) may differ, but
+        \step{<2>2}{The permissions (of $l_1$ and/or $l_2$) may differ, but
             \textsc{Op\_Gemm\_Match} universally quantifies over them and
@@
 
-        \step{}{Only the pointed to matrix value at $l_3$ changes.}
+        \step{<2>3}{Only the pointed to matrix value at $l_3$ changes.}
 
-        \step{}{\suffices{$l_3 \notin \pi_1 [ \sigma_r ]$.}}
+        \step{<2>4}{\suffices{$l_3 \notin \pi_1 [\sigma_r]$.}}
 
-        \step{}{By assumption of $\varsigma_s \star \varsigma_r$ defined, $l_3
+        \step{<2>5}{By assumption of $\varsigma_s \star \varsigma_r$ defined, $l_3
             \notin \dom(\varsigma_r)$.}
 
-        \step{}{Hence $l_3 \notin \pi_1 [ \sigma_r ]$.}
+        \step{<2>6}{Hence $l_3 \notin \pi_1 [\sigma_r]$.}
 
@@
 
-\subsection{$\forall k, t.\ \V{k}{t} \subseteq \C{k}{t}$}\label{subsetVC}
+\subsection{Semantically, values are expressions
+    \texorpdfstring{$\forall k, t.\ \V{k}{t} \subseteq \C{k}{t}$}{}
+}\label{subsetVC}
+
 Follows from definition of $\C{k}{t}$, $\rightarrow^j$ ($\forall n.\ \langle
@@
 
-\subsection{$\forall \theta, \delta, \gamma, v.\ \theta(\delta(\gamma(v)))\textrm{ is a value.}$}\label{valueSub}
+\subsection{Values remain values under all substitutions
+    \texorpdfstring{$\forall \theta, \delta, \gamma, v.\ \theta(\delta(\gamma(v)))\textrm{ is a value.}$}{}
+}\label{valueSub}
 
-$\theta$ is irrelevant because it only maps fractional capability variables to
-fractional capabilities. By construction, $\delta$ and $\gamma$ only map
+$\theta$ is irrelevant because it only maps fractional permission variables to
+fractional permissions. By construction, $\delta$ and $\gamma$ only map
 variables to values, and values are closed under substitution.
 
-\subsection{$
-    \forall k, \sigma, \sigma', e, e', t.  \ (\varsigma', e') \in \C{k}{t} \wedge
+\subsection{Stepping reduces the step-index
+\texorpdfstring{
+    $\forall k, \sigma, \sigma', e, e', t.  \ (\varsigma', e') \in \C{k}{t} \wedge
     \langle \sigma, e \rangle \rightarrow \langle \sigma', e' \rangle
     \Rightarrow (\varsigma, e) \in \C{k+1}{t}
-$}\label{stepInC}
+$}{}}\label{stepInC}
+
+In the lemma, and for the rest of its proof, $\varsigma = \den{H}{}{\sigma}$.
 
@@
 
-    \step{}{\case{$j=0$. Clearly $\sigma_f = \sigma_s + \sigma_r$ and $e' = e$.}{
+    \step{<1>1}{\case{$j=0$. Clearly $\sigma_f = \sigma_s + \sigma_r$ and $e' = e$.}{
         Remains to show that if $e$ is a value then $(\varsigma_s \star
@@
 
-    \step{}{\case{$j \geq 1$. We have $\langle \sigma, e \rangle
-        \rightarrow^j\, = \langle \sigma', e' \rangle \rightarrow^{j-1}$.}{
+    \step{<1>2}{\case{$j \geq 1$. We have $\langle \sigma, e \rangle
+        \rightarrow^j \, = \langle \sigma', e' \rangle \rightarrow^{j-1}$.}{
         Instantiate $(\varsigma', e') \in \C{k}{t}$, with $j-1 < k$ and
@@
 
-\subsection{$j \leq k \Rightarrow \den{\_\ }{k}{\cdot} \subseteq \den{\_\ }{j}{\cdot}$}\label{subsetKJ}
+\subsection{Monotonicity for step-index
+    \texorpdfstring{$j \leq k \Rightarrow \den{\_\ }{k}{\cdot} \subseteq \den{\_\ }{j}{\cdot}$}{}
+}\label{subsetKJ}
 
+For the rest of this proof, $\varsigma = \den{H}{}{\sigma}$.\\
 Lemma~\ref{stepInC} is the inductive step for this lemma for the $\C{}{}$ case.\\
@@
 
-    \step{}{$v'$ is of the correct syntactic form (lambda or fixpoint) by
+    \step{<1>1}{$v'$ is of the correct syntactic form (lambda or fixpoint) by
         assumption.}
 
-    \step{}{\assume{arbitrary $j' \leq j$ and $(\varsigma_v, v) \in \V{j'}{t}$
+    \step{<1>2}{\assume{arbitrary $j' \leq j$ and $(\varsigma_v, v) \in \V{j'}{t}$
         such that $\varsigma_{v'} \star \varsigma_v$ is defined.}}
 
-    \step{}{\suffices{to show $(\varsigma_{v'} \star \varsigma_v, v' v) \in
+    \step{<1>3}{\suffices{to show $(\varsigma_{v'} \star \varsigma_v, v' v) \in
         \C{j'}{t'}$.}}
 
-    \step{}{This is true by instantiating $(\varsigma_{v'}, v') \in \V{k}{t
+    \step{<1>4}{This is true by instantiating $(\varsigma_{v'}, v') \in \V{k}{t
         \multimap t'}$ with $j' \leq k$ and $(\varsigma_v, v) \in \V{j'}{t}$.}
@@
 
-\subsection{$\forall \Delta, \Gamma, t, k, \theta, \delta, \gamma.\ %
+\subsection{Domains match
+\texorpdfstring{$
+    \forall \Delta, \Gamma, t, k, \theta, \delta, \gamma.\ %
     \delta \in \den{I}{k}{\Delta}\theta \wedge \gamma \in \pi_2[\den{L}{k}{\Gamma}\theta]
-    \Rightarrow \dom(\Delta) = \dom(\delta)$ and $\dom(\Gamma) = \dom(\gamma)$}\label{samedom}
+    \Rightarrow \dom(\Delta) = \dom(\delta)$ and $\dom(\Gamma) = \dom(\gamma)
+$}{}}\label{samedom}
 
@@
 
-\subsection{$\forall k, \Gamma, \Gamma', \theta, \sigma_+, \gamma_+.\ %
+\subsection{Splitting up linear environments corresponds to splitting up heaps
+\texorpdfstring{$
+    \forall k, \Gamma, \Gamma', \theta, \sigma_+, \gamma_+.\ %
     (\varsigma_+, \gamma) \in \den{L}{k}{ \Gamma, \Gamma' }\theta
-    \wedge \Gamma, \Gamma' \textrm{ disjoint } \Rightarrow\\
+    \wedge \Gamma, \Gamma' \textrm{ disjoint } \Rightarrow
     \exists \sigma, \gamma, \sigma' , \gamma' .\ \sigma_+ = \sigma + \sigma'
     \wedge \gamma, \gamma' \textrm{ disjoint }
-    \wedge \gamma_+ = \gamma \cup \gamma'\\
+    \wedge \gamma_+ = \gamma \cup \gamma'
     \wedge (\varsigma, \gamma) \in \den{L}{k}{\Gamma}
-    \wedge (\varsigma', \gamma') \in \den{L}{k}{\Gamma'} $}\label{restriction}
+    \wedge (\varsigma', \gamma') \in \den{L}{k}{\Gamma'}
+$}{}}\label{restriction}
 
 \pf~By induction on $\Gamma'$.
+
+\subsection{Fractional permission substitutions preserve progress
+\texorpdfstring{$
+    \forall e, \sigma, e', \sigma', \theta.\
+    \langle \sigma, e \rangle \rightarrow \langle \sigma',  e' \rangle
+    \Rightarrow \langle \theta(\sigma), \theta(e) \rangle \rightarrow
+    \langle \theta(\sigma') , \theta(e') \rangle
+$}{}}\label{fracPermSub}
+
+\pf~By induction on $\rightarrow$.
+\begin{proof}
+
+    \step{<1>1}{\assume{Arbitrary $e, \sigma, e', \sigma', \theta$ such that 
+        $\langle \sigma, e \rangle \rightarrow \langle \sigma', e' \rangle$.}}
+
+    \step{<1>2}{\suffices{To consider only the following rules which mention
+        fractional permission variables: \textsc{Op\_Frac\_Perm},
+        \textsc{Op\_Share}, \textsc{Op\_Unshare\_(N)Eq} and
+        \textsc{Op\_Gemm\_(Mis)Match}.}}
+
+    \step{<1>3}{\case{\textsc{Op\_Frac\_Perm.}}{Because substitution avoids capture,
+        \\ $\langle \theta(\sigma), \theta((\ottkw{fun}\: '\!f\!c \rightarrow v) \, [f])
+         \rangle \rightarrow \langle \theta(\sigma' \, [f / f\!c]),
+        \theta(v \, [f / f\!c]) \rangle$.}}
+
+    \step{<1>4}{The rest of the cases are parametric in their use of fractional
+        permission variables and so will take the same step ater any substitution.}
+
+    \step{<1>5}{\textsc{Corollary:}
+        If $\langle \sigma \, [f_1 / f\!c], e \, [f_1 / f\!c] \rangle \rightarrow^n
+        \langle \sigma_2, e'_2 \rangle$ and
+        $\langle \sigma \, [f_2 / f\!c], e \, [f_2 / f\!c] \rangle \rightarrow^n
+        \langle \sigma_2, e'_2 \rangle$, then
+        $\exists \, \sigma, e'.\ \sigma_1 = \sigma \, [f_1 / f\!c] \wedge
+        \sigma_2 = \sigma \, [f_2 / f\!c] \wedge
+        e'_1 = e' \, [f_1 / f\!c] \wedge e'_2 = e' \, [f_2 / f\!c]$.}
+
+\end{proof}

### 2019-05-27 write-up/paper/formal_system.tex
Remove Appendix (cleanly) from rest of paper

@@
 
-The full typing rules are in Appendix \ref{subsec:static_sem}, but the key
+The full typing rules are in Appendix~\ref{subsec:static_sem}, but the key
 ideas are as follow.
@@
 
-The full, small-step transition relation is in Appendix \ref{subsec:dyn_sem},
+The full, small-step transition relation is in Appendix~\ref{subsec:dyn_sem},
 but the key ideas are as follow.

### 2019-06-29 write-up/paper/implementation.tex
Update implementation.tex

@@
     \begin{minted}[fontsize=\small]{ocaml}
-let rec f i n x0 row =
+let rec sum_array i n x0 row =
   if Prim.extract @@ Prim.eqI i n then (row, x0)
@@
     let row, x1 = Prim.get row i in
-    f (Prim.addI i (Many 1)) n (Prim.addE x0 x1) row
-in
-f
+    sum_array (Prim.addI i (Many 1)) n (Prim.addE x0 x1) row in
+sum_array
     \end{minted}

### 2019-08-14 write-up/paper/intro.tex
Tidy up intro

@@
 values. However, this power comes paired with the responsibility to
-manually manage the memory associated with each array object, and in
-addition to bringing in the familiar difficulties of reasoning about
-lifetimes, aliasing and sharing that plague low-level systems
-programming; this also moves the APIs away from the linear-algebraic,
+manually manage the memory, lifetimes, aliasing and sharing of each array
+object: concerns similar to those that plague low-level systems
+programming. It also moves the APIs away from the linear-algebraic,
 mathematical style of thinking that numerical programmers want to use.
